### 2023-09-01 Tula'i-fin.qmd
yet to finish. I have added cleaning, and Regression codes so far.

---
title: "Tulai"
author: "Setareh Shafizadeh, Ben Marwick"
date: now
format: 
  html:
    self-contained: true
execute: 
  warning: false
  fig-dpi: 600
title-block-published: "last updated"
editor: visual
---
## Introduction

Cleaning Data:

```{r}
# library
library(tidyverse)
library(ggplot2)
library(dplyr)
library(magrittr)
library(stringr)

```

```{r}
# Read the data
tl <- read_csv("Tulai_Lithic_Assamblage.csv")

# Create a new data frame with renamed columns
tl1 <- tl
names(tl1)[1:2] <- c("lithic_id", 
                     "TA")

# modify TA column
tl1 <- tl1 %>%
  # Remove everything within brackets
  mutate(TA = str_remove_all(TA, 
                             "\\(.*\\)")) %>%

  # Add slash between TA and numbers
  mutate(TA = str_replace_all(TA, 
                              "TA([0-9]+)", 
                              "TA/\\1")) %>%

  # Add slash between numbers
  mutate(TA = str_replace_all(TA, 
                              "([0-9]+)/([0-9]+)", "\\1/\\2")) %>% 
 # fill NA values in TA column
  tidyr::fill(TA, .direction = "down") %>%
  # Prepend 'TA/' to numbers
  mutate(TA = if_else(str_detect(TA, 
                                 "^[0-9]+$"), 
                      paste0("TA/",
                             TA), 
                      TA)) %>% 
  # Remove all non-alphanumeric characters
  mutate(TA = str_replace_all(TA, 
                              "[^[:alnum:] /]", 
                              "")) %>% 
  # Make all text lowercase
  mutate(TA = tolower(TA)) %>% 
  # Remove white space
  mutate(TA = str_replace_all(TA, " ", ""))  %>%

  # Separate 'TA' column into 'area' and 'depth' columns
  separate(TA, into = c("area", 
                        "depth"), 
           sep = "/",
           remove = FALSE,
           convert = TRUE) %>%

  # Convert 'depth' to numeric
  mutate(depth = as.numeric(depth)) %>%

  # Create depth ranges
  mutate(depth = case_when(
             depth == 3 ~ "0-30 cm",
             depth == 34 ~ "30-40 cm",
             depth == 45 ~ "40-50 cm",
             depth == 56 ~ "50-60 cm",
             depth == 67 ~ "60-70 cm",
             depth == 72 ~ "70-120 cm",
             depth == 125 ~ "120-150 cm",
             depth == 158 ~ "150-180 cm",
             depth == 189 ~ "180-190 cm",
             depth == 190 ~ "190-200 cm",
             depth == 201 ~ "200-210 cm",
             depth == 212 ~ "210-220 cm",
             depth == 223 ~ "220-230 cm",
             depth == 234 ~ "230-240 cm",
             depth == 235 ~ "230-250 cm",
             depth == 256 ~ "250-260 cm",
             depth == 267 ~ "260-270 cm",
             depth == 72 ~ "70-120 cm",
             depth == 1 ~ "0-10 cm",
             depth == 2 ~ "0-20 cm", 
             depth == 13 ~ "10-30 cm",
             depth == 19 ~ "10-90 cm",
             depth == 23 ~ "20-30 cm",
             depth == 34 ~ "30-40 cm",
             depth == 45 ~ "40-50 cm",
             depth == 51 ~ "50-100 cm",
             depth == 56 ~ "50-60 cm",
             depth == 57 ~ "50-70 cm",
             depth == 89 ~ "80-90 cm",
             depth == 91 ~ "90-100 cm",
             depth == 101 ~ "100-110 cm",
             depth == 112 ~ "110-120 cm",
             depth == 123 ~ "120-130 cm",
             depth == 124 ~ "120-140 cm",
             depth == 23 ~ "20-30 cm",
             depth == 46 ~ "40-60 cm",
             depth == 78 ~ "70-80 cm",
             depth == 235 ~ "230-250 cm",
             depth == 12 ~ "10-20 cm",
             depth == 90 ~ "90-100 cm",

             TRUE ~ as.character(depth)
  )) %>%

  separate(depth, 
           into = c("upper",
                    "lower"),
           sep = "-",
           remove = FALSE,
           convert = TRUE) %>% 
  mutate(lower = parse_number(lower)) %>% 
  rowwise() %>% 
  mutate(midpoint = mean(c(upper, lower))) %>%

# create 'layer' column
  mutate(layer = case_when(
      area == "tp1" & depth == "0-10 cm" ~ "c",
      area == "tp1" & depth == "0-20 cm" ~ "c",
      area == "tp1" & depth == "10-30 cm" ~ "c",
      area == "tp1" & depth == "10-90 cm" ~ "c",
    area == "tp1" & depth == "20-30 cm" ~ "c",
    area == "tp1" & depth == "30-40 cm" ~ "c",
    area == "tp1" & depth == "40-50 cm" ~ "c",
    area == "tp1" & depth == "50-100 cm" ~ "c",
    area == "tp1" & depth == "50-60 cm" ~ "c",
    area == "tp1" & depth == "50-70 cm" ~ "c",
    area == "tp1" & depth == "80-90 cm" ~ "c",
    area == "tp1" & depth == "90-100 cm" ~ "c",
    area == "tp1" & depth == "100-110 cm" ~ "c",
    area == "tp1" & depth == "110-120 cm" ~ "c",
    area == "tp1" & depth == "120-130 cm" ~ "c",
    area == "tp1" & depth == "120-140 cm" ~ "c",
    area == "d1" & depth == "0-20 cm" ~ "c",
    area == "d1" & depth == "20-30 cm" ~ "c",
    area == "d1" & depth == "40-60 cm" ~ "a",
    area == "d1" & depth == "50-60 cm" ~ "a",
    area == "d1" & depth == "60-70 cm" ~ "a",
    area == "d1" & depth == "70-80 cm" ~ "a",
    area == "ta" & depth == "0-30 cm" ~ "e",
    area == "ta" & depth == "30-40 cm" ~ "e",
    area == "ta" & depth == "40-50 cm" ~ "e or d",
    area == "ta" & depth == "50-60 cm" ~ "e or d",
    area == "ta" & depth == "60-70 cm" ~ "e or d",
    area == "ta" & depth == "70-120 cm" ~ "e or d",
    area == "ta" & depth == "120-150 cm" ~ "d",
    area == "ta" & depth == "150-180 cm" ~ "d",
    area == "ta" & depth == "180-190 cm" ~ "d",
    area == "ta" & depth == "190-200 cm" ~ "d",
    area == "ta" & depth == "200-210 cm" ~ "d",
    area == "ta" & depth == "210-220 cm" ~ "d",
    area == "ta" & depth == "220-230 cm" ~ "d",
    area == "ta" & depth == "230-240 cm" ~ "d",
    area == "ta" & depth == "230-250 cm" ~ "d",
    area == "ta" & depth == "250-260 cm" ~ "d",
    area == "ta" & depth == "260-270 cm" ~ "d",

      TRUE ~ NA_character_
  ))
# Remove specific columns
tl1 %>% 
  select(-c("AREA", 
            "DEPTH (cm)", 
            "Raw Material", 
            "Colour/Grain/Opacity", 
            "Pattern"))

```

```{r}
#exploring Data

tl1 %>% 
  group_by(midpoint) %>% 
  tally() %>% 
  drop_na(midpoint) %>% 
ggplot() + 
  aes(midpoint, n) +
  geom_col()
```

```{r}
#exploring Data
# barplot of lithic frequency based on layers

tl1 %>% 
  group_by(layer) %>% 
  tally() %>% 
  drop_na(layer) %>% 
ggplot() + 
  aes(layer, n) +
  geom_col()
```

```{r}
#exploring Data

tl2 <- tl1 %>%
  mutate(
    Blank = case_when(
      Breakage %in% c("0", "1", NA_character_) ~ Blank,  # If Breakage is 0, 1, or NA, keep Blank unchanged
      TRUE ~ Breakage  # If Breakage is other than 0, 1, or NA, transfer its value to Blank
    )
  )

```

```{r}
# Cleaning Blank column
library(dplyr)
library(stringr)
library(tidyr)

tl2 <- tl2 %>%
  mutate(
    Blank = str_to_lower(Blank),
    Blank = str_trim(Blank),
    Blank = str_replace_all(Blank, "[./]", "-"),
    Blank = str_replace_all(Blank, "\\s*-\\s*", "-")
  ) %>%
  mutate(Blank = case_when(
             Blank == "peo" ~ "bladelet-pro",
             Blank == "pro" ~ "bladelet-pro",
             Blank == "bladelet-dis" ~ "bladelet-dis",
             Blank == "microblade" ~ "bladelet-complete",
             Blank == "microblade-pro" ~ "bladelet-pro",
             Blank == "microblade-med" ~ "bladelet-med",
             Blank == "micrpblade-dis" ~ "bladelet-dis",
             Blank == "dis" ~ "bladelet-dis",
             Blank == "bladelrt-med" ~ "bladelet-med",
             Blank == "bladelet-mes" ~ "bladelet-med",
             Blank == "bladelet-,ed" ~ "bladelet-med",
             Blank == "bladelet-bladelet-med" ~ "bladelet-med",
             Blank == "medial" ~ "bladelet-med",
             Blank == "bladelet-bladelet-pro" ~ "bladelet-pro",
             Blank == "bladelert-pro" ~ "bladelet-pro",
             Blank == "microblade-medial" ~ "bladelet-med",
             Blank == "nicroblade-pro" ~ "bladelet-pro",
             Blank == "microblade-?" ~ "bladelet-complete",
             Blank == "med" ~ "bladelet-med",
             Blank == "bladeler-pro" ~ "bladelet-pro",
             Blank == "microblade-dis" ~ "bladelet-dis",
             Blank == "microblde-dis" ~ "bladelet-dis",
             Blank == "bladlet-pro" ~ "bladelet-pro",
             Blank == "indistinct" ~ "bladelet-med",
             Blank == "flake" ~ "flake-complete",
             Blank == "dladelet-pro" ~ "bladelet-pro",
             Blank == "dladelet-pro" ~ "bladelet-pro",
             Blank == "bldelet-bladelet-pro" ~ "bladelet-pro",
             Blank == "bldelet-pro" ~ "bladelet-pro",
             Blank == "nicroblade-pro" ~ "bladelet-pro",
             Blank == "microblade-?" ~ "bladelet-med",
             Blank == "bladelt-pro" ~ "bladelet-pro",
             Blank == "bladeket-med" ~ "bladelet-med",
             Blank == "nicroblade-pro" ~ "bladelet-pro",
             Blank == "blaelet" ~ "bladelet-complete",
             Blank == "bladele-med" ~ "bladelet-med",
             Blank == "bladelet-nearly complete" ~ "bladelet-complete",
             Blank == "blade-nearly complete" ~ "blade-complete",
             Blank == "bladelet-?" ~ "bladelet-complete",
             Blank == "flake?" ~ "flake-complete",
             Blank == "micrpblade-dis" ~ "bladelet-dis",
             Blank == "blaelet-med" ~ "bladelet-med",
             Blank == "blaedlet-pro" ~ "bladelet-pro",  
             Blank == "bladelt-pro" ~ "bladelet-pro",   
             Blank == "bladelet-nearly complete" ~ "bladelet-complete", 
             Blank == "microblade-med" ~ "bladelet-med", 
             Blank == "bldelet-bladelet-pro" ~ "bladelet-pro",
             Blank == "bldelet-flake" ~ "blade-complete",
             Blank == "mwdial" ~ "bladelet-med",
             Blank == "blade" ~ "blade-complete",
             Blank == "bladelt" ~ "bladelet-complete",
             Blank == "bladlet" ~ "bladelet-complete",
             Blank == "bladlet-med" ~ "bladelet-med",
             Blank == "bladelt-med" ~ "bladelet-med",
             Blank == "flaje" ~ "flake",
             Blank == "flke" ~ "flake",
             Blank == "bladelet" ~ "bladelet-complete",
             Blank == "NA" ~ "bladelet-",
             Blank == "thick flake" ~ "flake",
             Blank == "flke-mid" ~ "flake-med",
             Blank == "flake-blade" ~ "flake",
             Blank == "angular flake?" ~ "flake",
             Blank == "flake-prox" ~ "flake-pro",
             Blank == "fklae-dis" ~ "flake-dis",
             Blank == "NA" ~ NA_character_,
    TRUE ~ Blank
  )) %>%
  mutate(Blank = na_if(Blank, ""))

unique(tl2$Blank)

tl3 <- tl2 %>%
  separate(Blank, 
           into = c("Blank2", "Blank_part"),
           sep = "-",
           remove = FALSE,
           convert = TRUE)

na_rows <- which(is.na(tl2$Blank))
print(na_rows)

zero_rows <- which(tl2$Blank == "0")
print(zero_rows)

one_rows <- which(tl2$Blank == "1")
print(one_rows)

print(unique
      (tl3$Blank2))

```

```{r}

tl3 <- tl3 %>%
  mutate(
    Blank2 = case_when(
      Blank2 == "flake" ~ "flake",  # Keep "flake" as it is
      (`Core Typology` %in% c("0", "NA", "na", "-", ".") | 
       is.na(`Core Typology`) | 
       `Core Typology` == "") & `Width (mm)` < 12 ~ "bladelet",
      (`Core Typology` %in% c("0", "NA", "na", "-", ".") | 
       is.na(`Core Typology`) | 
       `Core Typology` == "") & `Width (mm)` >= 12 ~ "blade",
      TRUE ~ Blank2
    )
  )

```

```{r}
tl3 <- tl3 %>%
  mutate(
    Blank2 = case_when(
      Blank2 %in% c("blade", "bladelet", "flake") ~ Blank2,
      TRUE ~ NA_character_
    )
  )

```

```{r}
# Remove rows 
tl3 <- tl3[!(rownames(tl3) %in% c("1030", "644", "543")), ]
```

```{r}
tl4 <- tl3 %>%
  mutate(
    Utilization = if_else(Retouch == "1", NA_character_, as.character(Utilization))
  )

```

```{r}
library(dplyr)
library(stringr)

tl4 <- tl4 %>%
  mutate_all(~ ifelse(is.na(.x) | .x == "", NA, .x)) %>% 
  mutate_all(~ str_replace(., "(?<=\\D)\\*(?=\\D)|\\s+|[*?!]+", "")) %>%
  mutate(
    SHINE = str_trim(SHINE) %>% 
            str_to_title(),
    shine2 = SHINE
  ) %>%
  mutate(
    shine2 = case_when(
      SHINE %in% c("Subparallel", "Parallel", "Sub-Parallel", "Parallel-Subprarallel",
                   "Scaled", "Sub-Paralel", "Sub-Parallell", "Sub-Paallel",
                   "Sub-Paralell", "Sub-Prallel", "L:semiparallel/R: Scaled",
                   "Semiparallel", "Semi-Parallel", "Scaled/Sub-Parallel",
                   "Seb-Parallel", "Semi-Abrupt") ~ NA_character_,
      TRUE ~ SHINE
    ),
    Utilization = if_else(Utilization == "0", NA_character_, Utilization),
    Retouch = if_else(Retouch == "0", NA_character_, Retouch),
    SHINE = if_else(SHINE == "0", NA_character_, SHINE),
    shine2 = if_else(shine2 == "0", NA_character_, shine2),
    Blank2 = if_else(Blank2 == "0", NA_character_, Blank2)
  ) %>% 
  mutate(
    SHINE = na_if(SHINE, "")
  )

# Check
print(unique(tl4$SHINE))

```

```{r}
library(dplyr)
library(stringr)
# Convert 'NA' and 'na' strings to actual NA
tl4$`Core Typology` <- na_if(tl4$`Core Typology`, "NA")
tl4$`Core Typology` <- na_if(tl4$`Core Typology`, "na")

tl4 <- tl4 %>%
  mutate(
    `Core Typology` = str_to_lower(`Core Typology`),
    `Core Typology` = str_trim(`Core Typology`),
    `Core Typology` = str_replace_all(`Core Typology`, "^[-._/NA\\s]*$|^\\s*$", "0")
  ) %>%

  mutate(`Core Typology` = case_when(
    `Core Typology` == "flatcore" ~ "flat-pressure",
             `Core Typology` == "rejuvention" ~ "rejuvenation piece-NA",
             `Core Typology` == "pressureprymidal core" ~ "pyramid-pressure",
             `Core Typology` == "pressureflat core" ~ "flat-pressure",
             `Core Typology` == "pressurebullet core" ~ "bullet-pressure",
             `Core Typology` == "prussurebullet core" ~ "bullet-pressure",
             `Core Typology` == "prismaticcore" ~ "prismatic-percussion",
             `Core Typology` == "pyramidalpressure core" ~ "pyramid-pressure",
             `Core Typology` == "flatcore with one debitage surface/pressure" ~ "flat-pressure",
             `Core Typology` == "corefragment" ~ "core fragment-NA",

             `Core Typology` == "cilandrical/pressure" ~ "cylinder-pressure",
             `Core Typology` == "bullet/pressure" ~ "bullet-pressure",
             `Core Typology` == "pyramidal/pressure" ~ "pyramid-pressure",
             `Core Typology` == "percussion/pyramidal" ~ "pyramid-pressure",
             `Core Typology` == "flat/pressure" ~ "flat-pressure",
             `Core Typology` == "pyramidal/ percussion??" ~ "pyramid-pressure",
             `Core Typology` == "pyramidal/ peressure" ~ "pyramid-pressure",
             `Core Typology` == "pyramidal/ pressure" ~ "pyramid-pressure",
             `Core Typology` == "pressure" ~ "pyramid-pressure",
             `Core Typology` == "multidirectional/percussion" ~ "shapeless-percussion",
             `Core Typology` == "prismatic/ percussion" ~ "pyramid-pressure",
             `Core Typology` == "flatcore/ unidirectional/pressure" ~ "flat-pressure",
             `Core Typology` == "heavilyused" ~ "NA",
             `Core Typology` == "burnt" ~ "NA",
             `Core Typology` == "patinated" ~ "NA",
             `Core Typology` == "pyramidal(bullet)/pressure" ~ "bullet-pressure",
             `Core Typology` == "cylandrycal/ bidirectional pressure core" ~ "bullet-pressure",
             `Core Typology` == "pyramidal/unidirectional pressure bullet core" ~ "bullet-pressure",
             `Core Typology` == "pyramidal/unidirectional pressure  core" ~ "pyramid-pressure",
             `Core Typology` == "pyramidal/unidirectional pressure  core" ~ "pyramid-pressure",
             `Core Typology` == "bladelet" ~ "bullet-pressure",
             `Core Typology` == "lip" ~ "NA",
             `Core Typology` == "bladlet" ~ "bullet-pressure",
             `Core Typology` == "bladelet(30.56)" ~ "pyramid-pressure",
             `Core Typology` == "bladelet(28.55.6)" ~ "bullet-pressure",
             `Core Typology` == "bladelet(16.28.1)" ~ "bullet-pressure",
             `Core Typology` == "bladelet(40 5.7)" ~ "pyramid-pressure",
             `Core Typology` == "bladelet(23.2 2.3)" ~ "bullet-pressure",
             `Core Typology` == "bladelet-flakeblade(41 16)" ~ "bullet-pressure",
             `Core Typology` == "bladelet(28 6.7)" ~ "bullet-pressure",
             `Core Typology` == "bladelet(28 6.7)" ~ "bullet-pressure",
             `Core Typology` == "bladelet(3512.6/7)" ~ "pyramid-pressure",
             `Core Typology` == "bladelet(27.307.5)" ~ "bullet-pressure",
    `Core Typology` == "pressure/pyramidal" ~ "pyramid-pressure",
    `Core Typology` == "bladeletcore" ~ "bullet-pressure",
    `Core Typology` == "pyramidal? / pressure" ~ "pyramid-pressure",
    `Core Typology` == "prismatic" ~ "pyramid-pressure",

    `Core Typology` == "bullet" ~ "bullet-pressure",

    TRUE ~ `Core Typology`
  ))

# Separate 'Core Typology' into 'core-typology' and 'core-technology'
tl4 <- tl4 %>%
  separate(`Core Typology`, 
           into = c("core-typology", "core-technology"),
           sep = "-",
           remove = FALSE,
           convert = TRUE)

# Convert 'NA' and 'na' in new columns to actual NA
tl4$`core-typology` <- na_if(tl4$`core-typology`, "na")
tl4$`core-technology` <- na_if(tl4$`core-technology`, "na")

# Check 
unique(tl4$`core-typology`)
unique(tl4$`core-technology`)
```

```{r}

library(dplyr)
library(stringr)

tl4 <- tl4 %>%
  mutate(
    Typology = str_to_lower(Typology),
    Typology = str_trim(Typology),
    Typology = str_replace_all(Typology, "^[-._/NA\\s]*$|^\\s*$", "0")
  ) %>%
  mutate(Typology = case_when(

             `Typology` == "notch, truncated" ~ "tool/notch;truncated piece",

             `Typology` == "retouchpiece" ~ "tool/retouched piece",

             `Typology` == "endscraper" ~ "tool/end scraper",

             `Typology` == "notch" ~ "tool/notch",

             `Typology` == "sidescraper" ~ "tool/side scraper",

             `Typology` == "doubleside scraper" ~ "tool/double side scraper",

             `Typology` == "convergentscraper (?)" ~ "tool/convergent scraper",

             `Typology` == "doublenotch on retouched piece" ~ "tool/double notch",

             `Typology` == "retouchpieces" ~ "tool/retouched piece",

             `Typology` == "burin(?)/ notch/scraper" ~ "tool/burin-notch-scraper",

             `Typology` == "borer" ~ "tool/perforator-borer",

             `Typology` == "retouchedpiece" ~ "tool/retouched piece",

             `Typology` == "notch/side scraper" ~ "tool/notch-side scraper",

             `Typology` == "corefragment" ~ "core/core fragment",

             `Typology` == "backed" ~ "tool/backed knife",

             `Typology` == "doubleside scraper ?" ~ "tool/double side scraper",

             `Typology` == "inversedenticulate" ~ "tool/denticulate",

             `Typology` == "retouchpieces/ side scraper?" ~ "tool/side scraper",

             `Typology` == "sidescraper ?" ~ "tool/side scraper",

             `Typology` == "notch-denticulate" ~ "tool/notch-denticulate",

             `Typology` == "borer/drill" ~ "tool/perforator-drill",

             `Typology` == "retouchedpiece (side scraper?)" ~ "tool/side scraper",

             `Typology` == "retouchepiece (side scraper?)" ~ "tool/side scraper",

             `Typology` == "retouchedpiece (double side scraper)" ~ "tool/double side scraper",

             `Typology` == "retouchedpiece (double side scraper?)" ~ "tool/double side scraper",

             `Typology` == "retouchepiece" ~ "tool/retouched piece",

             `Typology` == "inversenotch" ~ "tool/notch",

             `Typology` == "retouchepiece (double side scaraper?)" ~ "tool/double side scraper",

             `Typology` == "retouchedpiece( double side scraper?)" ~ "tool/double side scraper",

             `Typology` == "notchon retouche piece" ~ "tool/notch",

             `Typology` == "retouchedpiece (souble side scraper)" ~ "tool/double side scraper",

             `Typology` == "point" ~ "tool/perforator-point",

             `Typology` == "chunk" ~ "NA/NA",

             `Typology` == "coretablet" ~ "core/core tablet",

             `Typology` == "core" ~ "core/shapeless",

             `Typology` == "denticulate" ~ "tool/denticulate",

             `Typology` == "doublenotch (haft)" ~ "tool/double notch",

             `Typology` == "retouchedpiece (fine retouch)" ~ "tool/retouched piece",

             `Typology` == "endscraper on retouched piece" ~ "tool/end scraper",

             `Typology` == "retouchedpiece (serrated scraper)" ~ "tool/serrated scraper",

             `Typology` == "truncation" ~ "tool/truncated piece",

             `Typology` == "rejuventionpiece" ~ "core/rejuvenation piece",

             `Typology` == "notch/denticulate" ~ "tool/notch-denticulate",

             `Typology` == "retoucheduse on breakage" ~ "tool/retouched piece",

             `Typology` == "inversenotch/endscraper" ~ "tool/notch-end scraper",

             `Typology` == "alternatingside scraper" ~ "tool/side scraper",

             `Typology` == "alternatedouble side scraper" ~ "tool/double side scraper",

             `Typology` == "truncation/sidescaraper" ~ "tool/truncated piece",

             `Typology` == "doublenotch" ~ "tool/double notch",

             `Typology` == "borer/awl" ~ "tool/perforator-awl",

             `Typology` == "roundscraper?" ~ "tool/round scraper",

             `Typology` == "doublenotch scraper" ~ "tool/double notch scraper",

             `Typology` == "corerejuvention (core tablet?)" ~ "core/core tablet",

             `Typology` == "truncationاریبب" ~ "tool/truncated piece",

             `Typology` == "retouchedpiece (side scraper)" ~ "tool/side scraper",

             `Typology` == "retochedpice" ~ "tool/retouched piece",

             `Typology` == "truncation/notch" ~ "tool/truncated piece-notch",

             `Typology` == "doublealternate scraper" ~ "tool/double side scraper",

             `Typology` == "awl/ inverse notch" ~ "tool/notch",

             `Typology` == "retouchedpiece (alternate scraper?)" ~ "tool/double side scraper",

             `Typology` == "inverseside scraper" ~ "tool/side scraper",

             `Typology` == "usedcore rejuvention" ~ "core/rejuvenation piece",

             `Typology` == "coreside rejuvention" ~ "core/rejuvenation piece",

             `Typology` == "coreplatform rejuvention flake" ~ "core/rejuvenation piece",

             `Typology` == "corerejuvention platform" ~ "core/rejuvenation piece",

             `Typology` == "coreside rejuvention blade" ~ "core/rejuvenation piece",

             `Typology` == "coreside rejuvention blade??" ~ "core/rejuvenation piece",

             `Typology` == "coreedge rejuvention" ~ "core/rejuvenation piece",

             `Typology` == "coreedge rejivention" ~ "core/rejuvenation piece",

             `Typology` == "retouchedpice (side scraper?)" ~ "tool/side scraper" ,

             `Typology` == "scraper" ~ "tool/side scraper",

             `Typology` == "sickleblade" ~ "tool/sickle shine",

             `Typology` == "coretool (scraper)" ~ "tool/scraper-on core piece",

             `Typology` == "retouchedpiece (scraper)" ~ "tool/side scraper",

             `Typology` == "inversedenticulate?" ~ "tool/denticulate",

             `Typology` == "retouchedpiece (scraper?)" ~ "tool/side scraper",

             `Typology` == "coreside rejuvention flake" ~ "core/rejuvenation piece",

             `Typology` == "truncation/endscraper" ~ "tool/endscraper on truncated piece",

             `Typology` == "retouchedpice(double side scraper?)" ~ "tool/double side scraper",

             `Typology` == "doublescraper" ~ "tool/double side scraper",

             `Typology` == "dendiculate" ~ "tool/denticulate",

             `Typology` == "double-scraper" ~ "tool/double side scraper",

             `Typology` == "pyramidalcore" ~ "core/pyramid",

             `Typology` == "bulletcore" ~ "core/bullet",

             `Typology` == "corerejuvention" ~ "core/rejuvenation piece",

             `Typology` == "pyrmidalcore" ~ "core/pyramid",

             `Typology` == "corepreperation" ~ "core/core preparation piece",

             `Typology` == "crested" ~ "core/crested bladelet",

             `Typology` == "preperationblade" ~ "core/core preperation",

             `Typology` == "burin" ~ "tool/burin",

             `Typology` == "serrateddenticulate" ~ "tool/serrated scraper",

             `Typology` == "saw:serrated denticulate" ~ "tool/serrated scraper",

             `Typology` == "serratedused" ~ "tool/serrated scraper",

             `Typology` == "corepreeration" ~ "tool/core preparation piece",

             `Typology` == "coreprepearation" ~ "tool/core preparation piece",

             `Typology` == "awl" ~ "tool/perforator-awl",

             `Typology` == "notch.inverse" ~ "tool/notch",

             `Typology` == "point.notch" ~ "tool/perforator-point;notch",

             `Typology` == "inversescraper" ~ "tool/side scraper",

             `Typology` == "roundendscraper" ~ "tool/round scraper",

             `Typology` == "alternate.scraper" ~ "tool/double side scraper",

             `Typology` == "convergentscraper/disc scraper" ~ "tool/scraper-convergent scraper",

             `Typology` == "serratedside scraper" ~ "tool/serrated scraper",

             `Typology` == "debri" ~ "NA/NA",

             `Typology` == "distalpart of a drill" ~ "tool/perforator-drill",

             `Typology` == "corepreperation?" ~ "core/core preparation piece",

             `Typology` == "preformend of a borer" ~ "tool/perforator-borer; preform",

             `Typology` == "used" ~ "tool/used",

             `Typology` == "distalend of a inverse sidescraper" ~ "tool/side scraper",

             `Typology` == "retouchedblade" ~ "tool/retouched piece",

             `Typology` == "drill" ~ "tool/perforator-drill",

             `Typology` == "corerejuvention tablet" ~ "core/core tablet",

             `Typology` == "bladeletcore" ~ "core/pyramid",

             `Typology` == "flake-bladecore" ~ "core/mixed",

             `Typology` == "bladelet-corerejuvention" ~ "core/rejuvenation piece",

             `Typology` == "bladelet-corefragment" ~ "core/core fragment",

             `Typology` == "pointedused" ~ "tool/perforator-point",

             `Typology` == "microburin" ~ "tool/microburin",

             `Typology` == "bladeletcore fragment" ~ "core/core fragment",

             `Typology` == "rejuventionface flake" ~ "core/rejuvenation piece",

             `Typology` == "corepreperation flake" ~ "core/core preparation piece",

             `Typology` == "corerejuvention flake"  ~ "core/rejuvenation piece",

             `Typology` == "unfinished  pyramid core" ~ "core/pyramid",

             `Typology` == "bladeletcoretablet" ~ "core/core tablet",

             `Typology` == "bladeletcore rejuvention" ~ "core/rejuvenation piece",

             `Typology` == "debitage" ~ "NA/NA",

             `Typology` == "bladeletcore  rejuvention" ~ "core/rejuvenation piece",

             `Typology` == "inverseserrated scraper" ~ "tool/serrated scraper",

             `Typology` == "backedknife" ~ "tool/backed knife",

             `Typology` == "corepreparation" ~ "core/core preparation piece",

             `Typology` == "primaryflake" ~ "core/primary flake",

             `Typology` == "pointedconvergent scraper on bladelet" ~ "tool/perforator-point",

             `Typology` == "trie" ~ "NA/NA",

             `Typology` == "part of core tablet" ~ "core/core tablet",

             `Typology` == "zaviedar" ~ "NA/NA",

             `Typology` == "convergentscraper" ~ "tool/convergent scraper",

             `Typology` == "preparationflake" ~ "core/core preparation piece",

             `Typology` == "atypicalborer" ~ "tool/perforator-borer",

             `Typology` == "naturalybacked knife" ~ "tool/backed knife-naturally backed knife",

             `Typology` == "pyramidbladelet core" ~ "core/pyramid",

             `Typology` == "flatbladelet core" ~ "core/flat",

             `Typology` == "bladecore" ~ "core/pyramid",

             `Typology` == "bulletbladelet core" ~ "core/bullet",

             `Typology` == "coerejuvention flake" ~ "core/rejuvenation piece",

             `Typology` == "finishedbladelet core" ~ "core/bullet",

             `Typology` == "failed pyramid core?" ~ "core/pyramid",

             `Typology` == "pyramidplain bladelet core" ~ "core/pyramid",

             `Typology` == "coeon flake" ~ "core/core on flake",

             `Typology` == "coeejuvention flake" ~ "core/rejuvenation piece",

             `Typology` == "bladeletcoe rejuvention tablet?" ~ "core/core tablet",

             `Typology` == "pyramidcore" ~ "core/pyramid",

             `Typology` == "bladeletcore tablet" ~ "core/core tablet",

             `Typology` == "pyramidbladeleet core" ~ "core/pyramid",

             `Typology` == "flake" ~ "NA/NA",

             `Typology` == "serrated" ~ "tool/serrated scraper",

             `Typology` == "curvedretouched piece" ~ "tool/retouched piece",

             `Typology` == "pyramidalbladelet core" ~ "core/pyramid",

             `Typology` == "multidirectionalcore fragment" ~ "core/multidirectional core fragment",

             `Typology` == "unidirectionalblade? core fragent" ~ "core/pyramid",

             `Typology` == "backed/corerejuvention" ~ "tool/backed knife",

             `Typology` == "unidirectionalbladelet core" ~ "core/pyramid",

             `Typology` == "unifacialbalde core fragment?" ~ "core/pyramid",

             `Typology` == "flakedetached from a bladelet core" ~ "core/core preparation piece",

             `Typology` == "corepreparation tablet" ~ "core/core tablet",

             `Typology` == "bladeletcore fragent" ~ "core/core fragment-pyramid",

             `Typology` == "alternatescraper??" ~ "tool/double side scraper",

             `Typology` == "endscraper.onrejuvention" ~ "core/rejuvenation piece",

             `Typology` == "inverseserrated" ~ "tool/serrated scraper",

             `Typology` == "patination.omitted" ~ "NA/NA",

             `Typology` == "trapze" ~ "tool/geometric-triangle",

             `Typology` == "trapzoid" ~ "tool/geometric-triangle",

             `Typology` == "borer.drill" ~ "tool/perforator-drill",

             `Typology` == "serraed" ~ "tool/serrated scraper",

             `Typology` == "bladeletcore on a flake" ~ "core/core on flake",

             `Typology` == "geofact" ~ "NA/NA",

             `Typology` == "denticulateborer?" ~ "tool/perforator-borer;denticulate",

             `Typology` == "sideborer-denticulate" ~ "tool/perforator-borer;denticulate",

             `Typology` == "borer.awl" ~ "tool/perforator-awl",

             `Typology` == "roundscraper" ~ "tool/round scraper",

             `Typology` == "truncationon a bladelet core" ~ "tool/truncated piece",

             `Typology` == "truncation.notch" ~ "tool/truncated piece;notch",

             `Typology` == "backed-denticulate" ~ "tool/backed;denticulate",

             `Typology` == "brokendrill" ~ "tool/perforator-drill",

             `Typology` == "drillbroken" ~ "tool/perforator-drill",

             `Typology` == "point/broken drill" ~ "tool/perforator-point",

             `Typology` == "brokendrill?" ~ "tool/perforator-drill",

             `Typology` == "partialyserrated" ~ "tool/serrated scraper",

             `Typology` == "naturallybacked" ~ "tool/backed knife-naturally backed knife",

             `Typology` == "transversescraper" ~ "tool/transverse scraper",

             `Typology` == "naturalybacked" ~ "tool/backed knife-naturally backed knife",

             `Typology` == "awl-notch" ~ "tool/perforator-awl",

             `Typology` == "psedulevalois" ~ "tool/psedulevalois flake",

             `Typology` == "alternatingdenticulate" ~ "tool/denticulate",

             `Typology` == "usedcore fragment" ~ "tool/core tool-used core fragment",

             `Typology` == "multidirectionalmicroflake core" ~ "core/multidirectional flake core",

             `Typology` == "mixedpyramidal core" ~ "core/pyramid",

             `Typology` == "usedbladecore fragment" ~ "tool/core tool-used core fragment",

             `Typology` == "multidirectionalbladelet core fragment" ~ "core/multidirectional bladelet core",

             `Typology` == "unidirectionalbladelet core.cylinder" ~ "core/prismatic",

             `Typology` == "pyrymedalmixed core" ~ "core/pyramid",

             `Typology` == "pyramedial/bullet core" ~ "core/bullet",

             `Typology` == "corefragent" ~ "core/core fragment",

             `Typology` == "pyramedialbladelet core fragment" ~ "core/pyramid",

             `Typology` == "unidirectionalbladelet core fragment" ~ "core/pyramid",

             `Typology` == "unifacialunidirectional bladelet core" ~ "core/pyramid",

             `Typology` == "unidirectionalblade core" ~ "core/pyramid",

             `Typology` == "pyramidalbladelet core fragment" ~ "core/pyramid",

             `Typology` == "corerejuvention?" ~ "core/rejuvenation piece",

             `Typology` == "flatburin? point?" ~ "tool/flat burin",

             `Typology` == "primaryblade" ~ "core/primary blade",

             `Typology` == "trihedralgeometric" ~ "tool/geometric-triangle",

             `Typology` == "lunate" ~ "tool/geometric-lunate",

             `Typology` == "corepaltform rejuvention flake" ~ "core/rejuvenation piece",

             `Typology` == "pressurebladelet core /semi flat" ~ "core/pyramid",

             `Typology` == "failedpercussion blade core/semi pyramedal" ~ "core/pyramid",

             `Typology` == "percussionbladelet core/unidirectional" ~ "core/pyramid-percussion",

             `Typology` == "percussion.unidirectional.blade-flakecore" ~ "core/mixed-percussion",

             `Typology` == "obliqueretouched bladelet" ~ "tool/retouched piece",

             `Typology` == "percussionunidirectional pyramidal  bladelet core" ~ "core/pyramid",

             `Typology` == "lunategeometric" ~ "tool/geometric-lunate",

             `Typology` == "awl on a core rejuvention" ~ "tool/perforator-awl on core rejuvenation piece",

             `Typology` == "truncated" ~ "tool/truncated piece",

             `Typology` == "multiplenotch" ~ "tool/notch",

             `Typology` == "dishedconvergent bladelet" ~ "tool/convergent scraper",

             `Typology` == "bipolarpercussion blade-bladelet core" ~ "core/bipolar-percussion",

             `Typology` == "bladeletbullet core" ~ "core/bullet",

             `Typology` == "alternatingserrated" ~ "tool/serrated scraper",

             `Typology` == "shaplesscore" ~ "core/shapeless",

             `Typology` == "corepreparation?" ~ "core/core preparation piece",

             `Typology` == "obliqueretouched" ~ "tool/retouched piece",

             `Typology` == "corepreparation flake" ~ "core/core preparation piece",

             `Typology` == "backeddenticulate" ~ "tool/backed;denticulate",

             `Typology` == "awlnotch" ~ "tool/perforator-awl;notch",

             `Typology` == "scraperburin?" ~ "tool/burin;scraper",

             `Typology` == "notchround scraper" ~ "tool/round scraper;notch",

             `Typology` == "corerejuvention tablet round scraper" ~ "tool/round scraper on core tablet",

             `Typology` == "alternatescraper" ~ "tool/double side scraper",

             `Typology` == "coreplatform preperation flake" ~ "core/rejuvenation piece",

             `Typology` == "coreplatform rejuvention tablet" ~ "core/core tablet",

             `Typology` == "brokenpyramidal core bladelet" ~ "core/pyramid",

             `Typology` == "pyramidalbladelet core/semi bullet?" ~ "core/bullet",

             `Typology` == "coreside rejuvention flake?" ~ "core/rejuvenation piece-side",

             `Typology` == "unidirectionalpressure bladelet core/unifacila" ~ "core/pyramid",

             `Typology` == "coreplatform rejuvention tablet/or multidirectional core" ~ "core/rejuvenation piece",

             `Typology` == "bulletbladlet core" ~ "core/bullet",

             `Typology` == "cylinderbipolar bladelet core" ~ "core/prismatic",

             `Typology` == "backedknife/core platform rejuvenation" ~ "tool/backed knife",

             `Typology` == "flatunifacial bladelet core" ~ "core/flat",

             `Typology` == "coreprepration flake" ~ "core/core preparation piece",

             `Typology` == "coreplatform preperation" ~ "core/core preparation piece-platform",

             `Typology` == "borerpreform" ~ "tool/perforator-borer",

             `Typology` == "rejuvention" ~ "core/rejuvenation piece",

             `Typology` == "borerpreform?" ~ "tool/perforator-borer",

             `Typology` == "inversnotch" ~ "tool/notch",

             `Typology` == "alternateconvergent" ~ "tool/convergent scraper",

             `Typology` == "unidirectionalbladelet/flake-blade core" ~ "core/mixed",

             `Typology` == "unidirectionalmixed core" ~ "core/mixed",

             `Typology` == "unidirectionalflat bladelet core" ~ "core/flat",

             `Typology` == "" ~ "NA-NA",
             `Typology` == "na" ~ "NA-NA",

    TRUE ~ `Typology`
  ))

tl4 <- tl4 %>%
  mutate(
    typology_split = str_split(Typology, "/", simplify = TRUE),
    `typology-tool-core` = typology_split[, 1],
    tool_or_core = ifelse(str_starts(`typology-tool-core`, "tool"), "tool", "core"),
    `tool-typology` = ifelse(tool_or_core == "tool", typology_split[, 2], NA),
    `core-fragment` = ifelse(tool_or_core == "core", typology_split[, 2], NA)
  ) %>%
  # remove temporary columns
  select(-typology_split, -tool_or_core)

unique(tl4$`tool-typology`)
unique(tl4$`core-fragment`)
```

```{r}
# Exploring datra

tl4 %>% 
  group_by(midpoint) %>% 
  tally() %>% 
  drop_na(midpoint) %>% 
ggplot() + 
  aes(midpoint, n) +
  geom_col()

```

```{r}
# First convert all columns to character type (cant remember why I did that, )
tl4 <- tl4 %>%
  mutate(across(everything(), as.character))

# Then replace 'NA', 'na', or '0' with R's NA
tl4 <- tl4 %>%
  mutate(across(everything(), ~na_if(., "NA"))) %>%
  mutate(across(everything(), ~na_if(., "na"))) %>%
  mutate(across(everything(), ~na_if(., "0")))

# Show the updated data frame
print(tl4)

```

```{r}
#refine the table

library(dplyr)

tl_final <- tl4 %>%

  rename(
    blank = Blank2,
    `sickle shine` = shine2,
    length = `Length (mm)` ,
    width = `Width (mm)`,
    thickness = `Thickness (mm)`, 
    typology = Typology,
    utilization = Utilization,
    retouch = Retouch,
    blank = Blank2, 
    `blank part` = `Blank_part`,
    `bulb thickness` = `Bulb Thickness (mm)`, 
    `platform thickness` = `Platform Thickness (mm)`, 
    `platform length` = `platform Length (mm)`, 
    eraillure = Eraillure , 
    cortex = `Cortex (%)`)

```

```{r}
tl_final <- tl_final  %>%

 select(lithic_id, 
         TA, 
         area, 
         layer,
        cortex,
         depth, upper, lower, midpoint, 
         `length`, `width`, `thickness`, 
         typology, 
        utilization, 
        `sickle shine`,
        retouch, 
        `typology-tool-core`,
        `tool-typology`,
         blank, `blank part`,

         `core-typology`, `core-fragment`,`core-technology`,

        cortex)
```

```{r}
# Update the 'typology-tool-core' column based on the condition in 'Retouch'
tl_final <- tl_final %>% 
  mutate(
    `typology-tool-core` = case_when(
      retouch == 1 ~ "tool",           
      utilization == 1 ~ "tool",        
      TRUE ~ `typology-tool-core`      
    )
  )

print(tl_final)

```

```{r}
tl_final <- tl_final %>% 
  mutate(
    `tool-typology` = case_when(
      utilization == 1 ~ "utilized tool",  
      TRUE ~ `tool-typology`               
    )
  )

print(tl_final)

```

```{r}
tl_final <- tl_final %>% 
  mutate(
    `tool-typology` = case_when(

     # If Retouch is 1 and tool-typology is NA
       retouch == 1 & is.na(`tool-typology`) ~ "retouched piece",

      # For all other cases 
      TRUE ~ `tool-typology`                                       
    )
  )

# Show updated data frame
print(tl_final)

```

```{r}

# Creating tooltypegroup based on the given criteria

tl_final <- tl_final %>%
  mutate(
    tooltypegroup = case_when(
      `tool-typology` %in% 
        c("end scraper", "side scraper", "double side scraper", "convergent scraper", "round scraper", "perforator-convergent scraper", "perforator-borer; preform", "scraper-on core piece", "endscraper on truncated piece", "round scraper", "transverse scraper", "core tool-used core fragment", "round scraper on core tablet") ~ "scraper",
      `tool-typology` %in% 
        c("perforator-borer", "perforator-drill", "perforator-point", "perforator-awl", "perforator-point;notch", "perforator-awl on core rejuvenation piece", "perforator-awl;notch", "perforator-borer;denticulate") ~ "Perforator",
      `tool-typology` %in% 
        c("backed knife", "backed knife-naturally backed knife", "backed;denticulate") ~ "backed pieces",
      `tool-typology` %in% c("double notch", "denticulate", " denticulate", "notch-denticulate", "notch") ~ "denticulate-notch",
      `tool-typology` %in% c("notch-side scraper", "notch-end scraper", "double notch scraper", "round scraper;notch") ~ "scraper-notch",
      `tool-typology` %in% c("geometric-triangle", "geometric-lunate") ~ "geometric",
      `tool-typology` == "serrated scraper" ~ "serrated scraper",
      `tool-typology` %in% c("truncated piece", "truncated piece-notch", "truncated piece;notch", "notch;truncated piece") ~ "truncated pieces",

      # Represents NA group
      `tool-typology` %in% c("core preparation piece", "used", "psedulevalois flake") ~ NA_character_, 
      `tool-typology` %in% c("burin", "flat burin", "burin;scraper", "burin-notch-scraper") ~ "burin",
      `tool-typology` == "microburin" ~ "microburin",
            `tool-typology` == "retouched piece" ~ "retouched piece",
      `tool-typology` == "utilized tool" ~ "utilized tool",
      `tool-typology` == "sickle shine" ~ "sickle shine",

      #classify any remaining values as NA
      TRUE ~ NA_character_ 
    )
  )

# Checking the unique values of tooltypegroup
unique(tl_final$tooltypegroup)

```

```{r}
tl_final <- tl_final %>%
  mutate(
    `core-fragment` = if_else(
      !is.na(`core-typology`),  # If core-typology is not NA
      NA_character_,            # Set core-fragment to NA
      `core-fragment`           # Otherwise, keep the original value
    )
  )

# Show updated data frame
print(tl_final)

```

```{r}

library(dplyr)
library(stringr)

tl_final <- tl_final %>%
  mutate(
    `core-fragment` = if_else(
      !is.na(`core-typology`),
      NA_character_,
      `core-fragment`

  ) 
  ) %>%

  mutate(`core-fragment` = case_when(
    `core-fragment` == "rejuvenation piece" ~ "rejuvenation piece",

             `core-fragment` == "rejuvention piece-side" ~ "rejuvenation piece",

             `core-fragment` == "core preparation piece-platform" ~ "core tablet",

             `core-fragment` == "core preparation" ~ "rejuvenation piece",

             `core-fragment` == "core preparation piece" ~ "rejuvenation piece",

             `core-fragment` == "core preperation" ~ "rejuvenation piece",

             `core-fragment` == "pyramid" ~ "pyramid",

             `core-fragment` == "pyramid core" ~ "pyramid",

             `core-fragment` == "pyramid-unidirectional" ~ "pyramid",

             `core-fragment` == "core fragment-pyramid" ~ "pyramid",

             `core-fragment` == "cylinder" ~ "cylinder/prismatic",

             `core-fragment` == "prismatic" ~ "cylinder/prismatic",

             `core-fragment` == "shapeless" ~ "cylinder/prismatic",

             `core-fragment` == "multidirectional core fragment" ~ "shapeless",

             `core-fragment` == "multidirectional flake core" ~ "shapeless",

             `core-fragment` == "multidirectional bladelet core" ~ "shapeless",

             `core-fragment` == "mixed core" ~ "shapeless",

             `core-fragment` == "mixed-percussion" ~ "shapeless",

             `core-fragment` == "mixed" ~ "shapeless",

             `core-fragment` == ("bipolar-percussion") ~ "bipolar",

             `core-fragment` == "flat" ~ "flat",

             `core-fragment` == "core on flake" ~ "core on flake",

             `core-fragment` == "bullet" ~ "bullet",

    TRUE ~ `core-fragment`

  ))

```

```{r}

tl_final <- tl_final %>%
  mutate(
    core = case_when(
      `core-fragment` %in% c("pyramid", "pyramid core", "pyramid-unidirectional", "core fragment-pyramid", "cylinder/prismatic", "shapeless", "multidirectional core fragment", "multidirectional flake core", "multidirectional bladelet core", "mixed core", "mixed-percussion", "mixed", "bipolar-percussion", "flat", "core on flake", "bullet") ~ `core-fragment`,
      TRUE ~ NA_character_
    ),
    corefrag = case_when(
      `core-fragment` %in% c("rejuvenation piece", "rejuvention piece-side", "core preparation piece-platform", "core preparation", "core preparation piece", "core preperation", "crested bladelet" , "primary flake" , "primary blade" ) ~ `core-fragment`,
      TRUE ~ NA_character_
    )
  )

unique(`tl_final`$`core`)
unique(`tl_final`$`corefrag`)
```

```{r}

library(dplyr)

tl_final <- tl_final %>%

  mutate(`core-fragment` = if_else(`core-fragment` == "", NA_character_, `core-fragment`)) %>%
  # Convert "pyramid-percussion" to "pyramid" in `core-fragment` column
  mutate(`core-fragment` = 
           if_else(
             `core-fragment` == "pyramid-percussion", "pyramid", `core-fragment`)) %>%
  # Update existing `core` column with selected values from `core-fragment` column
  mutate(core = if_else(`core-fragment` %in% 
                          c("pyramid", 
                            "cylinder/prismatic", 
                            "flat", 
                            "bullet", 
                            "core on flake", 
                            "pyramid"), 
                        `core-fragment`, 
                        core))

```

```{r}

library(dplyr)

tl_final <- tl_final %>%
  # Add 'core tablet' and 'core fragment' from `core-fragment` to `corefrag`
  mutate(corefrag = case_when(
    `core-fragment` == "core tablet" ~ "core tablet",
    `core-fragment` == "core fragment" ~ "core fragment",
    TRUE ~ corefrag
  ))

```

```{r}

library(dplyr)

tl_final <- tl_final %>%
  # Transfer 'rejuvenation piece' and 'core fragment' from `core-typology` to `corefrag`
  mutate(corefrag = case_when(
    `core-typology` == "rejuvenation piece" ~ "rejuvenation piece",
    `core-typology` == "core fragment" ~ "core fragment",
    TRUE ~ corefrag
  )) %>%
  # Remove 'rejuvenation piece' and 'core fragment' from `core-typology` 
  mutate(`core-typology` = if_else(`core-typology` %in% c("rejuvenation piece", "core fragment"), NA_character_, `core-typology`))

unique(tl_final$`core-typology`)
unique(tl_final$`corefrag`)

```

```{r}

# Update the 'blank' column based on conditions in the 'core-fragment' column

library(dplyr)

tl_final <- tl_final %>%
  mutate(blank = if_else(`core-fragment` %in% c(
    "core fragment", "core tablet", "rejuvenation piece", "pyramid", "crested bladelet", 
    "shapeless", "primary flake", "flat", "bullet", "core on flake", "cylinder/prismatic",
    "primary blade", "rejuvenation piece-side", "core platform rejuvention"),
    NA_character_, 
    blank))

print(tl_final)

```

Techno-Typo_exploration

```{r}
tl_final_area <- tl_final %>%
  filter(area %in% c("ta", "d1", "tp1"))

```

```{r}
library(dplyr)

# add al for regression plot area-layer

tl_final_area <- tl_final_area %>%
  mutate(al = paste(area, layer, sep = "_"))
```

```{r}
# Prepare data for Summary table-area
summary_table <- tl_final_area %>% 
  filter(area %in% c("ta", "d1", "tp1")) %>%
  group_by(area) %>% 
  summarise(
    blade_count = sum(blank == "blade", na.rm = TRUE),
    flake_count = sum(blank == "flake", na.rm = TRUE),
    bladelet_count = sum(blank == "bladelet", na.rm = TRUE),

    tool_count = sum(retouch == 1, na.rm = TRUE) + sum(utilization == 1, na.rm = TRUE),

    core_count = sum(!is.na(`core-typology`), na.rm = TRUE),
    core_rejuvenation_count = sum(!is.na(`corefrag`), na.rm = TRUE),
    retouched_tool_count = sum(retouch == 1, na.rm = TRUE),
    utilized_tool_count = sum(utilization == 1, na.rm = TRUE),

    # Total count of rows for each group
    total = n()  
  )

summary_table <- summary_table %>% 
  add_row(

    blade_count = sum(summary_table$blade_count, na.rm = TRUE),
    flake_count = sum(summary_table$flake_count, na.rm = TRUE),
    bladelet_count = sum(summary_table$bladelet_count, na.rm = TRUE),
    tool_count = sum(summary_table$tool_count, na.rm = TRUE),
    core_count = sum(summary_table$core_count, na.rm = TRUE),
    core_rejuvenation_count = sum(summary_table$core_rejuvenation_count, na.rm = TRUE),
    retouched_tool_count = sum(summary_table$retouched_tool_count, na.rm = TRUE),
    utilized_tool_count = sum(summary_table$utilized_tool_count, na.rm = TRUE),
    total = sum(
      summary_table$total, 
      na.rm = TRUE)
  )

```

```{r}
#Generate summary table-area

library(knitr)
library(dplyr)
# Rename columns
summary_table <- summary_table %>%
  rename(
    "Area" = area,
    "Blade" = blade_count,
    "Flake" = flake_count,
    "Bladelet" = bladelet_count,
    "Retouched Tools" = retouched_tool_count,
    "Utilized Tools" = utilized_tool_count,
    "Cores" = core_count,
    "Core rejuvenations pieces" = core_rejuvenation_count,
    "Total" = total
  ) %>%
  mutate(
    Area = if_else(row_number() == n() & is.na(Area), "Total", Area)
  ) %>%
  mutate(
    Area = recode(Area, "ta" = "TA", "d1" = "D1", "tp1" = "TP1", "Total" = "Total")
  )
library(kableExtra)

kable(summary_table, 
      caption = "Table 1: Count of debitage type and tools in three different main areas.", 
      align = 'c') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 12) %>%

  row_spec(0, font_size = 11)

```

```{r}
# Create the summary table based on layer
summary_table1_layer <- tl_final_area %>%
  group_by(layer) %>%
  summarise(
    numb_retouched = sum(retouch == 1, na.rm = TRUE),
    numb_cores = sum(!is.na(`core-typology`), na.rm = TRUE),
    total_lithics = n(),
    percent_retouch = round((numb_retouched / total_lithics) * 100, 2)
  )

```

```{r}

# Define the volumes for each area 
volumes <- data.frame(
  area = c("ta", "tp1", "d1"),
  volume = c(52, 5.76, 2.1375) # in m^3
)

# Count the number of artifacts by area
artifact_counts <- tl_final_area %>%
  filter(area %in% c("ta", "tp1", "d1")) %>%
  group_by(area) %>%
  tally()

# Join artifact_counts with volumes by area
lithic_density_df <- left_join(artifact_counts, volumes, by = "area")

# Compute lithic density for each area 
lithic_density_df <- lithic_density_df %>%
  mutate(lithic_density = round(n / volume, 2))

# Create the summary table
summary_table2_area <- tl_final_area %>%
  filter(area %in% c("ta", "tp1", "d1")) %>%
  group_by(area) %>%
  summarise(
    numb_retouched = sum(retouch == 1, na.rm = TRUE),
    total_lithics = n(),
    percent_retouch = round((numb_retouched / total_lithics) * 100, 2)
  )

# Merge the density data with the summary table
summary_table2_area <- left_join(summary_table2_area, lithic_density_df[, c("area", "lithic_density")], by = "area") %>%
  mutate(area = recode(area, 'ta' = 'TA', 'tp1' = 'TP1', 'd1' = 'D1'),
         lithic_volumetric_density = lithic_density)

# Remove the 'lithic_density' column
summary_table2_area <- summary_table2_area %>% 
  select(-lithic_density)

print(summary_table2_area)

```

```{r}
# layer-Create the summary table
summary_table2_layer <- tl_final_area %>%
  filter(layer %in% c(c("a", "b", "c" , "d", "e"))) %>%  
  group_by(layer) %>%
  summarise(
    numb_retouched = sum(retouch == 1, na.rm = TRUE),
    numb_cores = sum(!is.na(`core-typology`), na.rm = TRUE),
    total_lithics = n(),
    percent_retouch = round((numb_retouched / total_lithics) * 100, 2)
  )

# Create a data frame for volumetric density values
density_data_layer <- data.frame(
  layer = c("a", "b", "c" , "d", "e"),
  `lithic_volumetric_density` = round(c(1.575, 6.12, 2.115, 47, 13), 2)
)

summary_table2_layer <- merge(summary_table2_layer, density_data_layer, by = "layer")

```

```{r}
# generate table 2 - area

library(kableExtra)

kable(summary_table2_area, 
      caption = "Table 2. Lithic counts for each of the areas in Tula'i", 
      align = 'c',
      col.names = c("Area", "Number of Retouched Lithics", "Total Number of Lithics", "Retouch frequency (%)", "Lithic Volumetric Density(pcs/(m^3)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 11) %>%
  column_spec(1, width = "1cm") %>% 
  column_spec(2, width = "1cm") %>%

  column_spec(3, width = "1cm") %>% 
  column_spec(4, width = "1cm") %>% 
  column_spec(5, width = "1cm")

```

```{r}

library(ggplot2)
library(dplyr)
library(tidyr)

# Set the value for n
n <- 100

# Tally by area
tl_final_area_tally <- tl_final_area %>%
  group_by(area) %>%
  tally()

# Filter areas with more than n rows
tl_final_area_tally_greater_than_n <- tl_final_area_tally %>%
  filter(n >= n)

# Filter data based on areas with more than n rows
tl_final_area_filtered <- tl_final_area %>%
  filter(area %in% tl_final_area_tally_greater_than_n$area) %>%
  filter(!is.na(blank) & blank != "")

# Select and mutate dimensions
tl_final_area_selected <- tl_final_area_filtered %>%
  select(area, width, thickness, length) %>%
  mutate(
    width = as.numeric(width),
    thickness = as.numeric(thickness),
    length = as.numeric(length)
  )

# Pivot data to longer format
tl_final_area_long <- tl_final_area_selected %>%
  pivot_longer(cols = c(width, thickness, length), 
               names_to = "Dimension", 
               values_to = "Value")

# Create the box plot
ggplot(tl_final_area_long, aes(x = area, y = Value)) +
  geom_boxplot() +
  facet_wrap(~Dimension, scales = "free") +
  labs(x = "Area", y = "Value") +
  theme_minimal()

```

```{r}

# Dimension By Area-boxplot

library(ggplot2)
library(dplyr)
library(tidyr)
library(ggbeeswarm)

# Create boxplot
ggplot(tl_final_area_long, 
       aes(x = area, y = Value)) +
  geom_boxplot() +
  geom_quasirandom(alpha = 0.01) +
  facet_wrap(~Dimension, scales = "free_y") +
  scale_y_log10() +
  theme_minimal()

```

```{r}

#Dimension-layer-Boxplot for each Dimension"

n <- 100

# Tally count of each layer
tl_final_layer_tally <- tl_final_area %>% 
  group_by(layer) %>% 
  tally()

# Filter layers with more than n rows
tl_final_layer_tally_greater_than_n <- tl_final_layer_tally %>% 
  filter(n >= n)

# Filter data for layers with more than n rows
tl_final_layer_filtered <- tl_final_area %>% 
  filter(layer %in% tl_final_layer_tally_greater_than_n$layer)

# Select and mutate
tl_final_layer_selected <- tl_final_layer_filtered %>%
  select(layer, 
         `width`, 
         `thickness`, 
         `length`) %>%
  mutate(across(c
                (
           `width`, 
           `thickness`, 
           `length`), 
           as.numeric)
         )

# Pivot longer
tl_final_layer_long <- tl_final_layer_selected %>%
  pivot_longer(cols = 
                 c(width, 
                   thickness, 
                   length), 
               names_to = 
                 "Dimension", 
               values_to = 
                 "Value") %>%
  drop_na(Value)

# Create box plot
ggplot(tl_final_layer_long, 
       aes(layer, 
           Value)) +

  geom_boxplot() +
  facet_wrap(~Dimension, 
             scales = "free") +
  labs(x = "Layer", 
       y = "Value") +

   theme_minimal()

```

```{r}

#Dimension By Layer

# Create boxplot
ggplot(tl_final_layer_long, 
       aes(x = layer, y = Value)) +
  geom_boxplot() +
  geom_quasirandom(alpha = 0.01) +
  facet_wrap(~Dimension, scales = "free_y") +
  scale_y_log10() +
  theme_minimal()

```

```{r}

#Width of Artifacts (Area A, D1, TP1)"

tl_final_area %>%
  filter(area %in% 
           c("ta", "d1", "tp1")) %>%
  mutate(width = as.numeric(width)) %>%  # Explicitly convert to numeric

   select(area, 
          `width`) %>%
  ggplot() +

  aes(x = width) +  # Make sure the aesthetic mapping is correct

  geom_histogram(binwidth = 2) +

  labs(x = "Width of Debitage (mm)", 
       y = "Frequency" ) +

  facet_wrap(~area, 
             ncol = 1) +

  theme_minimal()

```

```{r}

#Width of Artifact_layer (a, c, d, e or d)"

tl_final_area %>%
  filter(layer %in% c("a", 
                      "c", 
                      "e or d", 
                      "e", 
                      "d")) %>%
  mutate(width = as.numeric(width)) %>%  
  select(layer, width) %>%
  ggplot() +
  aes(x = width) +
  geom_histogram(binwidth = 2) +
  labs(x = "Width of Debitage (mm)", y = "Frequency") +
  facet_wrap(~layer, ncol = 1) +
  theme_minimal()

```

```{r}
# Thickness of Artifacts (Layers A, C, E or D)"

tl_final_area %>%
 filter(layer %in% c("a", 
                     "c", 
                     "e or d", 
                     "d", 
                     "e")) %>%
 mutate(thickness = as.numeric(thickness)) %>% 
  select(layer, 
         thickness) %>%
  ggplot() +
  aes(x = thickness) +
  geom_histogram(binwidth = 2) +
  labs(x = "Thickness of Debitage (mm)", 
       y = "Frequency" )+
  facet_wrap(~layer , ncol = 1) +
  theme_minimal()

```

```{r}

#Length of Artifacts (Area A, D1, TP1)"

tl_final_area %>%
  filter(`blank part` == "complete", area %in% c("ta", "tp1", "d1")) %>%
  mutate(length = as.numeric(length)) %>%
  select(area, 
         length) %>%
  ggplot() +
  aes(x = length) +
  geom_histogram(binwidth = 2) +
  labs(x = "Length of Debitage (mm)", 
       y = "Frequency") +
  facet_wrap(~area, ncol = 1) +
  theme_minimal()

```

Should fix the error 
```{r}

#Length of Artefact based_layers

tl_final_area %>% 
  filter(layer %in% c("layer a" , "layer d" , "layer e" , "layer c" , "layer e or d")) %>% 
  mutate(length = as.numeric(length)) %>% 
  select(layer, length) %>% 
  ggplot(aes(x = length)) + 
  geom_histogram(binwidth = 2) +
  labs(x = "Length of Debitage (mm)" ,
       y = "Frequency") + 
  facet_wrap(~layer, ncol = 1) + 
  theme_minimal()

```

1.   (formal + informal) to debitages
2.  formal  vs. informal
3.  shine/ sickle  as a  used for processing plants

```{r}

library(dplyr)
library(tidyr)

# Create a summary table
summary_table_tool_area <- tl_final_area %>%
  filter(!is.na(tooltypegroup), tooltypegroup != "") %>%  
  group_by(area, tooltypegroup) %>%  
  tally() %>%

  arrange(area, 
          tooltypegroup) %>%  
  spread(key = area, 
         value = n, 
         fill = 0)

# Add a "Total" column that sums each row
summary_table_tool_area <- summary_table_tool_area %>%
  mutate(Total = rowSums(select(., -tooltypegroup)))

# Create a row that contains the column totals, and append it to the table
total_row <- summary_table_tool_area %>%
  select(-tooltypegroup) %>%
  summarise(across(everything(), sum, na.rm = TRUE)) %>%
  mutate(tooltypegroup = "Total")

# Combine the total_row with the original table
summary_table_tool_area <- bind_rows(summary_table_tool_area, total_row)

print(summary_table_tool_area)

```

```{r}
#plot_blank_type

library(dplyr)
library(ggplot2)
library(ggbeeswarm)

# Data filtering and plotting
tl_final_area %>% 
  filter(blank %in% c('bladelet', 'blade', 'flake')) %>%
  ggplot(aes(x = area, 
             fill = blank)) +
  geom_bar(position = "dodge") +
  labs(
    x = "Area",
    y = "Number",
    fill = "Blank Type"
  ) +
  theme_minimal()
```

WORK ON THIS PART LATER:

```{r}

                                                #DO NOT RUN#

tl_final_area %>%
  filter(blank %in% c("bladelet", "blade", "flake")) %>%
  select(blank, `sickle shine`) %>%
  ggplot() +
  aes(x = `sickle shine`) +
  geom_bar() +
  labs(x = "shine", y = "Frequency") +
  theme_minimal()

```

```{r}

                                         #DO NOT RUN#
tl_final_area %>%
  filter(blank %in% c("bladelet", "blade", "flake")) %>%
  select(blank, retouch = 1) %>%
  ggplot() +
  aes(x = retouch) +
  geom_bar() +
  labs(x = "Retouch", 
       y = "Frequency" )+
  theme_minimal()
```

```{r}

                                       #DO NOT RUN#

tl_final_area %>%
  filter(blank %in% c("bladelet", "blade", "flake")) %>%
  select(blank, utilization = 1) %>%
  ggplot() +
  aes(x = utilization) +
  geom_bar() +
  labs(x = "Utilization", 
       y = "Frequency" )+
  theme_minimal()

```

```{r}

                                       #DO NOT RUN#

library(ggplot2)
library(patchwork)

plot_shine <- tl4_tool_Blank %>%
  filter(Blank2 %in% c("bladelet", 
                       "blade", 
                       "flake")) %>%
  select(Blank2, shine2 = 1) %>%
  ggplot() +
  aes(x = shine2) +
  geom_bar() +
  labs(x = "shine", y = "Frequency") +
  theme_minimal()

plot_retouch <- tl4_tool_Blank %>%
  filter(Blank2 %in% c("bladelet", 
                       "blade", 
                       "flake")) %>%
  select(Blank2, 
         Retouch = 1) %>%
  ggplot() +
  aes(x = Retouch) +
  geom_bar() +
  labs(x = "Retouch", 
       y = "Frequency") +
  theme_minimal()

plot_utilization <- tl4_tool_Blank %>%
  filter(Blank2 %in% c("bladelet", "blade", "flake")) %>%
  select(Blank2, Utilization = 1) %>%
  ggplot() +
  aes(x = Utilization) +
  geom_bar() +
  labs(x = "Utilization", y = "Frequency") +
  theme_minimal()

# Combine the plots using patchwork
plot_combined <- plot_shine + plot_retouch + plot_utilization
plot_combined

```

------------

WABI_AREA

```{r}
# regression plot-area
library(ggplot2)
library(ggpmisc)

# Plotting
p <- ggplot(summary_table2_area, 
            aes(x = lithic_volumetric_density, 
                y = percent_retouch)) +

  # regression line 
  geom_smooth(method = "lm", 
              se = TRUE, 
              aes(group = 1), 
              color = "blue") +

  # Points 
  geom_point(color = "black") +

  # Labels points with area names
  geom_text(aes(label = area),
            vjust = -1, 
            hjust = -0.5) +

  # Regression equation, R-squared, and p-value
  stat_poly_eq(aes(label = paste(stat(eq.label), stat(rr.label), stat(p.value.label), sep = "~~~")),
               formula = y ~ x,
               parse = TRUE,
               size = 3) +

  # Titles and labels
  labs(title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
       x = "Lithic Volumetric Density",
       y = "Retouch Frequency (%)")

print(p)

```

### 2023-09-01 tulai-fin.qmd
Added tulai-fin.qmd and removed Tula'i-fin.qmd

--- 2023-09-01 Tula'i-fin.qmd
@@
 ---
+
 ## Introduction
@@
 Should fix the error
+
 ```{r}
@@
 
-------------
+------------------------------------------------------------------------
 

### 2023-09-05 Tulai.qmd
aa

--- 2023-09-01 Tula'i-fin.qmd
@@
 title: "Tulai"
-author: "Setareh Shafizadeh, Ben Marwick"
+author: "S.SH"
 date: now
@@
 ---
-## Introduction
-
-Cleaning Data:
 
@@
 ```
+
+Cleaning Data:
 
@@
            convert = TRUE) %>%
-
   # Convert 'depth' to numeric
@@
     area == "ta" & depth == "30-40 cm" ~ "e",
-    area == "ta" & depth == "40-50 cm" ~ "e or d",
-    area == "ta" & depth == "50-60 cm" ~ "e or d",
-    area == "ta" & depth == "60-70 cm" ~ "e or d",
-    area == "ta" & depth == "70-120 cm" ~ "e or d",
+    area == "ta" & depth == "40-50 cm" ~ "e",
+    area == "ta" & depth == "50-60 cm" ~ "e",
+    area == "ta" & depth == "60-70 cm" ~ "e",
+    area == "ta" & depth == "70-120 cm" ~ "d",
     area == "ta" & depth == "120-150 cm" ~ "d",
@@
              Blank == "fklae-dis" ~ "flake-dis",
+             Blank == "bladelete" ~ "bladelet",
              Blank == "NA" ~ NA_character_,
@@
 ```{r}
-# Remove rows 
+# Remove rows based on row names
 tl3 <- tl3[!(rownames(tl3) %in% c("1030", "644", "543")), ]
@@
 
-# Check
+# Check unique values in the modified columns
 print(unique(tl4$SHINE))
@@
 
-# Check 
+# Check the unique values again
 unique(tl4$`core-typology`)
@@
 ```{r}
-# First convert all columns to character type (cant remember why I did that, )
+# First convert all columns to character type
 tl4 <- tl4 %>%
@@
 tl_final <- tl_final %>%
+  mutate(al = paste0(toupper(area), "_", layer)) %>%
+  relocate(al, .after = layer)
+
+```
+
+```{r}
+tl_final <- tl_final  %>%
 
@@
          layer,
+        al,
         cortex,
@@
       utilization == 1 ~ "tool",
+      `sickle shine` == 1 ~ "tool",
+
       TRUE ~ `typology-tool-core`      
@@
 ```{r}
+```
+
+```{r}
+
 tl_final <- tl_final %>% 
@@
       utilization == 1 ~ "utilized tool",
+      `sickle shine` == 1 ~ "sickle shine",
       TRUE ~ `tool-typology`               
@@
        retouch == 1 & is.na(`tool-typology`) ~ "retouched piece",
+       `sickle shine` == 1 & is.na(`tool-typology`) ~ "sickle shine",
 
@@
 # Show updated data frame
-print(tl_final)
+unique(tl_final$`tool-typology`)
 
@@
 
+                   #DO NOT RUN#############
+
+library(dplyr)
+library(stringr)
+
+tl_final <- tl_final %>%
+  mutate(
+    `core-fragments` = str_to_lower(`core-fragment`),
+    `core-fragments` = str_trim(`core-fragment`),
+    `core-fragments` = str_replace_all(`core-fragment`, "^[-._/NA\\s]*$|^\\s*$", "0")
+  ) %>%
+  mutate(`core-fragment` = case_when(
+
+            `core-fragment` == "rejuvenation piece" ~ "rejuvenation piece"
+
+            `core-fragment` == "crested bladelet" ~ "crested bladelet",
+
+            `core-fragment` == "crested bladelet" ~ "crested bladelet",
+
+            `core-fragment` == "primaryflake" ~ "primary flake",
+            `core-fragment` == "primary flake" ~ "primary flake",
+
+            `core-fragment` == "primary blade" ~ "primary blade",
+
+            `core-fragment` == "rejuvenationpiece-side" ~ "rejuvenation piece-side",
+            `core-fragment` == "rejuvention piece-side" ~ "rejuvenation piece-side",
+
+            `core-fragment` == "core preparation piece-platform" ~ "core tablet",
+
+            `core-fragment` == "core preparation" ~ "rejuvenation piece",
+
+            `core-fragment` == "core preparation piece" ~ "rejuvenation piece",
+
+            `core-fragment` == "core preperation" ~ "rejuvenation piece",
+
+            `core-fragment` == "pyramid" ~ "pyramid",
+
+            `core-fragment` == "pyramid core" ~ "pyramid",
+
+            `core-fragment` == "pyramid-unidirectional" ~ "pyramid",
+
+            `core-fragment` == "core fragment-pyramid" ~ "pyramid",
+
+            `core-fragment` == "cylinder" ~ "cylinder/prismatic",
+
+            `core-fragment` == "prismatic" ~ "cylinder/prismatic",
+
+            `core-fragment` == "shapeless" ~ "cylinder/prismatic",
+
+            `core-fragment` == "multidirectional core fragment" ~ "shapeless",
+            `core-fragment` == "multidirectionalcore fragment" ~ "shapeless",
+            `core-fragment` == "multidirectionalbladelet core" ~ "shapeless",
+
+            `core-fragment` == "multidirectional flake core" ~ "shapeless",
+
+            `core-fragment` == "multidirectional bladelet core" ~ "shapeless",
+
+            `core-fragment` == "mixed core" ~ "shapeless",
+
+            `core-fragment` == "mixed-percussion" ~ "shapeless",
+
+            `core-fragment` == "mixed" ~ "shapeless",
+
+            `core-fragment` == ("bipolar-percussion") ~ "cylinder/prismatic-percussion",
+
+            `core-fragment` == "flat" ~ "flat",
+
+            `core-fragment` == "core on flake" ~ "core on flake",
+
+            `core-fragment` == "bullet" ~ "bullet",
+
+            `core-fragment` == "core platform rejuvention" ~ "core tablet",
+            `core-fragment` == "bipolar" ~ "cylinder/prismatic",
+             `core-fragment` == "coretablet" ~ "core tablet",
+            `core-fragment` == "crestedbladelet" ~ "crested piece",
+            `core-fragment` == "coreon flake" ~ "core on flake",
+            `core-fragment` == "corefragment-pyramid" ~ "pyramid",
+            `core-fragment` == "corepreparation piece-platform" ~ "core tablet",
+
+    TRUE ~ `core-fragment`
+  ))
+
+unique(`tl_final`$`core-fragment`)
+```
+
+```{r}
+
 tl_final <- tl_final %>%
@@
 tl_final <- tl_final %>%
-
+  # Convert empty strings to NA in `core-fragment` column
   mutate(`core-fragment` = if_else(`core-fragment` == "", NA_character_, `core-fragment`)) %>%
@@
 
+unique(tl_final$`core-fragment`)
+unique(tl_final$`corefrag`)
+unique(tl_final$`core`)
 ```
@@
 
+# Process the tl_final_area dataframe using pipe (%>%)
 tl_final <- tl_final %>%
@@
   )) %>%
-  # Remove 'rejuvenation piece' and 'core fragment' from `core-typology` 
+  # Remove 'rejuvenation piece' and 'core fragment' from `core-typology` but keep the column
   mutate(`core-typology` = if_else(`core-typology` %in% c("rejuvenation piece", "core fragment"), NA_character_, `core-typology`))
@@
 unique(tl_final$`corefrag`)
-
-```
-
-```{r}
+unique(tl_final$`core-fragment`)
+
+```
+
+```{r}
+
+library(dplyr)
 
 # Update the 'blank' column based on conditions in the 'core-fragment' column
-
-library(dplyr)
-
 tl_final <- tl_final %>%
@@
 
+# Show the updated data
 print(tl_final)
-
-```
-
-Techno-Typo_exploration
+unique(tl_final$`core-fragment`)
+unique(tl_final$`blank`)
+
+```
+
+Summary tables: 
+1.AREA
+```{r}
+# Create a summary table
+summary_table_area <- tl_final %>%
+  group_by(area) %>%
+  summarise(
+    `Cores (n)` = sum(!is.na(core) & core != ""),
+    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag) & corefrag != ""),
+    `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
+    `Flake (n)` = sum(blank == "flake", na.rm = TRUE),
+    `Bladelet (n)` = sum(blank == "bladelet", na.rm = TRUE),
+    `Retouched Tools (n)` = sum(retouch == 1, na.rm = TRUE),
+    `Utilized (n)` = sum(utilization == 1, na.rm = TRUE),
+    `Sickle Shine (n)` = sum(`sickle shine` == 1, na.rm = TRUE)
+  )
+
+# Show summary table
+print(summary_table_area)
+
+```
+
+```{r}
+library(kableExtra)
+
+kable(summary_table_area, 
+      caption = "Table: Summary by Area", 
+      align = 'c') %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
+                full_width = F, 
+                font_size = 12) %>%
+  row_spec(0, bold = TRUE, font_size = 14)
+
+```
+
+2.LAYER
+```{r}
+# Create a summary table
+summary_table_layer <- tl_final %>%
+  group_by(layer) %>%  # Note that 'area' is replaced with 'layer'
+  summarise(
+    `Cores (n)` = sum(!is.na(core)),  # Count non-NA core entries
+    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag)),  # Count non-NA corefrag entries
+    `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
+    `Flake (n)` = sum(blank == "flake", na.rm = TRUE),
+    `Bladelet (n)` = sum(blank == "bladelet", na.rm = TRUE),
+    `Retouched Tools (n)` = sum(retouch == 1, na.rm = TRUE),
+    `Utilized (n)` = sum(utilization == 1, na.rm = TRUE),
+    `Sickle Shine (n)` = sum(`sickle shine` == 1, na.rm = TRUE)
+  )
+
+```
+
+```{r}
+# Generate the table
+kable(summary_table_layer, 
+      caption = "Table: Summary by Layer", 
+      align = 'c') %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
+                full_width = F, 
+                font_size = 12) %>%
+  row_spec(0, bold = TRUE, font_size = 14)
+
+```
+
+AREA_LAYER:
+```{r}
+# Create a summary table based on the 'al' column
+summary_table_al <- tl_final %>%
+  group_by(al) %>%  
+  summarise(
+    `Cores (n)` = sum(!is.na(core)),  # Count non-NA core entries
+    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag)),  # Count non-NA corefrag entries
+    `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
+    `Flake (n)` = sum(blank == "flake", na.rm = TRUE),
+    `Bladelet (n)` = sum(blank == "bladelet", na.rm = TRUE),
+    `Retouched Tools (n)` = sum(retouch == 1, na.rm = TRUE),
+    `Utilized (n)` = sum(utilization == 1, na.rm = TRUE),
+    `Sickle Shine (n)` = sum(`sickle shine` == 1, na.rm = TRUE)
+  )
+
+```
+
+```{r}
+# Generate the table
+library(kableExtra)
+
+kable(summary_table_al, 
+      caption = "Table: Summary by Al", 
+      align = 'c') %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
+                full_width = F, 
+                font_size = 12) %>%
+  row_spec(0, bold = TRUE, font_size = 14)
+
+```
+
+DEPTH:
+
+```{r}
+tl_final <- tl_final %>%
+  mutate(depth = gsub("cm", "", depth))
+```
+
+```{r}
+# Create a summary table based on the 'depth' column
+summary_table_depth <- tl_final %>%
+  group_by(depth) %>%  
+  summarise(
+    `Cores (n)` = sum(!is.na(core)),  # Count non-NA core entries
+    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag)),  # Count non-NA corefrag entries
+    `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
+    `Flake (n)` = sum(blank == "flake", na.rm = TRUE),
+    `Bladelet (n)` = sum(blank == "bladelet", na.rm = TRUE),
+    `Retouched Tools (n)` = sum(retouch == 1, na.rm = TRUE),
+    `Utilized (n)` = sum(utilization == 1, na.rm = TRUE),
+    `Sickle Shine (n)` = sum(`sickle shine` == 1, na.rm = TRUE)
+  )
+
+```
+
+```{r}
+# Generate the table
+library(kableExtra)
+
+kable(summary_table_depth, 
+      caption = "Table: Summary by Depth", 
+      align = 'c') %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
+                full_width = F, 
+                font_size = 12) %>%
+  row_spec(0, bold = TRUE, font_size = 14)
+
+```
+
+```{r}
+# Create a summary table based on the 'depth', 'area', and 'layer' columns
+summary_table_depth_area_layer <- tl_final %>%
+  group_by(depth, area, layer, upper, lower) %>%  
+  summarise(
+    `Cores (n)` = sum(!is.na(core)),  # Count non-NA core entries
+    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag)),  # Count non-NA corefrag entries
+    `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
+    `Flake (n)` = sum(blank == "flake", na.rm = TRUE),
+    `Bladelet (n)` = sum(blank == "bladelet", na.rm = TRUE),
+    `Retouched Tools (n)` = sum(retouch == 1, na.rm = TRUE),
+    `Utilized (n)` = sum(utilization == 1, na.rm = TRUE),
+    `Sickle Shine (n)` = sum(`sickle shine` == 1, na.rm = TRUE)
+  )
+
+# Generate the table
+library(kableExtra)
+
+kable(summary_table_depth_area_layer, 
+      caption = "Table: Summary by Depth, Area, and Layer", 
+      align = 'c') %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
+                full_width = F, 
+                font_size = 12) %>%
+  row_spec(0, bold = TRUE, font_size = 14)
+
+```
+
+Techno-Typo:
 
@@
 ```{r}
+# Load dplyr library
 library(dplyr)
 
-# add al for regression plot area-layer
-
-tl_final_area <- tl_final_area %>%
-  mutate(al = paste(area, layer, sep = "_"))
-```
-
-```{r}
 # Prepare data for Summary table-area
-summary_table <- tl_final_area %>% 
+summary_table_area <- tl_final_area %>% 
   filter(area %in% c("ta", "d1", "tp1")) %>%
@@
     bladelet_count = sum(blank == "bladelet", na.rm = TRUE),
-
     tool_count = sum(retouch == 1, na.rm = TRUE) + sum(utilization == 1, na.rm = TRUE),
-
     core_count = sum(!is.na(`core-typology`), na.rm = TRUE),
@@
     utilized_tool_count = sum(utilization == 1, na.rm = TRUE),
-
-    # Total count of rows for each group
+    shine_tool_count = sum(`sickle shine` == 1, na.rm = TRUE),
     total = n()
@@
 
-summary_table <- summary_table %>% 
+# Adding a new row with total counts
+# Adding a new row with total counts
+summary_table_area <- summary_table_area %>% 
   add_row(
-
-    blade_count = sum(summary_table$blade_count, na.rm = TRUE),
-    flake_count = sum(summary_table$flake_count, na.rm = TRUE),
-    bladelet_count = sum(summary_table$bladelet_count, na.rm = TRUE),
-    tool_count = sum(summary_table$tool_count, na.rm = TRUE),
-    core_count = sum(summary_table$core_count, na.rm = TRUE),
-    core_rejuvenation_count = sum(summary_table$core_rejuvenation_count, na.rm = TRUE),
-    retouched_tool_count = sum(summary_table$retouched_tool_count, na.rm = TRUE),
-    utilized_tool_count = sum(summary_table$utilized_tool_count, na.rm = TRUE),
-    total = sum(
-      summary_table$total, 
-      na.rm = TRUE)
-  )
+    area = "Total",
+    blade_count = sum(summary_table_area$blade_count, na.rm = TRUE),
+    flake_count = sum(summary_table_area$flake_count, na.rm = TRUE),
+    bladelet_count = sum(summary_table_area$bladelet_count, na.rm = TRUE),
+    tool_count = sum(summary_table_area$tool_count, na.rm = TRUE),
+    core_count = sum(summary_table_area$core_count, na.rm = TRUE),
+    core_rejuvenation_count = sum(summary_table_area$core_rejuvenation_count, na.rm = TRUE),
+    retouched_tool_count = sum(summary_table_area$retouched_tool_count, na.rm = TRUE),
+    utilized_tool_count = sum(summary_table_area$utilized_tool_count, na.rm = TRUE),
+    shine_tool_count = sum(summary_table_area$shine_tool_count, na.rm = TRUE),
+    total = sum(summary_table_area$total, na.rm = TRUE)  # Explicitly setting the total
+  )
+
+```
+
+```{r}
+# Prepare data for Summary table-al
+summary_table_al <- tl_final_area %>% 
+  group_by(al) %>% 
+  summarise(
+    blade_count = sum(blank == "blade", na.rm = TRUE),
+    flake_count = sum(blank == "flake", na.rm = TRUE),
+    bladelet_count = sum(blank == "bladelet", na.rm = TRUE),
+    tool_count = sum(retouch == 1, na.rm = TRUE) + sum(utilization == 1, na.rm = TRUE),
+    core_count = sum(!is.na(`core-typology`), na.rm = TRUE),
+    core_rejuvenation_count = sum(!is.na(`corefrag`), na.rm = TRUE),
+    retouched_tool_count = sum(retouch == 1, na.rm = TRUE),
+    utilized_tool_count = sum(utilization == 1, na.rm = TRUE),
+    shine_tool_count = sum(`sickle shine` == 1, na.rm = TRUE),
+    total = n()
+  )
+
+# Adding a new row with total counts
+summary_table_al <- summary_table_al %>% 
+  add_row(
+    al = "Total",
+    blade_count = sum(summary_table_al$blade_count, na.rm = TRUE),
+    flake_count = sum(summary_table_al$flake_count, na.rm = TRUE),
+    bladelet_count = sum(summary_table_al$bladelet_count, na.rm = TRUE),
+    tool_count = sum(summary_table_al$tool_count, na.rm = TRUE),
+    core_count = sum(summary_table_al$core_count, na.rm = TRUE),
+    core_rejuvenation_count = sum(summary_table_al$core_rejuvenation_count, na.rm = TRUE),
+    retouched_tool_count = sum(summary_table_al$retouched_tool_count, na.rm = TRUE),
+    utilized_tool_count = sum(summary_table_al$utilized_tool_count, na.rm = TRUE),
+    shine_tool_count = sum(summary_table_al$shine_tool_count, na.rm = TRUE),
+    total = sum(summary_table_al$total, na.rm = TRUE)  # Explicitly setting the total
+  )
+
+# View the summary table
+print(summary_table_al)
+
+```
+
+```{r}
+summary_table_al <- summary_table_al %>%
+  filter(!(al %in% c("D1_NA", "TP1_NA")))
+
+library(kableExtra)
+
+kable(summary_table_al, 
+      col.names = c("Area-Layer", "Blade", "Flake", "Bladelet", "Tools", "Cores", "Core Rejuvenations", "Retouched Tools", "Utilized Tools", "Sickle Shine", "Total"), 
+      caption = "Table: Summary ", 
+      align = 'c') %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
+                full_width = FALSE, 
+                font_size = 12) %>%
+  row_spec(0, bold = TRUE, font_size = 14)
 
@@
 # Rename columns
-summary_table <- summary_table %>%
+summary_table_area <- summary_table_area %>%
   rename(
@@
     "Retouched Tools" = retouched_tool_count,
+    "Tools" = tool_count,
     "Utilized Tools" = utilized_tool_count,
+    "Sickle Shine" = shine_tool_count,
     "Cores" = core_count,
@@
 
-kable(summary_table, 
+kable(summary_table_area, 
       caption = "Table 1: Count of debitage type and tools in three different main areas.", 
@@
   row_spec(0, font_size = 11)
+
+```
+
+```{r}
+# summary table-core and core rejuvenation
+```
+
+```{r}
+# summary table for tools: retouched-utelized-sickle
+
+```
+
+```{r}
+
+library(dplyr)
+library(tidyr)
+
+# Create a summary table
+summary_table_tool_area <- tl_final_area %>%
+  filter(!is.na(tooltypegroup), tooltypegroup != "") %>%  # Exclude rows where 'tooltypegroup' is NA or empty
+  group_by(area, tooltypegroup) %>%  # Group the data by area and tooltypegroup
+  tally() %>%  # Count the number of rows in each group
+  arrange(area, tooltypegroup) %>%  # Arrange by area and tooltypegroup
+  spread(key = area, value = n, fill = 0)  # Spread the 'area' column into multiple columns
+
+# Add a "Total" column that sums each row
+summary_table_tool_area <- summary_table_tool_area %>%
+  mutate(Total = rowSums(select(., -tooltypegroup)))
+
+# Create a row that contains the column totals, and append it to the table
+total_row <- summary_table_tool_area %>%
+  select(-tooltypegroup) %>%
+  summarise(across(everything(), sum, na.rm = TRUE)) %>%
+  mutate(tooltypegroup = "Total")
+
+# Combine the total_row with the original table
+summary_table_tool_area <- bind_rows(summary_table_tool_area, total_row)
+
+# View the summary table
+print(summary_table_tool_area)
+
+```
+
+```{r}
+
+```
+
+```{r}
+library(kableExtra)
+
+kable(summary_table_tool_area, 
+      col.names = c("Tools", "D1", "TA", "TP1", "Total"),
+      caption = "Table: Tool Typology in three different main areas.", 
+      align = 'c') %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
+                full_width = F, 
+                font_size = 12) %>%
+  column_spec(1, extra_css = "text-align:left;") %>%  # Make the first column left-aligned
+  row_spec(0, bold = TRUE, font_size = 14)
 
@@
 
-```{r}
-
-# Define the volumes for each area 
+volume for each area-layer: 
+Ta-
+
+```{r}
+# Define the volumes for each area based on the given information
 volumes <- data.frame(
@@
 
-# Compute lithic density for each area 
+# Compute lithic density for each area and round it to 2 decimal places
 lithic_density_df <- lithic_density_df %>%
-  mutate(lithic_density = round(n / volume, 2))
+  mutate(lithic_density = round(n / volume, 2)) # number of artifacts / volume
 
@@
 
+# Print out the resulting data frame
 print(summary_table2_area)
@@
 ```{r}
+
+                        UniqueDO NOT RUN
+
+# Define the volumes for each area based on the given information
+volumes <- data.frame(
+  area = c("ta", "tp1", "d1"),
+  volume = c(52, 5.76, 2.1375) # in m^3
+)
+
+# Count the number of artifacts by area
+artifact_counts <- tl_final_area %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  tally()
+
+# Calculate retouch frequency with explicit NA check
+retouch_freq <- tl_final_area %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  summarise(retouchFreq = ifelse(all(is.na(retouch)), NA, mean(!is.na(retouch))), .groups = 'drop')
+
+# Compute lithic density for each area and round it to 2 decimal places
+lithic_density_df <- left_join(artifact_counts, volumes, by = "area") %>%
+  mutate(lithic_density = round(n / volume, 2))
+
+# Create the summary table
+summary_table2_area <- tl_final_area %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  summarise(
+    numb_retouched = sum(retouch == 1, na.rm = TRUE),
+    total_lithics = n(),
+    percent_retouch = round((numb_retouched / total_lithics) * 100, 2)
+  )
+
+# Merge all the calculated data
+summary_table2_area <- left_join(summary_table2_area, lithic_density_df[, c("area", "lithic_density")], by = "area")
+summary_table2_area <- left_join(summary_table2_area, retouch_freq, by = "area")
+
+# Final modifications and print
+summary_table2_area <- summary_table2_area %>%
+  mutate(
+    area = recode(area, 'ta' = 'TA', 'tp1' = 'TP1', 'd1' = 'D1'),
+    lithic_volumetric_density = lithic_density
+  ) %>%
+  select(-lithic_density)  # Remove the unwanted column
+
+print(summary_table2_area)
+
+```
+
+```{r}
+# generate table 2 - area
+
+library(kableExtra)
+
+kable(summary_table2_area, 
+      caption = "Table 2. Lithic counts for each of the areas in Tula'i", 
+      align = 'c',
+      col.names = c("Area", "Number of Retouched Lithics", "Total Number of Lithics", "Retouch frequency (%)", "Lithic Volumetric Density(pcs/(m^3)")) %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 11) %>%
+  column_spec(1, width = "1cm") %>% 
+  column_spec(2, width = "1cm") %>%
+
+  column_spec(3, width = "1cm") %>% 
+  column_spec(4, width = "1cm") %>% 
+  column_spec(5, width = "1cm")
+
+```
+
+```{r}
+
+                                                      ##### DO NOT RUN ######
+
 # layer-Create the summary table
@@
 
-```{r}
-# generate table 2 - area
-
-library(kableExtra)
-
-kable(summary_table2_area, 
-      caption = "Table 2. Lithic counts for each of the areas in Tula'i", 
-      align = 'c',
-      col.names = c("Area", "Number of Retouched Lithics", "Total Number of Lithics", "Retouch frequency (%)", "Lithic Volumetric Density(pcs/(m^3)")) %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 11) %>%
-  column_spec(1, width = "1cm") %>% 
-  column_spec(2, width = "1cm") %>%
-
-  column_spec(3, width = "1cm") %>% 
-  column_spec(4, width = "1cm") %>% 
-  column_spec(5, width = "1cm")
-
-```
-
-```{r}
+consistency in dimention:
+```{r}
+#|label: fig_2
+#|fig-cap: "Boxplot for each Dimention"
+#|fig-height: 5
 
@@
 ```{r}
-
-# Dimension By Area-boxplot
+#|label: fig_3
+#|fig-cap: Dimention By Area
+#|ffig-height: 5
 
@@
 
-#Dimension-layer-Boxplot for each Dimension"
+```
+
+```{r}
+#|label: fig_2-layer
+#|fig-cap: "Boxplot for each Dimention"
+#|fig-height: 5
 
@@
 ```{r}
-
-#Dimension By Layer
+#|label: fig_3-layer
+#|fig-cap: Dimension By Layer
+#|fig-height: 5
 
@@
 ```{r}
-
-#Width of Artifacts (Area A, D1, TP1)"
+#|label: fig_4-width
+#|fig-cap: "Width of Artifacts (Area A)"
+#| fig-height: 5
+tl_final_area %>%
+ filter(area == "ta") %>%
+  select(area, 
+         width) %>%
+  ggplot() +
+  aes(x = width) +
+  geom_histogram(binwidth = 2) +
+  labs(x = "Width of Debitage (mm)", 
+       y = "Frequency" )+
+  theme_minimal()
+```
+
+```{r}
+#|label: fig_5-width
+#|fig-cap: "Width of Artifacts (Area A, D1, TP1)"
+#|fig-height: 5
 
@@
 ```{r}
-
-#Width of Artifact_layer (a, c, d, e or d)"
+#|label: fig_5-width-layer
+#|fig-cap: "Width of Artifacts (Area A, D1, TP1)"
+#|fig-height: 5
 
@@
                       "d")) %>%
-  mutate(width = as.numeric(width)) %>%  
+  mutate(width = as.numeric(width)) %>%  # Convert to numeric
   select(layer, width) %>%
@@
 ```{r}
-# Thickness of Artifacts (Layers A, C, E or D)"
+
+```
+
+```{r}
+#|label: fig_6-thickness-layer
+#|fig-cap: "Thickness of Artifacts (Layers A, C, E or D)"
+#|fig-height: 5
 
@@
 ```{r}
-
-#Length of Artifacts (Area A, D1, TP1)"
-
-tl_final_area %>%
-  filter(`blank part` == "complete", area %in% c("ta", "tp1", "d1")) %>%
-  mutate(length = as.numeric(length)) %>%
+#|label: fig_7
+#|fig-cap: "Length of Artifacts (Area A, D1, TP1)"
+#|fig-height: 5
+
+tl3 %>%
+  filter(Blank_part == "complete", area %in% c("ta", "tp1", "d1")) %>%
+  mutate(`Length (mm)` = as.numeric(`Length (mm)`)) %>%
   select(area, 
-         length) %>%
+         `Length (mm)`) %>%
   ggplot() +
-  aes(x = length) +
+  aes(x = `Length (mm)`) +
   geom_histogram(binwidth = 2) +
@@
 
-Should fix the error 
-```{r}
-
-#Length of Artefact based_layers
-
-tl_final_area %>% 
+```{r}
+#| label: fig-7-length-layer
+#| fig-cap: Length of Artefact based on layers
+#| fig-height: 5
+tl3 %>% 
   filter(layer %in% c("layer a" , "layer d" , "layer e" , "layer c" , "layer e or d")) %>% 
-  mutate(length = as.numeric(length)) %>% 
-  select(layer, length) %>% 
-  ggplot(aes(x = length)) + 
+  mutate(`Length (mm)` = as.numeric(`Length (mm)`)) %>% 
+  select(layer, `Length (mm)`) %>% 
+  ggplot(aes(x = `Length (mm)`)) + 
   geom_histogram(binwidth = 2) +
@@
 ```{r}
-
-library(dplyr)
-library(tidyr)
-
-# Create a summary table
-summary_table_tool_area <- tl_final_area %>%
-  filter(!is.na(tooltypegroup), tooltypegroup != "") %>%  
-  group_by(area, tooltypegroup) %>%  
-  tally() %>%
-
-  arrange(area, 
-          tooltypegroup) %>%  
-  spread(key = area, 
-         value = n, 
-         fill = 0)
-
-# Add a "Total" column that sums each row
-summary_table_tool_area <- summary_table_tool_area %>%
-  mutate(Total = rowSums(select(., -tooltypegroup)))
-
-# Create a row that contains the column totals, and append it to the table
-total_row <- summary_table_tool_area %>%
-  select(-tooltypegroup) %>%
-  summarise(across(everything(), sum, na.rm = TRUE)) %>%
-  mutate(tooltypegroup = "Total")
-
-# Combine the total_row with the original table
-summary_table_tool_area <- bind_rows(summary_table_tool_area, total_row)
-
-print(summary_table_tool_area)
-
-```
-
-```{r}
-#plot_blank_type
-
-library(dplyr)
+#|label: NA 
+#|fig-cap: "tool typology"
+#|fig-height: 5
+
+```
+
+```{r}
 library(ggplot2)
-library(ggbeeswarm)
-
-# Data filtering and plotting
-tl_final_area %>% 
-  filter(blank %in% c('bladelet', 'blade', 'flake')) %>%
-  ggplot(aes(x = area, 
-             fill = blank)) +
-  geom_bar(position = "dodge") +
-  labs(
-    x = "Area",
-    y = "Number",
-    fill = "Blank Type"
-  ) +
+
+# Calculate the counts for debitage and 
+counts <- tl4_tool_Blank %>%
+  group_by(debitage, ) %>%
+  summarise(count = n())
+
+# Create the bar plot
+ggplot(counts, 
+       aes(x = factor(debitage), 
+           y = count, 
+           fill = factor(debitage))) +
+
+  geom_bar(stat = "identity", 
+           position = "dodge") +
+
+  # Move the labels slightly above the bars
+  geom_text(aes(label = count), 
+            vjust = -0.5) +
+
+  geom_hline(data = counts %>% 
+               filter( == 1), 
+             aes(yintercept = count),
+             linetype = "dashed", 
+             color = "red") +
+  labs(title = "Debitage vs. ",
+       x = "",
+       y = "Numbers") +
   theme_minimal()
-```
-
-WORK ON THIS PART LATER:
-
-```{r}
-
-                                                #DO NOT RUN#
-
-tl_final_area %>%
-  filter(blank %in% c("bladelet", "blade", "flake")) %>%
-  select(blank, `sickle shine`) %>%
+
+```
+
+```{r}
+# Calculate the counts for debitage and 
+counts <- tl4_tool_Blank %>%
+  group_by(Retouch = 1,  =1) %>%
+  summarise(count = n())
+
+# Create the bar plot
+ggplot(counts, 
+       aes(x = factor( = 1), 
+           y = count, 
+           fill = factor( = 1))) +
+
+  geom_bar(stat = "identity", 
+           position = "dodge") +
+
+  # Move the labels slightly above the bars
+  geom_text(aes(label = count), 
+            vjust = -0.5) +
+
+  geom_hline(data = counts %>% 
+               filter( == 1), 
+             aes(yintercept = count),
+             linetype = "dashed", 
+             color = "red") +
+  labs(title = " vs. formal ",
+       x = "",
+       y = "Numbers") +
+  theme_minimal()
+```
+
+```{r}
+#|label: fig_
+#|fig-cap: ""
+#| fig-height: 5
+tl4_tool_Blank %>%
+  filter(Blank2 %in% c("bladelet", "blade", "flake")) %>%
+  select(Blank2, shine2 = 1) %>%
   ggplot() +
-  aes(x = `sickle shine`) +
+  aes(x = shine2) +
   geom_bar() +
-  labs(x = "shine", y = "Frequency") +
+  labs(x = "shine", 
+       y = "Frequency" )+
   theme_minimal()
 
-```
-
-```{r}
-
-                                         #DO NOT RUN#
-tl_final_area %>%
-  filter(blank %in% c("bladelet", "blade", "flake")) %>%
-  select(blank, retouch = 1) %>%
+tl4_tool_Blank %>%
+  filter(Blank2 %in% c("bladelet", "blade", "flake")) %>%
+  select(Blank2, Retouch = 1) %>%
   ggplot() +
-  aes(x = retouch) +
+  aes(x = Retouch) +
   geom_bar() +
@@
   theme_minimal()
-```
-
-```{r}
-
-                                       #DO NOT RUN#
-
-tl_final_area %>%
-  filter(blank %in% c("bladelet", "blade", "flake")) %>%
-  select(blank, utilization = 1) %>%
+
+tl4_tool_Blank %>%
+  filter(Blank2 %in% c("bladelet", "blade", "flake")) %>%
+  select(Blank2, Utilization = 1) %>%
   ggplot() +
-  aes(x = utilization) +
+  aes(x = Utilization) +
   geom_bar() +
@@
 ```{r}
-
-                                       #DO NOT RUN#
-
 library(ggplot2)
@@
 
-------------
-
-WABI_AREA
-
-```{r}
-# regression plot-area
+WABI_AL
+Ta-e → 13 m3
+Ta-d → 45 m3
+D1-c → 0.675 m3
+Tp-c → 1.44 m3
+D1-a → 1.575 m3
+
+```{r}
+# Create a tibble with example data
+summary_table2_area <- tibble(
+  area = c("D1_a", "D1_c", "TA_d", "TA_e", "TP1_c"),
+  lithic_volumetric_density = c(51.2, 110.9, 45.1, 100.2, 88.3),
+  percent_retouch = c(4.9, 9.3, 15.3, 18.0, 9.8)
+)
+
+```
+
+```{r}
+# Load necessary libraries
+# Import required packages
 library(ggplot2)
@@
 
-# Plotting
-p <- ggplot(summary_table2_area, 
-            aes(x = lithic_volumetric_density, 
-                y = percent_retouch)) +
-
-  # regression line 
+# Create the data frame based on the example table
+summary_table2_area <- data.frame(
+  area = c("D1_a", "D1_c", "TA_d", "TA_e", "TP1_c"),
+  lithic_volumetric_density = c(51.43, 111.11, 45.13, 100.08, 212.50),
+  retouch_frequency = c(0.05, 0.09, 0.07, 0.07, 0.10)
+)
+
+# Generate the regression plot
+p <- ggplot(summary_table2_area, aes(x = lithic_volumetric_density, y = retouch_frequency)) +
+
+  # Blue regression line with shaded area
   geom_smooth(method = "lm", 
@@
 
-  # Points 
+  # Points colored by area
   geom_point(color = "black") +
@@
        x = "Lithic Volumetric Density",
+       y = "Retouch Frequency")
+
+print(p)
+
+```
+
+WABI_AREA
+
+```{r}
+# regression plot-area
+library(ggplot2)
+library(ggpmisc)
+
+# Plotting
+p <- ggplot(summary_table2_area, aes(x = lithic_volumetric_density, y = percent_retouch)) +
+
+  # Blue regression line with shaded area
+  geom_smooth(method = "lm", 
+              se = TRUE, 
+              aes(group = 1), 
+              color = "blue") +
+
+  # Points colored by area
+  geom_point(color = "black") +
+
+  # Labels points with area names
+  geom_text(aes(label = area),
+            vjust = -1, 
+            hjust = -0.5) +
+
+  # Regression equation, R-squared, and p-value
+  stat_poly_eq(aes(label = paste(stat(eq.label), stat(rr.label), stat(p.value.label), sep = "~~~")),
+               formula = y ~ x,
+               parse = TRUE,
+               size = 3) +
+
+  # Titles and labels
+  labs(title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
+       x = "Lithic Volumetric Density",
        y = "Retouch Frequency (%)")
@@
 ```
+
+```{r}
+# Fit the linear model
+linear_model <- lm(percent_retouch ~ lithic_volumetric_density, data = summary_table2_area)
+
+# Display the summary of the model to get coefficients, R-squared, and p-value
+summary_linear_model <- summary(linear_model)
+
+# Extract the coefficients (intercept and slope)
+coefficients <- summary_linear_model$coefficients
+
+# Extract the R-squared value
+r_squared <- summary_linear_model$r.squared
+
+# Extract the p-value for the slope (lithic_volumetric_density)
+p_value_slope <- coefficients["lithic_volumetric_density", "Pr(>|t|)"]
+
+cat("Coefficients:\n")
+print(coefficients)
+
+cat("\nR-squared value:", r_squared)
+
+cat("\nP-value for slope:", p_value_slope)
+
+```
+
+```{r}
+#whole assemblage
+
+  # Summarize data by Area
+summary_table_tech <- tl_final_area %>%
+  group_by(area) %>%
+
+  summarise(
+   blade_count = sum(blank == "blade", na.rm = TRUE),
+    flake_count = sum(blank == "flake", na.rm = TRUE),
+    bladelet_count = sum(blank == "bladelet", na.rm = TRUE),
+
+    bullet_core_count = sum(`core-typology` == "bullet", na.rm = TRUE),
+    flat_core_count = sum(`core-typology` == "flat", na.rm = TRUE),
+    pyramid_core_count = sum(`core-typology` == "pyramid", na.rm = TRUE),
+    cylinder_core_count = sum(`core-typology` %in% c("prismatic", "cylinder"), na.rm = TRUE),
+    shapeless_core_count = sum(`core-typology` == "shapeless", na.rm = TRUE),
+
+    core_fragment_count = sum(`corefrag` == "core fragment", na.rm = TRUE),
+    core_rejuvention_count = sum(`corefrag` == "rejuvenation piece", na.rm = TRUE),
+    core_tablet_count = sum(`corefrag` == "core tablet", na.rm = TRUE),
+    core_crested_count = sum(`corefrag` == "crested bladelet", na.rm = TRUE),
+    core_primarypiece_count = sum(`corefrag`  %in% c("primary flake", "primary blade"), na.rm = TRUE),
+
+   total = n()
+  )
+```
+
+```{r}
+#plot_blank_type
+# Libraries
+library(dplyr)
+library(ggplot2)
+library(ggbeeswarm)
+
+# Data filtering and plotting
+tl_final_area %>% 
+  filter(blank %in% c('bladelet', 'blade', 'flake')) %>%
+  ggplot(aes(x = area, fill = blank)) +
+  geom_bar(position = "dodge") +
+  labs(
+    x = "Area",
+    y = "Number",
+    fill = "Blank Type"
+  ) +
+  theme_minimal()
+
+```
+
+```{r}
+
+library(dplyr)
+library(ggplot2)
+
+# Filter the data for the relevant categories
+tl4_filtered <- tl4_tool_Blank %>%
+  filter(Blank2 %in% 
+           c("bladelet", 
+             "blade", 
+             "flake"))
+
+# Create separate bar plots for shine, Retouch, and Utilization
+plot_shine2 <- ggplot(tl4_filtered, aes(x = Blank2, fill = shine2)) +
+  geom_bar(position = "dodge") +
+  labs(title = "Shine2 by Blank2",
+       x = "Blank2",
+       y = "Frequency",
+       fill = "Shine2") +
+  theme_minimal()
+
+plot_retouch <- ggplot(tl4_filtered, aes(x = Blank2, fill = Retouch)) +
+  geom_bar(position = "dodge") +
+  labs(title = "Retouch by Blank2",
+       x = "Blank2",
+       y = "Frequency",
+       fill = "Retouch") +
+  theme_minimal()
+
+plot_utilization <- ggplot(tl4_filtered, aes(x = Blank2, fill = Utilization)) +
+  geom_bar(position = "dodge") +
+  labs(title = "Utilization by Blank2",
+       x = "Blank2",
+       y = "Frequency",
+       fill = "Utilization") +
+  theme_minimal()
+
+# Combine the three plots side by side
+library(patchwork)
+combined_plots <- plot_shine2 + plot_retouch + plot_utilization
+combined_plots
+
+```
+
+```{r}
+library(dplyr)
+library(knitr)
+library(kableExtra)
+
+# Assuming your data is stored in the variable tl4_tool_Blank
+# Filter the data to keep only bladelet, blade, and flake in the Blank column
+tl_final_aread <- tl_final_area %>%
+  filter(blank %in% c("bladelet", "blade", "flake"))  %>%
+
+# Group the data by Blank2 and calculate the counts for each combination of Utilization, shine2, and Retouch
+
+  group_by(blank, utilization, `sickle shine`, retouch) %>%
+  summarise(count = n())
+
+# Print the table using kable() and kableExtra for formatting
+table_kable <- kable(table_summary, format = "html", caption = "Summary Table", 
+                     col.names = c("Blank2", "Utilization", "shine2", "Retouch", "Count")) %>%
+  kable_styling(bootstrap_options = "striped")
+
+# Print the table
+print(table_kable)
+
+```
+
+```{r}
+#|label: fig_
+#|fig-cap: "Boxplot for each Dimention"
+#|fig-height: 5
+
+library(dplyr)
+library(ggplot2)
+
+# Filter the data to keep only bladelet, blade, and flake in the Blank column
+tl4_tool_Blank_filtered <- tl4_tool_Blank %>%
+  filter(Blank2 %in% c("bladelet", "blade", "flake"))
+
+tl4_tool_Blank_tally <- 
+  tl4_tool_Blank_filtered %>%
+  group_by(Blank2) %>%
+  tally()
+
+# Select dimensions and convert Thickness (mm) to numeric
+tl4_tool_Blank_selected <- 
+  tl4_tool_Blank_filtered %>% 
+  select(Blank2, 
+         Utilization, 
+         Retouch, 
+         shine2)
+
+# Plot
+# Pivot data to longer format
+tl4_tool_Blank_long <- 
+  tl4_tool_Blank_selected %>% 
+  pivot_longer(cols = c(Utilization, Retouch, shine2), 
+               names_to = "Blank2", 
+               values_to = "Value")
+
+# Create boxplot
+ggplot(tl4_tool_Blank_long, aes(x = Blank2, y = Value)) +
+  geom_boxplot() +
+  facet_wrap(~Blank2, scales = "free") +
+  labs(x = "Blank", 
+       y = "Frequency") +
+  theme_minimal()
+
+```
+
+```{r}
+library(dplyr)
+library(ggplot2)
+
+# Filter the data to keep only bladelet, blade, and flake in the Blank2 column
+tl4_tool_Blank_filtered <- tl4_tool_Blank %>%
+  filter(Blank2 %in% c("bladelet", "blade", "flake"))
+
+tl4_tool_Blank_tally <- 
+  tl4_tool_Blank_filtered %>%
+  group_by(Blank2) %>%
+  tally()
+
+# Select dimensions and convert Thickness (mm) to numeric
+tl4_tool_Blank_selected <- 
+  tl4_tool_Blank_filtered %>% 
+  select(Blank2, 
+         Utilization, 
+         Retouch, 
+         shine2)
+
+# Manually rename the duplicate columns
+names(tl4_tool_Blank_selected)[names(tl4_tool_Blank_selected) == "Blank2"] <- "Dimension"
+
+# Pivot data to longer format
+tl4_tool_Blank_long <- 
+  tl4_tool_Blank_selected %>% 
+  pivot_longer(cols = c(Utilization, Retouch, shine2), 
+               names_to = "", 
+               values_to = "Value")
+
+# Create boxplot
+ggplot(tl4_tool_Blank_long, aes(x = Dimension, y = Value)) +
+  geom_boxplot() +
+  facet_wrap(~, scales = "free") +
+  labs(x = "Blank", 
+       y = "Frequency") +
+  theme_minimal()
+
+```
+
+```{r}
+
+# Load required packages
+library(ggplot2)
+
+# Filter the data to exclude the value 0 and keep only blade, bladelet, and flake
+tl4_filtered <- tl4_tool_Blank %>%
+  filter(Blank2 %in% c("blade", "bladelet", "flake"))
+
+# Create the plot with facetting
+tl4_plots <- ggplot(tl4_filtered, aes(x = Blank2, fill = Blank2)) +
+  geom_bar(position = "dodge") +
+  labs(title = "Utilization, Retouch, and shine2 by Blank",
+       x = "Blank2",
+       y = "Count") +
+  facet_grid(. ~ Blank2, scales = "free") +
+  theme_minimal()
+
+# Print the plot
+print(tl4_plots)
+
+```
+
+```{r}
+# Load required packages
+library(ggplot2)
+library(cowplot)
+
+# Filter the data to exclude the value 0 and keep only blade, bladelet, and flake
+tl4_filtered <- tl4_tool_Blank %>%
+  filter(Blank2 %in% c("blade", "bladelet", "flake"))
+
+# Plot Utilization by Blank2
+plot_utilization <- ggplot(tl4_filtered, aes(Blank2, Utilization)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  labs(title = "Utilization by Blank",
+       x = "Blank2",
+       y = "Utilization") +
+  theme_minimal()
+
+# Plot Retouch by Blank
+plot_retouch <- ggplot(tl4_filtered, aes(Blank2, Retouch)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  labs(title = "Retouch by Blank",
+       x = "Blank2",
+       y = "Retouch") +
+  theme_minimal()
+
+# Plot shine by Blank
+plot_shine <- ggplot(tl4_filtered, aes(x = Blank2, y = shine2)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  labs(title = "shine by Blank",
+       x = "Blank2",
+       y = "shine2") +
+  theme_minimal()
+
+# Combine the plots using cowplot or patchwork
+plot_combined <- cowplot::plot_grid(plot_utilization, plot_retouch, plot_shine, ncol = 3)
+
+# Print the combined plot
+print(plot_combined)
+
+```
+
+```{r}
+
+#|label: fig_
+#|fig-cap:  vs Debitage
+#|fig-height: 5
+
+# Filter data
+tl4__blank %>% 
+  filter(Blank2 %in% c('tp1', 'ta', 'd1')) %>% 
+  select(Blank2, 
+         `Width (mm)`, 
+         `Thickness (mm)`, 
+         `Length (mm)`) %>%
+mutate(`Thickness (mm)` = as.numeric(`Thickness (mm)`))
+
+# Pivot data to longer format
+long_data <- tl3_dimention_blank %>% 
+  pivot_longer(cols = c(`Width (mm)`, `Thickness (mm)`, `Length (mm)`), 
+               names_to = "Dimension", 
+               values_to = "Value")
+
+# Create boxplot
+ggplot(long_data, aes(x = area, y = Value)) +
+  geom_boxplot() +
+  geom_quasirandom(alpha = 0.01) +
+  facet_wrap(~Dimension, scales = "free_y") +
+  scale_y_log10() +
+  theme_minimal()
+
+```
+
+```{r}
+library(ggbeeswarm)
+tl3_dimention_blank <- tl3 %>% 
+  filter(area %in% tl3_area_tally_greater_than_n$area) %>% 
+  pivot_longer(-area) %>% 
+  ggplot() +
+  aes(area,
+      value) +
+  geom_boxplot() +
+  geom_quasirandom(alpha = 0.01) +
+  facet_wrap( ~ name,
+              scales = "free_y") +
+  scale_y_log10() +
+  theme_minimal()
+
+```
+
+```{r}
+library(ggbeeswarm)
+tl3 %>% 
+  filter(area %in% tl3_area_tally_greater_than_n$area) %>% 
+  pivot_longer(-area) %>% 
+  ggplot() +
+  aes(area,
+      value) +
+  geom_boxplot() +
+  geom_quasirandom(alpha = 0.01) +
+  facet_wrap( ~ name,
+              scales = "free_y") +
+  scale_y_log10() +
+  theme_minimal()
+
+```
+
+```{r}
+#| label: fig-Width-of-Artifacts-Area-A-by-depth
+#| fig-cap: "Width of Artifacts by depth (Area A)"
+#| fig-height: 5
+tl %>%
+  filter(AREA == "A") %>%
+  select(
+         `Width (mm)`,
+         `DEPTH (cm)`) %>%
+  ggplot() +
+  aes(x = `DEPTH (cm)`,
+      y = `Width (mm)`) +
+  geom_boxplot() +
+  theme_minimal()
+```
+
+```{r}
+#| label: Width of Artifacts (Area A)
+#| fig-cap: "Width of Artifacts (Area A)"
+#| fig-height: 5
+
+tl1 %>%
+  filter(AREA == "TP1") %>%
+  select(AREA, `Width (mm)`) %>%
+  ggplot() +
+  aes(x = `Width (mm)`) +
+  geom_histogram(binwidth = 2) +
+  labs(x = "Width of Debitage (mm)", 
+       y = "Frequency", 
+       title = "Histogram of Width for Area TP1" )+
+  theme_minimal()
+```
+
+WABI: 
+```{r}
+# regression plot-layer
+library(ggplot2)
+library(ggpmisc)
+
+# Plotting
+p <- ggplot(summary_table2_layer, 
+            aes(lithic_volumetric_density, 
+                percent_retouch)) +
+
+  geom_smooth(method = "lm", 
+              se = TRUE, 
+              aes(group = 1), 
+              color = "blue") +
+
+  # Points colored by layer
+  geom_point(color = "black") +
+
+  # Labels points with layer names
+  geom_text(aes(label = layer),
+            vjust = -1, 
+            hjust = -0.5) +
+
+  # Regression equation, R-squared, and p-value
+  stat_poly_eq(aes(label = paste(stat(eq.label), stat(rr.label), stat(p.value.label), sep = "~~~")),
+               formula = y ~ x,
+               parse = TRUE,
+               size = 3) +
+
+  # Titles and labels
+  labs(title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
+       x = "Lithic Volumetric Density",
+       y = "Retouch Frequency (%)")
+
+print(p)
+```
+
+```{r}
+#barplot_layer
+
+tl_final_area %>%
+  filter(layer %in% c("a", "c", "d", "e", "eor d")) %>%
+  filter(!is.na(corefrag)) %>% 
+  select(layer, corefrag) %>%
+  ggplot() +
+  aes(x = corefrag) +
+  geom_bar() +  # Use geom_bar for categorical data
+  labs(x = "Core and Core fragments", y = "Frequency") +
+  facet_wrap(~layer, ncol = 1) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+```
+
+```{r}
+#Core-barplot_area
+
+tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1")) %>%
+  filter(!is.na(corefrag)) %>% # Exclude rows where corefrag is NA
+  select(area, corefrag) %>%
+  ggplot() +
+  aes(x = corefrag) +
+  geom_bar() +
+  labs(x = "Core and Core fragments",
+       y = "Frequency") +
+  facet_wrap(~area, ncol = 1) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+```
+
+```{r}
+
+library(ggplot2)
+library(dplyr)
+filtered_data <- tl_final %>%
+  filter(!is.na(`tool-typology`) & `tool-typology` != "" & `tool-typology` != "NA")
+ggplot(data = filtered_data) + 
+  aes(y = `tool-typology`) + 
+  geom_bar() +
+  labs(x = "Frequency",
+       y = "Tool Type") +
+  theme_minimal() +
+  theme(axis.text.y = element_text(angle = 20, hjust = 1, vjust = 0.5))
+
+```
+
+```{r}
+
+filtered_data <- tl_final %>%
+  filter(!is.na(``) & `` != "")
+
+# Create the plot using the filtered data
+ggplot(data = filtered_data) + 
+  aes(y = ``) + 
+  geom_bar() +
+  labs(y = "Tool Type",
+       x = "Frequency") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+```
+
+```{r}
+
+```
+
+PCA-LAYER
+
+```{r}
+#prepare the data for PCA-layer
+
+# Calculating Blank Frequency
+blank_freq <- tl_final_area %>%
+  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(blank)) %>%
+  group_by(layer) %>%
+  summarise(blankFreq = sum(blank %in% c("bladelet", "blade", "flake")) / sum(!is.na(blank)), .groups = 'drop')
+
+# Calculating Core Frequency
+core_freq <- tl_final_area %>%
+  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(corefrag)) %>%
+  group_by(layer) %>%
+  summarise(coreFreq = sum(corefrag %in% c("core fragment", "bullet", "flat", "pyramid", "cylinder/prismatic", "core rejuvenation piece", "shapeless", "core on flake", "bipolar")) / sum(!is.na(corefrag)), .groups = 'drop')
+
+# Calculating Tool Diversity using Menhinick's Index
+tool_div <- tl_final %>%
+  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(tooltypegroup)) %>%
+  group_by(layer) %>%
+  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(sum(!is.na(tooltypegroup))), .groups = 'drop')
+
+# Combining all the calculated frequencies into one tibble
+final_table <- blank_freq %>%
+  inner_join(core_freq, by = "layer") %>%
+  inner_join(tool_div, by = "layer")
+
+# Printing the final table
+print(final_table)
+
+```
+
+```{r}
+# Calculating Blank Frequency
+blank_freq <- tl_final %>%
+  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(blank)) %>%
+  group_by(layer) %>%
+  summarise(blankFreq = sum(blank %in% c("bladelet", "blade", "flake")) / sum(!is.na(blank)), .groups = 'drop')
+
+# Calculating Core Frequency
+core_freq <- tl_final %>%
+  filter(layer %in% c("a", "c", "d", "e", "e or d")) %>%
+  group_by(layer) %>%
+  summarise(coreFreq = sum(corefrag %in% c("core fragment", "bullet", "flat", "pyramid", "cylinder/prismatic", "core rejuvenation piece", "shapeless", "core on flake", "bipolar")) / n(), .groups = 'drop')
+
+# Calculating Tool Diversity using Menhinick's Index
+tool_div <- tl_final %>%
+  filter(layer %in% c("a", "c", "d", "e", "e or d"), !is.na(tooltypegroup)) %>%
+  group_by(layer) %>%
+  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(sum(!is.na(tooltypegroup))), .groups = 'drop')
+
+# Calculating Retouch Frequency
+retouch_freq <- tl_final %>%
+  filter(layer %in% c("a", "c", "d", "e", "eor d")) %>%
+  group_by(layer) %>%
+  summarise(retFreq = sum( == "retouched piece", na.rm = TRUE) / n(), .groups = 'drop')
+
+# Combining all the calculated frequencies, including retouch frequency, into one tibble
+final_table <- blank_freq %>%
+  inner_join(core_freq, by = "layer") %>%
+  inner_join(tool_div, by = "layer") %>%
+  inner_join(retouch_freq, by = "layer")
+
+# Printing the final table
+print(final_table)
+
+```
+
+```{r}
+# Joining calculated frequencies
+final_table <- blank_freq %>%
+  inner_join(core_freq, by = "layer") %>%
+  inner_join(tool_div, by = "layer") %>%
+  inner_join(retouch_freq, by = "layer")
+
+# Converting layer to row names
+final_table_rownames <- final_table %>%
+  column_to_rownames(var = "layer")
+
+```
+
+```{r}
+# Compute the PCA
+library(FactoMineR)
+
+res_pca <-
+  PCA(final_table_rownames,
+      graph = FALSE)
+
+```
+
+```{r}
+library(factoextra)
+
+# Visualize the contributions of variables to the first principal component
+fviz_contrib(res_pca,
+             axes = 1,   
+             choice = "var")
+
+```
+
+```{r}
+library(factoextra)
+
+# Visualize the contributions of variables to the second principal component
+fviz_contrib(res_pca,
+             axes = 2,   # Selecting the second axis
+             choice = "var") # Choosing to show contributions of variables
+
+```
+
+```{r}
+library(factoextra)
+
+# Creating a biplot
+fviz_pca_biplot(res_pca)
+```
+
+```{r}
+library(factoextra)
+fviz_screeplot(res_pca)
+```
+
+```{r}
+fviz_pca_var(res_pca)
+```
+
+```{r}
+
+library(FactoMineR)
+library(factoextra)
+
+# Run PCA
+res_pca <- PCA(final_table_rownames, graph = FALSE)
+
+# Visualize the contributions of variables to the first principal component
+fviz_contrib(res_pca, axes = 1, choice = "var")
+
+# Visualize the contributions of variables to the second principal component
+fviz_contrib(res_pca, axes = 2, choice = "var")
+
+# Visualize eigenvalues and explained variance
+fviz_screeplot(res_pca)
+
+# Create a biplot of the PCA
+fviz_pca_biplot(res_pca)
+
+# Visualize the variables on the principal components
+fviz_pca_var(res_pca)
+
+```
+
+```{r}
+# Load necessary libraries
+library(FactoMineR)
+library(factoextra)
+
+# Run PCA
+res_pca <- PCA(final_table_rownames, graph = FALSE)
+
+# Get the loadings (coefficients) of original variables for each principal component
+loadings_matrix <- res_pca$var$coord
+
+# Print the loadings matrix
+print(loadings_matrix)
+
+```
+
+```{r}
+#Table
+
+# Eigenvalues
+eigenvalues <- res_pca$eig[1:2, 1]
+
+# Variance percent
+variance_percent <- res_pca$eig[1:2, 2]
+
+# Cumulative variance percent
+cumulative_variance_percent <- res_pca$eig[1:2, 3]
+
+# Create the table
+df <- data.frame(Eigenvalue = eigenvalues,
+                 Variance_Percent = variance_percent,
+                 Cumulative_Variance_Percent = cumulative_variance_percent)
+
+rownames(df) <- c("Dim1", "Dim2")
+
+# Print the table
+print(df)
+
+```
+
+```{r}
+#Table
+
+library(gt)
+
+# Create data frame
+df <- data.frame(Eigenvalue = eigenvalues,
+                 Variance_Percent = variance_percent,
+                 Cumulative_Variance_Percent = cumulative_variance_percent)
+rownames(df) <- c("Dim1", "Dim2")
+
+# Create the table using gt
+gt(df) %>%
+  tab_header(
+    title = "",
+    subtitle = "Eigenvalues and percentage of variance for each dimension of PCA"
+  ) %>%
+  cols_label(
+    Eigenvalue = "Eigenvalue",
+    Variance_Percent = "Variance (%)",
+    Cumulative_Variance_Percent = "Cumulative Variance (%)"
+  ) %>%
+  fmt_number(
+    columns = vars(Eigenvalue, Variance_Percent, Cumulative_Variance_Percent),
+    decimals = 2
+  )
+
+```
+
+calculating PCA based on AREA: TA/ TP1/ D1:
+
+```{r}
+#4lithic density
+
+# A→ V=4×5×2.6 = 52 m3
+# TP → V=2×1.8×1.6 = 5.76 m 3
+# D1 → 1.5×1.5×0.95 = 2.1375 m3
+
+library(dplyr)
+
+# Define the volumes for each area based on the given information
+volumes <- data.frame(
+  area = c("ta", "tp1", "d1"),
+  volume = c(52, 5.76, 2.1375) # in m^3
+)
+
+# Count the number of artifacts by area
+artifact_counts <- tl_final_area %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  tally()
+
+# Join artifact_counts with volumes by area
+lithic_density_df <- left_join(artifact_counts, volumes, by = "area")
+
+# Compute lithic density for each area
+lithic_density_df <- lithic_density_df %>%
+  mutate(lithic_density = n / volume) # number of artifacts / volume
+
+# Print the resulting table
+print(lithic_density_df[, c("area", "lithic_density")])
+
+```
+
+```{r}
+
+```
+
+```{r}
+
+#2 Calculate the frequency of cores and core fragments by area
+
+library(dplyr)
+# Calculate the core frequency
+core_frequency_by_area <- left_join(number_of_cores_by_area, 
+                                    total_lithics_by_area, 
+                                    by = "area") %>%
+  mutate(core_frequency = (number_of_cores / total_lithics) * 100)
+
+# Display the result
+print(core_frequency_by_area)
+
+```
+
+```{r}
+# 1number of Blank
+
+# Calculate the total number of lithic artifacts
+total_lithics <- tl_final_area %>%
+  summarise(total_lithics = n()) %>%
+  pull(total_lithics)
+
+# Display the total number of lithic artifacts
+print(paste("The total number of lithic artifacts is:", total_lithics))
+
+```
+
+```{r}
+# 5Calculate the retouched frequency for each area
+library(dplyr)
+
+total_lithics_by_area <- tl_final_area %>%
+  group_by(area) %>%
+  summarise(total_lithics = n())
+
+number_of_retouched_by_area <- tl_final_area %>%
+  mutate(retouch = as.numeric(`retouch`)) %>% 
+  filter(!is.na(retouch), retouch == 1) %>%  
+  group_by(area) %>%  
+  summarise(number_of_retouched = sum(retouch))
+
+retouched_frequency_by_area <- left_join(number_of_retouched_by_area, total_lithics_by_area, by = "area") %>%
+  mutate(retouched_frequency = (number_of_retouched / total_lithics) * 100)
+
+# Display the retouched frequency by area
+print(retouched_frequency_by_area)
+
+print(number_of_retouched_by_area)
+
+```
+
+```{r}
+# 3Calculating Tool Diversity using Menhinick's Index
+tool_div <- tl_final_area %>%
+  filter(area %in% c("ta", "d1", "tp1"), !is.na(tooltypegroup)) %>%
+  group_by(area) %>%
+  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(sum(!is.na(tooltypegroup))), .groups = 'drop')
+```
+
+```{r}
+#feature frequency: 
+d1     1 
+ta    1
+tp1 	0)
+
+```
+
+```{r}
+#estimated area
+A→  52 m3
+TP → 5.76 m 3
+D1 → 2.1375 m3
+```
+
+```{r}
+library(dplyr)
+library(FactoMineR)
+library(tibble)
+
+# Define estimated area and feature frequency data
+area_data <- tibble(
+  area = c("d1", "ta", "tp1"),
+  estimated_area = c(2.1375, 52, 5.76),
+  feature_freq = c(1, 1, 0)
+)
+
+# Step 1: Number of artifacts by area
+artifact_counts <- tl_final_area %>%
+  group_by(area) %>%
+  tally()
+
+# Step 2: Calculate each variable
+
+# Retouch Frequency 
+retouch_freq <- tl_final_area %>%
+  group_by(area) %>%
+  summarise(retouchFreq = mean(!is.na(retouch)), .groups = 'drop')
+
+# Blank Frequency
+blank_freq <- tl_final_area %>%
+  group_by(area) %>%
+  summarise(blankFreq = mean(!is.na(blank)), .groups = 'drop')
+
+# Core Frequency
+core_freq <- tl_final_area %>%
+  group_by(area) %>%
+  summarise(coreFreq = mean(!is.na(`core-typology`)),.groups = 'drop')
+
+# Lithic density (Artifact counts / estimated area)
+lithic_density <- artifact_counts %>%
+  left_join(area_data, by = "area") %>%
+  mutate(lithic_density = n / estimated_area)
+
+# Tool Diversity using Menhinick's Index
+tool_div <- tl_final_area %>%
+  group_by(area) %>%
+  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(n()), .groups = 'drop')
+
+# Step 3: Combining all variables
+
+pca_final_table_area <- left_join(area_data, retouch_freq, by = "area") %>%
+  left_join(blank_freq, by = "area") %>%
+  left_join(core_freq, by = "area") %>%
+  left_join(lithic_density, by = "area") %>%
+  left_join(tool_div, by = "area") %>%
+  select(area, estimated_area, feature_freq, retouchFreq, blankFreq, coreFreq, lithic_density, toolDiv)
+
+```
+
+```{r}
+pca_final_table_area <- pca_final_table_area %>%
+    select(
+        area, 
+        estimated_area = estimated_area.x, 
+        feature_freq = feature_freq.x, 
+        retouchFreq, 
+        blankFreq, 
+        coreFreq, 
+        n, 
+        lithic_density, 
+        toolDiv
+    )
+
+```
+
+```{r}
+# Step 4: Perform PCA
+res_pca_area <- PCA(pca_final_table_area %>% select(-area, -n), graph = FALSE)
+
+# Show the summary of PCA
+print(res_pca_area)
+
+print(res_pca_area$eig)
+print(res_pca_area$var$coord)
+```
+
+```{r}
+# TABLE: Eigenvalues and percentage of variance for each dimension of PCA 
+library(kableExtra)
+
+# Define the values manually (you would get these from your PCA results)
+eigenvalues <- c(4.411234, 2.588766)
+prop_var_explained <- c(63.02, 36.98)
+cumulative_var_explained <- c(63.02, 100.00)
+
+# Create the data frame
+results_df <- data.frame(
+  Dimension = c("Dim. 1", "Dim. 2"),  # Only keep the dimensions you want to display
+  Eigenvalue = eigenvalues,
+  `Variance / percent` = prop_var_explained,
+  `Cumulative / variance percent` = cumulative_var_explained
+)
+
+# Generate the table
+kable(results_df,
+      caption = "Table 3. Eigenvalues and percentage of variance for each dimension of PCA",
+      align = 'c',
+      col.names = c("Dimension", "Eigenvalue", "Variance (%)", "Cumulative Variance (%)")) %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 11)
+
+```
+
+```{r}
+#plotting Dim-1 and Dim-2
+library(FactoMineR)
+library(factoextra)
+
+# Visualize the contributions of variables to the first principal component
+fviz_contrib(res_pca_area, axes = 1, choice = "var")
+
+# Visualize the contributions of variables to the second principal component
+fviz_contrib(res_pca_area, axes = 2, choice = "var")
+
+```
+
+```{r}
+#|fig-cap: Contribution of variables for each of the tow relevant PCA dimensions
+
+# having Dim1 and Dim2 plots next to each other
+
+library(gridExtra)
+
+# Create the first plot and assign it to p1
+p1 <- fviz_contrib(res_pca_area, axes = 1, choice = "var")
+
+# Create the second plot and assign it to p2
+p2 <- fviz_contrib(res_pca_area, axes = 2, choice = "var")
+
+# Use grid.arrange to place them next to each other
+grid.arrange(p1, p2, ncol = 2)
+
+```
+
+```{r}
+# Extract the loadings (coordinates of variables)
+loadings <- res_pca_area$var$coord
+
+# Extract the eigenvalues
+eigenvalues <- res_pca_area$eig
+
+# Calculate the contribution for Dim.1
+contrib_dim1 <- (loadings[,1]^2 / eigenvalues[1, 1]) * 100
+
+# Calculate the contribution for Dim.2
+contrib_dim2 <- (loadings[,2]^2 / eigenvalues[2, 1]) * 100
+
+# Create a data frame to hold the contributions
+contrib_df <- data.frame(
+  Variable = rownames(loadings),
+  Contrib_Dim1 = contrib_dim1,
+  Contrib_Dim2 = contrib_dim2
+)
+
+# Show the contributions
+print(contrib_df)
+
+```
+
+```{r}
+# screeplot
+
+library(FactoMineR)
+library(factoextra)
+
+# Generate the scree plot
+fviz_eig(res_pca_area)
+
+```
+
+```{r}
+# Correlation plots of variables for each of the tow relevant PCA dimensions
+
+fviz_pca_var(res_pca_area,
+
+             repel = TRUE) + 
+  theme(text = element_text(size = 10))
+
+```
+
+```{r}
+# Eigenvalues and percentage of variance
+eig.val <- res_pca_area$eig
+print(eig.val)
+
+# Variable coordinates (correlations with principal components)
+var.coord <- res_pca_area$var$coord
+print(var.coord)
+
+```
+
+```{r}
+#PCA biplot for Dimensions 1 and 2
+
+fviz_pca_biplot(res_pca_area, labelsize = 3, repel = TRUE)
+
+```
+
+```{r}
+# Create the initial biplot without individual labels
+p <- fviz_pca_biplot(res_pca_area, label = "none", 
+                     repel = TRUE, labelsize = 4)
+
+# Add the text labels for the areas using their coordinates from the PCA result
+p + geom_text(aes(x = res_pca_area$ind$coord[, 1], 
+                  y = res_pca_area$ind$coord[, 2], 
+                  label = pca_final_table_area$area), 
+             size = 3, nudge_x = 0.1, nudge_y = 0.1)
+
+```
+
+```{r}
+individual_coordinates = res_pca_area$ind$coord
+variable_coordinates = res_pca_area$var$coord
+
+```
+
+```{r}
+# For variables
+fviz_cos2(res_pca_area, choice = "var")
+
+# For individuals
+fviz_cos2(res_pca_area, choice = "ind")
+
+```
+
+DONE WITH PCA
+
+```{r}
+# Extract individual coordinates
+individual_coordinates <- res_pca_area$ind$coord
+
+# Convert to a data frame for better visibility and add the 'area' column back for context
+individual_coordinates_df <- as.data.frame(individual_coordinates)
+individual_coordinates_df$area <- unique(area_data$area)
+
+# Show the data frame
+print(individual_coordinates_df)
+
+```
+
+```{r}
+library(kableExtra)
+
+# Extract individual coordinates
+individual_coordinates <- res_pca_area$ind$coord
+
+# Convert to a data frame for better visibility and add the 'area' column back for context
+individual_coordinates_df <- as.data.frame(individual_coordinates)
+individual_coordinates_df$Area <- unique(area_data$area)  # Make sure the order matches
+
+# Use kable to create a beautiful table
+kable(individual_coordinates_df, 
+      col.names = c("Area", "Principal Component 1", "Principal Component 2"), 
+      caption = "PCA Coordinates for the Archaeological Areas excavation",
+      align = c('l', 'c', 'c'),
+      digits = 3) %>%
+  kable_styling("striped", full_width = F)
+
+```
+
+```{r}
+# Get the loadings (coefficients) of original variables for each principal component
+loadings_matrix <- res_pca$var$coord
+
+# Print the loadings matrix
+print(loadings_matrix)
+
+```
+
+```{r}
+# Eigenvalues
+eigenvalues <- res_pca$eig[1:2, 1]
+variance_percent <- res_pca$eig[1:2, 2]
+cumulative_variance_percent <- res_pca$eig[1:2, 3]
+
+# Create the table
+df <- data.frame(Eigenvalue = eigenvalues,
+                 Variance_Percent = variance_percent,
+                 Cumulative_Variance_Percent = cumulative_variance_percent)
+
+rownames(df) <- c("Dim1", "Dim2")
+
+# Print the table
+print(df)
+
+```
+
+```{r}
+library(gt)
+
+# Create the table using gt
+gt(df) %>%
+  tab_header(
+    title = "",
+    subtitle = "Eigenvalues and percentage of variance for each dimension of PCA"
+  ) %>%
+  cols_label(
+    Eigenvalue = "Eigenvalue",
+    Variance_Percent = "Variance (%)",
+    Cumulative_Variance_Percent = "Cumulative Variance (%)"
+  ) %>%
+  fmt_number(
+    columns = c("Eigenvalue", "Variance_Percent", "Cumulative_Variance_Percent"),
+    decimals = 2
+  )
+
+```
+
+```{r}
+# Count the number of artifacts by area and store it in a new data frame
+artifact_counts <- tl_final %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  tally()
+
+# Print the resulting table
+print(artifact_counts)
+```
+
+```{r}
+
+# Calculating retouch Frequency
+
+# Calculating Blank Frequency
+blank_freq <- tl_final_area %>%
+  filter(area %in% c("ta", "d1", "tp1"), !is.na(blank)) %>%
+  group_by(area) %>%
+  summarise(blankFreq = sum(blank %in% c("bladelet", "blade", "flake")) / sum(!is.na(blank)), .groups = 'drop')
+
+# Calculating Core Frequency
+core_freq <- tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1"), !is.na(corefrag)) %>%
+  group_by(area) %>%
+  summarise(coreFreq = sum(corefrag %in% c("core fragment", "bullet", "flat", "pyramid", "cylinder/prismatic", "core rejuvenation piece", "shapeless", "core on flake", "bipolar")) / sum(!is.na(corefrag)), .groups = 'drop')
+
+# Calculating Tool Diversity using Menhinick's Index
+tool_div <- tl_final_area %>%
+  filter(area %in% c("ta", "d1", "tp1"), !is.na(tooltypegroup)) %>%
+  group_by(area) %>%
+  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(sum(!is.na(tooltypegroup))), .groups = 'drop')
+
+# Combining all the calculated frequencies into one tibble
+final_table <- blank_freq %>%
+  inner_join(core_freq, by = "area") %>%
+  inner_join(tool_div, by = "area")
+
+# Printing the final table
+print(final_table)
+
+```
+
+```{r}
+# Joining calculated frequencies
+final_table <- blank_freq %>%
+  inner_join(core_freq, by = "area") %>%
+  inner_join(tool_div, by = "area") %>%
+  inner_join(retouch_freq, by = "area")
+
+# Converting area to row names
+final_table_rownames <- final_table %>%
+  column_to_rownames(var = "area")
+
+```
+
+```{r}
+# Load necessary libraries
+library(FactoMineR)
+library(factoextra)
+
+# Run PCA
+res_pca <- PCA(final_table_rownames, graph = FALSE)
+
+# Visualize the contributions of variables to the second principal component
+fviz_contrib(res_pca, axes = 2, choice = "var")
+
+```
+
+```{r}
+# Visualize the contributions of variables to the first principal component
+fviz_contrib(res_pca, axes = 1, choice = "var") 
+```
+
+```{r}
+# Visualize eigenvalues and explained variance
+fviz_screeplot(res_pca)
+```
+
+```{r}
+fviz_pca_biplot(res_pca, labelsize = 3, repel = TRUE)
+
+```
+
+```{r}
+# Visualize the variables on the principal components
+fviz_pca_var(res_pca)
+```
+
+```{r}
+# Get the loadings (coefficients) of original variables for each principal component
+loadings_matrix <- res_pca$var$coord
+
+# Print the loadings matrix
+print(loadings_matrix)
+
+```
+
+```{r}
+# Eigenvalues
+eigenvalues <- res_pca$eig[1:2, 1]
+variance_percent <- res_pca$eig[1:2, 2]
+cumulative_variance_percent <- res_pca$eig[1:2, 3]
+
+# Create the table
+df <- data.frame(Eigenvalue = eigenvalues,
+                 Variance_Percent = variance_percent,
+                 Cumulative_Variance_Percent = cumulative_variance_percent)
+
+rownames(df) <- c("Dim1", "Dim2")
+
+# Print the table
+print(df)
+
+```
+
+```{r}
+library(gt)
+
+# Create the table using gt
+gt(df) %>%
+  tab_header(
+    title = "",
+    subtitle = "Eigenvalues and percentage of variance for each dimension of PCA"
+  ) %>%
+  cols_label(
+    Eigenvalue = "Eigenvalue",
+    Variance_Percent = "Variance (%)",
+    Cumulative_Variance_Percent = "Cumulative Variance (%)"
+  ) %>%
+  fmt_number(
+    columns = c("Eigenvalue", "Variance_Percent", "Cumulative_Variance_Percent"),
+    decimals = 2
+  )
+
+```
+
+```{r}
+# Count the number of artifacts by area and store it in a new data frame
+artifact_counts <- tl_final %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  tally()
+
+# Print the resulting table
+print(artifact_counts)
+
+```
+
+```{r}
+# Define the volumes for each area based on the given information
+volumes <- data.frame(
+  area = c("ta", "tp1", "d1"),
+  volume = c(52, 5.76, 2.1375)
+)
+
+# Count the number of artifacts by area
+artifact_counts <- tl_final %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  tally()
+
+# Join artifact_counts with volumes by area
+lithic_density_df <- left_join(artifact_counts, volumes, by = "area")
+
+# Compute lithic density for each area
+lithic_density_df <- lithic_density_df %>%
+  mutate(lithic_density = n / volume)
+
+# Print the resulting table
+print(lithic_density_df[, c("area", "lithic_density")])
+
+```
+
+```{r}
+# Counting the number of cores by area
+core_counts <- tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1") & !is.na(`core and core fragments`)) %>%
+  group_by(area) %>%
+  tally()
+
+print(core_counts)
+
+```
+
+OK
+
+```{r}
+
+```
+
+```{r}
+#| label: fig-1-layer
+#| fig-cap: "Type of Removals"
+#| fig-height: 5
+
+#Type of blank based on Layers
+
+library(ggbeeswarm)
+
+tl4 %>% 
+  filter(!is.na(layer) & Blank2 %in% c('bladelet', 
+                                       'blade', 
+                                       'flake')) %>%
+
+  # Create bar plot
+  ggplot() +
+  aes(x = layer, 
+      fill = Blank2) +
+  geom_bar(position = "dodge") +
+  labs(x = "Area", 
+       y = "Number", 
+       fill = "Blank Type") +
+  theme_minimal()
+
+```
+
+```{r}
+#Type of core based on Layers
+
+library(ggbeeswarm)
+
+tl4 %>% 
+  filter(!is.na(layer) & 
+           `core-typology` %in% c('bullet','prismatic','pyramid')) %>%
+
+  # Create bar plot
+  ggplot() +
+  aes(x = layer, 
+      fill = `core-typology`) +
+  geom_bar(position = "dodge") +
+  labs(x = "Area", 
+       y = "Number", 
+       fill = "Blank Type") +
+  theme_minimal()
+```
+
+OK

### 2023-09-06 Tulai-trash.qmd
test

--- 2023-09-05 Tulai.qmd
@@
 ```{r}
-```
-
-```{r}
 
@@
 
-Summary tables: 
-1.AREA
+Summary tables: 1.AREA
+
 ```{r}
@@
 2.LAYER
+
 ```{r}
@@
 AREA_LAYER:
+
 ```{r}
@@
 
-# Adding a new row with total counts
 # Adding a new row with total counts
@@
 
-volume for each area-layer: 
-Ta-
+volume for each area-layer: Ta-
 
@@
 
-                        UniqueDO NOT RUN
+                      #########DO NOT RUN0
 
@@
 consistency in dimention:
+
 ```{r}
@@
 
-WABI_AL
-Ta-e → 13 m3
-Ta-d → 45 m3
-D1-c → 0.675 m3
-Tp-c → 1.44 m3
-D1-a → 1.575 m3
+WABI_AL Ta-e → 13 m3 Ta-d → 45 m3 D1-c → 0.675 m3 Tp-c → 1.44 m3 D1-a → 1.575 m3
 
@@
 # Load necessary libraries
-# Import required packages
-library(ggplot2)
+library(tidyverse)
 library(ggpmisc)
 
-# Create the data frame based on the example table
-summary_table2_area <- data.frame(
+# Create the data frame
+summary_table2_area <- tibble(
   area = c("D1_a", "D1_c", "TA_d", "TA_e", "TP1_c"),
@@
 
-# Generate the regression plot
-p <- ggplot(summary_table2_area, aes(x = lithic_volumetric_density, y = retouch_frequency)) +
-
-  # Blue regression line with shaded area
-  geom_smooth(method = "lm", 
-              se = TRUE, 
-              aes(group = 1), 
-              color = "blue") +
-
-  # Points colored by area
-  geom_point(color = "black") +
-
-  # Labels points with area names
+# Create the ggplot
+p <- ggplot(summary_table2_area, 
+            aes(x = lithic_volumetric_density,
+                                     y = retouch_frequency)) +
+
+  # Add points
+  geom_point(color = 
+               "black", 
+             size = 2) +
+
+  # line with log scale
+  stat_smooth(method = "lm", 
+              formula = y ~ log(x), 
+              se = TRUE, aes(group = 1), 
+              color = "blue", 
+              size = 0.5) +
+
+  # Add text labels for each point
   geom_text(aes(label = area), 
             vjust = -1, 
-            hjust = -0.5) +
-
-  # Regression equation, R-squared, and p-value
+            hjust = -0.5, 
+            size = 2.5) +
+
+  # Add regression equation, R-squared, and p-value
   stat_poly_eq(aes(label = paste(stat(eq.label), stat(rr.label), stat(p.value.label), sep = "~~~")),
-               formula = y ~ x,
+               formula = y ~ log(x), 
                parse = TRUE, 
-               size = 3) +
-
-  # Titles and labels
-  labs(title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
+               size = 2) +
+
+  # Add titles and labels
+  labs(
+    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
     x = "Lithic Volumetric Density",
-       y = "Retouch Frequency")
+    y = "Retouch Frequency"
+  ) +
+
+  scale_x_log10() +
+
+  theme(
+    axis.text = element_text(size = 8),
+    axis.title = element_text(size = 10),
+    plot.title = element_text(size = 12)
+  )
 
 print(p)
+
+```
+
+```{r}
+
+library(googlesheets4)
+
+data_sheet <- "https://docs.google.com/spreadsheets/d/16yLb-RyIQ7x2jFEH_Q7OZYTwlEAXJgFu/edit#gid=2014421097"
+
+sum_depth_sheet <- read_sheet(data_sheet, sheet = "sum_depth")
 
@@
 WABI:
+
 ```{r}

### 2023-09-07 Tulai.qmd
test

@@
 title: "Tulai"
-author: "S.SH"
+author: ""
 date: now
@@
 #exploring Data
-# barplot of lithic frequency based on layers
 
@@
     Blank = case_when(
-      Breakage %in% c("0", "1", NA_character_) ~ Blank,  # If Breakage is 0, 1, or NA, keep Blank unchanged
-      TRUE ~ Breakage  # If Breakage is other than 0, 1, or NA, transfer its value to Blank
+      Breakage %in% c("0", "1", NA_character_) ~ Blank,
+
+      TRUE ~ Breakage  
     )
@@
 ```{r}
-# First convert all columns to character type
+# convert all columns to character type (I DONT KNOW WHY I DID THIS!)
 tl4 <- tl4 %>%
@@
 ```{r}
-# Update the 'typology-tool-core' column based on the condition in 'Retouch'
+# Update the 'typology-tool-core' column based on 'Retouch'
 tl_final <- tl_final %>% 
@@
 
-# Creating tooltypegroup based on the given criteria
+# Creating tooltypegroup
 
@@
     `core-fragment` = if_else(
-      !is.na(`core-typology`),  # If core-typology is not NA
-      NA_character_,            # Set core-fragment to NA
-      `core-fragment`           # Otherwise, keep the original value
-    )
-  )
-
-# Show updated data frame
+      !is.na(`core-typology`),  
+      NA_character_,            
+      `core-fragment`           
+    )
+  )
+
 print(tl_final)
@@
 
-```
-
-```{r}
-
-                   #DO NOT RUN#############
-
-library(dplyr)
-library(stringr)
-
-tl_final <- tl_final %>%
-  mutate(
-    `core-fragments` = str_to_lower(`core-fragment`),
-    `core-fragments` = str_trim(`core-fragment`),
-    `core-fragments` = str_replace_all(`core-fragment`, "^[-._/NA\\s]*$|^\\s*$", "0")
-  ) %>%
-  mutate(`core-fragment` = case_when(
-
-            `core-fragment` == "rejuvenation piece" ~ "rejuvenation piece"
-
-            `core-fragment` == "crested bladelet" ~ "crested bladelet",
-
-            `core-fragment` == "crested bladelet" ~ "crested bladelet",
-
-            `core-fragment` == "primaryflake" ~ "primary flake",
-            `core-fragment` == "primary flake" ~ "primary flake",
-
-            `core-fragment` == "primary blade" ~ "primary blade",
-
-            `core-fragment` == "rejuvenationpiece-side" ~ "rejuvenation piece-side",
-            `core-fragment` == "rejuvention piece-side" ~ "rejuvenation piece-side",
-
-            `core-fragment` == "core preparation piece-platform" ~ "core tablet",
-
-            `core-fragment` == "core preparation" ~ "rejuvenation piece",
-
-            `core-fragment` == "core preparation piece" ~ "rejuvenation piece",
-
-            `core-fragment` == "core preperation" ~ "rejuvenation piece",
-
-            `core-fragment` == "pyramid" ~ "pyramid",
-
-            `core-fragment` == "pyramid core" ~ "pyramid",
-
-            `core-fragment` == "pyramid-unidirectional" ~ "pyramid",
-
-            `core-fragment` == "core fragment-pyramid" ~ "pyramid",
-
-            `core-fragment` == "cylinder" ~ "cylinder/prismatic",
-
-            `core-fragment` == "prismatic" ~ "cylinder/prismatic",
-
-            `core-fragment` == "shapeless" ~ "cylinder/prismatic",
-
-            `core-fragment` == "multidirectional core fragment" ~ "shapeless",
-            `core-fragment` == "multidirectionalcore fragment" ~ "shapeless",
-            `core-fragment` == "multidirectionalbladelet core" ~ "shapeless",
-
-            `core-fragment` == "multidirectional flake core" ~ "shapeless",
-
-            `core-fragment` == "multidirectional bladelet core" ~ "shapeless",
-
-            `core-fragment` == "mixed core" ~ "shapeless",
-
-            `core-fragment` == "mixed-percussion" ~ "shapeless",
-
-            `core-fragment` == "mixed" ~ "shapeless",
-
-            `core-fragment` == ("bipolar-percussion") ~ "cylinder/prismatic-percussion",
-
-            `core-fragment` == "flat" ~ "flat",
-
-            `core-fragment` == "core on flake" ~ "core on flake",
-
-            `core-fragment` == "bullet" ~ "bullet",
-
-            `core-fragment` == "core platform rejuvention" ~ "core tablet",
-            `core-fragment` == "bipolar" ~ "cylinder/prismatic",
-             `core-fragment` == "coretablet" ~ "core tablet",
-            `core-fragment` == "crestedbladelet" ~ "crested piece",
-            `core-fragment` == "coreon flake" ~ "core on flake",
-            `core-fragment` == "corefragment-pyramid" ~ "pyramid",
-            `core-fragment` == "corepreparation piece-platform" ~ "core tablet",
-
-    TRUE ~ `core-fragment`
-  ))
-
-unique(`tl_final`$`core-fragment`)
 ```
@@
 
-# Process the tl_final_area dataframe using pipe (%>%)
 tl_final <- tl_final %>%
@@
   )) %>%
-  # Remove 'rejuvenation piece' and 'core fragment' from `core-typology` but keep the column
+  # Remove 'rejuvenation piece' and 'core fragment' from `core-typology` 
   mutate(`core-typology` = if_else(`core-typology` %in% c("rejuvenation piece", "core fragment"), NA_character_, `core-typology`))
@@
 
-# Show the updated data
 print(tl_final)
@@
 
-# Show summary table
 print(summary_table_area)
@@
   summarise(
-    `Cores (n)` = sum(!is.na(core)),  # Count non-NA core entries
-    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag)),  # Count non-NA corefrag entries
+    `Cores (n)` = sum(!is.na(core)),  
+    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag)),  
     `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
@@
 ```{r}
-# Load dplyr library
+
 library(dplyr)
@@
 # Adding a new row with total counts
-# Adding a new row with total counts
 summary_table_area <- summary_table_area %>% 
@@
 ```{r}
-#Generate summary table-area
+#summary table-area
 
@@
 
-# Create a row that contains the column totals, and append it to the table
 total_row <- summary_table_tool_area %>%
@@
 
-# Combine the total_row with the original table
 summary_table_tool_area <- bind_rows(summary_table_tool_area, total_row)
 
-# View the summary table
 print(summary_table_tool_area)
-
-```
-
-```{r}
 
@@
                 font_size = 12) %>%
-  column_spec(1, extra_css = "text-align:left;") %>%  # Make the first column left-aligned
+  column_spec(1, extra_css = "text-align:left;") %>%  
   row_spec(0, bold = TRUE, font_size = 14)
@@
 
-volume for each area-layer: 
-Ta-
-
-```{r}
-# Define the volumes for each area based on the given information
+volume for each area:
+
+```{r}
+# Define the volumes for each area 
 volumes <- data.frame(
@@
 
+WABI_AL Ta-e → 13 m3 Ta-d → 45 m3 D1-c → 0.675 m3 Tp-c → 1.44 m3 D1-a → 1.575 m3
+
+```{r}
+# Create a tibble with example data
+summary_table2_area <- tibble(
+  area = c("D1_a", "D1_c", "TA_d", "TA_e", "TP1_c"),
+  lithic_volumetric_density = c(51.2, 110.9, 45.1, 100.2, 88.3),
+  percent_retouch = c(4.9, 9.3, 15.3, 18.0, 9.8)
+)
+
+```
+
+```{r}
+
+library(tidyverse)
+library(ggpmisc)
+
+# Create the data frame
+summary_table2_area <- tibble(
+  area = c("D1_a", "D1_c", "TA_d", "TA_e", "TP1_c"),
+  lithic_volumetric_density = c(51.43, 111.11, 45.13, 100.08, 212.50),
+  retouch_frequency = c(0.05, 0.09, 0.07, 0.07, 0.10)
+)
+
+# Create the ggplot
+p <- ggplot(summary_table2_area, 
+            aes(x = lithic_volumetric_density,
+                                     y = retouch_frequency)) +
+
+  # Add points
+  geom_point(color = 
+               "black", 
+             size = 2) +
+
+  # line with log scale
+  stat_smooth(method = "lm", 
+              formula = y ~ log(x), 
+              se = TRUE, aes(group = 1), 
+              color = "blue", 
+              size = 0.5) +
+
+  # Add text labels for each point
+  geom_text(aes(label = area), 
+            vjust = -1, 
+            hjust = -0.5, 
+            size = 2.5) +
+
+  # Add regression equation, R-squared, and p-value
+  stat_poly_eq(aes(label = paste(stat(eq.label), stat(rr.label), stat(p.value.label), sep = "~~~")),
+               formula = y ~ log(x), 
+               parse = TRUE, 
+               size = 2) +
+
+  # Add titles and labels
+  labs(
+    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
+    x = "Lithic Volumetric Density",
+    y = "Retouch Frequency"
+  ) +
+
+  scale_x_log10() +
+
+  theme(
+    axis.text = element_text(size = 8),
+    axis.title = element_text(size = 10),
+    plot.title = element_text(size = 12)
+  )
+
+print(p)
+
+```
+
+```{r}
+
+library(googlesheets4)
+
+data_sheet <- "https://docs.google.com/spreadsheets/d/16yLb-RyIQ7x2jFEH_Q7OZYTwlEAXJgFu/edit#gid=2014421097"
+
+sum_depth_sheet <- read_sheet(data_sheet, sheet = "sum_depth")
+
+```
+
+WABI_AREA
+
+```{r}
+# regression plot-area
+library(ggplot2)
+library(ggpmisc)
+
+# Plotting
+p <- ggplot(summary_table2_area, aes(x = lithic_volumetric_density, y = percent_retouch)) +
+
+  # Blue regression line with shaded area
+  geom_smooth(method = "lm", 
+              se = TRUE, 
+              aes(group = 1), 
+              color = "blue") +
+
+  # Points colored by area
+  geom_point(color = "black") +
+
+  # Labels points with area names
+  geom_text(aes(label = area),
+            vjust = -1, 
+            hjust = -0.5) +
+
+  # Regression equation, R-squared, and p-value
+  stat_poly_eq(aes(label = paste(stat(eq.label), stat(rr.label), stat(p.value.label), sep = "~~~")),
+               formula = y ~ x,
+               parse = TRUE,
+               size = 3) +
+
+  # Titles and labels
+  labs(title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
+       x = "Lithic Volumetric Density",
+       y = "Retouch Frequency (%)")
+
+print(p)
+
+```
+
+```{r}
+# Fit the linear model
+linear_model <- lm(percent_retouch ~ lithic_volumetric_density, data = summary_table2_area)
+
+# Display the summary of the model to get coefficients, R-squared, and p-value
+summary_linear_model <- summary(linear_model)
+
+# Extract the coefficients (intercept and slope)
+coefficients <- summary_linear_model$coefficients
+
+# Extract the R-squared value
+r_squared <- summary_linear_model$r.squared
+
+# Extract the p-value for the slope (lithic_volumetric_density)
+p_value_slope <- coefficients["lithic_volumetric_density", "Pr(>|t|)"]
+
+cat("Coefficients:\n")
+print(coefficients)
+
+cat("\nR-squared value:", r_squared)
+
+cat("\nP-value for slope:", p_value_slope)
+
+```
+
+```{r}
+#whole assemblage
+
+  # Summarize data by Area
+summary_table_tech <- tl_final_area %>%
+  group_by(area) %>%
+
+  summarise(
+   blade_count = sum(blank == "blade", na.rm = TRUE),
+    flake_count = sum(blank == "flake", na.rm = TRUE),
+    bladelet_count = sum(blank == "bladelet", na.rm = TRUE),
+
+    bullet_core_count = sum(`core-typology` == "bullet", na.rm = TRUE),
+    flat_core_count = sum(`core-typology` == "flat", na.rm = TRUE),
+    pyramid_core_count = sum(`core-typology` == "pyramid", na.rm = TRUE),
+    cylinder_core_count = sum(`core-typology` %in% c("prismatic", "cylinder"), na.rm = TRUE),
+    shapeless_core_count = sum(`core-typology` == "shapeless", na.rm = TRUE),
+
+    core_fragment_count = sum(`corefrag` == "core fragment", na.rm = TRUE),
+    core_rejuvention_count = sum(`corefrag` == "rejuvenation piece", na.rm = TRUE),
+    core_tablet_count = sum(`corefrag` == "core tablet", na.rm = TRUE),
+    core_crested_count = sum(`corefrag` == "crested bladelet", na.rm = TRUE),
+    core_primarypiece_count = sum(`corefrag`  %in% c("primary flake", "primary blade"), na.rm = TRUE),
+
+   total = n()
+  )
+```
+
+```{r}
+#plot_blank_type
+
+library(dplyr)
+library(ggplot2)
+library(ggbeeswarm)
+
+# Data filtering and plotting
+tl_final_area %>% 
+  filter(blank %in% c('bladelet', 'blade', 'flake')) %>%
+  ggplot(aes(x = area, fill = blank)) +
+  geom_bar(position = "dodge") +
+  labs(
+    x = "Area",
+    y = "Number",
+    fill = "Blank Type"
+  ) +
+  theme_minimal()
+
+```
+
+```{r}
+
+library(dplyr)
+library(ggplot2)
+
+# Filter the data for the relevant categories
+tl4_filtered <- tl4_tool_Blank %>%
+  filter(Blank2 %in% 
+           c("bladelet", 
+             "blade", 
+             "flake"))
+
+# Create separate bar plots for shine, Retouch, and Utilization
+plot_shine2 <- ggplot(tl4_filtered, aes(x = Blank2, fill = shine2)) +
+  geom_bar(position = "dodge") +
+  labs(title = "Shine2 by Blank2",
+       x = "Blank2",
+       y = "Frequency",
+       fill = "Shine2") +
+  theme_minimal()
+
+plot_retouch <- ggplot(tl4_filtered, aes(x = Blank2, fill = Retouch)) +
+  geom_bar(position = "dodge") +
+  labs(title = "Retouch by Blank2",
+       x = "Blank2",
+       y = "Frequency",
+       fill = "Retouch") +
+  theme_minimal()
+
+plot_utilization <- ggplot(tl4_filtered, aes(x = Blank2, fill = Utilization)) +
+  geom_bar(position = "dodge") +
+  labs(title = "Utilization by Blank2",
+       x = "Blank2",
+       y = "Frequency",
+       fill = "Utilization") +
+  theme_minimal()
+
+# Combine the three plots side by side
+library(patchwork)
+combined_plots <- plot_shine2 + plot_retouch + plot_utilization
+combined_plots
+
+```
+
+```{r}
+library(dplyr)
+library(knitr)
+library(kableExtra)
+
+#table
+
+tl_final_aread <- tl_final_area %>%
+  filter(blank %in% c("bladelet", "blade", "flake"))  %>%
+
+  group_by(blank, utilization, `sickle shine`, retouch) %>%
+  summarise(count = n())
+
+table_kable <- kable(table_summary, format = "html", caption = "Summary Table", 
+                     col.names = c("Blank2", "Utilization", "shine2", "Retouch", "Count")) %>%
+  kable_styling(bootstrap_options = "striped")
+
+print(table_kable)
+
+```
+
+```{r}
+#|label: fig_
+#|fig-cap: "Boxplot for each Dimention"
+#|fig-height: 5
+
+library(dplyr)
+library(ggplot2)
+
+# Filter the data to keep only bladelet, blade, and flake in the Blank column
+tl4_tool_Blank_filtered <- tl4_tool_Blank %>%
+  filter(Blank2 %in% c("bladelet", "blade", "flake"))
+
+tl4_tool_Blank_tally <- 
+  tl4_tool_Blank_filtered %>%
+  group_by(Blank2) %>%
+  tally()
+
+# Select dimensions and convert Thickness (mm) to numeric
+tl4_tool_Blank_selected <- 
+  tl4_tool_Blank_filtered %>% 
+  select(Blank2, 
+         Utilization, 
+         Retouch, 
+         shine2)
+
+tl4_tool_Blank_long <- 
+  tl4_tool_Blank_selected %>% 
+  pivot_longer(cols = c(Utilization, Retouch, shine2), 
+               names_to = "Blank2", 
+               values_to = "Value")
+
+# Create boxplot
+ggplot(tl4_tool_Blank_long, aes(x = Blank2, y = Value)) +
+  geom_boxplot() +
+  facet_wrap(~Blank2, scales = "free") +
+  labs(x = "Blank", 
+       y = "Frequency") +
+  theme_minimal()
+
+```
+
+```{r}
+library(dplyr)
+library(ggplot2)
+
+# Filter the data to keep only bladelet, blade, and flake in the Blank2 column
+tl4_tool_Blank_filtered <- tl4_tool_Blank %>%
+  filter(Blank2 %in% c("bladelet", "blade", "flake"))
+
+tl4_tool_Blank_tally <- 
+  tl4_tool_Blank_filtered %>%
+  group_by(Blank2) %>%
+  tally()
+
+# Select dimensions and convert Thickness (mm) to numeric
+tl4_tool_Blank_selected <- 
+  tl4_tool_Blank_filtered %>% 
+  select(Blank2, 
+         Utilization, 
+         Retouch, 
+         shine2)
+
+# Manually rename the duplicate columns
+names(tl4_tool_Blank_selected)[names(tl4_tool_Blank_selected) == "Blank2"] <- "Dimension"
+
+# Pivot data to longer format
+tl4_tool_Blank_long <- 
+  tl4_tool_Blank_selected %>% 
+  pivot_longer(cols = c(Utilization, Retouch, shine2), 
+               names_to = "", 
+               values_to = "Value")
+
+ggplot(tl4_tool_Blank_long, aes(x = Dimension, y = Value)) +
+  geom_boxplot() +
+  facet_wrap(~, scales = "free") +
+  labs(x = "Blank", 
+       y = "Frequency") +
+  theme_minimal()
+
+```
+
+```{r}
+
+# Load required packages
+library(ggplot2)
+
+# Filter the data to exclude the value 0 and keep only blade, bladelet, and flake
+tl4_filtered <- tl4_tool_Blank %>%
+  filter(Blank2 %in% c("blade", "bladelet", "flake"))
+
+# Create the plot with facetting
+tl4_plots <- ggplot(tl4_filtered, aes(x = Blank2, fill = Blank2)) +
+  geom_bar(position = "dodge") +
+  labs(title = "Utilization, Retouch, and shine2 by Blank",
+       x = "Blank2",
+       y = "Count") +
+  facet_grid(. ~ Blank2, scales = "free") +
+  theme_minimal()
+
+print(tl4_plots)
+
+```
+
+```{r}
+# Load required packages
+library(ggplot2)
+library(cowplot)
+
+# Filter the data to exclude the value 0 and keep only blade, bladelet, and flake
+tl4_filtered <- tl4_tool_Blank %>%
+  filter(Blank2 %in% c("blade", "bladelet", "flake"))
+
+# Plot Utilization by Blank2
+plot_utilization <- ggplot(tl4_filtered, aes(Blank2, Utilization)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  labs(title = "Utilization by Blank",
+       x = "Blank2",
+       y = "Utilization") +
+  theme_minimal()
+
+# Plot Retouch by Blank
+plot_retouch <- ggplot(tl4_filtered, aes(Blank2, Retouch)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  labs(title = "Retouch by Blank",
+       x = "Blank2",
+       y = "Retouch") +
+  theme_minimal()
+
+# Plot shine by Blank
+plot_shine <- ggplot(tl4_filtered, aes(x = Blank2, y = shine2)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  labs(title = "shine by Blank",
+       x = "Blank2",
+       y = "shine2") +
+  theme_minimal()
+
+# Combine the plots using cowplot or patchwork
+plot_combined <- cowplot::plot_grid(plot_utilization, plot_retouch, plot_shine, ncol = 3)
+
+print(plot_combined)
+
+```
+
+```{r}
+
+#|label: fig_
+#|fig-cap:  vs Debitage
+#|fig-height: 5
+
+# Filter data
+tl4__blank %>% 
+  filter(Blank2 %in% c('tp1', 'ta', 'd1')) %>% 
+  select(Blank2, 
+         `Width (mm)`, 
+         `Thickness (mm)`, 
+         `Length (mm)`) %>%
+mutate(`Thickness (mm)` = as.numeric(`Thickness (mm)`))
+
+# Pivot data to longer format
+long_data <- tl3_dimention_blank %>% 
+  pivot_longer(cols = c(`Width (mm)`, `Thickness (mm)`, `Length (mm)`), 
+               names_to = "Dimension", 
+               values_to = "Value")
+
+# Create boxplot
+ggplot(long_data, aes(x = area, y = Value)) +
+  geom_boxplot() +
+  geom_quasirandom(alpha = 0.01) +
+  facet_wrap(~Dimension, scales = "free_y") +
+  scale_y_log10() +
+  theme_minimal()
+
+```
+
+```{r}
+library(ggbeeswarm)
+tl3_dimention_blank <- tl3 %>% 
+  filter(area %in% tl3_area_tally_greater_than_n$area) %>% 
+  pivot_longer(-area) %>% 
+  ggplot() +
+  aes(area,
+      value) +
+  geom_boxplot() +
+  geom_quasirandom(alpha = 0.01) +
+  facet_wrap( ~ name,
+              scales = "free_y") +
+  scale_y_log10() +
+  theme_minimal()
+
+```
+
+```{r}
+library(ggbeeswarm)
+tl3 %>% 
+  filter(area %in% tl3_area_tally_greater_than_n$area) %>% 
+  pivot_longer(-area) %>% 
+  ggplot() +
+  aes(area,
+      value) +
+  geom_boxplot() +
+  geom_quasirandom(alpha = 0.01) +
+  facet_wrap( ~ name,
+              scales = "free_y") +
+  scale_y_log10() +
+  theme_minimal()
+
+```
+
+```{r}
+#| label: fig-Width-of-Artifacts-Area-A-by-depth
+#| fig-cap: "Width of Artifacts by depth (Area A)"
+#| fig-height: 5
+tl %>%
+  filter(AREA == "A") %>%
+  select(
+         `Width (mm)`,
+         `DEPTH (cm)`) %>%
+  ggplot() +
+  aes(x = `DEPTH (cm)`,
+      y = `Width (mm)`) +
+  geom_boxplot() +
+  theme_minimal()
+```
+
+```{r}
+#| label: Width of Artifacts (Area A)
+#| fig-cap: "Width of Artifacts (Area A)"
+#| fig-height: 5
+
+tl1 %>%
+  filter(AREA == "TP1") %>%
+  select(AREA, `Width (mm)`) %>%
+  ggplot() +
+  aes(x = `Width (mm)`) +
+  geom_histogram(binwidth = 2) +
+  labs(x = "Width of Debitage (mm)", 
+       y = "Frequency", 
+       title = "Histogram of Width for Area TP1" )+
+  theme_minimal()
+```
+
+WABI:
+
+```{r}
+# regression plot-layer
+library(ggplot2)
+library(ggpmisc)
+
+# Plotting
+p <- ggplot(summary_table2_layer, 
+            aes(lithic_volumetric_density, 
+                percent_retouch)) +
+
+  geom_smooth(method = "lm", 
+              se = TRUE, 
+              aes(group = 1), 
+              color = "blue") +
+
+  # Points colored by layer
+  geom_point(color = "black") +
+
+  # Labels points with layer names
+  geom_text(aes(label = layer),
+            vjust = -1, 
+            hjust = -0.5) +
+
+  # Regression equation, R-squared, and p-value
+  stat_poly_eq(aes(label = paste(stat(eq.label), stat(rr.label), stat(p.value.label), sep = "~~~")),
+               formula = y ~ x,
+               parse = TRUE,
+               size = 3) +
+
+  # Titles and labels
+  labs(title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
+       x = "Lithic Volumetric Density",
+       y = "Retouch Frequency (%)")
+
+print(p)
+```
+
+```{r}
+#barplot_layer
+
+tl_final_area %>%
+  filter(layer %in% c("a", "c", "d", "e", "eor d")) %>%
+  filter(!is.na(corefrag)) %>% 
+  select(layer, corefrag) %>%
+  ggplot() +
+  aes(x = corefrag) +
+  geom_bar() +  
+  labs(x = "Core and Core fragments", y = "Frequency") +
+  facet_wrap(~layer, ncol = 1) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+```
+
+```{r}
+#Core-barplot_area
+
+tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1")) %>%
+  filter(!is.na(corefrag)) %>% # Exclude rows where corefrag is NA
+  select(area, corefrag) %>%
+  ggplot() +
+  aes(x = corefrag) +
+  geom_bar() +
+  labs(x = "Core and Core fragments",
+       y = "Frequency") +
+  facet_wrap(~area, ncol = 1) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+```
+
+```{r}
+
+library(ggplot2)
+library(dplyr)
+filtered_data <- tl_final %>%
+  filter(!is.na(`tool-typology`) & `tool-typology` != "" & `tool-typology` != "NA")
+ggplot(data = filtered_data) + 
+  aes(y = `tool-typology`) + 
+  geom_bar() +
+  labs(x = "Frequency",
+       y = "Tool Type") +
+  theme_minimal() +
+  theme(axis.text.y = element_text(angle = 20, hjust = 1, vjust = 0.5))
+
+```
+
+```{r}
+
+filtered_data <- tl_final %>%
+  filter(!is.na(``) & `` != "")
+
+# Create the plot using the filtered data
+ggplot(data = filtered_data) + 
+  aes(y = ``) + 
+  geom_bar() +
+  labs(y = "Tool Type",
+       x = "Frequency") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+```
+
+```{r}
+
+```
+
+PCA-LAYER
+
+```{r}
+#prepare the data for PCA-layer
+
+# Calculating Blank Frequency
+blank_freq <- tl_final_area %>%
+  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(blank)) %>%
+  group_by(layer) %>%
+  summarise(blankFreq = sum(blank %in% c("bladelet", "blade", "flake")) / sum(!is.na(blank)), .groups = 'drop')
+
+# Calculating Core Frequency
+core_freq <- tl_final_area %>%
+  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(corefrag)) %>%
+  group_by(layer) %>%
+  summarise(coreFreq = sum(corefrag %in% c("core fragment", "bullet", "flat", "pyramid", "cylinder/prismatic", "core rejuvenation piece", "shapeless", "core on flake", "bipolar")) / sum(!is.na(corefrag)), .groups = 'drop')
+
+# Calculating Tool Diversity using Menhinick's Index
+tool_div <- tl_final %>%
+  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(tooltypegroup)) %>%
+  group_by(layer) %>%
+  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(sum(!is.na(tooltypegroup))), .groups = 'drop')
+
+# Combining all the calculated frequencies into one tibble
+final_table <- blank_freq %>%
+  inner_join(core_freq, by = "layer") %>%
+  inner_join(tool_div, by = "layer")
+
+# Printing the final table
+print(final_table)
+
+```
+
+```{r}
+# Blank Frequency
+blank_freq <- tl_final %>%
+  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(blank)) %>%
+  group_by(layer) %>%
+  summarise(blankFreq = sum(blank %in% c("bladelet", "blade", "flake")) / sum(!is.na(blank)), .groups = 'drop')
+
+# Calculating Core Frequency
+core_freq <- tl_final %>%
+  filter(layer %in% c("a", "c", "d", "e", "e or d")) %>%
+  group_by(layer) %>%
+  summarise(coreFreq = sum(corefrag %in% c("core fragment", "bullet", "flat", "pyramid", "cylinder/prismatic", "core rejuvenation piece", "shapeless", "core on flake", "bipolar")) / n(), .groups = 'drop')
+
+# Calculating Tool Diversity using Menhinick's Index
+tool_div <- tl_final %>%
+  filter(layer %in% c("a", "c", "d", "e", "e or d"), !is.na(tooltypegroup)) %>%
+  group_by(layer) %>%
+  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(sum(!is.na(tooltypegroup))), .groups = 'drop')
+
+# Calculating Retouch Frequency
+retouch_freq <- tl_final %>%
+  filter(layer %in% c("a", "c", "d", "e", "eor d")) %>%
+  group_by(layer) %>%
+  summarise(retFreq = sum( == "retouched piece", na.rm = TRUE) / n(), .groups = 'drop')
+
+# Combining all the calculated frequencies, including retouch frequency, into one tibble
+final_table <- blank_freq %>%
+  inner_join(core_freq, by = "layer") %>%
+  inner_join(tool_div, by = "layer") %>%
+  inner_join(retouch_freq, by = "layer")
+
+# Printing the final table
+print(final_table)
+
+```
+
+```{r}
+# Joining calculated frequencies
+final_table <- blank_freq %>%
+  inner_join(core_freq, by = "layer") %>%
+  inner_join(tool_div, by = "layer") %>%
+  inner_join(retouch_freq, by = "layer")
+
+# Converting layer to row names
+final_table_rownames <- final_table %>%
+  column_to_rownames(var = "layer")
+
+```
+
+```{r}
+# Compute the PCA
+library(FactoMineR)
+
+res_pca <-
+  PCA(final_table_rownames,
+      graph = FALSE)
+
+```
+
+```{r}
+library(factoextra)
+
+# Visualize the contributions of variables to the first principal component
+fviz_contrib(res_pca,
+             axes = 1,   
+             choice = "var")
+
+```
+
+```{r}
+library(factoextra)
+
+# Visualize the contributions of variables to the second principal component
+fviz_contrib(res_pca,
+             axes = 2,   
+             choice = "var")
+
+```
+
+```{r}
+library(factoextra)
+
+# Creating a biplot
+fviz_pca_biplot(res_pca)
+```
+
+```{r}
+library(factoextra)
+fviz_screeplot(res_pca)
+```
+
+```{r}
+fviz_pca_var(res_pca)
+```
+
+```{r}
+
+library(FactoMineR)
+library(factoextra)
+
+# Run PCA
+res_pca <- PCA(final_table_rownames, graph = FALSE)
+
+# Visualize the contributions of variables to the first principal component
+fviz_contrib(res_pca, axes = 1, choice = "var")
+
+# Visualize the contributions of variables to the second principal component
+fviz_contrib(res_pca, axes = 2, choice = "var")
+
+# Visualize eigenvalues and explained variance
+fviz_screeplot(res_pca)
+
+# Create a biplot of the PCA
+fviz_pca_biplot(res_pca)
+
+# Visualize the variables on the principal components
+fviz_pca_var(res_pca)
+
+```
+
+```{r}
+
+library(FactoMineR)
+library(factoextra)
+
+# Run PCA
+res_pca <- PCA(final_table_rownames, graph = FALSE)
+
+# Get the loadings (coefficients) of original variables for each principal component
+loadings_matrix <- res_pca$var$coord
+
+# Print the loadings matrix
+print(loadings_matrix)
+
+```
+
+```{r}
+#Table
+
+# Eigenvalues
+eigenvalues <- res_pca$eig[1:2, 1]
+
+# Variance percent
+variance_percent <- res_pca$eig[1:2, 2]
+
+# Cumulative variance percent
+cumulative_variance_percent <- res_pca$eig[1:2, 3]
+
+# Create the table
+df <- data.frame(Eigenvalue = eigenvalues,
+                 Variance_Percent = variance_percent,
+                 Cumulative_Variance_Percent = cumulative_variance_percent)
+
+rownames(df) <- c("Dim1", "Dim2")
+
+# Print the table
+print(df)
+
+```
+
+```{r}
+#Table
+
+library(gt)
+
+# Create data frame
+df <- data.frame(Eigenvalue = eigenvalues,
+                 Variance_Percent = variance_percent,
+                 Cumulative_Variance_Percent = cumulative_variance_percent)
+rownames(df) <- c("Dim1", "Dim2")
+
+# Create the table using gt
+gt(df) %>%
+  tab_header(
+    title = "",
+    subtitle = "Eigenvalues and percentage of variance for each dimension of PCA"
+  ) %>%
+  cols_label(
+    Eigenvalue = "Eigenvalue",
+    Variance_Percent = "Variance (%)",
+    Cumulative_Variance_Percent = "Cumulative Variance (%)"
+  ) %>%
+  fmt_number(
+    columns = vars(Eigenvalue, Variance_Percent, Cumulative_Variance_Percent),
+    decimals = 2
+  )
+
+```
+
+calculating PCA based on AREA: TA/ TP1/ D1:
+
+```{r}
+#4lithic density
+
+# A→ V=4×5×2.6 = 52 m3
+# TP → V=2×1.8×1.6 = 5.76 m 3
+# D1 → 1.5×1.5×0.95 = 2.1375 m3
+
+library(dplyr)
+
+# Define the volumes for each area based on the given information
+volumes <- data.frame(
+  area = c("ta", "tp1", "d1"),
+  volume = c(52, 5.76, 2.1375) # in m^3
+)
+
+# Count the number of artifacts by area
+artifact_counts <- tl_final_area %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  tally()
+
+# Join artifact_counts with volumes by area
+lithic_density_df <- left_join(artifact_counts, volumes, by = "area")
+
+# Compute lithic density for each area
+lithic_density_df <- lithic_density_df %>%
+  mutate(lithic_density = n / volume) # number of artifacts / volume
+
+# Print the resulting table
+print(lithic_density_df[, c("area", "lithic_density")])
+
+```
+
+```{r}
+
+#2 Calculate the frequency of cores and core fragments by area
+
+library(dplyr)
+# Calculate the core frequency
+core_frequency_by_area <- left_join(number_of_cores_by_area, 
+                                    total_lithics_by_area, 
+                                    by = "area") %>%
+  mutate(core_frequency = (number_of_cores / total_lithics) * 100)
+
+# Display the result
+print(core_frequency_by_area)
+
+```
+
+```{r}
+# 1number of Blank
+
+# Calculate the total number of lithic artifacts
+total_lithics <- tl_final_area %>%
+  summarise(total_lithics = n()) %>%
+  pull(total_lithics)
+
+# Display the total number of lithic artifacts
+print(paste("The total number of lithic artifacts is:", total_lithics))
+
+```
+
+```{r}
+# 5Calculate the retouched frequency for each area
+library(dplyr)
+
+total_lithics_by_area <- tl_final_area %>%
+  group_by(area) %>%
+  summarise(total_lithics = n())
+
+number_of_retouched_by_area <- tl_final_area %>%
+  mutate(retouch = as.numeric(`retouch`)) %>% 
+  filter(!is.na(retouch), retouch == 1) %>%  
+  group_by(area) %>%  
+  summarise(number_of_retouched = sum(retouch))
+
+retouched_frequency_by_area <- left_join(number_of_retouched_by_area, total_lithics_by_area, by = "area") %>%
+  mutate(retouched_frequency = (number_of_retouched / total_lithics) * 100)
+
+# Display the retouched frequency by area
+print(retouched_frequency_by_area)
+
+print(number_of_retouched_by_area)
+
+```
+
+```{r}
+# 3Calculating Tool Diversity using Menhinick's Index
+tool_div <- tl_final_area %>%
+  filter(area %in% c("ta", "d1", "tp1"), 
+         !is.na(tooltypegroup)) %>%
+  group_by(area) %>%
+  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(sum(!is.na(tooltypegroup))), .groups = 'drop')
+```
+
+feature frequency: d1 1 ta 1 tp1 0)
+
+estimated volumes (area) A→ 52 m3 TP → 5.76 m 3 D1 → 2.1375 m3
+
+```{r}
+library(dplyr)
+library(FactoMineR)
+library(tibble)
+
+# Define estimated area and feature frequency data
+area_data <- tibble(
+  area = c("d1", "ta", "tp1"),
+  estimated_area = c(2.1375, 52, 5.76),
+  feature_freq = c(1, 1, 0)
+)
+
+# Step 1: Number of artifacts by area
+artifact_counts <- tl_final_area %>%
+  group_by(area) %>%
+  tally()
+
+# Step 2: Calculate each variable
+
+# Retouch Frequency 
+retouch_freq <- tl_final_area %>%
+  group_by(area) %>%
+  summarise(retouchFreq = mean(!is.na(retouch)), .groups = 'drop')
+
+# Blank Frequency
+blank_freq <- tl_final_area %>%
+  group_by(area) %>%
+  summarise(blankFreq = mean(!is.na(blank)), .groups = 'drop')
+
+# Core Frequency
+core_freq <- tl_final_area %>%
+  group_by(area) %>%
+  summarise(coreFreq = mean(!is.na(`core-typology`)),.groups = 'drop')
+
+# Lithic density (Artifact counts / estimated area)
+lithic_density <- artifact_counts %>%
+  left_join(area_data, by = "area") %>%
+  mutate(lithic_density = n / estimated_area)
+
+# Tool Diversity using Menhinick's Index
+tool_div <- tl_final_area %>%
+  group_by(area) %>%
+  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(n()), .groups = 'drop')
+
+# Step 3: Combining all variables
+
+pca_final_table_area <- left_join(area_data, retouch_freq, by = "area") %>%
+  left_join(blank_freq, by = "area") %>%
+  left_join(core_freq, by = "area") %>%
+  left_join(lithic_density, by = "area") %>%
+  left_join(tool_div, by = "area") %>%
+  select(area, estimated_area, feature_freq, retouchFreq, blankFreq, coreFreq, lithic_density, toolDiv)
+
+```
+
+```{r}
+pca_final_table_area <- pca_final_table_area %>%
+    select(
+        area, 
+        estimated_area = estimated_area.x, 
+        feature_freq = feature_freq.x, 
+        retouchFreq, 
+        blankFreq, 
+        coreFreq, 
+        n, 
+        lithic_density, 
+        toolDiv
+    )
+
+```
+
+```{r}
+# Step 4: Perform PCA
+res_pca_area <- PCA(pca_final_table_area %>% select(-area, -n), graph = FALSE)
+
+# Show the summary of PCA
+print(res_pca_area)
+
+print(res_pca_area$eig)
+print(res_pca_area$var$coord)
+```
+
+```{r}
+# TABLE: Eigenvalues and percentage of variance for each dimension of PCA 
+library(kableExtra)
+
+# Define the values manually (you would get these from your PCA results)
+eigenvalues <- c(4.411234, 2.588766)
+prop_var_explained <- c(63.02, 36.98)
+cumulative_var_explained <- c(63.02, 100.00)
+
+# Create the data frame
+results_df <- data.frame(
+  Dimension = c("Dim. 1", "Dim. 2"),  # Only keep the dimensions you want to display
+  Eigenvalue = eigenvalues,
+  `Variance / percent` = prop_var_explained,
+  `Cumulative / variance percent` = cumulative_var_explained
+)
+
+# Generate the table
+kable(results_df,
+      caption = "Table 3. Eigenvalues and percentage of variance for each dimension of PCA",
+      align = 'c',
+      col.names = c("Dimension", "Eigenvalue", "Variance (%)", "Cumulative Variance (%)")) %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 11)
+
+```
+
+```{r}
+#plotting Dim-1 and Dim-2
+library(FactoMineR)
+library(factoextra)
+
+# Visualize the contributions of variables to the first principal component
+fviz_contrib(res_pca_area, axes = 1, choice = "var")
+
+# Visualize the contributions of variables to the second principal component
+fviz_contrib(res_pca_area, axes = 2, choice = "var")
+
+```
+
+```{r}
+#|fig-cap: Contribution of variables for each of the tow relevant PCA dimensions
+
+# having Dim1 and Dim2 plots next to each other
+
+library(gridExtra)
+
+# Create the first plot and assign it to p1
+p1 <- fviz_contrib(res_pca_area, axes = 1, choice = "var")
+
+# Create the second plot and assign it to p2
+p2 <- fviz_contrib(res_pca_area, axes = 2, choice = "var")
+
+# Use grid.arrange to place them next to each other
+grid.arrange(p1, p2, ncol = 2)
+
+```
+
+```{r}
+
+loadings <- res_pca_area$var$coord
+
+# 
+eigenvalues <- res_pca_area$eig
+
+#  Dim.1
+contrib_dim1 <- (loadings[,1]^2 / eigenvalues[1, 1]) * 100
+
+#  Dim.2
+contrib_dim2 <- (loadings[,2]^2 / eigenvalues[2, 1]) * 100
+
+contrib_df <- data.frame(
+  Variable = rownames(loadings),
+  Contrib_Dim1 = contrib_dim1,
+  Contrib_Dim2 = contrib_dim2
+)
+
+print(contrib_df)
+
+```
+
+```{r}
+# screeplot
+
+library(FactoMineR)
+library(factoextra)
+
+fviz_eig(res_pca_area)
+
+```
+
+```{r}
+# Correlation plots of variables for each of the tow relevant PCA dimensions
+
+fviz_pca_var(res_pca_area,
+
+             repel = TRUE) + 
+  theme(text = element_text(size = 10))
+
+```
+
+```{r}
+# Eigenvalues and percentage of variance
+eig.val <- res_pca_area$eig
+print(eig.val)
+
+# Variable coordinates (correlations with principal components)
+var.coord <- res_pca_area$var$coord
+print(var.coord)
+
+```
+
+```{r}
+#PCA biplot for Dimensions 1 and 2
+
+fviz_pca_biplot(res_pca_area, labelsize = 3, repel = TRUE)
+
+```
+
+```{r}
+# Create the initial biplot without individual labels
+p <- fviz_pca_biplot(res_pca_area, label = "none", 
+                     repel = TRUE, labelsize = 4)
+
+# Add the text labels for the areas using their coordinates from the PCA result
+p + geom_text(aes(x = res_pca_area$ind$coord[, 1], 
+                  y = res_pca_area$ind$coord[, 2], 
+                  label = pca_final_table_area$area), 
+             size = 3, nudge_x = 0.1, nudge_y = 0.1)
+
+```
+
+```{r}
+individual_coordinates = res_pca_area$ind$coord
+variable_coordinates = res_pca_area$var$coord
+
+```
+
+```{r}
+# For variables
+fviz_cos2(res_pca_area, choice = "var")
+
+# For individuals
+fviz_cos2(res_pca_area, choice = "ind")
+
+```
+
+DONE WITH PCA
+
+```{r}
+
+individual_coordinates <- res_pca_area$ind$coord
+
+individual_coordinates_df <- as.data.frame(individual_coordinates)
+individual_coordinates_df$area <- unique(area_data$area)
+
+print(individual_coordinates_df)
+
+```
+
+```{r}
+library(kableExtra)
+
+# Extract individual coordinates
+individual_coordinates <- res_pca_area$ind$coord
+
+# Convert to a data frame for better visibility and add the 'area' column back for context
+individual_coordinates_df <- as.data.frame(individual_coordinates)
+individual_coordinates_df$Area <- unique(area_data$area)  # Make sure the order matches
+
+# Use kable to create a beautiful table
+kable(individual_coordinates_df, 
+      col.names = c("Area", "Principal Component 1", "Principal Component 2"), 
+      caption = "PCA Coordinates for the Archaeological Areas excavation",
+      align = c('l', 'c', 'c'),
+      digits = 3) %>%
+  kable_styling("striped", full_width = F)
+
+```
+
+```{r}
+# Get the loadings (coefficients) of original variables for each principal component
+loadings_matrix <- res_pca$var$coord
+
+```
+
+```{r}
+# Eigenvalues
+eigenvalues <- res_pca$eig[1:2, 1]
+variance_percent <- res_pca$eig[1:2, 2]
+cumulative_variance_percent <- res_pca$eig[1:2, 3]
+
+# Create the table
+df <- data.frame(Eigenvalue = eigenvalues,
+                 Variance_Percent = variance_percent,
+                 Cumulative_Variance_Percent = cumulative_variance_percent)
+
+rownames(df) <- c("Dim1", "Dim2")
+
+# Print the table
+print(df)
+
+```
+
+```{r}
+library(gt)
+
+gt(df) %>%
+  tab_header(
+    title = "",
+    subtitle = "Eigenvalues and percentage of variance for each dimension of PCA"
+  ) %>%
+  cols_label(
+    Eigenvalue = "Eigenvalue",
+    Variance_Percent = "Variance (%)",
+    Cumulative_Variance_Percent = "Cumulative Variance (%)"
+  ) %>%
+  fmt_number(
+    columns = c("Eigenvalue", "Variance_Percent", "Cumulative_Variance_Percent"),
+    decimals = 2
+  )
+
+```
+
+```{r}
+
+artifact_counts <- tl_final %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  tally()
+
+# Print the resulting table
+print(artifact_counts)
+```
+
+```{r}
+
+blank_freq <- tl_final_area %>%
+  filter(area %in% c("ta", "d1", "tp1"), !is.na(blank)) %>%
+  group_by(area) %>%
+  summarise(blankFreq = sum(blank %in% c("bladelet", "blade", "flake")) / sum(!is.na(blank)), .groups = 'drop')
+
+# Calculating Core Frequency
+core_freq <- tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1"), !is.na(corefrag)) %>%
+  group_by(area) %>%
+  summarise(coreFreq = sum(corefrag %in% c("core fragment", "bullet", "flat", "pyramid", "cylinder/prismatic", "core rejuvenation piece", "shapeless", "core on flake", "bipolar")) / sum(!is.na(corefrag)), .groups = 'drop')
+
+# Calculating Tool Diversity using Menhinick's Index
+tool_div <- tl_final_area %>%
+  filter(area %in% c("ta", "d1", "tp1"), !is.na(tooltypegroup)) %>%
+  group_by(area) %>%
+  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(sum(!is.na(tooltypegroup))), .groups = 'drop')
+
+# Combining all the calculated frequencies into one tibble
+final_table <- blank_freq %>%
+  inner_join(core_freq, by = "area") %>%
+  inner_join(tool_div, by = "area")
+
+# Printing the final table
+print(final_table)
+
+```
+
+```{r}
+# Joining calculated frequencies
+final_table <- blank_freq %>%
+  inner_join(core_freq, by = "area") %>%
+  inner_join(tool_div, by = "area") %>%
+  inner_join(retouch_freq, by = "area")
+
+# Converting area to row names
+final_table_rownames <- final_table %>%
+  column_to_rownames(var = "area")
+
+```
+
+```{r}
+
+library(FactoMineR)
+library(factoextra)
+
+# Run PCA
+res_pca <- PCA(final_table_rownames, graph = FALSE)
+
+# Visualize the contributions of variables to the second principal component
+fviz_contrib(res_pca, axes = 2, choice = "var")
+
+```
+
+```{r}
+# Visualize the contributions of variables to the first principal component
+fviz_contrib(res_pca, axes = 1, choice = "var") 
+```
+
+```{r}
+# Visualize eigenvalues and explained variance
+fviz_screeplot(res_pca)
+```
+
+```{r}
+fviz_pca_biplot(res_pca, labelsize = 3, repel = TRUE)
+
+```
+
+```{r}
+# Visualize the variables on the principal components
+fviz_pca_var(res_pca)
+```
+
+```{r}
+# Get the loadings (coefficients) of original variables for each principal component
+loadings_matrix <- res_pca$var$coord
+
+# Print the loadings matrix
+print(loadings_matrix)
+
+```
+
+```{r}
+# Eigenvalues
+eigenvalues <- res_pca$eig[1:2, 1]
+variance_percent <- res_pca$eig[1:2, 2]
+cumulative_variance_percent <- res_pca$eig[1:2, 3]
+
+# Create the table
+df <- data.frame(Eigenvalue = eigenvalues,
+                 Variance_Percent = variance_percent,
+                 Cumulative_Variance_Percent = cumulative_variance_percent)
+
+rownames(df) <- c("Dim1", "Dim2")
+
+# Print the table
+print(df)
+
+```
+
+```{r}
+library(gt)
+
+# Create the table 
+gt(df) %>%
+  tab_header(
+    title = "",
+    subtitle = "Eigenvalues and percentage of variance for each dimension of PCA"
+  ) %>%
+  cols_label(
+    Eigenvalue = "Eigenvalue",
+    Variance_Percent = "Variance (%)",
+    Cumulative_Variance_Percent = "Cumulative Variance (%)"
+  ) %>%
+  fmt_number(
+    columns = c("Eigenvalue", "Variance_Percent", "Cumulative_Variance_Percent"),
+    decimals = 2
+  )
+
+```
+
+```{r}
+# Count the number of artifacts by area
+artifact_counts <- tl_final %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  tally()
+
+# Print the resulting table
+print(artifact_counts)
+
+```
+
+```{r}
+
+volumes <- data.frame(
+  area = c("ta", "tp1", "d1"),
+  volume = c(52, 5.76, 2.1375)
+)
+
+# Count the number of artifacts by area
+artifact_counts <- tl_final %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  tally()
+
+# Join artifact_counts with volumes by area
+lithic_density_df <- left_join(artifact_counts, volumes, by = "area")
+
+# Compute lithic density for each area
+lithic_density_df <- lithic_density_df %>%
+  mutate(lithic_density = n / volume)
+
+# Print the resulting table
+print(lithic_density_df[, c("area", "lithic_density")])
+
+```
+
+```{r}
+# Counting the number of cores by area
+core_counts <- tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1") & !is.na(`core and core fragments`)) %>%
+  group_by(area) %>%
+  tally()
+
+print(core_counts)
+
+```
+
+OK
+
+```{r}
+#| label: fig-1-layer
+#| fig-cap: "Type of Removals"
+#| fig-height: 5
+
+#Type of blank based on Layers
+
+library(ggbeeswarm)
+
+tl4 %>% 
+  filter(!is.na(layer) & Blank2 %in% c('bladelet', 
+                                       'blade', 
+                                       'flake')) %>%
+
+  # Create bar plot
+  ggplot() +
+  aes(x = layer, 
+      fill = Blank2) +
+  geom_bar(position = "dodge") +
+  labs(x = "Area", 
+       y = "Number", 
+       fill = "Blank Type") +
+  theme_minimal()
+
+```
+
+```{r}
+#Type of core based on Layers
+
+library(ggbeeswarm)
+
+tl4 %>% 
+  filter(!is.na(layer) & 
+           `core-typology` %in% c('bullet','prismatic','pyramid')) %>%
+
+  # Create bar plot
+  ggplot() +
+  aes(x = layer, 
+      fill = `core-typology`) +
+  geom_bar(position = "dodge") +
+  labs(x = "Area", 
+       y = "Number", 
+       fill = "Blank Type") +
+  theme_minimal()
+```
+
 consistency in dimention:
+
 ```{r}
@@
 
-1.   (formal + informal) to debitages
-2.  formal  vs. informal
-3.  shine/ sickle  as a  used for processing plants
-
 ```{r}
@@
 ```
-
-WABI_AL
-Ta-e → 13 m3
-Ta-d → 45 m3
-D1-c → 0.675 m3
-Tp-c → 1.44 m3
-D1-a → 1.575 m3
-
-```{r}
-# Create a tibble with example data
-summary_table2_area <- tibble(
-  area = c("D1_a", "D1_c", "TA_d", "TA_e", "TP1_c"),
-  lithic_volumetric_density = c(51.2, 110.9, 45.1, 100.2, 88.3),
-  percent_retouch = c(4.9, 9.3, 15.3, 18.0, 9.8)
-)
-
-```
-
-```{r}
-# Load necessary libraries
-# Import required packages
-library(ggplot2)
-library(ggpmisc)
-
-# Create the data frame based on the example table
-summary_table2_area <- data.frame(
-  area = c("D1_a", "D1_c", "TA_d", "TA_e", "TP1_c"),
-  lithic_volumetric_density = c(51.43, 111.11, 45.13, 100.08, 212.50),
-  retouch_frequency = c(0.05, 0.09, 0.07, 0.07, 0.10)
-)
-
-# Generate the regression plot
-p <- ggplot(summary_table2_area, aes(x = lithic_volumetric_density, y = retouch_frequency)) +
-
-  # Blue regression line with shaded area
-  geom_smooth(method = "lm", 
-              se = TRUE, 
-              aes(group = 1), 
-              color = "blue") +
-
-  # Points colored by area
-  geom_point(color = "black") +
-
-  # Labels points with area names
-  geom_text(aes(label = area),
-            vjust = -1, 
-            hjust = -0.5) +
-
-  # Regression equation, R-squared, and p-value
-  stat_poly_eq(aes(label = paste(stat(eq.label), stat(rr.label), stat(p.value.label), sep = "~~~")),
-               formula = y ~ x,
-               parse = TRUE,
-               size = 3) +
-
-  # Titles and labels
-  labs(title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
-       x = "Lithic Volumetric Density",
-       y = "Retouch Frequency")
-
-print(p)
-
-```
-
-WABI_AREA
-
-```{r}
-# regression plot-area
-library(ggplot2)
-library(ggpmisc)
-
-# Plotting
-p <- ggplot(summary_table2_area, aes(x = lithic_volumetric_density, y = percent_retouch)) +
-
-  # Blue regression line with shaded area
-  geom_smooth(method = "lm", 
-              se = TRUE, 
-              aes(group = 1), 
-              color = "blue") +
-
-  # Points colored by area
-  geom_point(color = "black") +
-
-  # Labels points with area names
-  geom_text(aes(label = area),
-            vjust = -1, 
-            hjust = -0.5) +
-
-  # Regression equation, R-squared, and p-value
-  stat_poly_eq(aes(label = paste(stat(eq.label), stat(rr.label), stat(p.value.label), sep = "~~~")),
-               formula = y ~ x,
-               parse = TRUE,
-               size = 3) +
-
-  # Titles and labels
-  labs(title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
-       x = "Lithic Volumetric Density",
-       y = "Retouch Frequency (%)")
-
-print(p)
-
-```
-
-```{r}
-# Fit the linear model
-linear_model <- lm(percent_retouch ~ lithic_volumetric_density, data = summary_table2_area)
-
-# Display the summary of the model to get coefficients, R-squared, and p-value
-summary_linear_model <- summary(linear_model)
-
-# Extract the coefficients (intercept and slope)
-coefficients <- summary_linear_model$coefficients
-
-# Extract the R-squared value
-r_squared <- summary_linear_model$r.squared
-
-# Extract the p-value for the slope (lithic_volumetric_density)
-p_value_slope <- coefficients["lithic_volumetric_density", "Pr(>|t|)"]
-
-cat("Coefficients:\n")
-print(coefficients)
-
-cat("\nR-squared value:", r_squared)
-
-cat("\nP-value for slope:", p_value_slope)
-
-```
-
-```{r}
-#whole assemblage
-
-  # Summarize data by Area
-summary_table_tech <- tl_final_area %>%
-  group_by(area) %>%
-
-  summarise(
-   blade_count = sum(blank == "blade", na.rm = TRUE),
-    flake_count = sum(blank == "flake", na.rm = TRUE),
-    bladelet_count = sum(blank == "bladelet", na.rm = TRUE),
-
-    bullet_core_count = sum(`core-typology` == "bullet", na.rm = TRUE),
-    flat_core_count = sum(`core-typology` == "flat", na.rm = TRUE),
-    pyramid_core_count = sum(`core-typology` == "pyramid", na.rm = TRUE),
-    cylinder_core_count = sum(`core-typology` %in% c("prismatic", "cylinder"), na.rm = TRUE),
-    shapeless_core_count = sum(`core-typology` == "shapeless", na.rm = TRUE),
-
-    core_fragment_count = sum(`corefrag` == "core fragment", na.rm = TRUE),
-    core_rejuvention_count = sum(`corefrag` == "rejuvenation piece", na.rm = TRUE),
-    core_tablet_count = sum(`corefrag` == "core tablet", na.rm = TRUE),
-    core_crested_count = sum(`corefrag` == "crested bladelet", na.rm = TRUE),
-    core_primarypiece_count = sum(`corefrag`  %in% c("primary flake", "primary blade"), na.rm = TRUE),
-
-   total = n()
-  )
-```
-
-```{r}
-#plot_blank_type
-# Libraries
-library(dplyr)
-library(ggplot2)
-library(ggbeeswarm)
-
-# Data filtering and plotting
-tl_final_area %>% 
-  filter(blank %in% c('bladelet', 'blade', 'flake')) %>%
-  ggplot(aes(x = area, fill = blank)) +
-  geom_bar(position = "dodge") +
-  labs(
-    x = "Area",
-    y = "Number",
-    fill = "Blank Type"
-  ) +
-  theme_minimal()
-
-```
-
-```{r}
-
-library(dplyr)
-library(ggplot2)
-
-# Filter the data for the relevant categories
-tl4_filtered <- tl4_tool_Blank %>%
-  filter(Blank2 %in% 
-           c("bladelet", 
-             "blade", 
-             "flake"))
-
-# Create separate bar plots for shine, Retouch, and Utilization
-plot_shine2 <- ggplot(tl4_filtered, aes(x = Blank2, fill = shine2)) +
-  geom_bar(position = "dodge") +
-  labs(title = "Shine2 by Blank2",
-       x = "Blank2",
-       y = "Frequency",
-       fill = "Shine2") +
-  theme_minimal()
-
-plot_retouch <- ggplot(tl4_filtered, aes(x = Blank2, fill = Retouch)) +
-  geom_bar(position = "dodge") +
-  labs(title = "Retouch by Blank2",
-       x = "Blank2",
-       y = "Frequency",
-       fill = "Retouch") +
-  theme_minimal()
-
-plot_utilization <- ggplot(tl4_filtered, aes(x = Blank2, fill = Utilization)) +
-  geom_bar(position = "dodge") +
-  labs(title = "Utilization by Blank2",
-       x = "Blank2",
-       y = "Frequency",
-       fill = "Utilization") +
-  theme_minimal()
-
-# Combine the three plots side by side
-library(patchwork)
-combined_plots <- plot_shine2 + plot_retouch + plot_utilization
-combined_plots
-
-```
-
-```{r}
-library(dplyr)
-library(knitr)
-library(kableExtra)
-
-# Assuming your data is stored in the variable tl4_tool_Blank
-# Filter the data to keep only bladelet, blade, and flake in the Blank column
-tl_final_aread <- tl_final_area %>%
-  filter(blank %in% c("bladelet", "blade", "flake"))  %>%
-
-# Group the data by Blank2 and calculate the counts for each combination of Utilization, shine2, and Retouch
-
-  group_by(blank, utilization, `sickle shine`, retouch) %>%
-  summarise(count = n())
-
-# Print the table using kable() and kableExtra for formatting
-table_kable <- kable(table_summary, format = "html", caption = "Summary Table", 
-                     col.names = c("Blank2", "Utilization", "shine2", "Retouch", "Count")) %>%
-  kable_styling(bootstrap_options = "striped")
-
-# Print the table
-print(table_kable)
-
-```
-
-```{r}
-#|label: fig_
-#|fig-cap: "Boxplot for each Dimention"
-#|fig-height: 5
-
-library(dplyr)
-library(ggplot2)
-
-# Filter the data to keep only bladelet, blade, and flake in the Blank column
-tl4_tool_Blank_filtered <- tl4_tool_Blank %>%
-  filter(Blank2 %in% c("bladelet", "blade", "flake"))
-
-tl4_tool_Blank_tally <- 
-  tl4_tool_Blank_filtered %>%
-  group_by(Blank2) %>%
-  tally()
-
-# Select dimensions and convert Thickness (mm) to numeric
-tl4_tool_Blank_selected <- 
-  tl4_tool_Blank_filtered %>% 
-  select(Blank2, 
-         Utilization, 
-         Retouch, 
-         shine2)
-
-# Plot
-# Pivot data to longer format
-tl4_tool_Blank_long <- 
-  tl4_tool_Blank_selected %>% 
-  pivot_longer(cols = c(Utilization, Retouch, shine2), 
-               names_to = "Blank2", 
-               values_to = "Value")
-
-# Create boxplot
-ggplot(tl4_tool_Blank_long, aes(x = Blank2, y = Value)) +
-  geom_boxplot() +
-  facet_wrap(~Blank2, scales = "free") +
-  labs(x = "Blank", 
-       y = "Frequency") +
-  theme_minimal()
-
-```
-
-```{r}
-library(dplyr)
-library(ggplot2)
-
-# Filter the data to keep only bladelet, blade, and flake in the Blank2 column
-tl4_tool_Blank_filtered <- tl4_tool_Blank %>%
-  filter(Blank2 %in% c("bladelet", "blade", "flake"))
-
-tl4_tool_Blank_tally <- 
-  tl4_tool_Blank_filtered %>%
-  group_by(Blank2) %>%
-  tally()
-
-# Select dimensions and convert Thickness (mm) to numeric
-tl4_tool_Blank_selected <- 
-  tl4_tool_Blank_filtered %>% 
-  select(Blank2, 
-         Utilization, 
-         Retouch, 
-         shine2)
-
-# Manually rename the duplicate columns
-names(tl4_tool_Blank_selected)[names(tl4_tool_Blank_selected) == "Blank2"] <- "Dimension"
-
-# Pivot data to longer format
-tl4_tool_Blank_long <- 
-  tl4_tool_Blank_selected %>% 
-  pivot_longer(cols = c(Utilization, Retouch, shine2), 
-               names_to = "", 
-               values_to = "Value")
-
-# Create boxplot
-ggplot(tl4_tool_Blank_long, aes(x = Dimension, y = Value)) +
-  geom_boxplot() +
-  facet_wrap(~, scales = "free") +
-  labs(x = "Blank", 
-       y = "Frequency") +
-  theme_minimal()
-
-```
-
-```{r}
-
-# Load required packages
-library(ggplot2)
-
-# Filter the data to exclude the value 0 and keep only blade, bladelet, and flake
-tl4_filtered <- tl4_tool_Blank %>%
-  filter(Blank2 %in% c("blade", "bladelet", "flake"))
-
-# Create the plot with facetting
-tl4_plots <- ggplot(tl4_filtered, aes(x = Blank2, fill = Blank2)) +
-  geom_bar(position = "dodge") +
-  labs(title = "Utilization, Retouch, and shine2 by Blank",
-       x = "Blank2",
-       y = "Count") +
-  facet_grid(. ~ Blank2, scales = "free") +
-  theme_minimal()
-
-# Print the plot
-print(tl4_plots)
-
-```
-
-```{r}
-# Load required packages
-library(ggplot2)
-library(cowplot)
-
-# Filter the data to exclude the value 0 and keep only blade, bladelet, and flake
-tl4_filtered <- tl4_tool_Blank %>%
-  filter(Blank2 %in% c("blade", "bladelet", "flake"))
-
-# Plot Utilization by Blank2
-plot_utilization <- ggplot(tl4_filtered, aes(Blank2, Utilization)) +
-  geom_bar(stat = "identity", position = "dodge") +
-  labs(title = "Utilization by Blank",
-       x = "Blank2",
-       y = "Utilization") +
-  theme_minimal()
-
-# Plot Retouch by Blank
-plot_retouch <- ggplot(tl4_filtered, aes(Blank2, Retouch)) +
-  geom_bar(stat = "identity", position = "dodge") +
-  labs(title = "Retouch by Blank",
-       x = "Blank2",
-       y = "Retouch") +
-  theme_minimal()
-
-# Plot shine by Blank
-plot_shine <- ggplot(tl4_filtered, aes(x = Blank2, y = shine2)) +
-  geom_bar(stat = "identity", position = "dodge") +
-  labs(title = "shine by Blank",
-       x = "Blank2",
-       y = "shine2") +
-  theme_minimal()
-
-# Combine the plots using cowplot or patchwork
-plot_combined <- cowplot::plot_grid(plot_utilization, plot_retouch, plot_shine, ncol = 3)
-
-# Print the combined plot
-print(plot_combined)
-
-```
-
-```{r}
-
-#|label: fig_
-#|fig-cap:  vs Debitage
-#|fig-height: 5
-
-# Filter data
-tl4__blank %>% 
-  filter(Blank2 %in% c('tp1', 'ta', 'd1')) %>% 
-  select(Blank2, 
-         `Width (mm)`, 
-         `Thickness (mm)`, 
-         `Length (mm)`) %>%
-mutate(`Thickness (mm)` = as.numeric(`Thickness (mm)`))
-
-# Pivot data to longer format
-long_data <- tl3_dimention_blank %>% 
-  pivot_longer(cols = c(`Width (mm)`, `Thickness (mm)`, `Length (mm)`), 
-               names_to = "Dimension", 
-               values_to = "Value")
-
-# Create boxplot
-ggplot(long_data, aes(x = area, y = Value)) +
-  geom_boxplot() +
-  geom_quasirandom(alpha = 0.01) +
-  facet_wrap(~Dimension, scales = "free_y") +
-  scale_y_log10() +
-  theme_minimal()
-
-```
-
-```{r}
-library(ggbeeswarm)
-tl3_dimention_blank <- tl3 %>% 
-  filter(area %in% tl3_area_tally_greater_than_n$area) %>% 
-  pivot_longer(-area) %>% 
-  ggplot() +
-  aes(area,
-      value) +
-  geom_boxplot() +
-  geom_quasirandom(alpha = 0.01) +
-  facet_wrap( ~ name,
-              scales = "free_y") +
-  scale_y_log10() +
-  theme_minimal()
-
-```
-
-```{r}
-library(ggbeeswarm)
-tl3 %>% 
-  filter(area %in% tl3_area_tally_greater_than_n$area) %>% 
-  pivot_longer(-area) %>% 
-  ggplot() +
-  aes(area,
-      value) +
-  geom_boxplot() +
-  geom_quasirandom(alpha = 0.01) +
-  facet_wrap( ~ name,
-              scales = "free_y") +
-  scale_y_log10() +
-  theme_minimal()
-
-```
-
-```{r}
-#| label: fig-Width-of-Artifacts-Area-A-by-depth
-#| fig-cap: "Width of Artifacts by depth (Area A)"
-#| fig-height: 5
-tl %>%
-  filter(AREA == "A") %>%
-  select(
-         `Width (mm)`,
-         `DEPTH (cm)`) %>%
-  ggplot() +
-  aes(x = `DEPTH (cm)`,
-      y = `Width (mm)`) +
-  geom_boxplot() +
-  theme_minimal()
-```
-
-```{r}
-#| label: Width of Artifacts (Area A)
-#| fig-cap: "Width of Artifacts (Area A)"
-#| fig-height: 5
-
-tl1 %>%
-  filter(AREA == "TP1") %>%
-  select(AREA, `Width (mm)`) %>%
-  ggplot() +
-  aes(x = `Width (mm)`) +
-  geom_histogram(binwidth = 2) +
-  labs(x = "Width of Debitage (mm)", 
-       y = "Frequency", 
-       title = "Histogram of Width for Area TP1" )+
-  theme_minimal()
-```
-
-WABI: 
-```{r}
-# regression plot-layer
-library(ggplot2)
-library(ggpmisc)
-
-# Plotting
-p <- ggplot(summary_table2_layer, 
-            aes(lithic_volumetric_density, 
-                percent_retouch)) +
-
-  geom_smooth(method = "lm", 
-              se = TRUE, 
-              aes(group = 1), 
-              color = "blue") +
-
-  # Points colored by layer
-  geom_point(color = "black") +
-
-  # Labels points with layer names
-  geom_text(aes(label = layer),
-            vjust = -1, 
-            hjust = -0.5) +
-
-  # Regression equation, R-squared, and p-value
-  stat_poly_eq(aes(label = paste(stat(eq.label), stat(rr.label), stat(p.value.label), sep = "~~~")),
-               formula = y ~ x,
-               parse = TRUE,
-               size = 3) +
-
-  # Titles and labels
-  labs(title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
-       x = "Lithic Volumetric Density",
-       y = "Retouch Frequency (%)")
-
-print(p)
-```
-
-```{r}
-#barplot_layer
-
-tl_final_area %>%
-  filter(layer %in% c("a", "c", "d", "e", "eor d")) %>%
-  filter(!is.na(corefrag)) %>% 
-  select(layer, corefrag) %>%
-  ggplot() +
-  aes(x = corefrag) +
-  geom_bar() +  # Use geom_bar for categorical data
-  labs(x = "Core and Core fragments", y = "Frequency") +
-  facet_wrap(~layer, ncol = 1) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-```
-
-```{r}
-#Core-barplot_area
-
-tl_final %>%
-  filter(area %in% c("ta", "d1", "tp1")) %>%
-  filter(!is.na(corefrag)) %>% # Exclude rows where corefrag is NA
-  select(area, corefrag) %>%
-  ggplot() +
-  aes(x = corefrag) +
-  geom_bar() +
-  labs(x = "Core and Core fragments",
-       y = "Frequency") +
-  facet_wrap(~area, ncol = 1) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-```
-
-```{r}
-
-library(ggplot2)
-library(dplyr)
-filtered_data <- tl_final %>%
-  filter(!is.na(`tool-typology`) & `tool-typology` != "" & `tool-typology` != "NA")
-ggplot(data = filtered_data) + 
-  aes(y = `tool-typology`) + 
-  geom_bar() +
-  labs(x = "Frequency",
-       y = "Tool Type") +
-  theme_minimal() +
-  theme(axis.text.y = element_text(angle = 20, hjust = 1, vjust = 0.5))
-
-```
-
-```{r}
-
-filtered_data <- tl_final %>%
-  filter(!is.na(``) & `` != "")
-
-# Create the plot using the filtered data
-ggplot(data = filtered_data) + 
-  aes(y = ``) + 
-  geom_bar() +
-  labs(y = "Tool Type",
-       x = "Frequency") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-```
-
-```{r}
-
-```
-
-PCA-LAYER
-
-```{r}
-#prepare the data for PCA-layer
-
-# Calculating Blank Frequency
-blank_freq <- tl_final_area %>%
-  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(blank)) %>%
-  group_by(layer) %>%
-  summarise(blankFreq = sum(blank %in% c("bladelet", "blade", "flake")) / sum(!is.na(blank)), .groups = 'drop')
-
-# Calculating Core Frequency
-core_freq <- tl_final_area %>%
-  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(corefrag)) %>%
-  group_by(layer) %>%
-  summarise(coreFreq = sum(corefrag %in% c("core fragment", "bullet", "flat", "pyramid", "cylinder/prismatic", "core rejuvenation piece", "shapeless", "core on flake", "bipolar")) / sum(!is.na(corefrag)), .groups = 'drop')
-
-# Calculating Tool Diversity using Menhinick's Index
-tool_div <- tl_final %>%
-  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(tooltypegroup)) %>%
-  group_by(layer) %>%
-  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(sum(!is.na(tooltypegroup))), .groups = 'drop')
-
-# Combining all the calculated frequencies into one tibble
-final_table <- blank_freq %>%
-  inner_join(core_freq, by = "layer") %>%
-  inner_join(tool_div, by = "layer")
-
-# Printing the final table
-print(final_table)
-
-```
-
-```{r}
-# Calculating Blank Frequency
-blank_freq <- tl_final %>%
-  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(blank)) %>%
-  group_by(layer) %>%
-  summarise(blankFreq = sum(blank %in% c("bladelet", "blade", "flake")) / sum(!is.na(blank)), .groups = 'drop')
-
-# Calculating Core Frequency
-core_freq <- tl_final %>%
-  filter(layer %in% c("a", "c", "d", "e", "e or d")) %>%
-  group_by(layer) %>%
-  summarise(coreFreq = sum(corefrag %in% c("core fragment", "bullet", "flat", "pyramid", "cylinder/prismatic", "core rejuvenation piece", "shapeless", "core on flake", "bipolar")) / n(), .groups = 'drop')
-
-# Calculating Tool Diversity using Menhinick's Index
-tool_div <- tl_final %>%
-  filter(layer %in% c("a", "c", "d", "e", "e or d"), !is.na(tooltypegroup)) %>%
-  group_by(layer) %>%
-  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(sum(!is.na(tooltypegroup))), .groups = 'drop')
-
-# Calculating Retouch Frequency
-retouch_freq <- tl_final %>%
-  filter(layer %in% c("a", "c", "d", "e", "eor d")) %>%
-  group_by(layer) %>%
-  summarise(retFreq = sum( == "retouched piece", na.rm = TRUE) / n(), .groups = 'drop')
-
-# Combining all the calculated frequencies, including retouch frequency, into one tibble
-final_table <- blank_freq %>%
-  inner_join(core_freq, by = "layer") %>%
-  inner_join(tool_div, by = "layer") %>%
-  inner_join(retouch_freq, by = "layer")
-
-# Printing the final table
-print(final_table)
-
-```
-
-```{r}
-# Joining calculated frequencies
-final_table <- blank_freq %>%
-  inner_join(core_freq, by = "layer") %>%
-  inner_join(tool_div, by = "layer") %>%
-  inner_join(retouch_freq, by = "layer")
-
-# Converting layer to row names
-final_table_rownames <- final_table %>%
-  column_to_rownames(var = "layer")
-
-```
-
-```{r}
-# Compute the PCA
-library(FactoMineR)
-
-res_pca <-
-  PCA(final_table_rownames,
-      graph = FALSE)
-
-```
-
-```{r}
-library(factoextra)
-
-# Visualize the contributions of variables to the first principal component
-fviz_contrib(res_pca,
-             axes = 1,   
-             choice = "var")
-
-```
-
-```{r}
-library(factoextra)
-
-# Visualize the contributions of variables to the second principal component
-fviz_contrib(res_pca,
-             axes = 2,   # Selecting the second axis
-             choice = "var") # Choosing to show contributions of variables
-
-```
-
-```{r}
-library(factoextra)
-
-# Creating a biplot
-fviz_pca_biplot(res_pca)
-```
-
-```{r}
-library(factoextra)
-fviz_screeplot(res_pca)
-```
-
-```{r}
-fviz_pca_var(res_pca)
-```
-
-```{r}
-
-library(FactoMineR)
-library(factoextra)
-
-# Run PCA
-res_pca <- PCA(final_table_rownames, graph = FALSE)
-
-# Visualize the contributions of variables to the first principal component
-fviz_contrib(res_pca, axes = 1, choice = "var")
-
-# Visualize the contributions of variables to the second principal component
-fviz_contrib(res_pca, axes = 2, choice = "var")
-
-# Visualize eigenvalues and explained variance
-fviz_screeplot(res_pca)
-
-# Create a biplot of the PCA
-fviz_pca_biplot(res_pca)
-
-# Visualize the variables on the principal components
-fviz_pca_var(res_pca)
-
-```
-
-```{r}
-# Load necessary libraries
-library(FactoMineR)
-library(factoextra)
-
-# Run PCA
-res_pca <- PCA(final_table_rownames, graph = FALSE)
-
-# Get the loadings (coefficients) of original variables for each principal component
-loadings_matrix <- res_pca$var$coord
-
-# Print the loadings matrix
-print(loadings_matrix)
-
-```
-
-```{r}
-#Table
-
-# Eigenvalues
-eigenvalues <- res_pca$eig[1:2, 1]
-
-# Variance percent
-variance_percent <- res_pca$eig[1:2, 2]
-
-# Cumulative variance percent
-cumulative_variance_percent <- res_pca$eig[1:2, 3]
-
-# Create the table
-df <- data.frame(Eigenvalue = eigenvalues,
-                 Variance_Percent = variance_percent,
-                 Cumulative_Variance_Percent = cumulative_variance_percent)
-
-rownames(df) <- c("Dim1", "Dim2")
-
-# Print the table
-print(df)
-
-```
-
-```{r}
-#Table
-
-library(gt)
-
-# Create data frame
-df <- data.frame(Eigenvalue = eigenvalues,
-                 Variance_Percent = variance_percent,
-                 Cumulative_Variance_Percent = cumulative_variance_percent)
-rownames(df) <- c("Dim1", "Dim2")
-
-# Create the table using gt
-gt(df) %>%
-  tab_header(
-    title = "",
-    subtitle = "Eigenvalues and percentage of variance for each dimension of PCA"
-  ) %>%
-  cols_label(
-    Eigenvalue = "Eigenvalue",
-    Variance_Percent = "Variance (%)",
-    Cumulative_Variance_Percent = "Cumulative Variance (%)"
-  ) %>%
-  fmt_number(
-    columns = vars(Eigenvalue, Variance_Percent, Cumulative_Variance_Percent),
-    decimals = 2
-  )
-
-```
-
-calculating PCA based on AREA: TA/ TP1/ D1:
-
-```{r}
-#4lithic density
-
-# A→ V=4×5×2.6 = 52 m3
-# TP → V=2×1.8×1.6 = 5.76 m 3
-# D1 → 1.5×1.5×0.95 = 2.1375 m3
-
-library(dplyr)
-
-# Define the volumes for each area based on the given information
-volumes <- data.frame(
-  area = c("ta", "tp1", "d1"),
-  volume = c(52, 5.76, 2.1375) # in m^3
-)
-
-# Count the number of artifacts by area
-artifact_counts <- tl_final_area %>%
-  filter(area %in% c("ta", "tp1", "d1")) %>%
-  group_by(area) %>%
-  tally()
-
-# Join artifact_counts with volumes by area
-lithic_density_df <- left_join(artifact_counts, volumes, by = "area")
-
-# Compute lithic density for each area
-lithic_density_df <- lithic_density_df %>%
-  mutate(lithic_density = n / volume) # number of artifacts / volume
-
-# Print the resulting table
-print(lithic_density_df[, c("area", "lithic_density")])
-
-```
-
-```{r}
-
-```
-
-```{r}
-
-#2 Calculate the frequency of cores and core fragments by area
-
-library(dplyr)
-# Calculate the core frequency
-core_frequency_by_area <- left_join(number_of_cores_by_area, 
-                                    total_lithics_by_area, 
-                                    by = "area") %>%
-  mutate(core_frequency = (number_of_cores / total_lithics) * 100)
-
-# Display the result
-print(core_frequency_by_area)
-
-```
-
-```{r}
-# 1number of Blank
-
-# Calculate the total number of lithic artifacts
-total_lithics <- tl_final_area %>%
-  summarise(total_lithics = n()) %>%
-  pull(total_lithics)
-
-# Display the total number of lithic artifacts
-print(paste("The total number of lithic artifacts is:", total_lithics))
-
-```
-
-```{r}
-# 5Calculate the retouched frequency for each area
-library(dplyr)
-
-total_lithics_by_area <- tl_final_area %>%
-  group_by(area) %>%
-  summarise(total_lithics = n())
-
-number_of_retouched_by_area <- tl_final_area %>%
-  mutate(retouch = as.numeric(`retouch`)) %>% 
-  filter(!is.na(retouch), retouch == 1) %>%  
-  group_by(area) %>%  
-  summarise(number_of_retouched = sum(retouch))
-
-retouched_frequency_by_area <- left_join(number_of_retouched_by_area, total_lithics_by_area, by = "area") %>%
-  mutate(retouched_frequency = (number_of_retouched / total_lithics) * 100)
-
-# Display the retouched frequency by area
-print(retouched_frequency_by_area)
-
-print(number_of_retouched_by_area)
-
-```
-
-```{r}
-# 3Calculating Tool Diversity using Menhinick's Index
-tool_div <- tl_final_area %>%
-  filter(area %in% c("ta", "d1", "tp1"), !is.na(tooltypegroup)) %>%
-  group_by(area) %>%
-  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(sum(!is.na(tooltypegroup))), .groups = 'drop')
-```
-
-```{r}
-#feature frequency: 
-d1     1 
-ta    1
-tp1 	0)
-
-```
-
-```{r}
-#estimated area
-A→  52 m3
-TP → 5.76 m 3
-D1 → 2.1375 m3
-```
-
-```{r}
-library(dplyr)
-library(FactoMineR)
-library(tibble)
-
-# Define estimated area and feature frequency data
-area_data <- tibble(
-  area = c("d1", "ta", "tp1"),
-  estimated_area = c(2.1375, 52, 5.76),
-  feature_freq = c(1, 1, 0)
-)
-
-# Step 1: Number of artifacts by area
-artifact_counts <- tl_final_area %>%
-  group_by(area) %>%
-  tally()
-
-# Step 2: Calculate each variable
-
-# Retouch Frequency 
-retouch_freq <- tl_final_area %>%
-  group_by(area) %>%
-  summarise(retouchFreq = mean(!is.na(retouch)), .groups = 'drop')
-
-# Blank Frequency
-blank_freq <- tl_final_area %>%
-  group_by(area) %>%
-  summarise(blankFreq = mean(!is.na(blank)), .groups = 'drop')
-
-# Core Frequency
-core_freq <- tl_final_area %>%
-  group_by(area) %>%
-  summarise(coreFreq = mean(!is.na(`core-typology`)),.groups = 'drop')
-
-# Lithic density (Artifact counts / estimated area)
-lithic_density <- artifact_counts %>%
-  left_join(area_data, by = "area") %>%
-  mutate(lithic_density = n / estimated_area)
-
-# Tool Diversity using Menhinick's Index
-tool_div <- tl_final_area %>%
-  group_by(area) %>%
-  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(n()), .groups = 'drop')
-
-# Step 3: Combining all variables
-
-pca_final_table_area <- left_join(area_data, retouch_freq, by = "area") %>%
-  left_join(blank_freq, by = "area") %>%
-  left_join(core_freq, by = "area") %>%
-  left_join(lithic_density, by = "area") %>%
-  left_join(tool_div, by = "area") %>%
-  select(area, estimated_area, feature_freq, retouchFreq, blankFreq, coreFreq, lithic_density, toolDiv)
-
-```
-
-```{r}
-pca_final_table_area <- pca_final_table_area %>%
-    select(
-        area, 
-        estimated_area = estimated_area.x, 
-        feature_freq = feature_freq.x, 
-        retouchFreq, 
-        blankFreq, 
-        coreFreq, 
-        n, 
-        lithic_density, 
-        toolDiv
-    )
-
-```
-
-```{r}
-# Step 4: Perform PCA
-res_pca_area <- PCA(pca_final_table_area %>% select(-area, -n), graph = FALSE)
-
-# Show the summary of PCA
-print(res_pca_area)
-
-print(res_pca_area$eig)
-print(res_pca_area$var$coord)
-```
-
-```{r}
-# TABLE: Eigenvalues and percentage of variance for each dimension of PCA 
-library(kableExtra)
-
-# Define the values manually (you would get these from your PCA results)
-eigenvalues <- c(4.411234, 2.588766)
-prop_var_explained <- c(63.02, 36.98)
-cumulative_var_explained <- c(63.02, 100.00)
-
-# Create the data frame
-results_df <- data.frame(
-  Dimension = c("Dim. 1", "Dim. 2"),  # Only keep the dimensions you want to display
-  Eigenvalue = eigenvalues,
-  `Variance / percent` = prop_var_explained,
-  `Cumulative / variance percent` = cumulative_var_explained
-)
-
-# Generate the table
-kable(results_df,
-      caption = "Table 3. Eigenvalues and percentage of variance for each dimension of PCA",
-      align = 'c',
-      col.names = c("Dimension", "Eigenvalue", "Variance (%)", "Cumulative Variance (%)")) %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 11)
-
-```
-
-```{r}
-#plotting Dim-1 and Dim-2
-library(FactoMineR)
-library(factoextra)
-
-# Visualize the contributions of variables to the first principal component
-fviz_contrib(res_pca_area, axes = 1, choice = "var")
-
-# Visualize the contributions of variables to the second principal component
-fviz_contrib(res_pca_area, axes = 2, choice = "var")
-
-```
-
-```{r}
-#|fig-cap: Contribution of variables for each of the tow relevant PCA dimensions
-
-# having Dim1 and Dim2 plots next to each other
-
-library(gridExtra)
-
-# Create the first plot and assign it to p1
-p1 <- fviz_contrib(res_pca_area, axes = 1, choice = "var")
-
-# Create the second plot and assign it to p2
-p2 <- fviz_contrib(res_pca_area, axes = 2, choice = "var")
-
-# Use grid.arrange to place them next to each other
-grid.arrange(p1, p2, ncol = 2)
-
-```
-
-```{r}
-# Extract the loadings (coordinates of variables)
-loadings <- res_pca_area$var$coord
-
-# Extract the eigenvalues
-eigenvalues <- res_pca_area$eig
-
-# Calculate the contribution for Dim.1
-contrib_dim1 <- (loadings[,1]^2 / eigenvalues[1, 1]) * 100
-
-# Calculate the contribution for Dim.2
-contrib_dim2 <- (loadings[,2]^2 / eigenvalues[2, 1]) * 100
-
-# Create a data frame to hold the contributions
-contrib_df <- data.frame(
-  Variable = rownames(loadings),
-  Contrib_Dim1 = contrib_dim1,
-  Contrib_Dim2 = contrib_dim2
-)
-
-# Show the contributions
-print(contrib_df)
-
-```
-
-```{r}
-# screeplot
-
-library(FactoMineR)
-library(factoextra)
-
-# Generate the scree plot
-fviz_eig(res_pca_area)
-
-```
-
-```{r}
-# Correlation plots of variables for each of the tow relevant PCA dimensions
-
-fviz_pca_var(res_pca_area,
-
-             repel = TRUE) + 
-  theme(text = element_text(size = 10))
-
-```
-
-```{r}
-# Eigenvalues and percentage of variance
-eig.val <- res_pca_area$eig
-print(eig.val)
-
-# Variable coordinates (correlations with principal components)
-var.coord <- res_pca_area$var$coord
-print(var.coord)
-
-```
-
-```{r}
-#PCA biplot for Dimensions 1 and 2
-
-fviz_pca_biplot(res_pca_area, labelsize = 3, repel = TRUE)
-
-```
-
-```{r}
-# Create the initial biplot without individual labels
-p <- fviz_pca_biplot(res_pca_area, label = "none", 
-                     repel = TRUE, labelsize = 4)
-
-# Add the text labels for the areas using their coordinates from the PCA result
-p + geom_text(aes(x = res_pca_area$ind$coord[, 1], 
-                  y = res_pca_area$ind$coord[, 2], 
-                  label = pca_final_table_area$area), 
-             size = 3, nudge_x = 0.1, nudge_y = 0.1)
-
-```
-
-```{r}
-individual_coordinates = res_pca_area$ind$coord
-variable_coordinates = res_pca_area$var$coord
-
-```
-
-```{r}
-# For variables
-fviz_cos2(res_pca_area, choice = "var")
-
-# For individuals
-fviz_cos2(res_pca_area, choice = "ind")
-
-```
-
-DONE WITH PCA
-
-```{r}
-# Extract individual coordinates
-individual_coordinates <- res_pca_area$ind$coord
-
-# Convert to a data frame for better visibility and add the 'area' column back for context
-individual_coordinates_df <- as.data.frame(individual_coordinates)
-individual_coordinates_df$area <- unique(area_data$area)
-
-# Show the data frame
-print(individual_coordinates_df)
-
-```
-
-```{r}
-library(kableExtra)
-
-# Extract individual coordinates
-individual_coordinates <- res_pca_area$ind$coord
-
-# Convert to a data frame for better visibility and add the 'area' column back for context
-individual_coordinates_df <- as.data.frame(individual_coordinates)
-individual_coordinates_df$Area <- unique(area_data$area)  # Make sure the order matches
-
-# Use kable to create a beautiful table
-kable(individual_coordinates_df, 
-      col.names = c("Area", "Principal Component 1", "Principal Component 2"), 
-      caption = "PCA Coordinates for the Archaeological Areas excavation",
-      align = c('l', 'c', 'c'),
-      digits = 3) %>%
-  kable_styling("striped", full_width = F)
-
-```
-
-```{r}
-# Get the loadings (coefficients) of original variables for each principal component
-loadings_matrix <- res_pca$var$coord
-
-# Print the loadings matrix
-print(loadings_matrix)
-
-```
-
-```{r}
-# Eigenvalues
-eigenvalues <- res_pca$eig[1:2, 1]
-variance_percent <- res_pca$eig[1:2, 2]
-cumulative_variance_percent <- res_pca$eig[1:2, 3]
-
-# Create the table
-df <- data.frame(Eigenvalue = eigenvalues,
-                 Variance_Percent = variance_percent,
-                 Cumulative_Variance_Percent = cumulative_variance_percent)
-
-rownames(df) <- c("Dim1", "Dim2")
-
-# Print the table
-print(df)
-
-```
-
-```{r}
-library(gt)
-
-# Create the table using gt
-gt(df) %>%
-  tab_header(
-    title = "",
-    subtitle = "Eigenvalues and percentage of variance for each dimension of PCA"
-  ) %>%
-  cols_label(
-    Eigenvalue = "Eigenvalue",
-    Variance_Percent = "Variance (%)",
-    Cumulative_Variance_Percent = "Cumulative Variance (%)"
-  ) %>%
-  fmt_number(
-    columns = c("Eigenvalue", "Variance_Percent", "Cumulative_Variance_Percent"),
-    decimals = 2
-  )
-
-```
-
-```{r}
-# Count the number of artifacts by area and store it in a new data frame
-artifact_counts <- tl_final %>%
-  filter(area %in% c("ta", "tp1", "d1")) %>%
-  group_by(area) %>%
-  tally()
-
-# Print the resulting table
-print(artifact_counts)
-```
-
-```{r}
-
-# Calculating retouch Frequency
-
-# Calculating Blank Frequency
-blank_freq <- tl_final_area %>%
-  filter(area %in% c("ta", "d1", "tp1"), !is.na(blank)) %>%
-  group_by(area) %>%
-  summarise(blankFreq = sum(blank %in% c("bladelet", "blade", "flake")) / sum(!is.na(blank)), .groups = 'drop')
-
-# Calculating Core Frequency
-core_freq <- tl_final %>%
-  filter(area %in% c("ta", "d1", "tp1"), !is.na(corefrag)) %>%
-  group_by(area) %>%
-  summarise(coreFreq = sum(corefrag %in% c("core fragment", "bullet", "flat", "pyramid", "cylinder/prismatic", "core rejuvenation piece", "shapeless", "core on flake", "bipolar")) / sum(!is.na(corefrag)), .groups = 'drop')
-
-# Calculating Tool Diversity using Menhinick's Index
-tool_div <- tl_final_area %>%
-  filter(area %in% c("ta", "d1", "tp1"), !is.na(tooltypegroup)) %>%
-  group_by(area) %>%
-  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(sum(!is.na(tooltypegroup))), .groups = 'drop')
-
-# Combining all the calculated frequencies into one tibble
-final_table <- blank_freq %>%
-  inner_join(core_freq, by = "area") %>%
-  inner_join(tool_div, by = "area")
-
-# Printing the final table
-print(final_table)
-
-```
-
-```{r}
-# Joining calculated frequencies
-final_table <- blank_freq %>%
-  inner_join(core_freq, by = "area") %>%
-  inner_join(tool_div, by = "area") %>%
-  inner_join(retouch_freq, by = "area")
-
-# Converting area to row names
-final_table_rownames <- final_table %>%
-  column_to_rownames(var = "area")
-
-```
-
-```{r}
-# Load necessary libraries
-library(FactoMineR)
-library(factoextra)
-
-# Run PCA
-res_pca <- PCA(final_table_rownames, graph = FALSE)
-
-# Visualize the contributions of variables to the second principal component
-fviz_contrib(res_pca, axes = 2, choice = "var")
-
-```
-
-```{r}
-# Visualize the contributions of variables to the first principal component
-fviz_contrib(res_pca, axes = 1, choice = "var") 
-```
-
-```{r}
-# Visualize eigenvalues and explained variance
-fviz_screeplot(res_pca)
-```
-
-```{r}
-fviz_pca_biplot(res_pca, labelsize = 3, repel = TRUE)
-
-```
-
-```{r}
-# Visualize the variables on the principal components
-fviz_pca_var(res_pca)
-```
-
-```{r}
-# Get the loadings (coefficients) of original variables for each principal component
-loadings_matrix <- res_pca$var$coord
-
-# Print the loadings matrix
-print(loadings_matrix)
-
-```
-
-```{r}
-# Eigenvalues
-eigenvalues <- res_pca$eig[1:2, 1]
-variance_percent <- res_pca$eig[1:2, 2]
-cumulative_variance_percent <- res_pca$eig[1:2, 3]
-
-# Create the table
-df <- data.frame(Eigenvalue = eigenvalues,
-                 Variance_Percent = variance_percent,
-                 Cumulative_Variance_Percent = cumulative_variance_percent)
-
-rownames(df) <- c("Dim1", "Dim2")
-
-# Print the table
-print(df)
-
-```
-
-```{r}
-library(gt)
-
-# Create the table using gt
-gt(df) %>%
-  tab_header(
-    title = "",
-    subtitle = "Eigenvalues and percentage of variance for each dimension of PCA"
-  ) %>%
-  cols_label(
-    Eigenvalue = "Eigenvalue",
-    Variance_Percent = "Variance (%)",
-    Cumulative_Variance_Percent = "Cumulative Variance (%)"
-  ) %>%
-  fmt_number(
-    columns = c("Eigenvalue", "Variance_Percent", "Cumulative_Variance_Percent"),
-    decimals = 2
-  )
-
-```
-
-```{r}
-# Count the number of artifacts by area and store it in a new data frame
-artifact_counts <- tl_final %>%
-  filter(area %in% c("ta", "tp1", "d1")) %>%
-  group_by(area) %>%
-  tally()
-
-# Print the resulting table
-print(artifact_counts)
-
-```
-
-```{r}
-# Define the volumes for each area based on the given information
-volumes <- data.frame(
-  area = c("ta", "tp1", "d1"),
-  volume = c(52, 5.76, 2.1375)
-)
-
-# Count the number of artifacts by area
-artifact_counts <- tl_final %>%
-  filter(area %in% c("ta", "tp1", "d1")) %>%
-  group_by(area) %>%
-  tally()
-
-# Join artifact_counts with volumes by area
-lithic_density_df <- left_join(artifact_counts, volumes, by = "area")
-
-# Compute lithic density for each area
-lithic_density_df <- lithic_density_df %>%
-  mutate(lithic_density = n / volume)
-
-# Print the resulting table
-print(lithic_density_df[, c("area", "lithic_density")])
-
-```
-
-```{r}
-# Counting the number of cores by area
-core_counts <- tl_final %>%
-  filter(area %in% c("ta", "d1", "tp1") & !is.na(`core and core fragments`)) %>%
-  group_by(area) %>%
-  tally()
-
-print(core_counts)
-
-```
-
-OK
-
-```{r}
-
-```
-
-```{r}
-#| label: fig-1-layer
-#| fig-cap: "Type of Removals"
-#| fig-height: 5
-
-#Type of blank based on Layers
-
-library(ggbeeswarm)
-
-tl4 %>% 
-  filter(!is.na(layer) & Blank2 %in% c('bladelet', 
-                                       'blade', 
-                                       'flake')) %>%
-
-  # Create bar plot
-  ggplot() +
-  aes(x = layer, 
-      fill = Blank2) +
-  geom_bar(position = "dodge") +
-  labs(x = "Area", 
-       y = "Number", 
-       fill = "Blank Type") +
-  theme_minimal()
-
-```
-
-```{r}
-#Type of core based on Layers
-
-library(ggbeeswarm)
-
-tl4 %>% 
-  filter(!is.na(layer) & 
-           `core-typology` %in% c('bullet','prismatic','pyramid')) %>%
-
-  # Create bar plot
-  ggplot() +
-  aes(x = layer, 
-      fill = `core-typology`) +
-  geom_bar(position = "dodge") +
-  labs(x = "Area", 
-       y = "Number", 
-       fill = "Blank Type") +
-  theme_minimal()
-```
-
-OK

### 2023-09-08 Tulai.qmd
test

@@
 
-# Check unique values in the modified columns
 print(unique(tl4$SHINE))
@@
 
-# Then replace 'NA', 'na', or '0' with R's NA
+# replace 'NA', 'na', or '0' with R's NA
 tl4 <- tl4 %>%
@@
 
-# Show the updated data frame
 print(tl4)
@@
         cortex)
+```
+
+```{r}
+#set the value of utilization to NA only for the rows where `sickle shine` == 1. 
+#All other rows should remain unchanged
+
+#unsuccessful, should try again later
+
 ```
@@
 
+```
+
+```{r}
+
 library(dplyr)
@@
 ```{r}
-# Create a tibble with example data
+# Create a tibble 
 summary_table2_area <- tibble(
@@
 # Create the ggplot
-p <- ggplot(summary_table2_area, 
+WABI_AL <- ggplot(summary_table2_area, 
             aes(x = lithic_volumetric_density,
@@
 
-print(p)
+print(WABI_AL)
 
@@
 
-data_sheet <- "https://docs.google.com/spreadsheets/d/16yLb-RyIQ7x2jFEH_Q7OZYTwlEAXJgFu/edit#gid=2014421097"
-
-sum_depth_sheet <- read_sheet(data_sheet, sheet = "sum_depth")
+sum_depth_sheet <- "https://docs.google.com/spreadsheets/d/1Q0QZESk412ZQLE24yPs6Rg-7Y9OMX63DHoXeIPNAIFM/edit#gid=0"
+
+sum_depth_sheet <- read_sheet(sum_depth_sheet)
+
+sum_depth_sheet <- sum_depth_sheet %>%
+  mutate(area_layer_level = paste0(toupper(area), "_", layer, "_", level)) %>%
+  relocate(area_layer_level, .after = level)
+
+sum_depth_sheet <- sum_depth_sheet %>%
+  mutate(retouch_freq = round(retouch_tool / total_lithics, 2),
+         lithic_dens = round(total_lithics / volume, 2)
+  )
+
+write_sheet(sum_depth_sheet, ss = sheet_url)
+
+```
+
+```{r}
+# Create the ggplot
+WABI_Level <- ggplot(sum_depth_sheet, 
+            aes(x = lithic_dens,
+                y = retouch_freq)) +
+
+  geom_point(color = "black", size = 2) +
+
+  geom_smooth(method = "lm", 
+              se = TRUE, 
+              color = "blue", 
+              size = 0.5) +
+
+  geom_text(aes(label = area_layer_level), 
+            vjust = 1.5, 
+            hjust = 0.5, 
+            size = 2,
+            check_overlap = TRUE) +
+
+  stat_poly_eq(aes(label = paste(stat(eq.label), 
+                                 stat(rr.label), 
+                                 stat(p.value.label), 
+                                 sep = "~~~")),
+               formula = y ~ x, 
+               parse = TRUE, 
+               size = 2) +
+
+  # Add titles and labels
+  labs(
+    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
+    x = "Lithic Volumetric Density",
+    y = "Retouch Frequency"
+  ) +
+
+  theme(
+    axis.text = element_text(size = 8),
+    axis.title = element_text(size = 10),
+    plot.title = element_text(size = 12)
+  )
+
+print(WABI_Level)
+
+```
+
+```{r}
+# Create the ggplot with y-axis on a log scale
+WABI_Level <- ggplot(sum_depth_sheet, 
+            aes(x = lithic_dens,
+                y = retouch_freq)) +
+
+  # Add points with different shapes based on area
+  geom_point(aes(shape = area), size = 2) +
+
+  # Add smoothing line
+  geom_smooth(method = "lm", 
+              se = TRUE, 
+              color = "blue", 
+              size = 0.5) +
+
+  # Add regression equation, R-squared, and p-value
+  stat_poly_eq(aes(label = paste(stat(eq.label), 
+                                 stat(rr.label), 
+                                 stat(p.value.label), 
+                                 sep = "~~~")),
+               formula = y ~ x, 
+               parse = TRUE, 
+               size = 2) +
+
+  # Add titles and labels
+  labs(
+    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
+    x = "Lithic Volumetric Density",
+    y = "Retouch Frequency"
+  ) +
+
+  # Apply log scale to y-axis
+  scale_y_log10() +
+
+  # Add shape legend
+  scale_shape_manual(values = c(16, 17, 25)) +
+
+  # Customize theme
+  theme(
+    axis.text = element_text(size = 8),
+    axis.title = element_text(size = 10),
+    plot.title = element_text(size = 12)
+  )
+
+# Print the plot
+print(WABI_Level)
+
+```
+
+```{r}
+# log-scale
+
+WABI_Level <- ggplot(sum_depth_sheet, 
+            aes(x = lithic_dens,
+                y = retouch_freq)) +
+
+  # Add points
+  geom_point(color = "black", size = 2) +
+
+  # Add smoothing line
+  geom_smooth(method = "lm", 
+              se = TRUE, 
+              color = "blue", 
+              size = 0.5) +
+
+  # Add text labels for each point
+  geom_text(aes(label = area_layer_level), 
+            vjust = 1.5, 
+            hjust = 0.5, 
+            size = 2,
+            check_overlap = TRUE) +
+
+  stat_poly_eq(aes(label = paste(stat(eq.label), 
+                                 stat(rr.label), 
+                                 stat(p.value.label), 
+                                 sep = "~~~")),
+               formula = y ~ x, 
+               parse = TRUE, 
+               size = 2) +
+
+  labs(
+    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
+    x = "Lithic Volumetric Density",
+    y = "Retouch Frequency"
+  ) +
+
+  scale_y_log10() +
+
+  theme(
+    axis.text = element_text(size = 8),
+    axis.title = element_text(size = 10),
+    plot.title = element_text(size = 12)
+  )
+
+print(WABI_Level)
+
+```
+
+```{r}
+
+# Create the ggplot
+WABI_Level <- ggplot(sum_depth_sheet, 
+            aes(x = lithic_dens,
+                y = retouch_freq)) +
+
+  # Add points
+  geom_point(color = 
+               "black", 
+             size = 2) +
+
+  # line with log scale
+
+stat_smooth(method = "lm", 
+            formula = y ~ log(x), 
+            se = TRUE, aes(group = 1), 
+            color = "blue", 
+            size = 0.5)
+
+  # Add text labels for each point
+  geom_text(aes(label = area_layer_level), 
+            vjust = -1, 
+            hjust = -0.5, 
+            size = 2.5) +
+
+  # Add regression equation, R-squared, and p-value
+  stat_poly_eq(aes(label = paste(stat(eq.label), stat(rr.label), stat(p.value.label), sep = "~~~")),
+               formula = y ~ log(x), 
+               parse = TRUE, 
+               size = 2) +
+
+  # Add titles and labels
+  labs(
+    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
+    x = "Lithic Volumetric Density",
+    y = "Retouch Frequency"
+  ) +
+
+  scale_x_log10() +
+
+  theme(
+    axis.text = element_text(size = 8),
+    axis.title = element_text(size = 10),
+    plot.title = element_text(size = 12)
+  )
+
+print(WABI_Level)
 
@@
 # Plotting
-p <- ggplot(summary_table2_area, aes(x = lithic_volumetric_density, y = percent_retouch)) +
+p_area <- ggplot(summary_table2_area, aes(x = lithic_volumetric_density, y = percent_retouch)) +
 
@@
 
-print(p)
+print(p_area)
 
@@
 
-library(dplyr)
-library(ggplot2)
-
-# Filter the data for the relevant categories
-tl4_filtered <- tl4_tool_Blank %>%
-  filter(Blank2 %in% 
-           c("bladelet", 
-             "blade", 
-             "flake"))
-
-# Create separate bar plots for shine, Retouch, and Utilization
-plot_shine2 <- ggplot(tl4_filtered, aes(x = Blank2, fill = shine2)) +
-  geom_bar(position = "dodge") +
-  labs(title = "Shine2 by Blank2",
-       x = "Blank2",
-       y = "Frequency",
-       fill = "Shine2") +
-  theme_minimal()
-
-plot_retouch <- ggplot(tl4_filtered, aes(x = Blank2, fill = Retouch)) +
-  geom_bar(position = "dodge") +
-  labs(title = "Retouch by Blank2",
-       x = "Blank2",
-       y = "Frequency",
-       fill = "Retouch") +
-  theme_minimal()
-
-plot_utilization <- ggplot(tl4_filtered, aes(x = Blank2, fill = Utilization)) +
-  geom_bar(position = "dodge") +
-  labs(title = "Utilization by Blank2",
-       x = "Blank2",
-       y = "Frequency",
-       fill = "Utilization") +
-  theme_minimal()
-
-# Combine the three plots side by side
-library(patchwork)
-combined_plots <- plot_shine2 + plot_retouch + plot_utilization
-combined_plots
-
-```
-
-```{r}
-library(dplyr)
-library(knitr)
-library(kableExtra)
-
-#table
-
-tl_final_aread <- tl_final_area %>%
-  filter(blank %in% c("bladelet", "blade", "flake"))  %>%
-
-  group_by(blank, utilization, `sickle shine`, retouch) %>%
-  summarise(count = n())
-
-table_kable <- kable(table_summary, format = "html", caption = "Summary Table", 
-                     col.names = c("Blank2", "Utilization", "shine2", "Retouch", "Count")) %>%
-  kable_styling(bootstrap_options = "striped")
-
-print(table_kable)
-
-```
-
-```{r}
-#|label: fig_
-#|fig-cap: "Boxplot for each Dimention"
-#|fig-height: 5
-
-library(dplyr)
-library(ggplot2)
-
-# Filter the data to keep only bladelet, blade, and flake in the Blank column
-tl4_tool_Blank_filtered <- tl4_tool_Blank %>%
-  filter(Blank2 %in% c("bladelet", "blade", "flake"))
-
-tl4_tool_Blank_tally <- 
-  tl4_tool_Blank_filtered %>%
-  group_by(Blank2) %>%
-  tally()
-
-# Select dimensions and convert Thickness (mm) to numeric
-tl4_tool_Blank_selected <- 
-  tl4_tool_Blank_filtered %>% 
-  select(Blank2, 
-         Utilization, 
-         Retouch, 
-         shine2)
-
-tl4_tool_Blank_long <- 
-  tl4_tool_Blank_selected %>% 
-  pivot_longer(cols = c(Utilization, Retouch, shine2), 
-               names_to = "Blank2", 
-               values_to = "Value")
-
-# Create boxplot
-ggplot(tl4_tool_Blank_long, aes(x = Blank2, y = Value)) +
-  geom_boxplot() +
-  facet_wrap(~Blank2, scales = "free") +
-  labs(x = "Blank", 
-       y = "Frequency") +
-  theme_minimal()
-
-```
-
-```{r}
-library(dplyr)
-library(ggplot2)
-
-# Filter the data to keep only bladelet, blade, and flake in the Blank2 column
-tl4_tool_Blank_filtered <- tl4_tool_Blank %>%
-  filter(Blank2 %in% c("bladelet", "blade", "flake"))
-
-tl4_tool_Blank_tally <- 
-  tl4_tool_Blank_filtered %>%
-  group_by(Blank2) %>%
-  tally()
-
-# Select dimensions and convert Thickness (mm) to numeric
-tl4_tool_Blank_selected <- 
-  tl4_tool_Blank_filtered %>% 
-  select(Blank2, 
-         Utilization, 
-         Retouch, 
-         shine2)
-
-# Manually rename the duplicate columns
-names(tl4_tool_Blank_selected)[names(tl4_tool_Blank_selected) == "Blank2"] <- "Dimension"
-
-# Pivot data to longer format
-tl4_tool_Blank_long <- 
-  tl4_tool_Blank_selected %>% 
-  pivot_longer(cols = c(Utilization, Retouch, shine2), 
-               names_to = "", 
-               values_to = "Value")
-
-ggplot(tl4_tool_Blank_long, aes(x = Dimension, y = Value)) +
-  geom_boxplot() +
-  facet_wrap(~, scales = "free") +
-  labs(x = "Blank", 
-       y = "Frequency") +
-  theme_minimal()
-
-```
-
-```{r}
-
-# Load required packages
-library(ggplot2)
-
-# Filter the data to exclude the value 0 and keep only blade, bladelet, and flake
-tl4_filtered <- tl4_tool_Blank %>%
-  filter(Blank2 %in% c("blade", "bladelet", "flake"))
-
-# Create the plot with facetting
-tl4_plots <- ggplot(tl4_filtered, aes(x = Blank2, fill = Blank2)) +
-  geom_bar(position = "dodge") +
-  labs(title = "Utilization, Retouch, and shine2 by Blank",
-       x = "Blank2",
-       y = "Count") +
-  facet_grid(. ~ Blank2, scales = "free") +
-  theme_minimal()
-
-print(tl4_plots)
+```
+
+```{r}
 

### 2023-09-08 paper_tulai/paper_tulai.qmd
tidy up
WABI-Removed zero values from dataset

--- 2023-09-08 Tulai.qmd
@@
   ))
+
 # Remove specific columns
@@
 
-sum_depth_sheet <- "https://docs.google.com/spreadsheets/d/1Q0QZESk412ZQLE24yPs6Rg-7Y9OMX63DHoXeIPNAIFM/edit#gid=0"
-
-sum_depth_sheet <- read_sheet(sum_depth_sheet)
+google_sheet_url <- "https://docs.google.com/spreadsheets/d/1Q0QZESk412ZQLE24yPs6Rg-7Y9OMX63DHoXeIPNAIFM/edit#gid=0"
+
+sum_depth_sheet <- read_sheet(google_sheet_url)
 
@@
   mutate(area_layer_level = paste0(toupper(area), "_", layer, "_", level)) %>%
-  relocate(area_layer_level, .after = level)
-
-sum_depth_sheet <- sum_depth_sheet %>%
-  mutate(retouch_freq = round(retouch_tool / total_lithics, 2),
+  relocate(area_layer_level, .after = last_col()) %>%
+  mutate(
+    retouch_freq = round(retouch_tool / total_lithics, 2),
     lithic_dens = round(total_lithics / volume, 2)
@@
 
-write_sheet(sum_depth_sheet, ss = sheet_url)
-
-```
-
-```{r}
-# Create the ggplot
-WABI_Level <- ggplot(sum_depth_sheet, 
+write_sheet(sum_depth_sheet, ss = google_sheet_url)
+
+```
+
+```{r}
+# WABI_LEVEL
+filtered_data <- sum_depth_sheet %>% filter(lithic_dens > 0 & retouch_freq > 0)
+
+WABI_Level <- ggplot(filtered_data, 
             aes(x = lithic_dens,
@@
 ```{r}
-# Create the ggplot with y-axis on a log scale
-WABI_Level <- ggplot(sum_depth_sheet, 
+#  WABI-level-log scale
+
+filtered_data <- sum_depth_sheet %>% filter(lithic_dens > 0 & retouch_freq > 0)
+
+WABI_Level <- ggplot(filtered_data,  
             aes(x = lithic_dens,
@@
 
-  # Add points with different shapes based on area
-  geom_point(aes(shape = area), size = 2) +
-
-  # Add smoothing line
+  geom_point(color = "black", size = 1.5) +
+
   geom_smooth(method = "lm", 
@@
 
-  # Add regression equation, R-squared, and p-value
+  # Add text labels 
+  geom_text(aes(label = area_layer_level), 
+            vjust = 1.5, 
+            hjust = 0.5, 
+            size = 2,
+            check_overlap = TRUE) +
+
+  #regression equation, R-squared, and p-value
   stat_poly_eq(aes(label = paste(stat(eq.label), 
@@
 
-  # Add titles and labels
+  # titles and labels
   labs(
@@
 
-  # Apply log scale to y-axis
+  # Apply log scales
+  scale_x_log10() +
   scale_y_log10() +
 
-  # Add shape legend
-  scale_shape_manual(values = c(16, 17, 25)) +
-
-  # Customize theme
   theme(
@@
 
-# Print the plot
 print(WABI_Level)
@@
 ```{r}
-# log-scale
-
-WABI_Level <- ggplot(sum_depth_sheet, 
-            aes(x = lithic_dens,
-                y = retouch_freq)) +
-
-  # Add points
-  geom_point(color = "black", size = 2) +
-
-  # Add smoothing line
-  geom_smooth(method = "lm", 
-              se = TRUE, 
-              color = "blue", 
-              size = 0.5) +
-
-  # Add text labels for each point
-  geom_text(aes(label = area_layer_level), 
-            vjust = 1.5, 
-            hjust = 0.5, 
-            size = 2,
-            check_overlap = TRUE) +
-
-  stat_poly_eq(aes(label = paste(stat(eq.label), 
-                                 stat(rr.label), 
-                                 stat(p.value.label), 
-                                 sep = "~~~")),
-               formula = y ~ x, 
-               parse = TRUE, 
-               size = 2) +
-
-  labs(
-    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
-    x = "Lithic Volumetric Density",
-    y = "Retouch Frequency"
-  ) +
-
-  scale_y_log10() +
-
-  theme(
-    axis.text = element_text(size = 8),
-    axis.title = element_text(size = 10),
-    plot.title = element_text(size = 12)
-  )
-
-print(WABI_Level)
-
-```
-
-```{r}
-
-# Create the ggplot
-WABI_Level <- ggplot(sum_depth_sheet, 
-            aes(x = lithic_dens,
-                y = retouch_freq)) +
-
-  # Add points
-  geom_point(color = 
-               "black", 
-             size = 2) +
-
-  # line with log scale
-
-stat_smooth(method = "lm", 
-            formula = y ~ log(x), 
-            se = TRUE, aes(group = 1), 
-            color = "blue", 
-            size = 0.5)
-
-  # Add text labels for each point
-  geom_text(aes(label = area_layer_level), 
-            vjust = -1, 
-            hjust = -0.5, 
-            size = 2.5) +
-
-  # Add regression equation, R-squared, and p-value
-  stat_poly_eq(aes(label = paste(stat(eq.label), stat(rr.label), stat(p.value.label), sep = "~~~")),
-               formula = y ~ log(x), 
-               parse = TRUE, 
-               size = 2) +
-
-  # Add titles and labels
-  labs(
-    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
-    x = "Lithic Volumetric Density",
-    y = "Retouch Frequency"
-  ) +
-
-  scale_x_log10() +
-
-  theme(
-    axis.text = element_text(size = 8),
-    axis.title = element_text(size = 10),
-    plot.title = element_text(size = 12)
-  )
-
-print(WABI_Level)
 
@@
 
-```
-
-```{r}
-
-```
-
-```{r}
-# Load required packages
-library(ggplot2)
-library(cowplot)
-
-# Filter the data to exclude the value 0 and keep only blade, bladelet, and flake
-tl4_filtered <- tl4_tool_Blank %>%
-  filter(Blank2 %in% c("blade", "bladelet", "flake"))
-
-# Plot Utilization by Blank2
-plot_utilization <- ggplot(tl4_filtered, aes(Blank2, Utilization)) +
-  geom_bar(stat = "identity", position = "dodge") +
-  labs(title = "Utilization by Blank",
-       x = "Blank2",
-       y = "Utilization") +
-  theme_minimal()
-
-# Plot Retouch by Blank
-plot_retouch <- ggplot(tl4_filtered, aes(Blank2, Retouch)) +
-  geom_bar(stat = "identity", position = "dodge") +
-  labs(title = "Retouch by Blank",
-       x = "Blank2",
-       y = "Retouch") +
-  theme_minimal()
-
-# Plot shine by Blank
-plot_shine <- ggplot(tl4_filtered, aes(x = Blank2, y = shine2)) +
-  geom_bar(stat = "identity", position = "dodge") +
-  labs(title = "shine by Blank",
-       x = "Blank2",
-       y = "shine2") +
-  theme_minimal()
-
-# Combine the plots using cowplot or patchwork
-plot_combined <- cowplot::plot_grid(plot_utilization, plot_retouch, plot_shine, ncol = 3)
-
-print(plot_combined)
-
-```
-
-```{r}
-
-#|label: fig_
-#|fig-cap:  vs Debitage
-#|fig-height: 5
-
-# Filter data
-tl4__blank %>% 
-  filter(Blank2 %in% c('tp1', 'ta', 'd1')) %>% 
-  select(Blank2, 
-         `Width (mm)`, 
-         `Thickness (mm)`, 
-         `Length (mm)`) %>%
-mutate(`Thickness (mm)` = as.numeric(`Thickness (mm)`))
-
-# Pivot data to longer format
-long_data <- tl3_dimention_blank %>% 
-  pivot_longer(cols = c(`Width (mm)`, `Thickness (mm)`, `Length (mm)`), 
-               names_to = "Dimension", 
-               values_to = "Value")
-
-# Create boxplot
-ggplot(long_data, aes(x = area, y = Value)) +
-  geom_boxplot() +
-  geom_quasirandom(alpha = 0.01) +
-  facet_wrap(~Dimension, scales = "free_y") +
-  scale_y_log10() +
-  theme_minimal()
-
-```
-
-```{r}
-library(ggbeeswarm)
-tl3_dimention_blank <- tl3 %>% 
-  filter(area %in% tl3_area_tally_greater_than_n$area) %>% 
-  pivot_longer(-area) %>% 
-  ggplot() +
-  aes(area,
-      value) +
-  geom_boxplot() +
-  geom_quasirandom(alpha = 0.01) +
-  facet_wrap( ~ name,
-              scales = "free_y") +
-  scale_y_log10() +
-  theme_minimal()
-
-```
-
-```{r}
-library(ggbeeswarm)
-tl3 %>% 
-  filter(area %in% tl3_area_tally_greater_than_n$area) %>% 
-  pivot_longer(-area) %>% 
-  ggplot() +
-  aes(area,
-      value) +
-  geom_boxplot() +
-  geom_quasirandom(alpha = 0.01) +
-  facet_wrap( ~ name,
-              scales = "free_y") +
-  scale_y_log10() +
-  theme_minimal()
-
-```
-
-```{r}
-#| label: fig-Width-of-Artifacts-Area-A-by-depth
-#| fig-cap: "Width of Artifacts by depth (Area A)"
-#| fig-height: 5
-tl %>%
-  filter(AREA == "A") %>%
-  select(
-         `Width (mm)`,
-         `DEPTH (cm)`) %>%
-  ggplot() +
-  aes(x = `DEPTH (cm)`,
-      y = `Width (mm)`) +
-  geom_boxplot() +
-  theme_minimal()
-```
-
-```{r}
-#| label: Width of Artifacts (Area A)
-#| fig-cap: "Width of Artifacts (Area A)"
-#| fig-height: 5
-
-tl1 %>%
-  filter(AREA == "TP1") %>%
-  select(AREA, `Width (mm)`) %>%
-  ggplot() +
-  aes(x = `Width (mm)`) +
-  geom_histogram(binwidth = 2) +
-  labs(x = "Width of Debitage (mm)", 
-       y = "Frequency", 
-       title = "Histogram of Width for Area TP1" )+
-  theme_minimal()
-```
-
-WABI:
-
-```{r}
+#WABI:
+
 # regression plot-layer
@@
 
-# Printing the final table
 print(final_table)
@@
 
-# Visualize the contributions of variables to the second principal component
+# contributions of variables to the second principal component
 fviz_contrib(res_pca,
@@
 
-# Creating a biplot
+#  biplot
 fviz_pca_biplot(res_pca)
@@
 ```{r}
-
+# Run PCA
 library(FactoMineR)
@@
 
-# Run PCA
 res_pca <- PCA(final_table_rownames, graph = FALSE)
 
-# Get the loadings (coefficients) of original variables for each principal component
+# Get the coefficients of original variables for each principal component
 loadings_matrix <- res_pca$var$coord
 
-# Print the loadings matrix
 print(loadings_matrix)
@@
 
-# Print the table
 print(df)
@@
 ```{r}
-#|fig-cap: Contribution of variables for each of the tow relevant PCA dimensions
 
@@
 ```{r}
-# Create the initial biplot without individual labels
+# biplot 
 p <- fviz_pca_biplot(res_pca_area, label = "none", 
@@
 
-# Add the text labels for the areas using their coordinates from the PCA result
 p + geom_text(aes(x = res_pca_area$ind$coord[, 1], 
@@
 ```
-
-OK
-
-```{r}
-#| label: fig-1-layer
-#| fig-cap: "Type of Removals"
-#| fig-height: 5
-
-#Type of blank based on Layers
-
-library(ggbeeswarm)
-
-tl4 %>% 
-  filter(!is.na(layer) & Blank2 %in% c('bladelet', 
-                                       'blade', 
-                                       'flake')) %>%
-
-  # Create bar plot
-  ggplot() +
-  aes(x = layer, 
-      fill = Blank2) +
-  geom_bar(position = "dodge") +
-  labs(x = "Area", 
-       y = "Number", 
-       fill = "Blank Type") +
-  theme_minimal()
-
-```
-
-```{r}
-#Type of core based on Layers
-
-library(ggbeeswarm)
-
-tl4 %>% 
-  filter(!is.na(layer) & 
-           `core-typology` %in% c('bullet','prismatic','pyramid')) %>%
-
-  # Create bar plot
-  ggplot() +
-  aes(x = layer, 
-      fill = `core-typology`) +
-  geom_bar(position = "dodge") +
-  labs(x = "Area", 
-       y = "Number", 
-       fill = "Blank Type") +
-  theme_minimal()
-```
-
-consistency in dimention:
-
-```{r}
-#|label: fig_2
-#|fig-cap: "Boxplot for each Dimention"
-#|fig-height: 5
-
-library(ggplot2)
-library(dplyr)
-library(tidyr)
-
-# Set the value for n
-n <- 100
-
-# Tally by area
-tl_final_area_tally <- tl_final_area %>%
-  group_by(area) %>%
-  tally()
-
-# Filter areas with more than n rows
-tl_final_area_tally_greater_than_n <- tl_final_area_tally %>%
-  filter(n >= n)
-
-# Filter data based on areas with more than n rows
-tl_final_area_filtered <- tl_final_area %>%
-  filter(area %in% tl_final_area_tally_greater_than_n$area) %>%
-  filter(!is.na(blank) & blank != "")
-
-# Select and mutate dimensions
-tl_final_area_selected <- tl_final_area_filtered %>%
-  select(area, width, thickness, length) %>%
-  mutate(
-    width = as.numeric(width),
-    thickness = as.numeric(thickness),
-    length = as.numeric(length)
-  )
-
-# Pivot data to longer format
-tl_final_area_long <- tl_final_area_selected %>%
-  pivot_longer(cols = c(width, thickness, length), 
-               names_to = "Dimension", 
-               values_to = "Value")
-
-# Create the box plot
-ggplot(tl_final_area_long, aes(x = area, y = Value)) +
-  geom_boxplot() +
-  facet_wrap(~Dimension, scales = "free") +
-  labs(x = "Area", y = "Value") +
-  theme_minimal()
-
-```
-
-```{r}
-#|label: fig_3
-#|fig-cap: Dimention By Area
-#|ffig-height: 5
-
-library(ggplot2)
-library(dplyr)
-library(tidyr)
-library(ggbeeswarm)
-
-# Create boxplot
-ggplot(tl_final_area_long, 
-       aes(x = area, y = Value)) +
-  geom_boxplot() +
-  geom_quasirandom(alpha = 0.01) +
-  facet_wrap(~Dimension, scales = "free_y") +
-  scale_y_log10() +
-  theme_minimal()
-
-```
-
-```{r}
-
-```
-
-```{r}
-#|label: fig_2-layer
-#|fig-cap: "Boxplot for each Dimention"
-#|fig-height: 5
-
-n <- 100
-
-# Tally count of each layer
-tl_final_layer_tally <- tl_final_area %>% 
-  group_by(layer) %>% 
-  tally()
-
-# Filter layers with more than n rows
-tl_final_layer_tally_greater_than_n <- tl_final_layer_tally %>% 
-  filter(n >= n)
-
-# Filter data for layers with more than n rows
-tl_final_layer_filtered <- tl_final_area %>% 
-  filter(layer %in% tl_final_layer_tally_greater_than_n$layer)
-
-# Select and mutate
-tl_final_layer_selected <- tl_final_layer_filtered %>%
-  select(layer, 
-         `width`, 
-         `thickness`, 
-         `length`) %>%
-  mutate(across(c
-                (
-           `width`, 
-           `thickness`, 
-           `length`), 
-           as.numeric)
-         )
-
-# Pivot longer
-tl_final_layer_long <- tl_final_layer_selected %>%
-  pivot_longer(cols = 
-                 c(width, 
-                   thickness, 
-                   length), 
-               names_to = 
-                 "Dimension", 
-               values_to = 
-                 "Value") %>%
-  drop_na(Value)
-
-# Create box plot
-ggplot(tl_final_layer_long, 
-       aes(layer, 
-           Value)) +
-
-  geom_boxplot() +
-  facet_wrap(~Dimension, 
-             scales = "free") +
-  labs(x = "Layer", 
-       y = "Value") +
-
-   theme_minimal()
-
-```
-
-```{r}
-#|label: fig_3-layer
-#|fig-cap: Dimension By Layer
-#|fig-height: 5
-
-# Create boxplot
-ggplot(tl_final_layer_long, 
-       aes(x = layer, y = Value)) +
-  geom_boxplot() +
-  geom_quasirandom(alpha = 0.01) +
-  facet_wrap(~Dimension, scales = "free_y") +
-  scale_y_log10() +
-  theme_minimal()
-
-```
-
-```{r}
-#|label: fig_4-width
-#|fig-cap: "Width of Artifacts (Area A)"
-#| fig-height: 5
-tl_final_area %>%
- filter(area == "ta") %>%
-  select(area, 
-         width) %>%
-  ggplot() +
-  aes(x = width) +
-  geom_histogram(binwidth = 2) +
-  labs(x = "Width of Debitage (mm)", 
-       y = "Frequency" )+
-  theme_minimal()
-```
-
-```{r}
-#|label: fig_5-width
-#|fig-cap: "Width of Artifacts (Area A, D1, TP1)"
-#|fig-height: 5
-
-tl_final_area %>%
-  filter(area %in% 
-           c("ta", "d1", "tp1")) %>%
-  mutate(width = as.numeric(width)) %>%  # Explicitly convert to numeric
-
-   select(area, 
-          `width`) %>%
-  ggplot() +
-
-  aes(x = width) +  # Make sure the aesthetic mapping is correct
-
-  geom_histogram(binwidth = 2) +
-
-  labs(x = "Width of Debitage (mm)", 
-       y = "Frequency" ) +
-
-  facet_wrap(~area, 
-             ncol = 1) +
-
-  theme_minimal()
-
-```
-
-```{r}
-#|label: fig_5-width-layer
-#|fig-cap: "Width of Artifacts (Area A, D1, TP1)"
-#|fig-height: 5
-
-tl_final_area %>%
-  filter(layer %in% c("a", 
-                      "c", 
-                      "e or d", 
-                      "e", 
-                      "d")) %>%
-  mutate(width = as.numeric(width)) %>%  # Convert to numeric
-  select(layer, width) %>%
-  ggplot() +
-  aes(x = width) +
-  geom_histogram(binwidth = 2) +
-  labs(x = "Width of Debitage (mm)", y = "Frequency") +
-  facet_wrap(~layer, ncol = 1) +
-  theme_minimal()
-
-```
-
-```{r}
-
-```
-
-```{r}
-#|label: fig_6-thickness-layer
-#|fig-cap: "Thickness of Artifacts (Layers A, C, E or D)"
-#|fig-height: 5
-
-tl_final_area %>%
- filter(layer %in% c("a", 
-                     "c", 
-                     "e or d", 
-                     "d", 
-                     "e")) %>%
- mutate(thickness = as.numeric(thickness)) %>% 
-  select(layer, 
-         thickness) %>%
-  ggplot() +
-  aes(x = thickness) +
-  geom_histogram(binwidth = 2) +
-  labs(x = "Thickness of Debitage (mm)", 
-       y = "Frequency" )+
-  facet_wrap(~layer , ncol = 1) +
-  theme_minimal()
-
-```
-
-```{r}
-#|label: fig_7
-#|fig-cap: "Length of Artifacts (Area A, D1, TP1)"
-#|fig-height: 5
-
-tl3 %>%
-  filter(Blank_part == "complete", area %in% c("ta", "tp1", "d1")) %>%
-  mutate(`Length (mm)` = as.numeric(`Length (mm)`)) %>%
-  select(area, 
-         `Length (mm)`) %>%
-  ggplot() +
-  aes(x = `Length (mm)`) +
-  geom_histogram(binwidth = 2) +
-  labs(x = "Length of Debitage (mm)", 
-       y = "Frequency") +
-  facet_wrap(~area, ncol = 1) +
-  theme_minimal()
-
-```
-
-```{r}
-#| label: fig-7-length-layer
-#| fig-cap: Length of Artefact based on layers
-#| fig-height: 5
-tl3 %>% 
-  filter(layer %in% c("layer a" , "layer d" , "layer e" , "layer c" , "layer e or d")) %>% 
-  mutate(`Length (mm)` = as.numeric(`Length (mm)`)) %>% 
-  select(layer, `Length (mm)`) %>% 
-  ggplot(aes(x = `Length (mm)`)) + 
-  geom_histogram(binwidth = 2) +
-  labs(x = "Length of Debitage (mm)" ,
-       y = "Frequency") + 
-  facet_wrap(~layer, ncol = 1) + 
-  theme_minimal()
-
-```
-
-```{r}
-#|label: NA 
-#|fig-cap: "tool typology"
-#|fig-height: 5
-
-```
-
-```{r}
-library(ggplot2)
-
-# Calculate the counts for debitage and 
-counts <- tl4_tool_Blank %>%
-  group_by(debitage, ) %>%
-  summarise(count = n())
-
-# Create the bar plot
-ggplot(counts, 
-       aes(x = factor(debitage), 
-           y = count, 
-           fill = factor(debitage))) +
-
-  geom_bar(stat = "identity", 
-           position = "dodge") +
-
-  # Move the labels slightly above the bars
-  geom_text(aes(label = count), 
-            vjust = -0.5) +
-
-  geom_hline(data = counts %>% 
-               filter( == 1), 
-             aes(yintercept = count),
-             linetype = "dashed", 
-             color = "red") +
-  labs(title = "Debitage vs. ",
-       x = "",
-       y = "Numbers") +
-  theme_minimal()
-
-```
-
-```{r}
-# Calculate the counts for debitage and 
-counts <- tl4_tool_Blank %>%
-  group_by(Retouch = 1,  =1) %>%
-  summarise(count = n())
-
-# Create the bar plot
-ggplot(counts, 
-       aes(x = factor( = 1), 
-           y = count, 
-           fill = factor( = 1))) +
-
-  geom_bar(stat = "identity", 
-           position = "dodge") +
-
-  # Move the labels slightly above the bars
-  geom_text(aes(label = count), 
-            vjust = -0.5) +
-
-  geom_hline(data = counts %>% 
-               filter( == 1), 
-             aes(yintercept = count),
-             linetype = "dashed", 
-             color = "red") +
-  labs(title = " vs. formal ",
-       x = "",
-       y = "Numbers") +
-  theme_minimal()
-```
-
-```{r}
-#|label: fig_
-#|fig-cap: ""
-#| fig-height: 5
-tl4_tool_Blank %>%
-  filter(Blank2 %in% c("bladelet", "blade", "flake")) %>%
-  select(Blank2, shine2 = 1) %>%
-  ggplot() +
-  aes(x = shine2) +
-  geom_bar() +
-  labs(x = "shine", 
-       y = "Frequency" )+
-  theme_minimal()
-
-tl4_tool_Blank %>%
-  filter(Blank2 %in% c("bladelet", "blade", "flake")) %>%
-  select(Blank2, Retouch = 1) %>%
-  ggplot() +
-  aes(x = Retouch) +
-  geom_bar() +
-  labs(x = "Retouch", 
-       y = "Frequency" )+
-  theme_minimal()
-
-tl4_tool_Blank %>%
-  filter(Blank2 %in% c("bladelet", "blade", "flake")) %>%
-  select(Blank2, Utilization = 1) %>%
-  ggplot() +
-  aes(x = Utilization) +
-  geom_bar() +
-  labs(x = "Utilization", 
-       y = "Frequency" )+
-  theme_minimal()
-
-```
-
-```{r}
-library(ggplot2)
-library(patchwork)
-
-plot_shine <- tl4_tool_Blank %>%
-  filter(Blank2 %in% c("bladelet", 
-                       "blade", 
-                       "flake")) %>%
-  select(Blank2, shine2 = 1) %>%
-  ggplot() +
-  aes(x = shine2) +
-  geom_bar() +
-  labs(x = "shine", y = "Frequency") +
-  theme_minimal()
-
-plot_retouch <- tl4_tool_Blank %>%
-  filter(Blank2 %in% c("bladelet", 
-                       "blade", 
-                       "flake")) %>%
-  select(Blank2, 
-         Retouch = 1) %>%
-  ggplot() +
-  aes(x = Retouch) +
-  geom_bar() +
-  labs(x = "Retouch", 
-       y = "Frequency") +
-  theme_minimal()
-
-plot_utilization <- tl4_tool_Blank %>%
-  filter(Blank2 %in% c("bladelet", "blade", "flake")) %>%
-  select(Blank2, Utilization = 1) %>%
-  ggplot() +
-  aes(x = Utilization) +
-  geom_bar() +
-  labs(x = "Utilization", y = "Frequency") +
-  theme_minimal()
-
-# Combine the plots using patchwork
-plot_combined <- plot_shine + plot_retouch + plot_utilization
-plot_combined
-
-```

### 2023-09-08 paper/paper_tulai.qmd
code review and edits up to PCA section

--- 2023-09-08 paper_tulai/paper_tulai.qmd
@@
 ---
-title: "Tulai"
+title: ""
 author: ""
@@
 ```{r}
-# library
+# load libraries
 library(tidyverse)
-library(ggplot2)
-library(dplyr)
-library(magrittr)
-library(stringr)
-
+library(here)
+library(ggpmisc)
+library(ggrepel)
+library(ggbeeswarm)
 ```
@@
 # Read the data
-tl <- read_csv("Tulai_Lithic_Assamblage.csv")
+tl <- read_csv(here("data/Tulai_Lithic_Assamblage.csv"))
 
@@
                               "TA/\\1")) %>% 
-
   # Add slash between numbers
@@
   mutate(TA = str_replace_all(TA, " ", ""))  %>%
-
   # Separate 'TA' column into 'area' and 'depth' columns
@@
   mutate(depth = as.numeric(depth)) %>% 
-
   # Create depth ranges
@@
   )) %>% 
-
   separate(depth, 
@@
   mutate(midpoint = mean(c(upper, lower))) %>%
-
 # create 'layer' column
@@
             "Pattern"))
-
-```
-
-```{r}
-#exploring Data
+```
+
+```{r}
+# exploring Data
 
@@
 ```{r}
-#exploring Data
+# exploring Data
 
@@
 # Cleaning Blank column
-library(dplyr)
-library(stringr)
-library(tidyr)
 
@@
              Blank == "bladelt-pro" ~              "bladelet-pro",   
-             Blank == "bladelet-nearly complete" ~ "bladelet-complete", 
+             Blank == "bladelet-nearly complete"~  "bladelet-complete", 
              Blank == "microblade-med" ~           "bladelet-med", 
@@
 
-unique(tl2$Blank)
+rev(sort(table(tl2$Blank)))
+
+# BM: still many typos and unusual categories in here, why is that? Can they be fixed?
 
@@
 
-print(unique
-      (tl3$Blank2))
-
-```
-
-```{r}
-
+rev(sort(table(tl3$Blank2))) # BM: still a lot of typos on the categories here, can you fix them?
+
+```
+
+```{r}
+# BM: this is an unusual coding pattern, I'm curious about where you got it from?
 tl3 <- tl3 %>%
@@
     )
-  )
-
-```
-
-```{r}
-tl3 <- tl3 %>%
+  )  %>%
   mutate(
@@
 # Remove rows based on row names
-tl3 <- tl3[!(rownames(tl3) %in% c("1030", "644", "543")), ]
+# tl3 <- tl3[!(rownames(tl3) %in% c("1030", "644", "543")), ]
+
+# BM: why are we removing these rows? A comment here would be good to explain. A simpler way, by referencing the row numbers directly:
+
+tl3 <- tl3[-c(1030, 644, 543), ]
 ```
@@
   mutate(
-    Utilization = if_else(Retouch == "1", NA_character_, as.character(Utilization))
-  )
-
-```
-
-```{r}
-library(dplyr)
-library(stringr)
+    Utilization = if_else(Retouch == "1", 
+                          NA_character_, 
+                          as.character(Utilization))
+  )
+
+```
+
+```{r}
 
 tl4 <- tl4 %>%
+  # BM: the problem with this combination of mutate_all and ifelse that you use below is that they 
+  # coerce all columns to be the same data type, so you'll see that all the numeric
+  # columns are now character columns after you run these lines. So you need to
+  # coerce those number columns back to numeric type so you can plot and compute
+  # with them
   mutate_all(~ ifelse(is.na(.x) | .x == "", NA, .x)) %>% 
+  # BM: what does this next line do?
   mutate_all(~ str_replace(., "(?<=\\D)\\*(?=\\D)|\\s+|[*?!]+", "")) %>%
+  # BM: what is this SHINE variable? I'm not familiar with it
   mutate(
@@
     shine2 = case_when(
-      SHINE %in% c("Subparallel", "Parallel", "Sub-Parallel", "Parallel-Subprarallel",
-                   "Scaled", "Sub-Paralel", "Sub-Parallell", "Sub-Paallel",
-                   "Sub-Paralell", "Sub-Prallel", "L:semiparallel/R: Scaled",
-                   "Semiparallel", "Semi-Parallel", "Scaled/Sub-Parallel",
-                   "Seb-Parallel", "Semi-Abrupt") ~ NA_character_,
+      SHINE %in% c("Subparallel", 
+                   "Parallel", 
+                   "Sub-Parallel", 
+                   "Parallel-Subprarallel",
+                   "Scaled", 
+                   "Sub-Paralel", 
+                   "Sub-Parallell", 
+                   "Sub-Paallel",
+                   "Sub-Paralell", 
+                   "Sub-Prallel", 
+                   "L:semiparallel/R:scaled",
+                   "Semiparallel", 
+                   "Semi-Parallel", 
+                   "Scaled/Sub-Parallel",
+                   "Seb-Parallel", 
+                   "Semi-Abrupt") ~ NA_character_,
       TRUE ~ SHINE
@@
 
-print(unique(tl4$SHINE))
-
-```
-
-```{r}
-library(dplyr)
-library(stringr)
+# BM: I think this is a better way to inspect:
+rev(sort(table(tl4$SHINE)))
+rev(sort(table(tl4$shine2)))
+
+```
+
+```{r}
+
 # Convert 'NA' and 'na' strings to actual NA
@@
     `Core Typology` = str_trim(`Core Typology`),
+    # BM: what does this regex do in the next line?
     `Core Typology` = str_replace_all(`Core Typology`, "^[-._/NA\\s]*$|^\\s*$", "0")
   ) %>%
-
   mutate(`Core Typology` = case_when(
@@
              `Core Typology` == "corefragment" ~ "core fragment-NA",
-
              `Core Typology` == "cilandrical/pressure" ~ "cylinder-pressure",
@@
     TRUE ~ `Core Typology`
-  ))
-
+  )) %>%
 # Separate 'Core Typology' into 'core-typology' and 'core-technology'
-tl4 <- tl4 %>%
   separate(`Core Typology`, 
@@
 # Check the unique values again
-unique(tl4$`core-typology`)
-unique(tl4$`core-technology`)
-```
-
-```{r}
-
-library(dplyr)
-library(stringr)
+# BM: still many typos, are they important?
+rev(sort(table((tl4$`core-typology`))))
+rev(sort(table((tl4$`core-technology`))))
+
+```
+
+```{r}
 
@@
   ) %>%
-  mutate(Typology = case_when(
-
+  mutate(
+    # BM: wow, seems like almost each artefact is its own type! 
+    # BM: That is not ideal for statistical analysis
+    # BM: This looks like it was very time-consuming to simplify. 
+    Typology = case_when(
       `Typology` == "notch, truncated" ~ "tool/notch;truncated piece",
-
       `Typology` == "retouchpiece" ~ "tool/retouched piece",
-
       `Typology` == "endscraper" ~ "tool/end scraper",
-
       `Typology` == "notch" ~ "tool/notch",
-
       `Typology` == "sidescraper" ~ "tool/side scraper",
-
       `Typology` == "doubleside scraper" ~ "tool/double side scraper",
-
       `Typology` == "convergentscraper (?)" ~ "tool/convergent scraper",
-
       `Typology` == "doublenotch on retouched piece" ~ "tool/double notch",
-
       `Typology` == "retouchpieces" ~ "tool/retouched piece",
-
       `Typology` == "burin(?)/ notch/scraper" ~ "tool/burin-notch-scraper",
-
       `Typology` == "borer" ~ "tool/perforator-borer",
-
       `Typology` == "retouchedpiece" ~ "tool/retouched piece",
-
       `Typology` == "notch/side scraper" ~ "tool/notch-side scraper",
-
       `Typology` == "corefragment" ~ "core/core fragment",
-
       `Typology` == "backed" ~ "tool/backed knife",
-
       `Typology` == "doubleside scraper ?" ~ "tool/double side scraper",
-
       `Typology` == "inversedenticulate" ~ "tool/denticulate",
-
       `Typology` == "retouchpieces/ side scraper?" ~ "tool/side scraper",
-
       `Typology` == "sidescraper ?" ~ "tool/side scraper",
-
       `Typology` == "notch-denticulate" ~ "tool/notch-denticulate",
-
       `Typology` == "borer/drill" ~ "tool/perforator-drill",
-
       `Typology` == "retouchedpiece (side scraper?)" ~ "tool/side scraper",
-
       `Typology` == "retouchepiece (side scraper?)" ~ "tool/side scraper",
-
       `Typology` == "retouchedpiece (double side scraper)" ~ "tool/double side scraper",
-
       `Typology` == "retouchedpiece (double side scraper?)" ~ "tool/double side scraper",
-
       `Typology` == "retouchepiece" ~ "tool/retouched piece",
-
       `Typology` == "inversenotch" ~ "tool/notch",
-
       `Typology` == "retouchepiece (double side scaraper?)" ~ "tool/double side scraper",
-
       `Typology` == "retouchedpiece( double side scraper?)" ~ "tool/double side scraper",
-
       `Typology` == "notchon retouche piece" ~ "tool/notch",
-
       `Typology` == "retouchedpiece (souble side scraper)" ~ "tool/double side scraper",
-
       `Typology` == "point" ~ "tool/perforator-point",
-
       `Typology` == "chunk" ~ "NA/NA",
-
       `Typology` == "coretablet" ~ "core/core tablet",
-
       `Typology` == "core" ~ "core/shapeless",
-
       `Typology` == "denticulate" ~ "tool/denticulate",
-
       `Typology` == "doublenotch (haft)" ~ "tool/double notch",
-
       `Typology` == "retouchedpiece (fine retouch)" ~ "tool/retouched piece",
-
       `Typology` == "endscraper on retouched piece" ~ "tool/end scraper",
-
       `Typology` == "retouchedpiece (serrated scraper)" ~ "tool/serrated scraper",
-
       `Typology` == "truncation" ~ "tool/truncated piece",
-
       `Typology` == "rejuventionpiece" ~ "core/rejuvenation piece",
-
       `Typology` == "notch/denticulate" ~ "tool/notch-denticulate",
-
       `Typology` == "retoucheduse on breakage" ~ "tool/retouched piece",
-
       `Typology` == "inversenotch/endscraper" ~ "tool/notch-end scraper",
-
       `Typology` == "alternatingside scraper" ~ "tool/side scraper",
-
       `Typology` == "alternatedouble side scraper" ~ "tool/double side scraper",
-
       `Typology` == "truncation/sidescaraper" ~ "tool/truncated piece",
-
       `Typology` == "doublenotch" ~ "tool/double notch",
-
       `Typology` == "borer/awl" ~ "tool/perforator-awl",
-
       `Typology` == "roundscraper?" ~ "tool/round scraper",
-
       `Typology` == "doublenotch scraper" ~ "tool/double notch scraper",
-
       `Typology` == "corerejuvention (core tablet?)" ~ "core/core tablet",
-
       `Typology` == "truncationاریبب" ~ "tool/truncated piece",
-
       `Typology` == "retouchedpiece (side scraper)" ~ "tool/side scraper",
-
       `Typology` == "retochedpice" ~ "tool/retouched piece",
-
       `Typology` == "truncation/notch" ~ "tool/truncated piece-notch",
-
       `Typology` == "doublealternate scraper" ~ "tool/double side scraper",
-
       `Typology` == "awl/ inverse notch" ~ "tool/notch",
-
       `Typology` == "retouchedpiece (alternate scraper?)" ~ "tool/double side scraper",
-
       `Typology` == "inverseside scraper" ~ "tool/side scraper",
-
       `Typology` == "usedcore rejuvention" ~ "core/rejuvenation piece",
-
       `Typology` == "coreside rejuvention" ~ "core/rejuvenation piece",
-
       `Typology` == "coreplatform rejuvention flake" ~ "core/rejuvenation piece",
-
       `Typology` == "corerejuvention platform" ~ "core/rejuvenation piece",
-
       `Typology` == "coreside rejuvention blade" ~ "core/rejuvenation piece",
-
       `Typology` == "coreside rejuvention blade??" ~ "core/rejuvenation piece",
-
       `Typology` == "coreedge rejuvention" ~ "core/rejuvenation piece",
-
       `Typology` == "coreedge rejivention" ~ "core/rejuvenation piece",
-
       `Typology` == "retouchedpice (side scraper?)" ~ "tool/side scraper" ,
-
       `Typology` == "scraper" ~ "tool/side scraper",
-
       `Typology` == "sickleblade" ~ "tool/sickle shine",
-
       `Typology` == "coretool (scraper)" ~ "tool/scraper-on core piece",
-
       `Typology` == "retouchedpiece (scraper)" ~ "tool/side scraper",
-
       `Typology` == "inversedenticulate?" ~ "tool/denticulate",
-
       `Typology` == "retouchedpiece (scraper?)" ~ "tool/side scraper",
-
       `Typology` == "coreside rejuvention flake" ~ "core/rejuvenation piece",
-
       `Typology` == "truncation/endscraper" ~ "tool/endscraper on truncated piece",
-
       `Typology` == "retouchedpice(double side scraper?)" ~ "tool/double side scraper",
-
       `Typology` == "doublescraper" ~ "tool/double side scraper",
-
       `Typology` == "dendiculate" ~ "tool/denticulate",
-
       `Typology` == "double-scraper" ~ "tool/double side scraper",
-
       `Typology` == "pyramidalcore" ~ "core/pyramid",
-
       `Typology` == "bulletcore" ~ "core/bullet",
-
       `Typology` == "corerejuvention" ~ "core/rejuvenation piece",
-
       `Typology` == "pyrmidalcore" ~ "core/pyramid",
-
       `Typology` == "corepreperation" ~ "core/core preparation piece",
-
       `Typology` == "crested" ~ "core/crested bladelet",
-
       `Typology` == "preperationblade" ~ "core/core preperation",
-
       `Typology` == "burin" ~ "tool/burin",
-
       `Typology` == "serrateddenticulate" ~ "tool/serrated scraper",
-
       `Typology` == "saw:serrated denticulate" ~ "tool/serrated scraper",
-
       `Typology` == "serratedused" ~ "tool/serrated scraper",
-
       `Typology` == "corepreeration" ~ "tool/core preparation piece",
-
       `Typology` == "coreprepearation" ~ "tool/core preparation piece",
-
       `Typology` == "awl" ~ "tool/perforator-awl",
-
       `Typology` == "notch.inverse" ~ "tool/notch",
-
       `Typology` == "point.notch" ~ "tool/perforator-point;notch",
-
       `Typology` == "inversescraper" ~ "tool/side scraper",
-
       `Typology` == "roundendscraper" ~ "tool/round scraper",
-
       `Typology` == "alternate.scraper" ~ "tool/double side scraper",
-
       `Typology` == "convergentscraper/disc scraper" ~ "tool/scraper-convergent scraper",
-
       `Typology` == "serratedside scraper" ~ "tool/serrated scraper",
-
       `Typology` == "debri" ~ "NA/NA",
-
       `Typology` == "distalpart of a drill" ~ "tool/perforator-drill",
-
       `Typology` == "corepreperation?" ~ "core/core preparation piece",
-
       `Typology` == "preformend of a borer" ~ "tool/perforator-borer; preform",
-
       `Typology` == "used" ~ "tool/used",
-
       `Typology` == "distalend of a inverse sidescraper" ~ "tool/side scraper",
-
       `Typology` == "retouchedblade" ~ "tool/retouched piece",
-
       `Typology` == "drill" ~ "tool/perforator-drill",
-
       `Typology` == "corerejuvention tablet" ~ "core/core tablet",
-
       `Typology` == "bladeletcore" ~ "core/pyramid",
-
       `Typology` == "flake-bladecore" ~ "core/mixed",
-
+      `Typology` == "notch, truncated" ~ "tool/notch;truncated piece",
+      `Typology` == "retouchpiece" ~ "tool/retouched piece",
+      `Typology` == "endscraper" ~ "tool/end scraper",
+      `Typology` == "notch" ~ "tool/notch",
+      `Typology` == "sidescraper" ~ "tool/side scraper",
+      `Typology` == "doubleside scraper" ~ "tool/double side scraper",
+      `Typology` == "convergentscraper (?)" ~ "tool/convergent scraper",
+      `Typology` == "doublenotch on retouched piece" ~ "tool/double notch",
+      `Typology` == "retouchpieces" ~ "tool/retouched piece",
+      `Typology` == "burin(?)/ notch/scraper" ~ "tool/burin-notch-scraper",
+      `Typology` == "borer" ~ "tool/perforator-borer",
+      `Typology` == "retouchedpiece" ~ "tool/retouched piece",
+      `Typology` == "notch/side scraper" ~ "tool/notch-side scraper",
+      `Typology` == "corefragment" ~ "core/core fragment",
+      `Typology` == "backed" ~ "tool/backed knife",
+      `Typology` == "doubleside scraper ?" ~ "tool/double side scraper",
+      `Typology` == "inversedenticulate" ~ "tool/denticulate",
+      `Typology` == "retouchpieces/ side scraper?" ~ "tool/side scraper",
+      `Typology` == "sidescraper ?" ~ "tool/side scraper",
+      `Typology` == "notch-denticulate" ~ "tool/notch-denticulate",
+      `Typology` == "borer/drill" ~ "tool/perforator-drill",
+      `Typology` == "retouchedpiece (side scraper?)" ~ "tool/side scraper",
+      `Typology` == "retouchepiece (side scraper?)" ~ "tool/side scraper",
+      `Typology` == "retouchedpiece (double side scraper)" ~ "tool/double side scraper",
+      `Typology` == "retouchedpiece (double side scraper?)" ~ "tool/double side scraper",
+      `Typology` == "retouchepiece" ~ "tool/retouched piece",
+      `Typology` == "inversenotch" ~ "tool/notch",
+      `Typology` == "retouchepiece (double side scaraper?)" ~ "tool/double side scraper",
+      `Typology` == "retouchedpiece( double side scraper?)" ~ "tool/double side scraper",
+      `Typology` == "notchon retouche piece" ~ "tool/notch",
+      `Typology` == "retouchedpiece (souble side scraper)" ~ "tool/double side scraper",
+      `Typology` == "point" ~ "tool/perforator-point",
+      `Typology` == "chunk" ~ "NA/NA",
+      `Typology` == "coretablet" ~ "core/core tablet",
+      `Typology` == "core" ~ "core/shapeless",
+      `Typology` == "denticulate" ~ "tool/denticulate",
+      `Typology` == "doublenotch (haft)" ~ "tool/double notch",
+      `Typology` == "retouchedpiece (fine retouch)" ~ "tool/retouched piece",
+      `Typology` == "endscraper on retouched piece" ~ "tool/end scraper",
+      `Typology` == "retouchedpiece (serrated scraper)" ~ "tool/serrated scraper",
+      `Typology` == "truncation" ~ "tool/truncated piece",
+      `Typology` == "rejuventionpiece" ~ "core/rejuvenation piece",
+      `Typology` == "notch/denticulate" ~ "tool/notch-denticulate",
+      `Typology` == "retoucheduse on breakage" ~ "tool/retouched piece",
+      `Typology` == "inversenotch/endscraper" ~ "tool/notch-end scraper",
+      `Typology` == "alternatingside scraper" ~ "tool/side scraper",
+      `Typology` == "alternatedouble side scraper" ~ "tool/double side scraper",
+      `Typology` == "truncation/sidescaraper" ~ "tool/truncated piece",
+      `Typology` == "doublenotch" ~ "tool/double notch",
+      `Typology` == "borer/awl" ~ "tool/perforator-awl",
+      `Typology` == "roundscraper?" ~ "tool/round scraper",
+      `Typology` == "doublenotch scraper" ~ "tool/double notch scraper",
+      `Typology` == "corerejuvention (core tablet?)" ~ "core/core tablet",
+      `Typology` == "truncationاریبب" ~ "tool/truncated piece",
+      `Typology` == "retouchedpiece (side scraper)" ~ "tool/side scraper",
+      `Typology` == "retochedpice" ~ "tool/retouched piece",
+      `Typology` == "truncation/notch" ~ "tool/truncated piece-notch",
+      `Typology` == "doublealternate scraper" ~ "tool/double side scraper",
+      `Typology` == "awl/ inverse notch" ~ "tool/notch",
+      `Typology` == "retouchedpiece (alternate scraper?)" ~ "tool/double side scraper",
+      `Typology` == "inverseside scraper" ~ "tool/side scraper",
+      `Typology` == "usedcore rejuvention" ~ "core/rejuvenation piece",
+      `Typology` == "coreside rejuvention" ~ "core/rejuvenation piece",
+      `Typology` == "coreplatform rejuvention flake" ~ "core/rejuvenation piece",
+      `Typology` == "corerejuvention platform" ~ "core/rejuvenation piece",
+      `Typology` == "coreside rejuvention blade" ~ "core/rejuvenation piece",
+      `Typology` == "coreside rejuvention blade??" ~ "core/rejuvenation piece",
+      `Typology` == "coreedge rejuvention" ~ "core/rejuvenation piece",
+      `Typology` == "coreedge rejivention" ~ "core/rejuvenation piece",
+      `Typology` == "retouchedpice (side scraper?)" ~ "tool/side scraper" ,
+      `Typology` == "scraper" ~ "tool/side scraper",
+      `Typology` == "sickleblade" ~ "tool/sickle shine",
+      `Typology` == "coretool (scraper)" ~ "tool/scraper-on core piece",
+      `Typology` == "retouchedpiece (scraper)" ~ "tool/side scraper",
+      `Typology` == "inversedenticulate?" ~ "tool/denticulate",
+      `Typology` == "retouchedpiece (scraper?)" ~ "tool/side scraper",
+      `Typology` == "coreside rejuvention flake" ~ "core/rejuvenation piece",
+      `Typology` == "truncation/endscraper" ~ "tool/endscraper on truncated piece",
+      `Typology` == "retouchedpice(double side scraper?)" ~ "tool/double side scraper",
+      `Typology` == "doublescraper" ~ "tool/double side scraper",
+      `Typology` == "dendiculate" ~ "tool/denticulate",
+      `Typology` == "double-scraper" ~ "tool/double side scraper",
+      `Typology` == "pyramidalcore" ~ "core/pyramid",
+      `Typology` == "bulletcore" ~ "core/bullet",
+      `Typology` == "corerejuvention" ~ "core/rejuvenation piece",
+      `Typology` == "pyrmidalcore" ~ "core/pyramid",
+      `Typology` == "corepreperation" ~ "core/core preparation piece",
+      `Typology` == "crested" ~ "core/crested bladelet",
+      `Typology` == "preperationblade" ~ "core/core preperation",
+      `Typology` == "burin" ~ "tool/burin",
+      `Typology` == "serrateddenticulate" ~ "tool/serrated scraper",
+      `Typology` == "saw:serrated denticulate" ~ "tool/serrated scraper",
+      `Typology` == "serratedused" ~ "tool/serrated scraper",
+      `Typology` == "corepreeration" ~ "tool/core preparation piece",
+      `Typology` == "coreprepearation" ~ "tool/core preparation piece",
+      `Typology` == "awl" ~ "tool/perforator-awl",
+      `Typology` == "notch.inverse" ~ "tool/notch",
+      `Typology` == "point.notch" ~ "tool/perforator-point;notch",
+      `Typology` == "inversescraper" ~ "tool/side scraper",
+      `Typology` == "roundendscraper" ~ "tool/round scraper",
+      `Typology` == "alternate.scraper" ~ "tool/double side scraper",
+      `Typology` == "convergentscraper/disc scraper" ~ "tool/scraper-convergent scraper",
+      `Typology` == "serratedside scraper" ~ "tool/serrated scraper",
+      `Typology` == "debri" ~ "NA/NA",
+      `Typology` == "distalpart of a drill" ~ "tool/perforator-drill",
+      `Typology` == "corepreperation?" ~ "core/core preparation piece",
+      `Typology` == "preformend of a borer" ~ "tool/perforator-borer; preform",
+      `Typology` == "used" ~ "tool/used",
+      `Typology` == "distalend of a inverse sidescraper" ~ "tool/side scraper",
+      `Typology` == "retouchedblade" ~ "tool/retouched piece",
+      `Typology` == "drill" ~ "tool/perforator-drill",
+      `Typology` == "corerejuvention tablet" ~ "core/core tablet",
+      `Typology` == "bladeletcore" ~ "core/pyramid",
+      `Typology` == "flake-bladecore" ~ "core/mixed",
       `Typology` == "bladelet-corerejuvention" ~ "core/rejuvenation piece",
-
       `Typology` == "bladelet-corefragment" ~ "core/core fragment",
-
       `Typology` == "pointedused" ~ "tool/perforator-point",
-
       `Typology` == "microburin" ~ "tool/microburin",
-
       `Typology` == "bladeletcore fragment" ~ "core/core fragment",
-
       `Typology` == "rejuventionface flake" ~ "core/rejuvenation piece",
-
       `Typology` == "corepreperation flake" ~ "core/core preparation piece",
-
       `Typology` == "corerejuvention flake"  ~ "core/rejuvenation piece",
-
       `Typology` == "unfinished  pyramid core" ~ "core/pyramid",
-
       `Typology` == "bladeletcoretablet" ~ "core/core tablet",
-
       `Typology` == "bladeletcore rejuvention" ~ "core/rejuvenation piece",
-
       `Typology` == "debitage" ~ "NA/NA",
-
       `Typology` == "bladeletcore  rejuvention" ~ "core/rejuvenation piece",
-
       `Typology` == "inverseserrated scraper" ~ "tool/serrated scraper",
-
       `Typology` == "backedknife" ~ "tool/backed knife",
-
       `Typology` == "corepreparation" ~ "core/core preparation piece",
-
       `Typology` == "primaryflake" ~ "core/primary flake",
-
       `Typology` == "pointedconvergent scraper on bladelet" ~ "tool/perforator-point",
-
       `Typology` == "trie" ~ "NA/NA",
-
       `Typology` == "part of core tablet" ~ "core/core tablet",
-
       `Typology` == "zaviedar" ~ "NA/NA",
-
       `Typology` == "convergentscraper" ~ "tool/convergent scraper",
-
       `Typology` == "preparationflake" ~ "core/core preparation piece",
-
       `Typology` == "atypicalborer" ~ "tool/perforator-borer",
-
       `Typology` == "naturalybacked knife" ~ "tool/backed knife-naturally backed knife",
-
       `Typology` == "pyramidbladelet core" ~ "core/pyramid",
-
       `Typology` == "flatbladelet core" ~ "core/flat",
-
       `Typology` == "bladecore" ~ "core/pyramid",
-
       `Typology` == "bulletbladelet core" ~ "core/bullet",
-
       `Typology` == "coerejuvention flake" ~ "core/rejuvenation piece",
-
       `Typology` == "finishedbladelet core" ~ "core/bullet",
-
       `Typology` == "failed pyramid core?" ~ "core/pyramid",
-
       `Typology` == "pyramidplain bladelet core" ~ "core/pyramid",
-
       `Typology` == "coeon flake" ~ "core/core on flake",
-
       `Typology` == "coeejuvention flake" ~ "core/rejuvenation piece",
-
       `Typology` == "bladeletcoe rejuvention tablet?" ~ "core/core tablet",
-
       `Typology` == "pyramidcore" ~ "core/pyramid",
-
       `Typology` == "bladeletcore tablet" ~ "core/core tablet",
-
       `Typology` == "pyramidbladeleet core" ~ "core/pyramid",
-
       `Typology` == "flake" ~ "NA/NA",
-
       `Typology` == "serrated" ~ "tool/serrated scraper",
-
       `Typology` == "curvedretouched piece" ~ "tool/retouched piece",
-
       `Typology` == "pyramidalbladelet core" ~ "core/pyramid",
-
       `Typology` == "multidirectionalcore fragment" ~ "core/multidirectional core fragment",
-
       `Typology` == "unidirectionalblade? core fragent" ~ "core/pyramid",
-
       `Typology` == "backed/corerejuvention" ~ "tool/backed knife",
-
       `Typology` == "unidirectionalbladelet core" ~ "core/pyramid",
-
       `Typology` == "unifacialbalde core fragment?" ~ "core/pyramid",
-
       `Typology` == "flakedetached from a bladelet core" ~ "core/core preparation piece",
-
       `Typology` == "corepreparation tablet" ~ "core/core tablet",
-
       `Typology` == "bladeletcore fragent" ~ "core/core fragment-pyramid",
-
       `Typology` == "alternatescraper??" ~ "tool/double side scraper",
-
       `Typology` == "endscraper.onrejuvention" ~ "core/rejuvenation piece",
-
       `Typology` == "inverseserrated" ~ "tool/serrated scraper",
-
       `Typology` == "patination.omitted" ~ "NA/NA",
-
       `Typology` == "trapze" ~ "tool/geometric-triangle",
-
       `Typology` == "trapzoid" ~ "tool/geometric-triangle",
-
       `Typology` == "borer.drill" ~ "tool/perforator-drill",
-
       `Typology` == "serraed" ~ "tool/serrated scraper",
-
       `Typology` == "bladeletcore on a flake" ~ "core/core on flake",
-
       `Typology` == "geofact" ~ "NA/NA",
-
       `Typology` == "denticulateborer?" ~ "tool/perforator-borer;denticulate",
-
       `Typology` == "sideborer-denticulate" ~ "tool/perforator-borer;denticulate",
-
       `Typology` == "borer.awl" ~ "tool/perforator-awl",
-
       `Typology` == "roundscraper" ~ "tool/round scraper",
-
       `Typology` == "truncationon a bladelet core" ~ "tool/truncated piece",
-
       `Typology` == "truncation.notch" ~ "tool/truncated piece;notch",
-
       `Typology` == "backed-denticulate" ~ "tool/backed;denticulate",
-
       `Typology` == "brokendrill" ~ "tool/perforator-drill",
-
       `Typology` == "drillbroken" ~ "tool/perforator-drill",
-
       `Typology` == "point/broken drill" ~ "tool/perforator-point",
-
       `Typology` == "brokendrill?" ~ "tool/perforator-drill",
-
       `Typology` == "partialyserrated" ~ "tool/serrated scraper",
-
       `Typology` == "naturallybacked" ~ "tool/backed knife-naturally backed knife",
-
       `Typology` == "transversescraper" ~ "tool/transverse scraper",
-
       `Typology` == "naturalybacked" ~ "tool/backed knife-naturally backed knife",
-
       `Typology` == "awl-notch" ~ "tool/perforator-awl",
-
       `Typology` == "psedulevalois" ~ "tool/psedulevalois flake",
-
       `Typology` == "alternatingdenticulate" ~ "tool/denticulate",
-
       `Typology` == "usedcore fragment" ~ "tool/core tool-used core fragment",
-
       `Typology` == "multidirectionalmicroflake core" ~ "core/multidirectional flake core",
-
       `Typology` == "mixedpyramidal core" ~ "core/pyramid",
-
       `Typology` == "usedbladecore fragment" ~ "tool/core tool-used core fragment",
-
       `Typology` == "multidirectionalbladelet core fragment" ~ "core/multidirectional bladelet core",
-
       `Typology` == "unidirectionalbladelet core.cylinder" ~ "core/prismatic",
-
       `Typology` == "pyrymedalmixed core" ~ "core/pyramid",
-
       `Typology` == "pyramedial/bullet core" ~ "core/bullet",
-
       `Typology` == "corefragent" ~ "core/core fragment",
-
       `Typology` == "pyramedialbladelet core fragment" ~ "core/pyramid",
-
       `Typology` == "unidirectionalbladelet core fragment" ~ "core/pyramid",
-
       `Typology` == "unifacialunidirectional bladelet core" ~ "core/pyramid",
-
       `Typology` == "unidirectionalblade core" ~ "core/pyramid",
-
       `Typology` == "pyramidalbladelet core fragment" ~ "core/pyramid",
-
       `Typology` == "corerejuvention?" ~ "core/rejuvenation piece",
-
       `Typology` == "flatburin? point?" ~ "tool/flat burin",
-
       `Typology` == "primaryblade" ~ "core/primary blade",
-
       `Typology` == "trihedralgeometric" ~ "tool/geometric-triangle",
-
       `Typology` == "lunate" ~ "tool/geometric-lunate",
-
       `Typology` == "corepaltform rejuvention flake" ~ "core/rejuvenation piece",
-
       `Typology` == "pressurebladelet core /semi flat" ~ "core/pyramid",
-
       `Typology` == "failedpercussion blade core/semi pyramedal" ~ "core/pyramid",
-
       `Typology` == "percussionbladelet core/unidirectional" ~ "core/pyramid-percussion",
-
       `Typology` == "percussion.unidirectional.blade-flakecore" ~ "core/mixed-percussion",
-
       `Typology` == "obliqueretouched bladelet" ~ "tool/retouched piece",
-
       `Typology` == "percussionunidirectional pyramidal  bladelet core" ~ "core/pyramid",
-
       `Typology` == "lunategeometric" ~ "tool/geometric-lunate",
-
       `Typology` == "awl on a core rejuvention" ~ "tool/perforator-awl on core rejuvenation piece",
-
       `Typology` == "truncated" ~ "tool/truncated piece",
-
       `Typology` == "multiplenotch" ~ "tool/notch",
-
       `Typology` == "dishedconvergent bladelet" ~ "tool/convergent scraper",
-
       `Typology` == "bipolarpercussion blade-bladelet core" ~ "core/bipolar-percussion",
-
       `Typology` == "bladeletbullet core" ~ "core/bullet",
-
       `Typology` == "alternatingserrated" ~ "tool/serrated scraper",
-
       `Typology` == "shaplesscore" ~ "core/shapeless",
-
       `Typology` == "corepreparation?" ~ "core/core preparation piece",
-
       `Typology` == "obliqueretouched" ~ "tool/retouched piece",
-
       `Typology` == "corepreparation flake" ~ "core/core preparation piece",
-
       `Typology` == "backeddenticulate" ~ "tool/backed;denticulate",
-
       `Typology` == "awlnotch" ~ "tool/perforator-awl;notch",
-
       `Typology` == "scraperburin?" ~ "tool/burin;scraper",
-
       `Typology` == "notchround scraper" ~ "tool/round scraper;notch",
-
       `Typology` == "corerejuvention tablet round scraper" ~ "tool/round scraper on core tablet",
-
       `Typology` == "alternatescraper" ~ "tool/double side scraper",
-
       `Typology` == "coreplatform preperation flake" ~ "core/rejuvenation piece",
-
       `Typology` == "coreplatform rejuvention tablet" ~ "core/core tablet",
-
       `Typology` == "brokenpyramidal core bladelet" ~ "core/pyramid",
-
       `Typology` == "pyramidalbladelet core/semi bullet?" ~ "core/bullet",
-
       `Typology` == "coreside rejuvention flake?" ~ "core/rejuvenation piece-side",
-
       `Typology` == "unidirectionalpressure bladelet core/unifacila" ~ "core/pyramid",
-
       `Typology` == "coreplatform rejuvention tablet/or multidirectional core" ~ "core/rejuvenation piece",
-
       `Typology` == "bulletbladlet core" ~ "core/bullet",
-
       `Typology` == "cylinderbipolar bladelet core" ~ "core/prismatic",
-
       `Typology` == "backedknife/core platform rejuvenation" ~ "tool/backed knife",
-
       `Typology` == "flatunifacial bladelet core" ~ "core/flat",
-
       `Typology` == "coreprepration flake" ~ "core/core preparation piece",
-
       `Typology` == "coreplatform preperation" ~ "core/core preparation piece-platform",
-
       `Typology` == "borerpreform" ~ "tool/perforator-borer",
-
       `Typology` == "rejuvention" ~ "core/rejuvenation piece",
-
       `Typology` == "borerpreform?" ~ "tool/perforator-borer",
-
       `Typology` == "inversnotch" ~ "tool/notch",
-
       `Typology` == "alternateconvergent" ~ "tool/convergent scraper",
-
       `Typology` == "unidirectionalbladelet/flake-blade core" ~ "core/mixed",
-
       `Typology` == "unidirectionalmixed core" ~ "core/mixed",
-
       `Typology` == "unidirectionalflat bladelet core" ~ "core/flat",
-
       `Typology` == "" ~ "NA-NA",
       `Typology` == "na" ~ "NA-NA",
-
+      `Typology` == "bladelet-corerejuvention" ~ "core/rejuvenation piece",
+      `Typology` == "bladelet-corefragment" ~ "core/core fragment",
+      `Typology` == "pointedused" ~ "tool/perforator-point",
+      `Typology` == "microburin" ~ "tool/microburin",
+      `Typology` == "bladeletcore fragment" ~ "core/core fragment",
+      `Typology` == "rejuventionface flake" ~ "core/rejuvenation piece",
+      `Typology` == "corepreperation flake" ~ "core/core preparation piece",
+      `Typology` == "corerejuvention flake"  ~ "core/rejuvenation piece",
+      `Typology` == "unfinished  pyramid core" ~ "core/pyramid",
+      `Typology` == "bladeletcoretablet" ~ "core/core tablet",
+      `Typology` == "bladeletcore rejuvention" ~ "core/rejuvenation piece",
+      `Typology` == "debitage" ~ "NA/NA",
+      `Typology` == "bladeletcore  rejuvention" ~ "core/rejuvenation piece",
+      `Typology` == "inverseserrated scraper" ~ "tool/serrated scraper",
+      `Typology` == "backedknife" ~ "tool/backed knife",
+      `Typology` == "corepreparation" ~ "core/core preparation piece",
+      `Typology` == "primaryflake" ~ "core/primary flake",
+      `Typology` == "pointedconvergent scraper on bladelet" ~ "tool/perforator-point",
+      `Typology` == "trie" ~ "NA/NA",
+      `Typology` == "part of core tablet" ~ "core/core tablet",
+      `Typology` == "zaviedar" ~ "NA/NA",
+      `Typology` == "convergentscraper" ~ "tool/convergent scraper",
+      `Typology` == "preparationflake" ~ "core/core preparation piece",
+      `Typology` == "atypicalborer" ~ "tool/perforator-borer",
+      `Typology` == "naturalybacked knife" ~ "tool/backed knife-naturally backed knife",
+      `Typology` == "pyramidbladelet core" ~ "core/pyramid",
+      `Typology` == "flatbladelet core" ~ "core/flat",
+      `Typology` == "bladecore" ~ "core/pyramid",
+      `Typology` == "bulletbladelet core" ~ "core/bullet",
+      `Typology` == "coerejuvention flake" ~ "core/rejuvenation piece",
+      `Typology` == "finishedbladelet core" ~ "core/bullet",
+      `Typology` == "failed pyramid core?" ~ "core/pyramid",
+      `Typology` == "pyramidplain bladelet core" ~ "core/pyramid",
+      `Typology` == "coeon flake" ~ "core/core on flake",
+      `Typology` == "coeejuvention flake" ~ "core/rejuvenation piece",
+      `Typology` == "bladeletcoe rejuvention tablet?" ~ "core/core tablet",
+      `Typology` == "pyramidcore" ~ "core/pyramid",
+      `Typology` == "bladeletcore tablet" ~ "core/core tablet",
+      `Typology` == "pyramidbladeleet core" ~ "core/pyramid",
+      `Typology` == "flake" ~ "NA/NA",
+      `Typology` == "serrated" ~ "tool/serrated scraper",
+      `Typology` == "curvedretouched piece" ~ "tool/retouched piece",
+      `Typology` == "pyramidalbladelet core" ~ "core/pyramid",
+      `Typology` == "multidirectionalcore fragment" ~ "core/multidirectional core fragment",
+      `Typology` == "unidirectionalblade? core fragent" ~ "core/pyramid",
+      `Typology` == "backed/corerejuvention" ~ "tool/backed knife",
+      `Typology` == "unidirectionalbladelet core" ~ "core/pyramid",
+      `Typology` == "unifacialbalde core fragment?" ~ "core/pyramid",
+      `Typology` == "flakedetached from a bladelet core" ~ "core/core preparation piece",
+      `Typology` == "corepreparation tablet" ~ "core/core tablet",
+      `Typology` == "bladeletcore fragent" ~ "core/core fragment-pyramid",
+      `Typology` == "alternatescraper??" ~ "tool/double side scraper",
+      `Typology` == "endscraper.onrejuvention" ~ "core/rejuvenation piece",
+      `Typology` == "inverseserrated" ~ "tool/serrated scraper",
+      `Typology` == "patination.omitted" ~ "NA/NA",
+      `Typology` == "trapze" ~ "tool/geometric-triangle",
+      `Typology` == "trapzoid" ~ "tool/geometric-triangle",
+      `Typology` == "borer.drill" ~ "tool/perforator-drill",
+      `Typology` == "serraed" ~ "tool/serrated scraper",
+      `Typology` == "bladeletcore on a flake" ~ "core/core on flake",
+      `Typology` == "geofact" ~ "NA/NA",
+      `Typology` == "denticulateborer?" ~ "tool/perforator-borer;denticulate",
+      `Typology` == "sideborer-denticulate" ~ "tool/perforator-borer;denticulate",
+      `Typology` == "borer.awl" ~ "tool/perforator-awl",
+      `Typology` == "roundscraper" ~ "tool/round scraper",
+      `Typology` == "truncationon a bladelet core" ~ "tool/truncated piece",
+      `Typology` == "truncation.notch" ~ "tool/truncated piece;notch",
+      `Typology` == "backed-denticulate" ~ "tool/backed;denticulate",
+      `Typology` == "brokendrill" ~ "tool/perforator-drill",
+      `Typology` == "drillbroken" ~ "tool/perforator-drill",
+      `Typology` == "point/broken drill" ~ "tool/perforator-point",
+      `Typology` == "brokendrill?" ~ "tool/perforator-drill",
+      `Typology` == "partialyserrated" ~ "tool/serrated scraper",
+      `Typology` == "naturallybacked" ~ "tool/backed knife-naturally backed knife",
+      `Typology` == "transversescraper" ~ "tool/transverse scraper",
+      `Typology` == "naturalybacked" ~ "tool/backed knife-naturally backed knife",
+      `Typology` == "awl-notch" ~ "tool/perforator-awl",
+      `Typology` == "psedulevalois" ~ "tool/psedulevalois flake",
+      `Typology` == "alternatingdenticulate" ~ "tool/denticulate",
+      `Typology` == "usedcore fragment" ~ "tool/core tool-used core fragment",
+      `Typology` == "multidirectionalmicroflake core" ~ "core/multidirectional flake core",
+      `Typology` == "mixedpyramidal core" ~ "core/pyramid",
+      `Typology` == "usedbladecore fragment" ~ "tool/core tool-used core fragment",
+      `Typology` == "multidirectionalbladelet core fragment" ~ "core/multidirectional bladelet core",
+      `Typology` == "unidirectionalbladelet core.cylinder" ~ "core/prismatic",
+      `Typology` == "pyrymedalmixed core" ~ "core/pyramid",
+      `Typology` == "pyramedial/bullet core" ~ "core/bullet",
+      `Typology` == "corefragent" ~ "core/core fragment",
+      `Typology` == "pyramedialbladelet core fragment" ~ "core/pyramid",
+      `Typology` == "unidirectionalbladelet core fragment" ~ "core/pyramid",
+      `Typology` == "unifacialunidirectional bladelet core" ~ "core/pyramid",
+      `Typology` == "unidirectionalblade core" ~ "core/pyramid",
+      `Typology` == "pyramidalbladelet core fragment" ~ "core/pyramid",
+      `Typology` == "corerejuvention?" ~ "core/rejuvenation piece",
+      `Typology` == "flatburin? point?" ~ "tool/flat burin",
+      `Typology` == "primaryblade" ~ "core/primary blade",
+      `Typology` == "trihedralgeometric" ~ "tool/geometric-triangle",
+      `Typology` == "lunate" ~ "tool/geometric-lunate",
+      `Typology` == "corepaltform rejuvention flake" ~ "core/rejuvenation piece",
+      `Typology` == "pressurebladelet core /semi flat" ~ "core/pyramid",
+      `Typology` == "failedpercussion blade core/semi pyramedal" ~ "core/pyramid",
+      `Typology` == "percussionbladelet core/unidirectional" ~ "core/pyramid-percussion",
+      `Typology` == "percussion.unidirectional.blade-flakecore" ~ "core/mixed-percussion",
+      `Typology` == "obliqueretouched bladelet" ~ "tool/retouched piece",
+      `Typology` == "percussionunidirectional pyramidal  bladelet core" ~ "core/pyramid",
+      `Typology` == "lunategeometric" ~ "tool/geometric-lunate",
+      `Typology` == "awl on a core rejuvention" ~ "tool/perforator-awl on core rejuvenation piece",
+      `Typology` == "truncated" ~ "tool/truncated piece",
+      `Typology` == "multiplenotch" ~ "tool/notch",
+      `Typology` == "dishedconvergent bladelet" ~ "tool/convergent scraper",
+      `Typology` == "bipolarpercussion blade-bladelet core" ~ "core/bipolar-percussion",
+      `Typology` == "bladeletbullet core" ~ "core/bullet",
+      `Typology` == "alternatingserrated" ~ "tool/serrated scraper",
+      `Typology` == "shaplesscore" ~ "core/shapeless",
+      `Typology` == "corepreparation?" ~ "core/core preparation piece",
+      `Typology` == "obliqueretouched" ~ "tool/retouched piece",
+      `Typology` == "corepreparation flake" ~ "core/core preparation piece",
+      `Typology` == "backeddenticulate" ~ "tool/backed;denticulate",
+      `Typology` == "awlnotch" ~ "tool/perforator-awl;notch",
+      `Typology` == "scraperburin?" ~ "tool/burin;scraper",
+      `Typology` == "notchround scraper" ~ "tool/round scraper;notch",
+      `Typology` == "corerejuvention tablet round scraper" ~ "tool/round scraper on core tablet",
+      `Typology` == "alternatescraper" ~ "tool/double side scraper",
+      `Typology` == "coreplatform preperation flake" ~ "core/rejuvenation piece",
+      `Typology` == "coreplatform rejuvention tablet" ~ "core/core tablet",
+      `Typology` == "brokenpyramidal core bladelet" ~ "core/pyramid",
+      `Typology` == "pyramidalbladelet core/semi bullet?" ~ "core/bullet",
+      `Typology` == "coreside rejuvention flake?" ~ "core/rejuvenation piece-side",
+      `Typology` == "unidirectionalpressure bladelet core/unifacila" ~ "core/pyramid",
+      `Typology` == "coreplatform rejuvention tablet/or multidirectional core" ~ "core/rejuvenation piece",
+      `Typology` == "bulletbladlet core" ~ "core/bullet",
+      `Typology` == "cylinderbipolar bladelet core" ~ "core/prismatic",
+      `Typology` == "backedknife/core platform rejuvenation" ~ "tool/backed knife",
+      `Typology` == "flatunifacial bladelet core" ~ "core/flat",
+      `Typology` == "coreprepration flake" ~ "core/core preparation piece",
+      `Typology` == "coreplatform preperation" ~ "core/core preparation piece-platform",
+      `Typology` == "borerpreform" ~ "tool/perforator-borer",
+      `Typology` == "rejuvention" ~ "core/rejuvenation piece",
+      `Typology` == "borerpreform?" ~ "tool/perforator-borer",
+      `Typology` == "inversnotch" ~ "tool/notch",
+      `Typology` == "alternateconvergent" ~ "tool/convergent scraper",
+      `Typology` == "unidirectionalbladelet/flake-blade core" ~ "core/mixed",
+      `Typology` == "unidirectionalmixed core" ~ "core/mixed",
+      `Typology` == "unidirectionalflat bladelet core" ~ "core/flat",
+      `Typology` == "" ~ "NA-NA",
+      `Typology` == "na" ~ "NA-NA",
       TRUE ~ `Typology`
-  ))
-
-tl4 <- tl4 %>%
+    )
+  )  %>%
   mutate(
@@
   # remove temporary columns
-  select(-typology_split, -tool_or_core)
-
-unique(tl4$`tool-typology`)
-unique(tl4$`core-fragment`)
-```
-
-```{r}
-# Exploring datra
+  select(-typology_split,-tool_or_core)
+
+# BM: still some typos in there
+rev(sort(table((tl4$`tool-typology`))))
+rev(sort(table((tl4$`core-fragment`))))
+```
+
+```{r}
+# Exploring data
 
 tl4 %>% 
+  # BM: we need this because of the mutate_all ifelse combo you have above
+  mutate(midpoint = parse_number(midpoint)) %>% 
   group_by(midpoint) %>% 
@@
 # convert all columns to character type (I DONT KNOW WHY I DID THIS!)
+# BM: my preference is not to see comments and code like this. Go ahead and delete it.
 tl4 <- tl4 %>%
@@
 tl4 <- tl4 %>%
+  # BM: in my opinion this is nicer than mutate_all and ifelse
   mutate(across(everything(), ~na_if(., "NA"))) %>%
@@
 
-print(tl4)
-
-```
-
-```{r}
-#refine the table
-
-library(dplyr)
+```
+
+```{r}
+
+# refine the table
 
 tl_final <- tl4 %>%
-
   rename(
@@
     eraillure =            Eraillure , 
-    cortex = `Cortex (%)`)
-
-```
-
-```{r}
-tl_final <- tl_final %>%
+    cortex =               `Cortex (%)`)  %>%
   mutate(al = paste0(toupper(area), "_", layer)) %>%
-  relocate(al, .after = layer)
-
-```
-
-```{r}
-tl_final <- tl_final  %>%
-
+  relocate(al, .after = layer)   %>%
   select(lithic_id, 
@@
          blank, `blank part`, 
-
         `core-typology`, `core-fragment`,`core-technology`, 
-
-        cortex)
-```
-
-```{r}
-#set the value of utilization to NA only for the rows where `sickle shine` == 1. 
-#All other rows should remain unchanged
-
-#unsuccessful, should try again later
-
-```
-
-```{r}
+        cortex)  %>% 
 # Update the 'typology-tool-core' column based on 'Retouch'
-tl_final <- tl_final %>% 
   mutate(
@@
     )
-  )
-
-print(tl_final)
-
-```
-
-```{r}
-
-tl_final <- tl_final %>% 
+  ) %>% 
   mutate(
@@
     )
-  )
-
-print(tl_final)
-
-```
-
-```{r}
-tl_final <- tl_final %>% 
+  ) %>% 
   mutate(
     `tool-typology` = case_when(
-
      # If Retouch is 1 and tool-typology is NA
@@
 
-# Show updated data frame
-unique(tl_final$`tool-typology`)
+# BM: take a look at tool typology
+rev(sort(table(tl_final$`tool-typology`)))
+
+```
+
+```{r}
+#set the value of utilization to NA only for the rows where `sickle shine` == 1. 
+#All other rows should remain unchanged
+
+#unsuccessful, should try again later
 
@@
       `tool-typology` %in%
-        c("end scraper", "side scraper", "double side scraper", "convergent scraper", "round scraper", "perforator-convergent scraper", "perforator-borer; preform", "scraper-on core piece", "endscraper on truncated piece", "round scraper", "transverse scraper", "core tool-used core fragment", "round scraper on core tablet") ~ "scraper",
+        c(
+          "end scraper",
+          "side scraper",
+          "double side scraper",
+          "convergent scraper",
+          "round scraper",
+          "perforator-convergent scraper",
+          "perforator-borer; preform",
+          "scraper-on core piece",
+          "endscraper on truncated piece",
+          "round scraper",
+          "transverse scraper",
+          "core tool-used core fragment",
+          "round scraper on core tablet"
+        ) ~ "scraper",
       `tool-typology` %in%
-        c("perforator-borer", "perforator-drill", "perforator-point", "perforator-awl", "perforator-point;notch", "perforator-awl on core rejuvenation piece", "perforator-awl;notch", "perforator-borer;denticulate") ~ "Perforator",
+        c(
+          "perforator-borer",
+          "perforator-drill",
+          "perforator-point",
+          "perforator-awl",
+          "perforator-point;notch",
+          "perforator-awl on core rejuvenation piece",
+          "perforator-awl;notch",
+          "perforator-borer;denticulate"
+        ) ~ "Perforator",
       `tool-typology` %in%
-        c("backed knife", "backed knife-naturally backed knife", "backed;denticulate") ~ "backed pieces",
-      `tool-typology` %in% c("double notch", "denticulate", " denticulate", "notch-denticulate", "notch") ~ "denticulate-notch",
-      `tool-typology` %in% c("notch-side scraper", "notch-end scraper", "double notch scraper", "round scraper;notch") ~ "scraper-notch",
+        c(
+          "backed knife",
+          "backed knife-naturally backed knife",
+          "backed;denticulate"
+        ) ~ "backed pieces",
+      `tool-typology` %in% c(
+        "double notch",
+        "denticulate",
+        " denticulate",
+        "notch-denticulate",
+        "notch"
+      ) ~ "denticulate-notch",
+      `tool-typology` %in% c(
+        "notch-side scraper",
+        "notch-end scraper",
+        "double notch scraper",
+        "round scraper;notch"
+      ) ~ "scraper-notch",
       `tool-typology` %in% c("geometric-triangle", "geometric-lunate") ~ "geometric",
       `tool-typology` == "serrated scraper" ~ "serrated scraper",
-      `tool-typology` %in% c("truncated piece", "truncated piece-notch", "truncated piece;notch", "notch;truncated piece") ~ "truncated pieces",
+      `tool-typology` %in% c(
+        "truncated piece",
+        "truncated piece-notch",
+        "truncated piece;notch",
+        "notch;truncated piece"
+      ) ~ "truncated pieces",
 
@@
 # Checking the unique values of tooltypegroup
-unique(tl_final$tooltypegroup)
+rev(sort(table(tl_final$tooltypegroup)))
 
@@
 tl_final <- tl_final %>%
+  mutate(`core-fragment` = if_else(!is.na(`core-typology`),
+                                   NA_character_,
+                                   `core-fragment`))  %>%
+  mutate(`core-fragment` = if_else(!is.na(`core-typology`),
+                                   NA_character_,
+                                   `core-fragment`)) %>%
   mutate(
-    `core-fragment` = if_else(
-      !is.na(`core-typology`),  
-      NA_character_,            
-      `core-fragment`           
-    )
-  )
-
-print(tl_final)
-
-```
-
-```{r}
-
-library(dplyr)
-library(stringr)
-
-tl_final <- tl_final %>%
-  mutate(
-    `core-fragment` = if_else(
-      !is.na(`core-typology`),
-      NA_character_,
-      `core-fragment`
+    `core-fragment` = case_when(
+      `core-fragment` == "rejuvenation piece" ~ "rejuvenation piece",
+      `core-fragment` == "rejuvention piece-side" ~ "rejuvenation piece",
+      `core-fragment` == "core preparation piece-platform" ~ "core tablet",
+      `core-fragment` == "core preparation" ~ "rejuvenation piece",
+      `core-fragment` == "core preparation piece" ~ "rejuvenation piece",
+      `core-fragment` == "core preperation" ~ "rejuvenation piece",
+      `core-fragment` == "pyramid" ~ "pyramid",
+      `core-fragment` == "pyramid core" ~ "pyramid",
+      `core-fragment` == "pyramid-unidirectional" ~ "pyramid",
+      `core-fragment` == "core fragment-pyramid" ~ "pyramid",
+      `core-fragment` == "cylinder" ~ "cylinder/prismatic",
+      `core-fragment` == "prismatic" ~ "cylinder/prismatic",
+      `core-fragment` == "shapeless" ~ "cylinder/prismatic",
+      `core-fragment` == "multidirectional core fragment" ~ "shapeless",
+      `core-fragment` == "multidirectional flake core" ~ "shapeless",
+      `core-fragment` == "multidirectional bladelet core" ~ "shapeless",
+      `core-fragment` == "mixed core" ~ "shapeless",
+      `core-fragment` == "mixed-percussion" ~ "shapeless",
+      `core-fragment` == "mixed" ~ "shapeless",
+      `core-fragment` == ("bipolar-percussion") ~ "bipolar",
+      `core-fragment` == "flat" ~ "flat",
+      `core-fragment` == "core on flake" ~ "core on flake",
+      `core-fragment` == "bullet" ~ "bullet",
+      TRUE ~ `core-fragment`
 
@@
   )  %>%
-
-  mutate(`core-fragment` = case_when(
-    `core-fragment` == "rejuvenation piece" ~ "rejuvenation piece",
-
-             `core-fragment` == "rejuvention piece-side" ~ "rejuvenation piece",
-
-             `core-fragment` == "core preparation piece-platform" ~ "core tablet",
-
-             `core-fragment` == "core preparation" ~ "rejuvenation piece",
-
-             `core-fragment` == "core preparation piece" ~ "rejuvenation piece",
-
-             `core-fragment` == "core preperation" ~ "rejuvenation piece",
-
-             `core-fragment` == "pyramid" ~ "pyramid",
-
-             `core-fragment` == "pyramid core" ~ "pyramid",
-
-             `core-fragment` == "pyramid-unidirectional" ~ "pyramid",
-
-             `core-fragment` == "core fragment-pyramid" ~ "pyramid",
-
-             `core-fragment` == "cylinder" ~ "cylinder/prismatic",
-
-             `core-fragment` == "prismatic" ~ "cylinder/prismatic",
-
-             `core-fragment` == "shapeless" ~ "cylinder/prismatic",
-
-             `core-fragment` == "multidirectional core fragment" ~ "shapeless",
-
-             `core-fragment` == "multidirectional flake core" ~ "shapeless",
-
-             `core-fragment` == "multidirectional bladelet core" ~ "shapeless",
-
-             `core-fragment` == "mixed core" ~ "shapeless",
-
-             `core-fragment` == "mixed-percussion" ~ "shapeless",
-
-             `core-fragment` == "mixed" ~ "shapeless",
-
-             `core-fragment` == ("bipolar-percussion") ~ "bipolar",
-
-             `core-fragment` == "flat" ~ "flat",
-
-             `core-fragment` == "core on flake" ~ "core on flake",
-
-             `core-fragment` == "bullet" ~ "bullet",
-
-    TRUE ~ `core-fragment`
-
-  ))
-
-```
-
-```{r}
-
-tl_final <- tl_final %>%
   mutate(
     core = case_when(
-      `core-fragment` %in% c("pyramid", "pyramid core", "pyramid-unidirectional", "core fragment-pyramid", "cylinder/prismatic", "shapeless", "multidirectional core fragment", "multidirectional flake core", "multidirectional bladelet core", "mixed core", "mixed-percussion", "mixed", "bipolar-percussion", "flat", "core on flake", "bullet") ~ `core-fragment`,
+      `core-fragment` %in% c(
+        "pyramid",
+        "pyramid core",
+        "pyramid-unidirectional",
+        "core fragment-pyramid",
+        "cylinder/prismatic",
+        "shapeless",
+        "multidirectional core fragment",
+        "multidirectional flake core",
+        "multidirectional bladelet core",
+        "mixed core",
+        "mixed-percussion",
+        "mixed",
+        "bipolar-percussion",
+        "flat",
+        "core on flake",
+        "bullet"
+      ) ~ `core-fragment`,
       TRUE ~ NA_character_
@@
     corefrag = case_when(
-      `core-fragment` %in% c("rejuvenation piece", "rejuvention piece-side", "core preparation piece-platform", "core preparation", "core preparation piece", "core preperation", "crested bladelet" , "primary flake" , "primary blade" ) ~ `core-fragment`,
+      `core-fragment` %in% c(
+        "rejuvenation piece",
+        "rejuvention piece-side",
+        "core preparation piece-platform",
+        "core preparation",
+        "core preparation piece",
+        "core preperation",
+        "crested bladelet" ,
+        "primary flake" ,
+        "primary blade"
+      ) ~ `core-fragment`,
       TRUE ~ NA_character_
     )
-  )
-
-unique(`tl_final`$`core`)
-unique(`tl_final`$`corefrag`)
-```
-
-```{r}
-
-library(dplyr)
-
-tl_final <- tl_final %>%
+  ) %>%
   # Convert empty strings to NA in `core-fragment` column
@@
   # Convert "pyramid-percussion" to "pyramid" in `core-fragment` column
-  mutate(`core-fragment` = 
+  mutate(
+    `core-fragment` =
       if_else(
-             `core-fragment` == "pyramid-percussion", "pyramid", `core-fragment`)) %>%
+        `core-fragment` == "pyramid-percussion",
+        "pyramid",
+        `core-fragment`
+      )
+  ) %>%
   # Update existing `core` column with selected values from `core-fragment` column
-  mutate(core = if_else(`core-fragment` %in% 
-                          c("pyramid", 
+  mutate(core = if_else(
+    `core-fragment` %in%
+      c(
+        "pyramid",
         "cylinder/prismatic",
@@
         "core on flake",
-                            "pyramid"), 
+        "pyramid"
+      ),
     `core-fragment`,
-                        core))
-
-unique(tl_final$`core-fragment`)
-unique(tl_final$`corefrag`)
-unique(tl_final$`core`)
-```
-
-```{r}
-
-library(dplyr)
-
-tl_final <- tl_final %>%
+    core
+  )) %>%
   # Add 'core tablet' and 'core fragment' from `core-fragment` to `corefrag`
-  mutate(corefrag = case_when(
+  mutate(
+    corefrag = case_when(
       `core-fragment` == "core tablet" ~ "core tablet",
@@
       TRUE ~ corefrag
-  ))
-
-unique(tl_final$`core-fragment`)
-unique(tl_final$`corefrag`)
-unique(tl_final$`core`)
-```
-
-```{r}
-
-```
-
-```{r}
-
-library(dplyr)
-
-tl_final <- tl_final %>%
+    )
+  ) %>%
   # Transfer 'rejuvenation piece' and 'core fragment' from `core-typology` to `corefrag`
-  mutate(corefrag = case_when(
+  mutate(
+    corefrag = case_when(
       `core-typology` == "rejuvenation piece" ~ "rejuvenation piece",
@@
       TRUE ~ corefrag
+    )
+  ) %>%
+  # Remove 'rejuvenation piece' and 'core fragment' from `core-typology`
+  mutate(`core-typology` = if_else(
+    `core-typology` %in% c("rejuvenation piece", "core fragment"),
+    NA_character_,
+    `core-typology`
   ))  %>% 
-  # Remove 'rejuvenation piece' and 'core fragment' from `core-typology` 
-  mutate(`core-typology` = if_else(`core-typology` %in% c("rejuvenation piece", "core fragment"), NA_character_, `core-typology`))
-
-unique(tl_final$`core-typology`)
-unique(tl_final$`corefrag`)
-unique(tl_final$`core-fragment`)
-
-```
-
-```{r}
-
-library(dplyr)
-
 # Update the 'blank' column based on conditions in the 'core-fragment' column
-tl_final <- tl_final %>%
-  mutate(blank = if_else(`core-fragment` %in% c(
-    "core fragment", "core tablet", "rejuvenation piece", "pyramid", "crested bladelet", 
-    "shapeless", "primary flake", "flat", "bullet", "core on flake", "cylinder/prismatic",
-    "primary blade", "rejuvenation piece-side", "core platform rejuvention"),
+  mutate(blank = if_else(
+    `core-fragment` %in% c(
+      "core fragment",
+      "core tablet",
+      "rejuvenation piece",
+      "pyramid",
+      "crested bladelet",
+      "shapeless",
+      "primary flake",
+      "flat",
+      "bullet",
+      "core on flake",
+      "cylinder/prismatic",
+      "primary blade",
+      "rejuvenation piece-side",
+      "core platform rejuvention"
+    ),
     NA_character_,
-    blank))
-
-print(tl_final)
-unique(tl_final$`core-fragment`)
-unique(tl_final$`blank`)
-
-```
-
-Summary tables: 1.AREA
+    blank
+  ))
+
+# take a look
+rev(sort(table(tl_final$core)))
+rev(sort(table(tl_final$corefrag)))
+rev(sort(table(tl_final$`core-fragment`)))
+rev(sort(table(tl_final$blank)))
+```
+
+Summary tables: 1. AREA
 
@@
     `Cores (n)` = sum(!is.na(core) & core != ""),
-    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag) & corefrag != ""),
+    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag) &
+                                            corefrag != ""),
     `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
@@
 
-print(summary_table_area)
+# BM: looks like this is area and layer, is that right? I see c1, d1, d2, d3, d4 etc.
+knitr::kable(summary_table_area)
 
@@
 
-AREA_LAYER:
+AREA-LAYER:
 
@@
 # Generate the table
-library(kableExtra)
 
@@
 ```{r}
-tl_final <- tl_final %>%
-  mutate(depth = gsub("cm", "", depth))
-```
-
-```{r}
 # Create a summary table based on the 'depth' column
 summary_table_depth <- tl_final %>%
+  mutate(depth = gsub("cm", "", depth)) %>%
   group_by(depth) %>%  
@@
 ```{r}
-tl_final_area <- tl_final %>%
-  filter(area %in% c("ta", "d1", "tp1"))
-
-```
-
-```{r}
-
-library(dplyr)
 
 # Prepare data for Summary table-area
-summary_table_area <- tl_final_area %>% 
+summary_table_area <- tl_final %>%
   filter(area %in% c("ta", "d1", "tp1")) %>%
@@
 
+summary_table_area <- summary_table_area %>%
+  add_row(
+    area = "Total",
+    blade_count =              sum(.$blade_count, na.rm = TRUE),
+    flake_count =              sum(.$flake_count, na.rm = TRUE),
+    bladelet_count =           sum(.$bladelet_count, na.rm = TRUE),
+    tool_count =               sum(.$tool_count, na.rm = TRUE),
+    core_count =               sum(.$core_count, na.rm = TRUE),
+    core_rejuvenation_count =  sum(.$core_rejuvenation_count, na.rm = TRUE),
+    retouched_tool_count =     sum(.$retouched_tool_count, na.rm = TRUE),
+    utilized_tool_count =      sum(.$utilized_tool_count, na.rm = TRUE),
+    shine_tool_count =         sum(.$shine_tool_count, na.rm = TRUE),
+    total =                    sum(.$total, na.rm = TRUE)  # Explicitly setting the total
+  )
+
 # Adding a new row with total counts
@@
     area = "Total",
-    blade_count = sum(summary_table_area$blade_count, na.rm = TRUE),
-    flake_count = sum(summary_table_area$flake_count, na.rm = TRUE),
-    bladelet_count = sum(summary_table_area$bladelet_count, na.rm = TRUE),
-    tool_count = sum(summary_table_area$tool_count, na.rm = TRUE),
-    core_count = sum(summary_table_area$core_count, na.rm = TRUE),
-    core_rejuvenation_count = sum(summary_table_area$core_rejuvenation_count, na.rm = TRUE),
-    retouched_tool_count = sum(summary_table_area$retouched_tool_count, na.rm = TRUE),
-    utilized_tool_count = sum(summary_table_area$utilized_tool_count, na.rm = TRUE),
-    shine_tool_count = sum(summary_table_area$shine_tool_count, na.rm = TRUE),
-    total = sum(summary_table_area$total, na.rm = TRUE)  # Explicitly setting the total
+    blade_count =              sum(.$blade_count, na.rm = TRUE),
+    flake_count =              sum(.$flake_count, na.rm = TRUE),
+    bladelet_count =           sum(.$bladelet_count, na.rm = TRUE),
+    tool_count =               sum(.$tool_count, na.rm = TRUE),
+    core_count =               sum(.$core_count, na.rm = TRUE),
+    core_rejuvenation_count =  sum(.$core_rejuvenation_count, na.rm = TRUE),
+    retouched_tool_count =     sum(.$retouched_tool_count, na.rm = TRUE),
+    utilized_tool_count =      sum(.$utilized_tool_count, na.rm = TRUE),
+    shine_tool_count =         sum(.$shine_tool_count, na.rm = TRUE),
+    total =                    sum(.$total, na.rm = TRUE)  # Explicitly setting the total
   )
@@
 # Prepare data for Summary table-al
-summary_table_al <- tl_final_area %>% 
+summary_table_al <- tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1")) %>% 
   group_by(al) %>% 
@@
     al = "Total",
-    blade_count = sum(summary_table_al$blade_count, na.rm = TRUE),
-    flake_count = sum(summary_table_al$flake_count, na.rm = TRUE),
-    bladelet_count = sum(summary_table_al$bladelet_count, na.rm = TRUE),
-    tool_count = sum(summary_table_al$tool_count, na.rm = TRUE),
-    core_count = sum(summary_table_al$core_count, na.rm = TRUE),
-    core_rejuvenation_count = sum(summary_table_al$core_rejuvenation_count, na.rm = TRUE),
-    retouched_tool_count = sum(summary_table_al$retouched_tool_count, na.rm = TRUE),
-    utilized_tool_count = sum(summary_table_al$utilized_tool_count, na.rm = TRUE),
-    shine_tool_count = sum(summary_table_al$shine_tool_count, na.rm = TRUE),
-    total = sum(summary_table_al$total, na.rm = TRUE)  # Explicitly setting the total
-  )
-
-# View the summary table
-print(summary_table_al)
+    blade_count =              sum(.$blade_count, na.rm = TRUE),
+    flake_count =              sum(.$flake_count, na.rm = TRUE),
+    bladelet_count =           sum(.$bladelet_count, na.rm = TRUE),
+    tool_count =               sum(.$tool_count, na.rm = TRUE),
+    core_count =               sum(.$core_count, na.rm = TRUE),
+    core_rejuvenation_count =  sum(.$core_rejuvenation_count, na.rm = TRUE),
+    retouched_tool_count =     sum(.$retouched_tool_count, na.rm = TRUE),
+    utilized_tool_count =      sum(.$utilized_tool_count, na.rm = TRUE),
+    shine_tool_count =         sum(.$shine_tool_count, na.rm = TRUE),
+    total =                    sum(.$total, na.rm = TRUE)  # Explicitly setting the total
+  )
 
@@
 
-library(kableExtra)
-
-kable(summary_table_al, 
-      col.names = c("Area-Layer", "Blade", "Flake", "Bladelet", "Tools", "Cores", "Core Rejuvenations", "Retouched Tools", "Utilized Tools", "Sickle Shine", "Total"), 
+kable(
+  summary_table_al,
+  col.names = c(
+    "Area-Layer",
+    "Blade",
+    "Flake",
+    "Bladelet",
+    "Tools",
+    "Cores",
+    "Core Rejuvenations",
+    "Retouched Tools",
+    "Utilized Tools",
+    "Sickle Shine",
+    "Total"
+  ),
   caption = "Table: Summary ",
-      align = 'c') %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
+  align = 'c'
+) %>%
+  kable_styling(
+    bootstrap_options = c("striped", "hover", "condensed"),
     full_width = FALSE,
-                font_size = 12) %>%
+    font_size = 12
+  ) %>%
   row_spec(0, bold = TRUE, font_size = 14)
@@
 
-library(knitr)
-library(dplyr)
 # Rename columns
@@
   )
-library(kableExtra)
 
@@
       align = 'c') %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 12) %>%
-
+  kable_styling(
+    bootstrap_options = c("striped", "hover", "condensed"),
+    font_size = 12
+  ) %>%
   row_spec(0, font_size = 11)
-
 ```
@@
 
-library(dplyr)
-library(tidyr)
-
 # Create a summary table
-summary_table_tool_area <- tl_final_area %>%
+summary_table_tool_area <- tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1")) %>%
   filter(!is.na(tooltypegroup), tooltypegroup != "") %>%  # Exclude rows where 'tooltypegroup' is NA or empty
@@
 summary_table_tool_area <- summary_table_tool_area %>%
-  mutate(Total = rowSums(select(., -tooltypegroup)))
+  mutate(Total = rowSums(select(.,-tooltypegroup)))
 
@@
   select(-tooltypegroup) %>%
-  summarise(across(everything(), sum, na.rm = TRUE)) %>%
+  summarise(across(everything(), \(x) sum(x, na.rm = TRUE))) %>%
   mutate(tooltypegroup = "Total")
 
-summary_table_tool_area <- bind_rows(summary_table_tool_area, total_row)
-
-print(summary_table_tool_area)
-
-```
-
-```{r}
-library(kableExtra)
-
-kable(summary_table_tool_area, 
+summary_table_tool_area <-
+  bind_rows(summary_table_tool_area, total_row)
+
+```
+
+```{r}
+
+kable(
+  summary_table_tool_area,
   col.names = c("Tools", "D1", "TA", "TP1", "Total"),
   caption = "Table: Tool Typology in three different main areas.",
-      align = 'c') %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
+  align = 'c'
+) %>%
+  kable_styling(
+    bootstrap_options = c("striped", "hover", "condensed"),
     full_width = F,
-                font_size = 12) %>%
+    font_size = 12
+  ) %>%
   column_spec(1, extra_css = "text-align:left;") %>%
@@
 # Create the summary table based on layer
-summary_table1_layer <- tl_final_area %>%
+summary_table1_layer <- tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1")) %>%
   group_by(layer) %>%
@@
 
+```{r}
+
+# BM: looks like you need this data starting from around here in the document
+library(googlesheets4)
+
+google_sheet_url <- 
+  "https://docs.google.com/spreadsheets/d/1Q0QZESk412ZQLE24yPs6Rg-7Y9OMX63DHoXeIPNAIFM/edit#gid=0"
+
+sum_depth_sheet <- read_sheet(google_sheet_url)
+
+sum_depth_sheet <- sum_depth_sheet %>%
+  mutate(`depth (cm)` = unlist(`depth (cm)`),
+         level = unlist(level)) %>% 
+  mutate(area_layer_level = paste0(toupper(area), 
+                                   "_", 
+                                   layer, 
+                                   "_", 
+                                   level)) %>%
+  relocate(area_layer_level, 
+           .after = last_col()) %>%
+  mutate(
+    retouch_freq = round(retouch_tool / total_lithics, 2),
+    lithic_dens = round(total_lithics / volume, 2)
+  )
+
+# BM: from my perspective, what would be ideal here is to have the 
+# excavation area and volume calculations done here in R. I feel anxious about
+# the possibilities of typos when these values are calculated somewhere I can't 
+# see and pasted into the google sheet. I prefer for the workflow of calculations
+# to be as transparent and traceable as possible. From this point in the qmd I
+# think we can use the volume values from this google sheet. Another option is to 
+# export the google sheet as a CSV, and put that in the data folder in this project. 
+# Then we can use read_csv to import it.
+
+```
+
 volume for each area:
@@
 # Define the volumes for each area 
+volumes <- data.frame(
+  area = c("ta", "tp1", "d1"),
+  volume = c(52, 5.76, 2.1375) # in m^3
+  # BM: My preference would be to show the calculation of these 
+  # values from the raw data, rather than type them in 
+  # here, where we can't see where they came from, we can't easily
+  # verify these values as they are currently 
+)
+
+# Count the number of artifacts by area
+artifact_counts <- tl_final %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  tally()
+
+# Join artifact_counts with volumes by area
+lithic_density_df <- left_join(artifact_counts, volumes, by = "area")
+
+# Compute lithic density for each area and round it to 2 decimal places
+lithic_density_df <- lithic_density_df %>%
+  mutate(lithic_density = round(n / volume, 2)) # number of artifacts / volume
+
+# Create the summary table
+summary_table2_area <- tl_final %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  summarise(
+    numb_retouched = sum(retouch == 1, na.rm = TRUE),
+    total_lithics = n(),
+    percent_retouch = round((numb_retouched / total_lithics) * 100, 2)
+  )
+
+# Merge the density data with the summary table
+summary_table2_area <- 
+  left_join(summary_table2_area, 
+            lithic_density_df[, c("area", "lithic_density")], 
+            by = "area") %>%
+  mutate(area = recode(area, 'ta' = 'TA', 
+                       'tp1' = 'TP1', 
+                       'd1' = 'D1'),
+         lithic_volumetric_density = lithic_density)
+
+# Remove the 'lithic_density' column
+summary_table2_area <- summary_table2_area %>% 
+  select(-lithic_density)
+
+# Print out the resulting data frame
+summary_table2_area
+
+```
+
+```{r}
+
+                      #########DO NOT RUN0
+# BM: My preference is to delete this, or move it to a file called 'scraps.R' if you think you might need it later.
+
+# Define the volumes for each area based on the given information
 volumes <- data.frame(
@@
 
-# Join artifact_counts with volumes by area
-lithic_density_df <- left_join(artifact_counts, volumes, by = "area")
+# Calculate retouch frequency with explicit NA check
+retouch_freq <- tl_final_area %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  summarise(retouchFreq = ifelse(all(is.na(retouch)), NA, mean(!is.na(retouch))), .groups = 'drop')
 
 # Compute lithic density for each area and round it to 2 decimal places
-lithic_density_df <- lithic_density_df %>%
-  mutate(lithic_density = round(n / volume, 2)) # number of artifacts / volume
+lithic_density_df <- left_join(artifact_counts, volumes, by = "area") %>%
+  mutate(lithic_density = round(n / volume, 2))
 
@@
 
-# Merge the density data with the summary table
-summary_table2_area <- left_join(summary_table2_area, lithic_density_df[, c("area", "lithic_density")], by = "area") %>%
-  mutate(area = recode(area, 'ta' = 'TA', 'tp1' = 'TP1', 'd1' = 'D1'),
-         lithic_volumetric_density = lithic_density)
-
-# Remove the 'lithic_density' column
-summary_table2_area <- summary_table2_area %>% 
-  select(-lithic_density)
-
-# Print out the resulting data frame
-print(summary_table2_area)
-
-```
-
-```{r}
-
-                      #########DO NOT RUN0
-
-# Define the volumes for each area based on the given information
-volumes <- data.frame(
-  area = c("ta", "tp1", "d1"),
-  volume = c(52, 5.76, 2.1375) # in m^3
-)
-
-# Count the number of artifacts by area
-artifact_counts <- tl_final_area %>%
-  filter(area %in% c("ta", "tp1", "d1")) %>%
-  group_by(area) %>%
-  tally()
-
-# Calculate retouch frequency with explicit NA check
-retouch_freq <- tl_final_area %>%
-  filter(area %in% c("ta", "tp1", "d1")) %>%
-  group_by(area) %>%
-  summarise(retouchFreq = ifelse(all(is.na(retouch)), NA, mean(!is.na(retouch))), .groups = 'drop')
-
-# Compute lithic density for each area and round it to 2 decimal places
-lithic_density_df <- left_join(artifact_counts, volumes, by = "area") %>%
-  mutate(lithic_density = round(n / volume, 2))
-
-# Create the summary table
-summary_table2_area <- tl_final_area %>%
-  filter(area %in% c("ta", "tp1", "d1")) %>%
-  group_by(area) %>%
-  summarise(
-    numb_retouched = sum(retouch == 1, na.rm = TRUE),
-    total_lithics = n(),
-    percent_retouch = round((numb_retouched / total_lithics) * 100, 2)
-  )
-
 # Merge all the calculated data
@@
                                             ##### DO NOT RUN ######
+
+# BM: My preference is to delete this, or move it to a file called 'scraps.R' if you think you might need it later.
 
@@
 
+BM: what does this line below mean?
 WABI_AL Ta-e → 13 m3 Ta-d → 45 m3 D1-c → 0.675 m3 Tp-c → 1.44 m3 D1-a → 1.575 m3
@@
 ```{r}
-# Create a tibble 
-summary_table2_area <- tibble(
-  area = c("D1_a", "D1_c", "TA_d", "TA_e", "TP1_c"),
-  lithic_volumetric_density = c(51.2, 110.9, 45.1, 100.2, 88.3),
-  percent_retouch = c(4.9, 9.3, 15.3, 18.0, 9.8)
-)
-
-```
-
-```{r}
-
-library(tidyverse)
-library(ggpmisc)
-
 # Create the data frame
+
+# BM: why are you creating this data frame like this?
+It's not good practice to store data in code.
+It's very hard to keep track of changes to the data if you write it among the lines of code.
+Better practice is to export the table as a CSV file, then import it again when you need it.
 summary_table2_area <- tibble(
@@
                       y = retouch_frequency)) +
-
   # Add points
@@
              size = 2) +
-
   # line with log scale
-  stat_smooth(method = "lm", 
+  stat_smooth(
+    method = "lm",
     formula = y ~ log(x),
-              se = TRUE, aes(group = 1), 
+    se = TRUE,
+    aes(group = 1),
     color = "blue",
-              size = 0.5) +
-
+    size = 0.5
+  ) +
   # Add text labels for each point
-  geom_text(aes(label = area), 
-            vjust = -1, 
-            hjust = -0.5, 
-            size = 2.5) +
-
+  geom_text_repel(
+    aes(label = area),
+    size = 2.5
+  ) +
   # Add regression equation, R-squared, and p-value
-  stat_poly_eq(aes(label = paste(stat(eq.label), stat(rr.label), stat(p.value.label), sep = "~~~")),
+  stat_poly_eq(
+    aes(label = paste(
+      stat(eq.label), 
+      stat(rr.label), 
+      stat(p.value.label), 
+      sep = "~~~"
+    )),
     formula = y ~ log(x),
     parse = TRUE,
-               size = 2) +
-
+    size = 4
+  ) +
   # Add titles and labels
-  labs(
-    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
+  labs(title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
        x = "Lithic Volumetric Density",
-    y = "Retouch Frequency"
-  ) +
-
+       y = "Retouch Frequency") +
+  # BM: I think we need to have both axes on the log scale, not just one 
   scale_x_log10() + 
-
+  scale_y_log10() +
   theme(
@@
 
-print(WABI_AL)
-
-```
-
-```{r}
-
-library(googlesheets4)
-
-google_sheet_url <- "https://docs.google.com/spreadsheets/d/1Q0QZESk412ZQLE24yPs6Rg-7Y9OMX63DHoXeIPNAIFM/edit#gid=0"
-
-sum_depth_sheet <- read_sheet(google_sheet_url)
-
-sum_depth_sheet <- sum_depth_sheet %>%
-  mutate(area_layer_level = paste0(toupper(area), "_", layer, "_", level)) %>%
-  relocate(area_layer_level, .after = last_col()) %>%
-  mutate(
-    retouch_freq = round(retouch_tool / total_lithics, 2),
-    lithic_dens = round(total_lithics / volume, 2)
-  )
-
-write_sheet(sum_depth_sheet, ss = google_sheet_url)
+WABI_AL
 
@@
 # WABI_LEVEL
+
+# BM: I am confused why we are getting the counts of artefacts 
+# and retouch pieces from the google sheet, after you wrote so many lines of
+# code above to prepare the data from the CSV file. Where did the lithic data
+# in the google sheet come from? My preference is to have 
+# only copy of the lithic data in one place, and then join it with 
+# other data using code. Currently it seems like we have the 
+# lithic data in both the google sheet and the CSV file here. I'm worried that 
+# will be confusing and be difficult to update and keep correct. I prefer that we 
+# omit the lithic data from the google sheet, then import the google sheet and 
+# join it to the lithic data from the CSV file. Then we can have a data frame
+# with the area/layer/level data and lithic data to work on for plotting.
+
 filtered_data <- sum_depth_sheet %>% filter(lithic_dens > 0 & retouch_freq > 0)
@@
                 y = retouch_freq)) +
-
   geom_point(color = "black", size = 2) +
-
   geom_smooth(method = "lm", 
@@
               size = 0.5) +
-
   geom_text(aes(label = area_layer_level), 
@@
             check_overlap = TRUE) +
-
   stat_poly_eq(aes(label = paste(stat(eq.label), 
@@
                size = 2) +
-
   # Add titles and labels
@@
 
-print(WABI_Level)
+WABI_Level
 
@@
                 y = retouch_freq)) +
-
   geom_point(color = "black", size = 1.5) +
-
   geom_smooth(method = "lm", 
@@
               size = 0.5) +
-
   # Add text labels 
@@
             check_overlap = TRUE) +
-
   #regression equation, R-squared, and p-value
@@
                size = 2) +
-
   # titles and labels
@@
   ) +
-
   # Apply log scales
@@
   scale_y_log10() +
-
   theme(
@@
 
-print(WABI_Level)
-
-```
-
-```{r}
-
+WABI_Level
 ```
@@
 ```{r}
+# Create a tibble 
+# BM: why are these values written in here?
+They are they same as summary_table2_area generated by the code above.
+My preference is to avoid hard-coding values like this, I think it increases the chance of errors, so I'm uncomfortable with tables like this.
+summary_table2_area <- tibble(
+  area = c("D1_a", "D1_c", "TA_d", "TA_e", "TP1_c"),
+  lithic_volumetric_density = c(51.2, 110.9, 45.1, 100.2, 88.3),
+  percent_retouch = c(4.9, 9.3, 15.3, 18.0, 9.8)
+)
+
+```
+
+```{r}
 # regression plot-area
-library(ggplot2)
-library(ggpmisc)
 
 # Plotting
-p_area <- ggplot(summary_table2_area, aes(x = lithic_volumetric_density, y = percent_retouch)) +
-
+p_area <- 
+  ggplot(summary_table2_area, 
+         aes(x = lithic_volumetric_density,
+             y = percent_retouch)) +
   # Blue regression line with shaded area
@@
               color = "blue") +
-
   # Points colored by area
   geom_point(color = "black") + 
-
   # Labels points with area names
@@
             hjust = -0.5) +  
-
   # Regression equation, R-squared, and p-value
-  stat_poly_eq(aes(label = paste(stat(eq.label), stat(rr.label), stat(p.value.label), sep = "~~~")),
+  stat_poly_eq(aes(label = paste(stat(eq.label), 
+                                 stat(rr.label), 
+                                 stat(p.value.label), 
+                                 sep = "~~~")),
                formula = y ~ x,
@@
 
-print(p_area)
+p_area
 
@@
 ```{r}
-#whole assemblage
+# whole assemblage
 
   # Summarize data by Area
-summary_table_tech <- tl_final_area %>%
+summary_table_tech <- tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1")) %>%
   group_by(area) %>% 
-
   summarise(
@@
     bladelet_count = sum(blank == "bladelet", na.rm = TRUE),
-
     bullet_core_count = sum(`core-typology` == "bullet", na.rm = TRUE),
@@
 ```{r}
-#plot_blank_type
-
-library(dplyr)
-library(ggplot2)
-library(ggbeeswarm)
+# plot_blank_type
 
 # Data filtering and plotting
-tl_final_area %>% 
+tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1")) %>% 
   filter(blank %in% c('bladelet', 'blade', 'flake')) %>%
@@
 
-#WABI:
+# WABI:
 
 # regression plot-layer
-library(ggplot2)
-library(ggpmisc)
 
 # Plotting
+# BM: this summary_table2_layer object is made in a code block above that has 
+# comment "do not run". So I didn't run it, and now I'm confused about what to 
+# do here. 
 p <- ggplot(summary_table2_layer, 
@@
 
-print(p)
+p
 ```
@@
 
-tl_final_area %>%
+tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1")) %>%
   filter(layer %in% c("a", "c", "d", "e", "eor d")) %>%
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
 ```
@@
 
-library(ggplot2)
-library(dplyr)
 filtered_data <- tl_final %>%
@@
 filtered_data <- tl_final %>%
+  # BM: this line below gives me an error, so I can't run this chunk at all
   filter(!is.na(``) & `` != "")
@@
 
-```{r}
-
-```
-
 PCA-LAYER
@@
 ```{r}
-#prepare the data for PCA-layer
+# prepare the data for PCA-layer
 
 # Calculating Blank Frequency
-blank_freq <- tl_final_area %>%
+blank_freq <- tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1")) %>%
   filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(blank)) %>%
@@
 # Calculating Core Frequency
-core_freq <- tl_final_area %>%
+core_freq <- tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1")) %>%
   filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(corefrag)) %>%
@@
 # Printing the final table
-print(final_table)
+final_table
 
@@
 
+#------------------------------------------------------------------------------
+# BM: I wasn't able to run any code below this line
+# -----------------------------------------------------------------------------
+
 # Calculating Retouch Frequency
@@
   group_by(layer) %>%
+  # BM: something missing in the line below so I get an error
+  # I can't run anything beyond this line
   summarise(retFreq = sum( == "retouched piece", na.rm = TRUE) / n(), .groups = 'drop')
@@
 
-print(final_table)
+final_table
 
@@
 # Define the values manually (you would get these from your PCA results)
+# BM: I prefer not to write values in the code like this. I think
+# this is bad practice and leads to mistakes
 eigenvalues <- c(4.411234, 2.588766)

### 2023-09-12 paper/paper_tulai.qmd
edited most parts-made main figures for our purpose

@@
   )) %>%
+
   separate(depth, 
@@
 # BM: this is an unusual coding pattern, I'm curious about where you got it from?
+
 tl3 <- tl3 %>%
@@
 # BM: why are we removing these rows? A comment here would be good to explain. A simpler way, by referencing the row numbers directly:
+# SS: they are natural rocks, and don't want to count them as lithic
 
@@
   # BM: what does this next line do?
+  #SS: I wanted to remove any asterisks, whitespace, or *, ?, or !. is that wrong?
+
   mutate_all(~ str_replace(., "(?<=\\D)\\*(?=\\D)|\\s+|[*?!]+", "")) %>%
   # BM: what is this SHINE variable? I'm not familiar with it
+
+  ##SS: we meant sickle shine, this is one of the criteria based on which Hole claimed the site is not related to agriculture (similar to ALi Kosh located in Dehluran)
   mutate(
@@
 ```{r}
-# convert all columns to character type (I DONT KNOW WHY I DID THIS!)
-# BM: my preference is not to see comments and code like this. Go ahead and delete it.
-tl4 <- tl4 %>%
-  mutate(across(everything(), as.character))
-
-# replace 'NA', 'na', or '0' with R's NA
-tl4 <- tl4 %>%
-  # BM: in my opinion this is nicer than mutate_all and ifelse
-  mutate(across(everything(), ~na_if(., "NA"))) %>%
-  mutate(across(everything(), ~na_if(., "na"))) %>%
-  mutate(across(everything(), ~na_if(., "0")))
-
-```
-
-```{r}
 
@@
 ```{r}
+# List of columns that should be numeric
+columns_to_convert <- c("cortex", 
+                        "upper", 
+                        "lower", 
+                        "midpoint", 
+                        "length", 
+                        "width", 
+                        "utilization", 
+                        "retouch",
+                        "sickle shine")
+
+# Convert each column to numeric
+for (col_name in columns_to_convert) {
+  # Replace non-numeric characters with NA
+  tl_final[[col_name]][tl_final[[col_name]] %in% c("", "NA", "N/A")] <- NA
+  # Convert to numeric
+  tl_final[[col_name]] <- as.numeric(tl_final[[col_name]])
+}
+
+# check the structure to see if they have been converted
+str(tl_final)
+```
+
+```{r}
 #set the value of utilization to NA only for the rows where `sickle shine` == 1.
-#All other rows should remain unchanged
-
-#unsuccessful, should try again later
-
+
+tl_final <- tl_final %>%
+  rowwise() %>%
+  mutate(utilization = ifelse(is.na(`sickle shine`) | is.na(utilization), utilization,
+                              ifelse(`sickle shine` == 1 & utilization == 1, NA_real_, utilization))) %>%
+  ungroup()
+
+```
+
+```{r}
+#number of `sickle shine` in clumn `sickle shine`=1 and tooltypegroup are not match: 
+rev(sort(table(tl_final$tooltypegroup)))
+rev(sort(table(tl_final$`sickle shine`)))
+
+#troubleshoot:
+tl_final %>% filter(`sickle shine` == 1) %>% 
+  select(`tool-typology`, 
+         `sickle shine`)
+
+tl_final %>% filter(`tool-typology` == "sickle shine") %>% 
+  select(`tool-typology`, `sickle shine`)
+
+tl_final <- tl_final %>%
+  mutate(
+    `tool-typology` = case_when(
+      `sickle shine` == 1 ~ "sickle shine",
+      utilization == 1 ~ "utilized tool",
+      TRUE ~ `tool-typology`               
+    )
+  )
+
+rev(sort(table(tl_final$`tool-typology`)))
 ```
@@
 
+```{r}
+library(dplyr)
+
+# Filter the data to include only the areas 'ta', 'd1', and 'tp1'
+tl_final_area <- tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1"))
+
+```
+
 Summary tables: 1. AREA
@@
 # BM: looks like this is area and layer, is that right? I see c1, d1, d2, d3, d4 etc.
+# SS: those are different excavated areas, we mainly consider ta, tp1, and d1.
 knitr::kable(summary_table_area)
@@
 summary_table_layer <- tl_final %>%
-  group_by(layer) %>%  # Note that 'area' is replaced with 'layer'
+  group_by(layer) %>% 
   summarise(
@@
       caption = "Table: Summary by Al", 
-      align = 'c') %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
-                full_width = F, 
-                font_size = 12) %>%
-  row_spec(0, bold = TRUE, font_size = 14)
-
-```
-
-DEPTH:
-
-```{r}
-# Create a summary table based on the 'depth' column
-summary_table_depth <- tl_final %>%
-  mutate(depth = gsub("cm", "", depth)) %>%
-  group_by(depth) %>%  
-  summarise(
-    `Cores (n)` = sum(!is.na(core)),  
-    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag)),  
-    `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
-    `Flake (n)` = sum(blank == "flake", na.rm = TRUE),
-    `Bladelet (n)` = sum(blank == "bladelet", na.rm = TRUE),
-    `Retouched Tools (n)` = sum(retouch == 1, na.rm = TRUE),
-    `Utilized (n)` = sum(utilization == 1, na.rm = TRUE),
-    `Sickle Shine (n)` = sum(`sickle shine` == 1, na.rm = TRUE)
-  )
-
-```
-
-```{r}
-# Generate the table
-
-kable(summary_table_depth, 
-      caption = "Table: Summary by Depth", 
-      align = 'c') %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
-                full_width = F, 
-                font_size = 12) %>%
-  row_spec(0, bold = TRUE, font_size = 14)
-
-```
-
-```{r}
-# Create a summary table based on the 'depth', 'area', and 'layer' columns
-summary_table_depth_area_layer <- tl_final %>%
-  group_by(depth, area, layer, upper, lower) %>%  
-  summarise(
-    `Cores (n)` = sum(!is.na(core)),  
-    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag)),  
-    `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
-    `Flake (n)` = sum(blank == "flake", na.rm = TRUE),
-    `Bladelet (n)` = sum(blank == "bladelet", na.rm = TRUE),
-    `Retouched Tools (n)` = sum(retouch == 1, na.rm = TRUE),
-    `Utilized (n)` = sum(utilization == 1, na.rm = TRUE),
-    `Sickle Shine (n)` = sum(`sickle shine` == 1, na.rm = TRUE)
-  )
-
-# Generate the table
-library(kableExtra)
-
-kable(summary_table_depth_area_layer, 
-      caption = "Table: Summary by Depth, Area, and Layer", 
       align = 'c') %>%
@@
 ```{r}
-# Prepare data for Summary table-al
+# Prepare data for Summary table-area_layer(al)
 summary_table_al <- tl_final %>%
@@
     shine_tool_count =         sum(.$shine_tool_count, na.rm = TRUE),
-    total =                    sum(.$total, na.rm = TRUE)  # Explicitly setting the total
-  )
-
-```
-
-```{r}
+    total =                    sum(.$total, na.rm = TRUE)  
+  )
+
+rev(sort(table(tl_final$tooltypegroup)))
+```
+
+```{r}
+#explore tool type group
+
+summary_table_tooltype <- tl_final %>%
+  filter(tooltypegroup != 'utilized tool') %>%  # Exclude 'utilized tool'
+  group_by(tooltypegroup) %>% 
+  summarise(
+
+  )
+
+```
+
+```{r}
+#generate a table
+
 summary_table_al <- summary_table_al %>%
@@
   kable_styling(
-    bootstrap_options = c("striped", "hover", "condensed"),
+    bootstrap_options = c("striped", 
+                          "hover", 
+                          "condensed"),
     full_width = FALSE,
@@
   ) %>%
-  row_spec(0, bold = TRUE, font_size = 14)
-
-```
-
-```{r}
-#summary table-area
-
-# Rename columns
-summary_table_area <- summary_table_area %>%
-  rename(
-    "Area" = area,
-    "Blade" = blade_count,
-    "Flake" = flake_count,
-    "Bladelet" = bladelet_count,
-    "Retouched Tools" = retouched_tool_count,
-    "Tools" = tool_count,
-    "Utilized Tools" = utilized_tool_count,
-    "Sickle Shine" = shine_tool_count,
-    "Cores" = core_count,
-    "Core rejuvenations pieces" = core_rejuvenation_count,
-    "Total" = total
-  ) %>%
-  mutate(
-    Area = if_else(row_number() == n() & is.na(Area), "Total", Area)
-  ) %>%
-  mutate(
-    Area = recode(Area, "ta" = "TA", "d1" = "D1", "tp1" = "TP1", "Total" = "Total")
-  )
-
-kable(summary_table_area,
-      caption = "Table 1: Count of debitage type and tools in three different main areas.",
-      align = 'c') %>%
-  kable_styling(
-    bootstrap_options = c("striped", "hover", "condensed"),
-    font_size = 12
-  ) %>%
-  row_spec(0, font_size = 11)
-```
-
-```{r}
-# summary table-core and core rejuvenation
-```
-
-```{r}
-# summary table for tools: retouched-utelized-sickle
-
-```
-
-```{r}
-
-# Create a summary table
+  row_spec(0, 
+           bold = TRUE, 
+           font_size = 14)
+
+```
+
+```{r}
+# Create a summary table for typology based on area
+
 summary_table_tool_area <- tl_final %>%
@@
 ```{r}
+# Create a summary table based on the 'depth', 'area', and 'layer' columns
+tl_final <- tl_final %>%
+  mutate(depth = gsub("cm", "", depth))
+
+summary_table_depth_area_layer <- tl_final %>%
+  group_by(TA, depth, area, layer, upper, lower) %>%  
+  summarise(
+    `Cores (n)` = sum(!is.na(core)),  
+    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag)),  
+    `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
+    `Flake (n)` = sum(blank == "flake", na.rm = TRUE),
+    `Bladelet (n)` = sum(blank == "bladelet", na.rm = TRUE),
+    `Retouched Tools (n)` = sum(retouch == 1, na.rm = TRUE),
+    `Utilized (n)` = sum(utilization == 1, na.rm = TRUE),
+    `Sickle Shine (n)` = sum(`sickle shine` == 1, na.rm = TRUE)
+  )
+
+# Generate the table
+library(kableExtra)
+
+kable(summary_table_depth_area_layer, 
+      caption = "Table: Summary by Depth, Area, and Layer", 
+      align = 'c') %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
+                full_width = F, 
+                font_size = 12) %>%
+  row_spec(0, bold = TRUE, font_size = 14)
+```
+
+```{r}
+#Here, I've imported the table 'Summary by Depth, Area, Layer' into Google Sheets for easier management.
+#I thought this format would simplify the process of those depths needed to be splitting up.
 
 # BM: looks like you need this data starting from around here in the document
+
+library(tidyverse)
 library(googlesheets4)
@@
 sum_depth_sheet <- read_sheet(google_sheet_url)
-
-sum_depth_sheet <- sum_depth_sheet %>%
-  mutate(`depth (cm)` = unlist(`depth (cm)`),
-         level = unlist(level)) %>% 
-  mutate(area_layer_level = paste0(toupper(area), 
-                                   "_", 
-                                   layer, 
-                                   "_", 
-                                   level)) %>%
-  relocate(area_layer_level, 
-           .after = last_col()) %>%
-  mutate(
-    retouch_freq = round(retouch_tool / total_lithics, 2),
-    lithic_dens = round(total_lithics / volume, 2)
-  )
 
@@
 
-```
-
-volume for each area:
-
-```{r}
-# Define the volumes for each area 
-volumes <- data.frame(
-  area = c("ta", "tp1", "d1"),
-  volume = c(52, 5.76, 2.1375) # in m^3
+#SS: is that true?: (I have a weird number in row 5, cl Depth and couldn't find out what happened)
+
+# Separate `excavation area` into `length` and `width` and calculate `volume`
+sum_depth_sheet <- sum_depth_sheet %>%
+  separate(`excavation area`, into = c("length", "width"), sep = "\\*") %>%
+  mutate(
+    length = as.numeric(length),
+    width = as.numeric(width),
+    `thickness of deposit` = `thickness of deposit` / 100  # Convert from cm to m
+  ) %>%
+  mutate(
+    volume = length * width * `thickness of deposit`  # Calculate volume in m^3
+  )
+
+# Calculate lithic volumetric density 
+sum_depth_sheet <- sum_depth_sheet %>%
+  mutate(
+    `lithic_dens` = round(total_lithics / volume, 2)
+  )
+
+# Calculate retouch frequency 
+sum_depth_sheet <- sum_depth_sheet %>%
+  mutate(
+    retouch_freq = round(retouch_tool / volume, 2)  
+  )
+
+```
+
+```{r}
+#WABI plot: here I do not consider those levels with a value 0 of retouch frequency.
+
+library(ggpmisc)
+
+# Filter the data (get out those level with Retouch frequency=0)
+filtered_data <- sum_depth_sheet %>% filter(lithic_dens > 0 & retouch_freq > 0)
+
+# Create the plot: WABI_Level_logscale
+WABI_Level <- ggplot(filtered_data,  
+            aes(x = lithic_dens,
+                y = retouch_freq)) +
+  geom_point(color = "black", size = 1.5) +
+  geom_smooth(method = "lm", 
+              se = TRUE, 
+              color = "blue", 
+              linewidth = 0.5) +
+  # Add text labels 
+  geom_text(aes(label = paste(area, layer, level, sep = "_")), 
+            vjust = 1.5, 
+            hjust = 0.5, 
+            check_overlap = TRUE) +
+
+  stat_poly_eq(aes(label = paste(stat(eq.label), 
+                                  stat(rr.label), 
+                                  stat(p.value.label), 
+                                  sep = "~~~")),
+               formula = y ~ x, 
+               parse = TRUE, 
+               size = 2) +
+  labs(
+    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
+    x = "Lithic Volumetric Density",
+    y = "Retouch Frequency"
+  ) +
+  scale_x_log10() +
+  scale_y_log10() +
+  theme(
+    axis.text = element_text(size = 8),
+    axis.title = element_text(size = 10),
+    plot.title = element_text(size = 12)
+  )
+
+# Print the plot to display it
+print(WABI_Level)
+
+```
+
+```{r}
+#PCA_Level: preparing data
+
+#PCA_level: for PCA Bicho and Cascalheira considered these variables: estimated area (as we calculated based on levels, should we not consider it?), core frequency, blank frequency, chip frequency: artefacts smaller than 1 cm (as we work on a neolithic asssemblage, we have so many small artefacts; so, I do not consider it), feature frequency (we have some stone alignments, but we do not know the exact level, we know the area of those features), retouch frequency, tool diversity (diversity of tool types within each assemblage, calculated using Menhinick’s index);
+##core frequency
+##blank frequency
+##retouch frequency
+##tool diversity
+
+#Calculate Core Frequency
+sum_depth_sheet <- sum_depth_sheet %>%
+  mutate(
+    core_freq = round(cores / volume, 2)  
+  )
+
+# calculate Blank Frequency
+sum_depth_sheet <- sum_depth_sheet %>%
+  mutate(
+    blank_freq = round((flake + blade + bladelet) / volume, 2)  
+  )
+
+# calculate Tool Diversity
+##calculating tool diversity is a bit tricky here; as I have some spitted depth, which we assign the number of artefacts based on the calculation of splitting, so how should I consider the tool diversity here?
+```
+
+PCA_al:
+
+```{r}
+#PCA: area_layer (al)
+#I think as we are going to calculate PCA to understand other factors that might be affected the duration of the occupation, it would be logical to calculate it for each layer, instead of level.
+we have 5 stratigraphical layers: e, d, c, b, a
+#d1:1.5*1.5/ ta: 4*5/ tp1:2*1.8
+#e: excavated area is A/ depth of excavation is 0-75(mean)
+#d: excavated area is A/ 0.75m-3m→ 2.25
+#c: d1 and TP → 30 cm and 40cm
+#b: TP/ 170 cm 
+#a: D1: 70cm
+
+# Add 'dimension' columns and relocate it after 'al' (based on Hole (1974))
+library(dplyr)
+
+summary_table_al_pca <- summary_table_al %>%
+  mutate(
+    length = case_when(
+      al == "D1_a" ~ 1.5,
+      al == "D1_c" ~ 1.5,
+      al == "TA_d" ~ 4,
+      al == "TP1_c" ~ 2,
+      al == "TA_e" ~ 4,
+      TRUE ~ NA_real_
+    ),
+    width = case_when(
+      al == "D1_a" ~ 1.5,
+      al == "D1_c" ~ 1.5,
+      al == "TA_d" ~ 5,
+      al == "TP1_c" ~ 1.8,
+       al == "TA_e" ~ 5,
+      TRUE ~ NA_real_
+    ),
+    thickness = case_when(
+      al == "D1_a" ~ 0.7,
+      al == "D1_c" ~ 0.3,
+      al == "TA_d" ~ 2.25,
+      al == "TP1_c" ~ 0.4,
+      al == "TA_e" ~ 0.75,
+      TRUE ~ NA_real_)
+  ) %>%
+
+  mutate(volume = length * width * thickness) %>%
+  relocate(length, width, thickness, volume, .after = al)
+# Show the updated data frame
+summary_table_al_pca
+
+```
+
+```{r}
+# Load required library
+library(dplyr)
+
+# Add a new column called 'blank_count'
+summary_table_al_pca <- summary_table_al_pca %>%
+  mutate(blank_count = blade_count + flake_count + bladelet_count) %>%
+  select(al, length, width, thickness, volume, blade_count, flake_count, bladelet_count, blank_count, everything())
+
+# Calculate core frequency, blank frequency, retouch frequency, lithic density 
+summary_table_al_pca <- summary_table_al_pca %>%
+  mutate(
+    lithic_density = round((core_count + blank_count + core_rejuvenation_count) / volume, 2),
+    core_freq = round(core_count / volume, 2),
+    blank_freq = round(blank_count / volume, 2),
+    retouch_freq = round(retouched_tool_count / volume, 2)
+  )
+
+# Calculate tool diversity using Menhinick's index
+summary_table_al_with_diversity <- tl_final %>% 
+  filter(retouch == 1) %>%
+  group_by(al, `tool-typology`) %>%  
+  summarise(n = n(), .groups = 'drop') %>% 
+  group_by(al) %>%  
+  summarise(
+    num_unique_tool_types = n_distinct(`tool-typology`),
+    total_retouched_tools = sum(n),
+    tool_diversity = round(num_unique_tool_types / sqrt(total_retouched_tools), 2),
+    .groups = 'drop'
+  )
+
+# Merge this new information into the existing summary_table_al_pca
+summary_table_al_pca <- left_join(summary_table_al_pca, summary_table_al_with_diversity, by = "al")
+
+# Remove the 'Total' row and select only the columns needed for PCA
+fin_summary_table_al_pca <- summary_table_al_pca %>%
+  filter(al != "Total") %>%
+  select(al, core_freq, blank_freq, lithic_density, retouch_freq, tool_diversity)
+
+# Show the final table
+print(fin_summary_table_al_pca)
+
+```
+
+```{r}
+library(FactoMineR)
+# Perform PCA
+res.pca <- PCA(fin_summary_table_al_pca[, -1], graph = FALSE)  # Exclude the 'al' column
+
+```
+
+```{r}
+## Got error: couldnt figure out the reason of error
+
+# Create a data frame to hold the rounded eigenvalues and the rounded percentage variance
+eigen_df <- data.frame(
+  Dimension = paste("Dim", 1:nrow(eigenvalues), sep = "."),
+  Eigenvalue = round(eigenvalues[, 1], 3),
+  `Variance percent` = round(eigenvalues[, 2], 3),
+  `Cumulative variance percent` = round(cumsum(eigenvalues[, 2]), 3)
+)
+
+# Print the table
+print(eigen_df)
+
+```
+
+```{r}
+# variables to PC1
+fviz_contrib(res.pca,  
+             choice = "var",  
+             axes = 1,  
+             top = 10)
+```
+
+```{r}
+fviz_contrib(res.pca,  
+             choice = "var",  
+             axes = 2,  
+             top = 10)
+
+```
+
+```{r}
+fviz_contrib(res.pca,  
+             choice = "var",  
+             axes = 3,  
+             top = 10)
+```
+
+```{r}
+fviz_contrib(res.pca,  
+             choice = "var",  
+             axes = 4,  
+             top = 10)
+```
+
+```{r}
+#Plot: Contribution of variables for each of the four relevant PCA dimensions
+library(factoextra)
+library(ggplot2)
+
+#save contribution plots into one objects
+p1 <- fviz_contrib(res.pca, 
+                   choice = "var", 
+                   axes = 1, 
+                   top = 10)
+p2 <- fviz_contrib(res.pca, 
+                   choice = "var",
+                   axes = 2, 
+                   top = 10)
+p3 <- fviz_contrib(res.pca, 
+                   choice = "var", 
+                   axes = 3, 
+                   top = 10)
+p4 <- fviz_contrib(res.pca, 
+                   choice = "var", 
+                   axes = 4, 
+                   top = 10)
+
+#combine individual plots into one plot 
+library(gridExtra)
+grid.arrange(p1, p2, p3, p4, ncol = 2)
+
+```
+
+```{r}
+# Extract contributions of variables to each principal component
+contributions <- res.pca$var$contrib
+
+# Compute average contribution for each dimension
+avg_contrib_dim1 <- mean(contributions[,1], na.rm = TRUE)
+avg_contrib_dim2 <- mean(contributions[,2], na.rm = TRUE)
+avg_contrib_dim3 <- mean(contributions[,3], na.rm = TRUE)
+avg_contrib_dim4 <- mean(contributions[,4], na.rm = TRUE)
+
+# Find variables that contribute more than average to each dimension
+high_contrib_dim1 <- contributions[contributions[,1] > avg_contrib_dim1, 1]
+high_contrib_dim2 <- contributions[contributions[,2] > avg_contrib_dim2, 2]
+high_contrib_dim3 <- contributions[contributions[,3] > avg_contrib_dim3, 3]
+high_contrib_dim4 <- contributions[contributions[,4] > avg_contrib_dim4, 4]
+
+# Sort them in descending order
+sorted_high_contrib_dim1 <- sort(high_contrib_dim1, decreasing = TRUE)
+sorted_high_contrib_dim2 <- sort(high_contrib_dim2, decreasing = TRUE)
+sorted_high_contrib_dim3 <- sort(high_contrib_dim3, decreasing = TRUE)
+sorted_high_contrib_dim4 <- sort(high_contrib_dim4, decreasing = TRUE)
+
+# Print or use for further analysis
+print("Significant contributions to Dim1:")
+print(sorted_high_contrib_dim1)
+
+print("Significant contributions to Dim2:")
+print(sorted_high_contrib_dim2)
+
+print("Significant contributions to Dim3:")
+print(sorted_high_contrib_dim3)
+
+print("Significant contributions to Dim4:")
+print(sorted_high_contrib_dim4)
+
+```
+
+```{r}
+# screeplot; inspect distribution of PCs
+library(factoextra)  
+fviz_screeplot(res.pca)
+```
+
+```{r}
+# Inspect eigenvalues
+print(res.pca$eig)
+
+```
+
+```{r}
+# Load the required package
+library(kableExtra)
+
+# Create a table with kable and kableExtra
+eigen_df %>%
+  kable("html") %>%
+  kable_styling("striped", full_width = F) %>%
+  column_spec(1, bold = T)
+```
+
+```{r}
+
+# Rounding the numerical columns to 3 decimal places
+eigen_df[, -1] <- round(eigen_df[, -1], 3)
+
+# Replacing 'comp' with 'Dim'
+eigen_df$Dimension <- gsub("comp", "Dim", eigen_df$Dimension)
+
+# Removing row names
+rownames(eigen_df) <- NULL
+
+# Creating the kable output
+kable_output <- kable(eigen_df, 
+                      "html", 
+                      align = 'c', 
+                      col.names = c("Dimension", 
+                                    "Eigenvalue", 
+                                    "Variance percent", 
+                                    "Cumulative variance percent")) %>%
+  kable_styling("striped", 
+                full_width = F) %>%
+  add_header_above(c(" " = 1, "Eigenvalues and percentage of variance for each dimension of PCA" = 3))
+
+htmltools::browsable(kable_output)
+
+```
+
+```{r}
+# Visualize biplot:1
+fviz_pca_biplot(res.pca,
+                axes = c(1, 2))
+rownames(res.pca$ind$coord) <- fin_summary_table_al_pca$al
+# Visualize biplot with row names
+fviz_pca_biplot(res.pca)
+
+```
+
+```{r}
+# Visualize biplot: 2
+fviz_pca_biplot(res.pca,
+                axes = c(3, 4))
+rownames(res.pca$ind$coord) <- fin_summary_table_al_pca$al
+# Visualize biplot with row names
+fviz_pca_biplot(res.pca)
+
+print(fviz_pca_biplot(res.pca, axes = c(3, 4)))
+
+```
+
+```{r}
 # BM: My preference would be to show the calculation of these 
@@
   # verify these values as they are currently 
-)
+##SS: fixed
+
+# Define the volumes for each area
+
+# add the dimensions for each area
+dimensions <- data.frame(
+  area = c("ta", "tp1", "d1"),
+  length = c(4, 2, 1.5),
+  width = c(5, 1.8, 1.5),
+  thickness = c(3, 1.6, 0.95)
+)
+
+# Calculate volume for each area based on the dimensions
+dimensions <- dimensions %>%
+  mutate(volume = length * width * thickness)
 
@@
 
-# Join artifact_counts with volumes by area
-lithic_density_df <- left_join(artifact_counts, volumes, by = "area")
+# Join artifact_counts with dimensions (which now includes volume) by area
+lithic_density_df <- left_join(artifact_counts, dimensions, by = "area")
 
@@
     total_lithics = n(),
-    percent_retouch = round((numb_retouched / total_lithics) * 100, 2)
+    percent_retouch = round((numb_retouched / total_lithics), 2)
   )
@@
 # Print out the resulting data frame
-summary_table2_area
-
-```
-
-```{r}
-
-                      #########DO NOT RUN0
-# BM: My preference is to delete this, or move it to a file called 'scraps.R' if you think you might need it later.
-
-# Define the volumes for each area based on the given information
-volumes <- data.frame(
-  area = c("ta", "tp1", "d1"),
-  volume = c(52, 5.76, 2.1375) # in m^3
-)
-
-# Count the number of artifacts by area
-artifact_counts <- tl_final_area %>%
-  filter(area %in% c("ta", "tp1", "d1")) %>%
-  group_by(area) %>%
-  tally()
-
-# Calculate retouch frequency with explicit NA check
-retouch_freq <- tl_final_area %>%
-  filter(area %in% c("ta", "tp1", "d1")) %>%
-  group_by(area) %>%
-  summarise(retouchFreq = ifelse(all(is.na(retouch)), NA, mean(!is.na(retouch))), .groups = 'drop')
-
-# Compute lithic density for each area and round it to 2 decimal places
-lithic_density_df <- left_join(artifact_counts, volumes, by = "area") %>%
-  mutate(lithic_density = round(n / volume, 2))
-
-# Create the summary table
-summary_table2_area <- tl_final_area %>%
-  filter(area %in% c("ta", "tp1", "d1")) %>%
-  group_by(area) %>%
-  summarise(
-    numb_retouched = sum(retouch == 1, na.rm = TRUE),
-    total_lithics = n(),
-    percent_retouch = round((numb_retouched / total_lithics) * 100, 2)
-  )
-
-# Merge all the calculated data
-summary_table2_area <- left_join(summary_table2_area, lithic_density_df[, c("area", "lithic_density")], by = "area")
-summary_table2_area <- left_join(summary_table2_area, retouch_freq, by = "area")
-
-# Final modifications and print
-summary_table2_area <- summary_table2_area %>%
-  mutate(
-    area = recode(area, 'ta' = 'TA', 'tp1' = 'TP1', 'd1' = 'D1'),
-    lithic_volumetric_density = lithic_density
-  ) %>%
-  select(-lithic_density)  # Remove the unwanted column
-
 print(summary_table2_area)
@@
       align = 'c',
-      col.names = c("Area", "Number of Retouched Lithics", "Total Number of Lithics", "Retouch frequency (%)", "Lithic Volumetric Density(pcs/(m^3)")) %>%
+      col.names = c("Area", 
+                    "Number of Retouched Lithics", 
+                    "Total Number of Lithics", 
+                    "Retouch frequency", 
+                    "Lithic Volumetric Density(pcs/(m^3)")) %>%
   kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 11) %>%
@@
 
-```{r}
-
-                                            ##### DO NOT RUN ######
-
-# BM: My preference is to delete this, or move it to a file called 'scraps.R' if you think you might need it later.
-
-# layer-Create the summary table
-summary_table2_layer <- tl_final_area %>%
-  filter(layer %in% c(c("a", "b", "c" , "d", "e"))) %>%  
-  group_by(layer) %>%
-  summarise(
-    numb_retouched = sum(retouch == 1, na.rm = TRUE),
-    numb_cores = sum(!is.na(`core-typology`), na.rm = TRUE),
-    total_lithics = n(),
-    percent_retouch = round((numb_retouched / total_lithics) * 100, 2)
-  )
-
-# Create a data frame for volumetric density values
-density_data_layer <- data.frame(
-  layer = c("a", "b", "c" , "d", "e"),
-  `lithic_volumetric_density` = round(c(1.575, 6.12, 2.115, 47, 13), 2)
-)
-
-summary_table2_layer <- merge(summary_table2_layer, density_data_layer, by = "layer")
-
-```
-
-BM: what does this line below mean?
-WABI_AL Ta-e → 13 m3 Ta-d → 45 m3 D1-c → 0.675 m3 Tp-c → 1.44 m3 D1-a → 1.575 m3
-
-```{r}
-# Create the data frame
-
-# BM: why are you creating this data frame like this?
-It's not good practice to store data in code.
-It's very hard to keep track of changes to the data if you write it among the lines of code.
-Better practice is to export the table as a CSV file, then import it again when you need it.
-summary_table2_area <- tibble(
-  area = c("D1_a", "D1_c", "TA_d", "TA_e", "TP1_c"),
-  lithic_volumetric_density = c(51.43, 111.11, 45.13, 100.08, 212.50),
-  retouch_frequency = c(0.05, 0.09, 0.07, 0.07, 0.10)
-)
-
-# Create the ggplot
-WABI_AL <- ggplot(summary_table2_area,
-                  aes(x = lithic_volumetric_density,
-                      y = retouch_frequency)) +
-  # Add points
-  geom_point(color =
-               "black",
-             size = 2) +
-  # line with log scale
-  stat_smooth(
-    method = "lm",
-    formula = y ~ log(x),
-    se = TRUE,
-    aes(group = 1),
-    color = "blue",
-    size = 0.5
-  ) +
-  # Add text labels for each point
-  geom_text_repel(
-    aes(label = area),
-    size = 2.5
-  ) +
-  # Add regression equation, R-squared, and p-value
-  stat_poly_eq(
-    aes(label = paste(
-      stat(eq.label), 
-      stat(rr.label), 
-      stat(p.value.label), 
-      sep = "~~~"
-    )),
-    formula = y ~ log(x),
-    parse = TRUE,
-    size = 4
-  ) +
-  # Add titles and labels
-  labs(title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
-       x = "Lithic Volumetric Density",
-       y = "Retouch Frequency") +
-  # BM: I think we need to have both axes on the log scale, not just one 
-  scale_x_log10() + 
-  scale_y_log10() +
-  theme(
-    axis.text = element_text(size = 8),
-    axis.title = element_text(size = 10),
-    plot.title = element_text(size = 12)
-  )
-
-WABI_AL
-
-```
+------------------------------
+Done with the main plots
+------------------------------
 
@@
 # with the area/layer/level data and lithic data to work on for plotting.
+
+#SS: the googlesheet is the last thing I created.
+I created the summary table in R, and then for splitting up the depth (we have few depth range that was not possible to assign the layer to them, so, we splited them), I found it is easier to do so in Google sheet.
 
@@
 ```{r}
-#  WABI-level-log scale
-
-filtered_data <- sum_depth_sheet %>% filter(lithic_dens > 0 & retouch_freq > 0)
-
-WABI_Level <- ggplot(filtered_data,  
-            aes(x = lithic_dens,
-                y = retouch_freq)) +
-  geom_point(color = "black", size = 1.5) +
-  geom_smooth(method = "lm", 
-              se = TRUE, 
-              color = "blue", 
-              size = 0.5) +
-  # Add text labels 
-  geom_text(aes(label = area_layer_level), 
-            vjust = 1.5, 
-            hjust = 0.5, 
-            size = 2,
-            check_overlap = TRUE) +
-  #regression equation, R-squared, and p-value
-  stat_poly_eq(aes(label = paste(stat(eq.label), 
-                                  stat(rr.label), 
-                                  stat(p.value.label), 
-                                  sep = "~~~")),
-               formula = y ~ x, 
-               parse = TRUE, 
-               size = 2) +
-  # titles and labels
-  labs(
-    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
-    x = "Lithic Volumetric Density",
-    y = "Retouch Frequency"
-  ) +
-  # Apply log scales
-  scale_x_log10() +
-  scale_y_log10() +
-  theme(
-    axis.text = element_text(size = 8),
-    axis.title = element_text(size = 10),
-    plot.title = element_text(size = 12)
-  )
-
-WABI_Level
-```
-
-WABI_AREA
-
-```{r}
-# Create a tibble 
-# BM: why are these values written in here?
-They are they same as summary_table2_area generated by the code above.
-My preference is to avoid hard-coding values like this, I think it increases the chance of errors, so I'm uncomfortable with tables like this.
-summary_table2_area <- tibble(
-  area = c("D1_a", "D1_c", "TA_d", "TA_e", "TP1_c"),
-  lithic_volumetric_density = c(51.2, 110.9, 45.1, 100.2, 88.3),
-  percent_retouch = c(4.9, 9.3, 15.3, 18.0, 9.8)
-)
-
-```
-
-```{r}
-# regression plot-area
-
-# Plotting
-p_area <- 
-  ggplot(summary_table2_area, 
-         aes(x = lithic_volumetric_density,
-             y = percent_retouch)) +
-  # Blue regression line with shaded area
-  geom_smooth(method = "lm", 
-              se = TRUE, 
-              aes(group = 1), 
-              color = "blue") +
-  # Points colored by area
-  geom_point(color = "black") + 
-  # Labels points with area names
-  geom_text(aes(label = area),
-            vjust = -1, 
-            hjust = -0.5) +  
-  # Regression equation, R-squared, and p-value
-  stat_poly_eq(aes(label = paste(stat(eq.label), 
-                                 stat(rr.label), 
-                                 stat(p.value.label), 
-                                 sep = "~~~")),
-               formula = y ~ x,
-               parse = TRUE,
-               size = 3) +
-
-  # Titles and labels
-  labs(title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
-       x = "Lithic Volumetric Density",
-       y = "Retouch Frequency (%)")
-
-p_area
-
-```
-
-```{r}
-# Fit the linear model
-linear_model <- lm(percent_retouch ~ lithic_volumetric_density, data = summary_table2_area)
-
-# Display the summary of the model to get coefficients, R-squared, and p-value
-summary_linear_model <- summary(linear_model)
-
-# Extract the coefficients (intercept and slope)
-coefficients <- summary_linear_model$coefficients
-
-# Extract the R-squared value
-r_squared <- summary_linear_model$r.squared
-
-# Extract the p-value for the slope (lithic_volumetric_density)
-p_value_slope <- coefficients["lithic_volumetric_density", "Pr(>|t|)"]
-
-cat("Coefficients:\n")
-print(coefficients)
-
-cat("\nR-squared value:", r_squared)
-
-cat("\nP-value for slope:", p_value_slope)
-
-```
-
-```{r}
 # whole assemblage
@@
 ```{r}
-
-# WABI:
-
-# regression plot-layer
-
-# Plotting
-# BM: this summary_table2_layer object is made in a code block above that has 
-# comment "do not run". So I didn't run it, and now I'm confused about what to 
-# do here. 
-p <- ggplot(summary_table2_layer, 
-            aes(lithic_volumetric_density, 
-                percent_retouch)) +
-
-  geom_smooth(method = "lm", 
-              se = TRUE, 
-              aes(group = 1), 
-              color = "blue") +
-
-  # Points colored by layer
-  geom_point(color = "black") +
-
-  # Labels points with layer names
-  geom_text(aes(label = layer),
-            vjust = -1, 
-            hjust = -0.5) +
-
-  # Regression equation, R-squared, and p-value
-  stat_poly_eq(aes(label = paste(stat(eq.label), stat(rr.label), stat(p.value.label), sep = "~~~")),
-               formula = y ~ x,
-               parse = TRUE,
-               size = 3) +
-
-  # Titles and labels
-  labs(title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
-       x = "Lithic Volumetric Density",
-       y = "Retouch Frequency (%)")
-
-p
-```
-
-```{r}
 #barplot_layer
@@
 
-```{r}
-
-filtered_data <- tl_final %>%
-  # BM: this line below gives me an error, so I can't run this chunk at all
-  filter(!is.na(``) & `` != "")
-
-# Create the plot using the filtered data
-ggplot(data = filtered_data) + 
-  aes(y = ``) + 
-  geom_bar() +
-  labs(y = "Tool Type",
-       x = "Frequency") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-```
-
 PCA-LAYER
@@
 # BM: I wasn't able to run any code below this line
+##SS: fixed
 # -----------------------------------------------------------------------------
@@
   group_by(layer) %>%
-  # BM: something missing in the line below so I get an error
-  # I can't run anything beyond this line
-  summarise(retFreq = sum( == "retouched piece", na.rm = TRUE) / n(), .groups = 'drop')
+  summarise(retFreq = sum(retouch == 1, na.rm = TRUE) / n(), .groups = 'drop')
 
@@
 
-calculating PCA based on AREA: TA/ TP1/ D1:
-
-```{r}
-#4lithic density
-
-# A→ V=4×5×2.6 = 52 m3
-# TP → V=2×1.8×1.6 = 5.76 m 3
-# D1 → 1.5×1.5×0.95 = 2.1375 m3
-
-library(dplyr)
-
-# Define the volumes for each area based on the given information
-volumes <- data.frame(
-  area = c("ta", "tp1", "d1"),
-  volume = c(52, 5.76, 2.1375) # in m^3
-)
-
-# Count the number of artifacts by area
-artifact_counts <- tl_final_area %>%
-  filter(area %in% c("ta", "tp1", "d1")) %>%
-  group_by(area) %>%
-  tally()
-
-# Join artifact_counts with volumes by area
-lithic_density_df <- left_join(artifact_counts, volumes, by = "area")
-
-# Compute lithic density for each area
-lithic_density_df <- lithic_density_df %>%
-  mutate(lithic_density = n / volume) # number of artifacts / volume
-
-# Print the resulting table
-print(lithic_density_df[, c("area", "lithic_density")])
-
-```
-
-```{r}
-
-#2 Calculate the frequency of cores and core fragments by area
-
-library(dplyr)
-# Calculate the core frequency
-core_frequency_by_area <- left_join(number_of_cores_by_area, 
-                                    total_lithics_by_area, 
-                                    by = "area") %>%
-  mutate(core_frequency = (number_of_cores / total_lithics) * 100)
-
-# Display the result
-print(core_frequency_by_area)
-
-```
-
-```{r}
-# 1number of Blank
-
-# Calculate the total number of lithic artifacts
-total_lithics <- tl_final_area %>%
-  summarise(total_lithics = n()) %>%
-  pull(total_lithics)
-
-# Display the total number of lithic artifacts
-print(paste("The total number of lithic artifacts is:", total_lithics))
-
-```
-
-```{r}
-# 5Calculate the retouched frequency for each area
-library(dplyr)
-
-total_lithics_by_area <- tl_final_area %>%
-  group_by(area) %>%
-  summarise(total_lithics = n())
-
-number_of_retouched_by_area <- tl_final_area %>%
-  mutate(retouch = as.numeric(`retouch`)) %>% 
-  filter(!is.na(retouch), retouch == 1) %>%  
-  group_by(area) %>%  
-  summarise(number_of_retouched = sum(retouch))
-
-retouched_frequency_by_area <- left_join(number_of_retouched_by_area, total_lithics_by_area, by = "area") %>%
-  mutate(retouched_frequency = (number_of_retouched / total_lithics) * 100)
-
-# Display the retouched frequency by area
-print(retouched_frequency_by_area)
-
-print(number_of_retouched_by_area)
-
-```
-
-```{r}
-# 3Calculating Tool Diversity using Menhinick's Index
-tool_div <- tl_final_area %>%
-  filter(area %in% c("ta", "d1", "tp1"), 
-         !is.na(tooltypegroup)) %>%
-  group_by(area) %>%
-  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(sum(!is.na(tooltypegroup))), .groups = 'drop')
-```
-
 feature frequency: d1 1 ta 1 tp1 0)
-
-estimated volumes (area) A→ 52 m3 TP → 5.76 m 3 D1 → 2.1375 m3
 
@@
   area = c("d1", "ta", "tp1"),
-  estimated_area = c(2.1375, 52, 5.76),
   feature_freq = c(1, 1, 0)
@@
 
-pca_final_table_area <- left_join(area_data, retouch_freq, by = "area") %>%
+pca_final_table_area <- left_join(dimensions, retouch_freq, by = "area") %>%
   left_join(blank_freq, by = "area") %>%
@@
   left_join(tool_div, by = "area") %>%
-  select(area, estimated_area, feature_freq, retouchFreq, blankFreq, coreFreq, lithic_density, toolDiv)
+  select(area, new_dimensions, feature_freq, retouchFreq, blankFreq, coreFreq, lithic_density, toolDiv)
 

### 2023-09-12 scraps.R

@@ -1,0 +1,134 @@
+
+#########DO NOT RUN0
+# BM: My preference is to delete this, or move it to a file called 'scraps.R' if you think you might need it later.
+
+# Define the volumes for each area based on the given information
+volumes <- data.frame(
+  area = c("ta", "tp1", "d1"),
+  volume = c(52, 5.76, 2.1375) # in m^3
+)
+
+tl_final_area <- tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1"))
+
+# Count the number of artifacts by area
+artifact_counts <- tl_final_area %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  tally()
+
+# Calculate retouch frequency with explicit NA check
+retouch_freq <- tl_final_area %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  summarise(retouchFreq = ifelse(all(is.na(retouch)), NA, mean(!is.na(retouch))), .groups = 'drop')
+
+# Compute lithic density for each area and round it to 2 decimal places
+lithic_density_df <- left_join(artifact_counts, volumes, by = "area") %>%
+  mutate(lithic_density = round(n / volume, 2))
+
+# Create the summary table
+summary_table2_area <- tl_final_area %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  summarise(
+    numb_retouched = sum(retouch == 1, na.rm = TRUE),
+    total_lithics = n(),
+    percent_retouch = round((numb_retouched / total_lithics) * 100, 2)
+  )
+
+# Merge all the calculated data
+summary_table2_area <- left_join(summary_table2_area, lithic_density_df[, c("area", "lithic_density")], by = "area")
+summary_table2_area <- left_join(summary_table2_area, retouch_freq, by = "area")
+
+# Final modifications and print
+summary_table2_area <- summary_table2_area %>%
+  mutate(
+    area = recode(area, 'ta' = 'TA', 'tp1' = 'TP1', 'd1' = 'D1'),
+    lithic_volumetric_density = lithic_density
+  ) %>%
+  select(-lithic_density)  # Remove the unwanted column
+
+print(summary_table2_area)
+
+```
+
+
+
+
+
+
+
+```{r}
+
+
+##### DO NOT RUN ######
+
+# BM: My preference is to delete this, or move it to a file called 'scraps.R' if you think you might need it later.
+
+
+# layer-Create the summary table
[…]

### 2023-09-15 paper/paper_tulai.qmd
some editions on WAIB and PCA plots.

@@
     # BM: This looks like it was very time-consuming to simplify. 
+    ##SS: this column was so messy with too many mistakes; I needed to have all typos correct then transfer them in their corresponding columns.  
     Typology = case_when(
@@
 ```{r}
+
+```
+
+```{r}
 # Exploring data
@@
 ```{r}
+
+library(dplyr)
+library(stringr)
+
+# Define the patterns to search for
+patterns <- c("denticulate-notch", "retouched piece", "scraper",
+              "Perforator", "serrated scraper", "backed pieces",
+              "truncated pieces", "burin", "geometric", "scraper-notch",
+              "microburin")
+
+# Update the `retouch` column based on the `tooltypegroup` column
+tl_final <- tl_final %>%
+  mutate(retouch = ifelse(str_detect(tooltypegroup, str_c(patterns, collapse = "|")), 1, retouch))
+
+# View the first few rows to confirm
+head(tl_final)
+rev(sort(table(tl_final$tooltypegroup)))
+rev(sort(table(tl_final$retouch)))
+```
+
+```{r}
 tl_final <- tl_final %>%
@@
 ```{r}
+# Generate a summary table by grouping by 'depth'
+summary_table_depth_tool_type <- tl_final %>%
+  group_by(depth) %>%
+  summarise(tool_type = n_distinct(tooltypegroup))
+
+# Create a full summary table by joining it back to your original summary table based on 'depth'
+summary_table_depth_area_layer <- summary_table_depth_area_layer %>%
+  left_join(summary_table_depth_tool_type, by = "depth")
+
+# Print column names 
+print(names(summary_table_depth_area_layer))
+
+# Generate the table
+library(kableExtra)
+
+kable(summary_table_depth_area_layer, 
+      caption = "Table: Summary by Depth, Area, and Layer", 
+      align = 'c') %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
+                full_width = F, 
+                font_size = 12) %>%
+  row_spec(0, bold = TRUE, font_size = 14)
+
+```
+
+```{r}
+# Specify the exact columns want to keep
+desired_columns <- c("TA", "depth", "area", "layer", "upper", "lower", 
+                     "Cores (n)", "Core rejuvenations pieces (n)", 
+                     "Blade (n)", "Flake (n)", "Bladelet (n)", 
+                     "Retouched Tools (n)", "Utilized (n)", "Sickle Shine (n)",
+                     "tool_type")  # Only keep this 'tool_type'
+
+# Filter the tl_final dataframe
+summary_table_depth_area_layer <- summary_table_depth_area_layer %>%
+  filter(area %in% c('ta', 'd1', 'tp1'))
+
+# Generate the table
+library(kableExtra)
+
+kable(summary_table_depth_area_layer, 
+      caption = "Table: Summary by Depth, Area, and Layer", 
+      align = 'c') %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
+                full_width = F, 
+                font_size = 12) %>%
+  row_spec(0, bold = TRUE, font_size = 14)
+
+```
+
+```{r}
+
 #Here, I've imported the table 'Summary by Depth, Area, Layer' into Google Sheets for easier management.
@@
 
-sum_depth_sheet <- read_sheet(google_sheet_url)
+# Create the 'label' column
+sum_depth_sheet <- sum_depth_sheet %>%
+  mutate(label = paste(area, layer, level, sep = "_"))
+
+# Add a suffix only to duplicates
+sum_depth_sheet <- sum_depth_sheet %>%
+  group_by(label) %>%
+  mutate(suffix = case_when(
+    n() == 1 ~ NA_character_,
+    TRUE ~ letters[row_number()]
+  )) %>%
+  ungroup()
+
+# Append the suffix to 'label'
+sum_depth_sheet <- sum_depth_sheet %>%
+  mutate(label = if_else(is.na(suffix), label, paste0(label, "_", suffix))) %>%
+  select(-suffix)
+
+# Relocate 'label' column after 'level'
+sum_depth_sheet <- sum_depth_sheet %>%
+  relocate(label, .after = level)
 
@@
 
+#### previously I had a column'excavation area'; I seprated them in 2 columns in order to calculate the colume; but now when I wanted to run the code it said "!
+object 'excavation area' not found"; would it be OK to delete these lines?
+
 # Separate `excavation area` into `length` and `width` and calculate `volume`
@@
     width = as.numeric(width),
-    `thickness of deposit` = `thickness of deposit` / 100  # Convert from cm to m
+    `thickness` = `thickness` / 100  # Convert from cm to m
   ) %>%
   mutate(
-    volume = length * width * `thickness of deposit`  # Calculate volume in m^3
+    volume = length * width * `thickness`  # Calculate volume in m^3
   )
@@
 ```{r}
-#WABI plot: here I do not consider those levels with a value 0 of retouch frequency.
-
-library(ggpmisc)
-
-# Filter the data (get out those level with Retouch frequency=0)
-filtered_data <- sum_depth_sheet %>% filter(lithic_dens > 0 & retouch_freq > 0)
-
 # Create the plot: WABI_Level_logscale
+# here I do not consider those levels with a value 0 of retouch frequency. 
 WABI_Level <- ggplot(filtered_data,  
@@
               linewidth = 0.5) +
-  # Add text labels 
-  geom_text(aes(label = paste(area, layer, level, sep = "_")), 
+  # Add text labels based on 'label' column
+  geom_text(aes(label = label), 
             vjust = 1.5, 
@@
 
-#PCA_level: for PCA Bicho and Cascalheira considered these variables: estimated area (as we calculated based on levels, should we not consider it?), core frequency, blank frequency, chip frequency: artefacts smaller than 1 cm (as we work on a neolithic asssemblage, we have so many small artefacts; so, I do not consider it), feature frequency (we have some stone alignments, but we do not know the exact level, we know the area of those features), retouch frequency, tool diversity (diversity of tool types within each assemblage, calculated using Menhinick’s index);
+#PCA_level: for PCA Bicho and Cascalheira considered these variables: estimated area (as we calculated based on levels, I consider the thickness; is that OK?), core frequency, blank frequency, chip frequency: artefacts smaller than 1 cm (as we work on a neolithic asssemblage, we have so many small artefacts; so, I do not consider it), feature frequency (we have some stone alignments, but we do not know the exact level, we know the area of those features), retouch frequency, tool diversity (diversity of tool types within each assemblage, calculated using Menhinick’s index);
 ##core frequency
@@
 ##tool diversity
+##thickness of deposit
+##lithic density
 
@@
 
-# calculate Tool Diversity
-##calculating tool diversity is a bit tricky here; as I have some spitted depth, which we assign the number of artefacts based on the calculation of splitting, so how should I consider the tool diversity here?
-```
+# Calculate tool diversity using Menhinick's index
+sum_depth_sheet <- sum_depth_sheet %>%
+  mutate(
+    total_tools = retouch_tool + utilized + sickle_shine,
+    tool_diversity = ifelse(total_tools > 0, n_distinct(tool_type) / sqrt(total_tools), NA)
+  )
+
+```
+
+```{r}
+pca_data <- sum_depth_sheet %>% 
+  select(thickness, lithic_dens, core_freq, blank_freq, retouch_freq, tool_diversity) %>%
+  na.omit()  # Remove rows with NA values
+
+```
+
+```{r}
+library(FactoMineR)
+library(factoextra)
+# Assuming 'sum_depth_sheet' is your dataframe and you have added 'm_index' as your Menhinick index
+pca_data <- sum_depth_sheet[, c('thickness', 'lithic_dens', 'core_freq', 'blank_freq', 'retouch_freq', 'tool_diversity')]
+
+# Perform PCA
+res.pca <- PCA(pca_data, graph = FALSE)
+
+# View the summary results
+summary(res.pca)
+
+```
+
+```{r}
+#Plot: Contribution of variables for each of the four relevant PCA dimensions
+library(factoextra)
+library(ggplot2)
+
+#save contribution plots into one objects
+p1 <- fviz_contrib(res.pca, 
+                   choice = "var", 
+                   axes = 1, 
+                   top = 10)
+p2 <- fviz_contrib(res.pca, 
+                   choice = "var",
+                   axes = 2, 
+                   top = 10)
+p3 <- fviz_contrib(res.pca, 
+                   choice = "var", 
+                   axes = 3, 
+                   top = 10)
+p4 <- fviz_contrib(res.pca, 
+                   choice = "var", 
+                   axes = 4, 
+                   top = 10)
+
+#combine individual plots into one plot 
+library(gridExtra)
+grid.arrange(p1, p2, p3, p4, ncol = 2)
+
+```
+
+```{r}
+# Extract contributions of variables to each principal component
+contributions <- res.pca$var$contrib
+
+# Compute average contribution for each dimension
+avg_contrib_dim1 <- mean(contributions[,1], na.rm = TRUE)
+avg_contrib_dim2 <- mean(contributions[,2], na.rm = TRUE)
+avg_contrib_dim3 <- mean(contributions[,3], na.rm = TRUE)
+avg_contrib_dim4 <- mean(contributions[,4], na.rm = TRUE)
+
+# Find variables that contribute more than average to each dimension
+high_contrib_dim1 <- contributions[contributions[,1] > avg_contrib_dim1, 1]
+high_contrib_dim2 <- contributions[contributions[,2] > avg_contrib_dim2, 2]
+high_contrib_dim3 <- contributions[contributions[,3] > avg_contrib_dim3, 3]
+high_contrib_dim4 <- contributions[contributions[,4] > avg_contrib_dim4, 4]
+
+# Sort them in descending order
+sorted_high_contrib_dim1 <- sort(high_contrib_dim1, decreasing = TRUE)
+sorted_high_contrib_dim2 <- sort(high_contrib_dim2, decreasing = TRUE)
+sorted_high_contrib_dim3 <- sort(high_contrib_dim3, decreasing = TRUE)
+sorted_high_contrib_dim4 <- sort(high_contrib_dim4, decreasing = TRUE)
+
+# Print or use for further analysis
+print("Significant contributions to Dim1:")
+print(sorted_high_contrib_dim1)
+
+print("Significant contributions to Dim2:")
+print(sorted_high_contrib_dim2)
+
+print("Significant contributions to Dim3:")
+print(sorted_high_contrib_dim3)
+
+print("Significant contributions to Dim4:")
+print(sorted_high_contrib_dim4)
+
+```
+
+```{r}
+# screeplot; inspect distribution of PCs
+library(factoextra)  
+fviz_screeplot(res.pca)
+```
+
+```{r}
+# Inspect eigenvalues
+print(res.pca$eig)
+
+```
+
+```{r}
+
+library(kableExtra)
+library(htmltools)
+
+# Create a new column named "Dimension" 
+eigen_df$Dimension <- paste("Dim", seq_len(nrow(eigen_df)), sep = "")
+
+eigen_df <- eigen_df[, c("Dimension", "eigenvalue", "percentage of variance", "cumulative percentage of variance")]
+
+# Round the numerical columns to 3 decimal places
+eigen_df[, 2:4] <- round(eigen_df[, 2:4], 3)
+
+# Removing row names
+rownames(eigen_df) <- NULL
+
+# Creating the kable output
+kable_output <- kable(eigen_df, 
+                      "html", 
+                      align = 'c', 
+                      col.names = c("Dimension", 
+                                    "Eigenvalue", 
+                                    "Variance Percent", 
+                                    "Cumulative Variance Percent")) %>%
+  kable_styling("striped", 
+                full_width = F) %>%
+  add_header_above(c(" " = 1, "Eigenvalues and percentage of variance for each dimension of PCA" = 3))
+
+# Display the table
+htmltools::browsable(kable_output)
+
+```
+
+```{r}
+# Visualize biplot:1
+fviz_pca_biplot(res.pca,
+                axes = c(1, 2))
+rownames(res.pca$ind$coord) <- sum_depth_sheet$label
+# Visualize biplot with row names
+fviz_pca_biplot(res.pca)
+
+```
+
+```{r}
+# Visualize biplot: 2
+fviz_pca_biplot(res.pca,
+                axes = c(3, 4))
+rownames(res.pca$ind$coord) <- sum_depth_sheet$label
+# Visualize biplot with row names
+fviz_pca_biplot(res.pca)
+
+print(fviz_pca_biplot(res.pca, axes = c(3, 4)))
+
+```
+
+-----------------------------------------------
+DONE UPTO HERE
+-----------------------------------------------
 

### 2023-09-16 paper/paper_tulai.qmd
update on n_distinct with na.rm

@@
 ```{r}
-
-```
-
-```{r}
 # Exploring data
@@
   # Replace non-numeric characters with NA
+  # BM: this is quite an unusual method for doing this, 
+  # where did you find it?
   tl_final[[col_name]][tl_final[[col_name]] %in% c("", "NA", "N/A")] <- NA
@@
 }
+
+# BM: this is how I would convert those columns to numeric, 
+# I think it's easier to read and requires less typing
+# tl_final %>% 
+#  mutate(across(all_of(columns_to_convert), parse_number))
 
@@
 tl_final <- tl_final %>%
-  mutate(retouch = ifelse(str_detect(tooltypegroup, str_c(patterns, collapse = "|")), 1, retouch))
+  mutate(retouch = ifelse(str_detect(tooltypegroup, 
+                                     str_c(patterns, collapse = "|")), 
+                          1, retouch))
 
@@
   group_by(depth) %>%
-  summarise(tool_type = n_distinct(tooltypegroup))
+  summarise(tool_type = n_distinct(tooltypegroup, na.rm = TRUE))
 
 # Create a full summary table by joining it back to your original summary table based on 'depth'
-summary_table_depth_area_layer <- summary_table_depth_area_layer %>%
-  left_join(summary_table_depth_tool_type, by = "depth")
+summary_table_depth_area_layer <- 
+  summary_table_depth_area_layer %>%
+  left_join(summary_table_depth_tool_type, 
+            by = "depth")
 

### 2023-09-16 paper/paper_tulai.qmd
summary table edited

--- 2023-09-15 paper/paper_tulai.qmd
@@
 ```{r}
-# Generate a summary table by grouping by 'depth'
+# Generate a summary table by grouping by 'depth' and 'area'
 summary_table_depth_tool_type <- tl_final %>%
-  group_by(depth) %>%
-  summarise(tool_type = n_distinct(tooltypegroup))
-
-# Create a full summary table by joining it back to your original summary table based on 'depth'
+  group_by(area, depth) %>%
+  summarise(tool_type = n_distinct(tooltypegroup, na.rm = TRUE))
+
+# Create a full summary table by joining it back to original summary table based on 'depth' and 'area'
 summary_table_depth_area_layer <- summary_table_depth_area_layer %>%
-  left_join(summary_table_depth_tool_type, by = "depth")
-
-# Print column names 
-print(names(summary_table_depth_area_layer))
+  left_join(summary_table_depth_tool_type, by = c("area", "depth"))
 

### 2023-09-18 paper/paper_tulai.qmd
edited and polished the codes. Good to render now.

--- 2023-09-16 paper/paper_tulai.qmd
@@
 ```{r}
-# Generate a summary table by grouping by 'depth'
+# Generate a summary table by grouping by 'depth' and 'area'
 summary_table_depth_tool_type <- tl_final %>%
-  group_by(depth) %>%
-  summarise(tool_type = n_distinct(tooltypegroup, na.rm = TRUE))
-
-# Create a full summary table by joining it back to your original summary table based on 'depth'
+  group_by(area, depth) %>%
+  summarise(tool_type = n_distinct(tooltypegroup, na.rm = TRUE), .groups = 'drop')
+
+# Create a full summary table by joining it back to the original summary table based on 'depth' and 'area'
+summary_table_depth_area_layer <- summary_table_depth_area_layer %>%
+  left_join(summary_table_depth_tool_type, by = c("area", "depth"))
+
+# Remove the existing 'tool_type' if it exists
+if ("tool_type" %in% names(summary_table_depth_area_layer)) {
+  summary_table_depth_area_layer <- summary_table_depth_area_layer %>%
+    select(-tool_type)
+}
+
+# Join again based on 'depth', this time being specific about what we want
 summary_table_depth_area_layer <- 
   summary_table_depth_area_layer %>%
-  left_join(summary_table_depth_tool_type, 
+  left_join(summary_table_depth_tool_type %>% select(depth, tool_type), 
             by = "depth")
-
-# Print column names 
-print(names(summary_table_depth_area_layer))
 
@@
   "https://docs.google.com/spreadsheets/d/1Q0QZESk412ZQLE24yPs6Rg-7Y9OMX63DHoXeIPNAIFM/edit#gid=0"
+sum_depth_sheet <- read_sheet(google_sheet_url)
 
@@
 # here I do not consider those levels with a value 0 of retouch frequency. 
+filtered_data <- sum_depth_sheet %>% 
+  filter(lithic_dens > 0 & retouch_freq > 0)
+
 WABI_Level <- ggplot(filtered_data,  
@@
                parse = TRUE, 
-               size = 2) +
+               size = 4) +
   labs(
@@
   theme(
-    axis.text = element_text(size = 8),
+    axis.text = element_text(size = 7),
     axis.title = element_text(size = 10),
@@
 ##tool diversity
-##thickness of deposit
+##thickness of deposit (???)
 ##lithic density
@@
 ```{r}
+
 pca_data <- sum_depth_sheet %>% 
-  select(thickness, lithic_dens, core_freq, blank_freq, retouch_freq, tool_diversity) %>%
+  select(lithic_dens, core_freq, blank_freq, retouch_freq, tool_diversity) %>%
   na.omit()  # Remove rows with NA values
@@
 ```{r}
+# performin PCA
+
 library(FactoMineR)
 library(factoextra)
-# Assuming 'sum_depth_sheet' is your dataframe and you have added 'm_index' as your Menhinick index
-pca_data <- sum_depth_sheet[, c('thickness', 'lithic_dens', 'core_freq', 'blank_freq', 'retouch_freq', 'tool_diversity')]
+
+pca_data <- sum_depth_sheet[, c('lithic_dens', 'core_freq', 'blank_freq', 'retouch_freq', 'tool_diversity')]
 
@@
 ```{r}
+eigen_df <- data.frame(
+  eigenvalue = res.pca$eig[,1],
+  `percentage of variance` = res.pca$eig[,2],
+  `cumulative percentage of variance` = res.pca$eig[,3]
+)
+
+```
+
+```{r}
 
@@
 
-eigen_df <- eigen_df[, c("Dimension", "eigenvalue", "percentage of variance", "cumulative percentage of variance")]
+# Re-order the columns based on the corrected names
+eigen_df <- eigen_df[, c("Dimension", "eigenvalue", "percentage.of.variance", "cumulative.percentage.of.variance")]
 
@@
 fviz_pca_biplot(res.pca,
-                axes = c(1, 2))
+                axes = c(1, 2),
+                labelsize = 2
+                )
 rownames(res.pca$ind$coord) <- sum_depth_sheet$label
-# Visualize biplot with row names
-fviz_pca_biplot(res.pca)
 
@@
 fviz_pca_biplot(res.pca,
-                axes = c(3, 4))
+                axes = c(3, 4),
+                labelsize = 2
+                )
 rownames(res.pca$ind$coord) <- sum_depth_sheet$label
-# Visualize biplot with row names
-fviz_pca_biplot(res.pca)
-
-print(fviz_pca_biplot(res.pca, axes = c(3, 4)))
-
-```
-
------------------------------------------------
-DONE UPTO HERE
------------------------------------------------
-
-PCA_al:
-
-```{r}
-#PCA: area_layer (al)
-#I think as we are going to calculate PCA to understand other factors that might be affected the duration of the occupation, it would be logical to calculate it for each layer, instead of level.
-we have 5 stratigraphical layers: e, d, c, b, a
-#d1:1.5*1.5/ ta: 4*5/ tp1:2*1.8
-#e: excavated area is A/ depth of excavation is 0-75(mean)
-#d: excavated area is A/ 0.75m-3m→ 2.25
-#c: d1 and TP → 30 cm and 40cm
-#b: TP/ 170 cm 
-#a: D1: 70cm
-
-# Add 'dimension' columns and relocate it after 'al' (based on Hole (1974))
-library(dplyr)
-
-summary_table_al_pca <- summary_table_al %>%
-  mutate(
-    length = case_when(
-      al == "D1_a" ~ 1.5,
-      al == "D1_c" ~ 1.5,
-      al == "TA_d" ~ 4,
-      al == "TP1_c" ~ 2,
-      al == "TA_e" ~ 4,
-      TRUE ~ NA_real_
-    ),
-    width = case_when(
-      al == "D1_a" ~ 1.5,
-      al == "D1_c" ~ 1.5,
-      al == "TA_d" ~ 5,
-      al == "TP1_c" ~ 1.8,
-       al == "TA_e" ~ 5,
-      TRUE ~ NA_real_
-    ),
-    thickness = case_when(
-      al == "D1_a" ~ 0.7,
-      al == "D1_c" ~ 0.3,
-      al == "TA_d" ~ 2.25,
-      al == "TP1_c" ~ 0.4,
-      al == "TA_e" ~ 0.75,
-      TRUE ~ NA_real_)
-  ) %>%
-
-  mutate(volume = length * width * thickness) %>%
-  relocate(length, width, thickness, volume, .after = al)
-# Show the updated data frame
-summary_table_al_pca
-
-```
-
-```{r}
-# Load required library
-library(dplyr)
-
-# Add a new column called 'blank_count'
-summary_table_al_pca <- summary_table_al_pca %>%
-  mutate(blank_count = blade_count + flake_count + bladelet_count) %>%
-  select(al, length, width, thickness, volume, blade_count, flake_count, bladelet_count, blank_count, everything())
-
-# Calculate core frequency, blank frequency, retouch frequency, lithic density 
-summary_table_al_pca <- summary_table_al_pca %>%
-  mutate(
-    lithic_density = round((core_count + blank_count + core_rejuvenation_count) / volume, 2),
-    core_freq = round(core_count / volume, 2),
-    blank_freq = round(blank_count / volume, 2),
-    retouch_freq = round(retouched_tool_count / volume, 2)
-  )
-
-# Calculate tool diversity using Menhinick's index
-summary_table_al_with_diversity <- tl_final %>% 
-  filter(retouch == 1) %>%
-  group_by(al, `tool-typology`) %>%  
-  summarise(n = n(), .groups = 'drop') %>% 
-  group_by(al) %>%  
-  summarise(
-    num_unique_tool_types = n_distinct(`tool-typology`),
-    total_retouched_tools = sum(n),
-    tool_diversity = round(num_unique_tool_types / sqrt(total_retouched_tools), 2),
-    .groups = 'drop'
-  )
-
-# Merge this new information into the existing summary_table_al_pca
-summary_table_al_pca <- left_join(summary_table_al_pca, summary_table_al_with_diversity, by = "al")
-
-# Remove the 'Total' row and select only the columns needed for PCA
-fin_summary_table_al_pca <- summary_table_al_pca %>%
-  filter(al != "Total") %>%
-  select(al, core_freq, blank_freq, lithic_density, retouch_freq, tool_diversity)
-
-# Show the final table
-print(fin_summary_table_al_pca)
-
-```
-
-```{r}
-library(FactoMineR)
-# Perform PCA
-res.pca <- PCA(fin_summary_table_al_pca[, -1], graph = FALSE)  # Exclude the 'al' column
-
-```
-
-```{r}
-## Got error: couldnt figure out the reason of error
-
-# Create a data frame to hold the rounded eigenvalues and the rounded percentage variance
-eigen_df <- data.frame(
-  Dimension = paste("Dim", 1:nrow(eigenvalues), sep = "."),
-  Eigenvalue = round(eigenvalues[, 1], 3),
-  `Variance percent` = round(eigenvalues[, 2], 3),
-  `Cumulative variance percent` = round(cumsum(eigenvalues[, 2]), 3)
-)
-
-# Print the table
-print(eigen_df)
-
-```
-
-```{r}
-# variables to PC1
-fviz_contrib(res.pca,  
-             choice = "var",  
-             axes = 1,  
-             top = 10)
-```
-
-```{r}
-fviz_contrib(res.pca,  
-             choice = "var",  
-             axes = 2,  
-             top = 10)
-
-```
-
-```{r}
-fviz_contrib(res.pca,  
-             choice = "var",  
-             axes = 3,  
-             top = 10)
-```
-
-```{r}
-fviz_contrib(res.pca,  
-             choice = "var",  
-             axes = 4,  
-             top = 10)
-```
-
-```{r}
-#Plot: Contribution of variables for each of the four relevant PCA dimensions
-library(factoextra)
-library(ggplot2)
-
-#save contribution plots into one objects
-p1 <- fviz_contrib(res.pca, 
-                   choice = "var", 
-                   axes = 1, 
-                   top = 10)
-p2 <- fviz_contrib(res.pca, 
-                   choice = "var",
-                   axes = 2, 
-                   top = 10)
-p3 <- fviz_contrib(res.pca, 
-                   choice = "var", 
-                   axes = 3, 
-                   top = 10)
-p4 <- fviz_contrib(res.pca, 
-                   choice = "var", 
-                   axes = 4, 
-                   top = 10)
-
-#combine individual plots into one plot 
-library(gridExtra)
-grid.arrange(p1, p2, p3, p4, ncol = 2)
-
-```
-
-```{r}
-# Extract contributions of variables to each principal component
-contributions <- res.pca$var$contrib
-
-# Compute average contribution for each dimension
-avg_contrib_dim1 <- mean(contributions[,1], na.rm = TRUE)
-avg_contrib_dim2 <- mean(contributions[,2], na.rm = TRUE)
-avg_contrib_dim3 <- mean(contributions[,3], na.rm = TRUE)
-avg_contrib_dim4 <- mean(contributions[,4], na.rm = TRUE)
-
-# Find variables that contribute more than average to each dimension
-high_contrib_dim1 <- contributions[contributions[,1] > avg_contrib_dim1, 1]
-high_contrib_dim2 <- contributions[contributions[,2] > avg_contrib_dim2, 2]
-high_contrib_dim3 <- contributions[contributions[,3] > avg_contrib_dim3, 3]
-high_contrib_dim4 <- contributions[contributions[,4] > avg_contrib_dim4, 4]
-
-# Sort them in descending order
-sorted_high_contrib_dim1 <- sort(high_contrib_dim1, decreasing = TRUE)
-sorted_high_contrib_dim2 <- sort(high_contrib_dim2, decreasing = TRUE)
-sorted_high_contrib_dim3 <- sort(high_contrib_dim3, decreasing = TRUE)
-sorted_high_contrib_dim4 <- sort(high_contrib_dim4, decreasing = TRUE)
-
-# Print or use for further analysis
-print("Significant contributions to Dim1:")
-print(sorted_high_contrib_dim1)
-
-print("Significant contributions to Dim2:")
-print(sorted_high_contrib_dim2)
-
-print("Significant contributions to Dim3:")
-print(sorted_high_contrib_dim3)
-
-print("Significant contributions to Dim4:")
-print(sorted_high_contrib_dim4)
-
-```
-
-```{r}
-# screeplot; inspect distribution of PCs
-library(factoextra)  
-fviz_screeplot(res.pca)
-```
-
-```{r}
-# Inspect eigenvalues
-print(res.pca$eig)
-
-```
-
-```{r}
-# Load the required package
-library(kableExtra)
-
-# Create a table with kable and kableExtra
-eigen_df %>%
-  kable("html") %>%
-  kable_styling("striped", full_width = F) %>%
-  column_spec(1, bold = T)
-```
-
-```{r}
-
-# Rounding the numerical columns to 3 decimal places
-eigen_df[, -1] <- round(eigen_df[, -1], 3)
-
-# Replacing 'comp' with 'Dim'
-eigen_df$Dimension <- gsub("comp", "Dim", eigen_df$Dimension)
-
-# Removing row names
-rownames(eigen_df) <- NULL
-
-# Creating the kable output
-kable_output <- kable(eigen_df, 
-                      "html", 
-                      align = 'c', 
-                      col.names = c("Dimension", 
-                                    "Eigenvalue", 
-                                    "Variance percent", 
-                                    "Cumulative variance percent")) %>%
-  kable_styling("striped", 
-                full_width = F) %>%
-  add_header_above(c(" " = 1, "Eigenvalues and percentage of variance for each dimension of PCA" = 3))
-
-htmltools::browsable(kable_output)
-
-```
-
-```{r}
-# Visualize biplot:1
-fviz_pca_biplot(res.pca,
-                axes = c(1, 2))
-rownames(res.pca$ind$coord) <- fin_summary_table_al_pca$al
-# Visualize biplot with row names
-fviz_pca_biplot(res.pca)
-
-```
-
-```{r}
-# Visualize biplot: 2
-fviz_pca_biplot(res.pca,
-                axes = c(3, 4))
-rownames(res.pca$ind$coord) <- fin_summary_table_al_pca$al
-# Visualize biplot with row names
-fviz_pca_biplot(res.pca)
-
-print(fviz_pca_biplot(res.pca, axes = c(3, 4)))
-
-```
-
-```{r}
-# BM: My preference would be to show the calculation of these 
-  # values from the raw data, rather than type them in 
-  # here, where we can't see where they came from, we can't easily
-  # verify these values as they are currently 
-##SS: fixed
-
-# Define the volumes for each area
-
-# add the dimensions for each area
-dimensions <- data.frame(
-  area = c("ta", "tp1", "d1"),
-  length = c(4, 2, 1.5),
-  width = c(5, 1.8, 1.5),
-  thickness = c(3, 1.6, 0.95)
-)
-
-# Calculate volume for each area based on the dimensions
-dimensions <- dimensions %>%
-  mutate(volume = length * width * thickness)
-
-# Count the number of artifacts by area
-artifact_counts <- tl_final %>%
-  filter(area %in% c("ta", "tp1", "d1")) %>%
-  group_by(area) %>%
-  tally()
-
-# Join artifact_counts with dimensions (which now includes volume) by area
-lithic_density_df <- left_join(artifact_counts, dimensions, by = "area")
-
-# Compute lithic density for each area and round it to 2 decimal places
-lithic_density_df <- lithic_density_df %>%
-  mutate(lithic_density = round(n / volume, 2)) # number of artifacts / volume
-
-# Create the summary table
-summary_table2_area <- tl_final %>%
-  filter(area %in% c("ta", "tp1", "d1")) %>%
-  group_by(area) %>%
-  summarise(
-    numb_retouched = sum(retouch == 1, na.rm = TRUE),
-    total_lithics = n(),
-    percent_retouch = round((numb_retouched / total_lithics), 2)
-  )
-
-# Merge the density data with the summary table
-summary_table2_area <- 
-  left_join(summary_table2_area, 
-            lithic_density_df[, c("area", "lithic_density")], 
-            by = "area") %>%
-  mutate(area = recode(area, 'ta' = 'TA', 
-                       'tp1' = 'TP1', 
-                       'd1' = 'D1'),
-         lithic_volumetric_density = lithic_density)
-
-# Remove the 'lithic_density' column
-summary_table2_area <- summary_table2_area %>% 
-  select(-lithic_density)
-
-# Print out the resulting data frame
-print(summary_table2_area)
-
-```
-
-```{r}
-# generate table 2 - area
-
-library(kableExtra)
-
-kable(summary_table2_area, 
-      caption = "Table 2. Lithic counts for each of the areas in Tula'i", 
-      align = 'c',
-      col.names = c("Area", 
-                    "Number of Retouched Lithics", 
-                    "Total Number of Lithics", 
-                    "Retouch frequency", 
-                    "Lithic Volumetric Density(pcs/(m^3)")) %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 11) %>%
-  column_spec(1, width = "1cm") %>% 
-  column_spec(2, width = "1cm") %>%
-
-  column_spec(3, width = "1cm") %>% 
-  column_spec(4, width = "1cm") %>% 
-  column_spec(5, width = "1cm")
-
-```
-
-------------------------------
-Done with the main plots
-------------------------------
-
-```{r}
-# WABI_LEVEL
-
-# BM: I am confused why we are getting the counts of artefacts 
-# and retouch pieces from the google sheet, after you wrote so many lines of
-# code above to prepare the data from the CSV file. Where did the lithic data
-# in the google sheet come from? My preference is to have 
-# only copy of the lithic data in one place, and then join it with 
-# other data using code. Currently it seems like we have the 
-# lithic data in both the google sheet and the CSV file here. I'm worried that 
-# will be confusing and be difficult to update and keep correct. I prefer that we 
-# omit the lithic data from the google sheet, then import the google sheet and 
-# join it to the lithic data from the CSV file. Then we can have a data frame
-# with the area/layer/level data and lithic data to work on for plotting.
-
-#SS: the googlesheet is the last thing I created.
-I created the summary table in R, and then for splitting up the depth (we have few depth range that was not possible to assign the layer to them, so, we splited them), I found it is easier to do so in Google sheet.
-
-filtered_data <- sum_depth_sheet %>% filter(lithic_dens > 0 & retouch_freq > 0)
-
-WABI_Level <- ggplot(filtered_data, 
-            aes(x = lithic_dens,
-                y = retouch_freq)) +
-  geom_point(color = "black", size = 2) +
-  geom_smooth(method = "lm", 
-              se = TRUE, 
-              color = "blue", 
-              size = 0.5) +
-  geom_text(aes(label = area_layer_level), 
-            vjust = 1.5, 
-            hjust = 0.5, 
-            size = 2,
-            check_overlap = TRUE) +
-  stat_poly_eq(aes(label = paste(stat(eq.label), 
-                                 stat(rr.label), 
-                                 stat(p.value.label), 
-                                 sep = "~~~")),
-               formula = y ~ x, 
-               parse = TRUE, 
-               size = 2) +
-  # Add titles and labels
-  labs(
-    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
-    x = "Lithic Volumetric Density",
-    y = "Retouch Frequency"
-  ) +
-
-  theme(
-    axis.text = element_text(size = 8),
-    axis.title = element_text(size = 10),
-    plot.title = element_text(size = 12)
-  )
-
-WABI_Level
-
-```
-
-```{r}
-# whole assemblage
-
-  # Summarize data by Area
-summary_table_tech <- tl_final %>%
-  filter(area %in% c("ta", "d1", "tp1")) %>%
-  group_by(area) %>% 
-  summarise(
-   blade_count = sum(blank == "blade", na.rm = TRUE),
-    flake_count = sum(blank == "flake", na.rm = TRUE),
-    bladelet_count = sum(blank == "bladelet", na.rm = TRUE),
-    bullet_core_count = sum(`core-typology` == "bullet", na.rm = TRUE),
-    flat_core_count = sum(`core-typology` == "flat", na.rm = TRUE),
-    pyramid_core_count = sum(`core-typology` == "pyramid", na.rm = TRUE),
-    cylinder_core_count = sum(`core-typology` %in% c("prismatic", "cylinder"), na.rm = TRUE),
-    shapeless_core_count = sum(`core-typology` == "shapeless", na.rm = TRUE),
-
-    core_fragment_count = sum(`corefrag` == "core fragment", na.rm = TRUE),
-    core_rejuvention_count = sum(`corefrag` == "rejuvenation piece", na.rm = TRUE),
-    core_tablet_count = sum(`corefrag` == "core tablet", na.rm = TRUE),
-    core_crested_count = sum(`corefrag` == "crested bladelet", na.rm = TRUE),
-    core_primarypiece_count = sum(`corefrag`  %in% c("primary flake", "primary blade"), na.rm = TRUE),
-
-   total = n()
-  )
-```
-
-```{r}
-# plot_blank_type
-
-# Data filtering and plotting
-tl_final %>%
-  filter(area %in% c("ta", "d1", "tp1")) %>% 
-  filter(blank %in% c('bladelet', 'blade', 'flake')) %>%
-  ggplot(aes(x = area, fill = blank)) +
-  geom_bar(position = "dodge") +
-  labs(
-    x = "Area",
-    y = "Number",
-    fill = "Blank Type"
-  ) +
-  theme_minimal()
-
-```
-
-```{r}
-#barplot_layer
-
-tl_final %>%
-  filter(area %in% c("ta", "d1", "tp1")) %>%
-  filter(layer %in% c("a", "c", "d", "e", "eor d")) %>%
-  filter(!is.na(corefrag)) %>% 
-  select(layer, corefrag) %>%
-  ggplot() +
-  aes(x = corefrag) +
-  geom_bar() +  
-  labs(x = "Core and Core fragments", y = "Frequency") +
-  facet_wrap(~layer, ncol = 1) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-```
-
-```{r}
-#Core-barplot_area
-
-tl_final %>%
-  filter(area %in% c("ta", "d1", "tp1")) %>%
-  filter(!is.na(corefrag)) %>% # Exclude rows where corefrag is NA
-  select(area, corefrag) %>%
-  ggplot() +
-  aes(x = corefrag) +
-  geom_bar() +
-  labs(x = "Core and Core fragments",
-       y = "Frequency") +
-  facet_wrap(~area, ncol = 1) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-```
-
-```{r}
-
-filtered_data <- tl_final %>%
-  filter(!is.na(`tool-typology`) & `tool-typology` != "" & `tool-typology` != "NA")
-ggplot(data = filtered_data) + 
-  aes(y = `tool-typology`) + 
-  geom_bar() +
-  labs(x = "Frequency",
-       y = "Tool Type") +
-  theme_minimal() +
-  theme(axis.text.y = element_text(angle = 20, hjust = 1, vjust = 0.5))
-
-```
-
-PCA-LAYER
-
-```{r}
-# prepare the data for PCA-layer
-
-# Calculating Blank Frequency
-blank_freq <- tl_final %>%
-  filter(area %in% c("ta", "d1", "tp1")) %>%
-  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(blank)) %>%
-  group_by(layer) %>%
-  summarise(blankFreq = sum(blank %in% c("bladelet", "blade", "flake")) / sum(!is.na(blank)), .groups = 'drop')
-
-# Calculating Core Frequency
-core_freq <- tl_final %>%
-  filter(area %in% c("ta", "d1", "tp1")) %>%
-  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(corefrag)) %>%
-  group_by(layer) %>%
-  summarise(coreFreq = sum(corefrag %in% c("core fragment", "bullet", "flat", "pyramid", "cylinder/prismatic", "core rejuvenation piece", "shapeless", "core on flake", "bipolar")) / sum(!is.na(corefrag)), .groups = 'drop')
-
-# Calculating Tool Diversity using Menhinick's Index
-tool_div <- tl_final %>%
-  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(tooltypegroup)) %>%
-  group_by(layer) %>%
-  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(sum(!is.na(tooltypegroup))), .groups = 'drop')
-
-# Combining all the calculated frequencies into one tibble
-final_table <- blank_freq %>%
-  inner_join(core_freq, by = "layer") %>%
-  inner_join(tool_div, by = "layer")
-
-# Printing the final table
-final_table
-
-```
-
-```{r}
-# Blank Frequency
-blank_freq <- tl_final %>%
-  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(blank)) %>%
-  group_by(layer) %>%
-  summarise(blankFreq = sum(blank %in% c("bladelet", "blade", "flake")) / sum(!is.na(blank)), .groups = 'drop')
-
-# Calculating Core Frequency
-core_freq <- tl_final %>%
-  filter(layer %in% c("a", "c", "d", "e", "e or d")) %>%
-  group_by(layer) %>%
-  summarise(coreFreq = sum(corefrag %in% c("core fragment", "bullet", "flat", "pyramid", "cylinder/prismatic", "core rejuvenation piece", "shapeless", "core on flake", "bipolar")) / n(), .groups = 'drop')
-
-# Calculating Tool Diversity using Menhinick's Index
-tool_div <- tl_final %>%
-  filter(layer %in% c("a", "c", "d", "e", "e or d"), !is.na(tooltypegroup)) %>%
-  group_by(layer) %>%
-  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(sum(!is.na(tooltypegroup))), .groups = 'drop')
-
-#------------------------------------------------------------------------------
-# BM: I wasn't able to run any code below this line
-##SS: fixed
-# -----------------------------------------------------------------------------
-
-# Calculating Retouch Frequency
-retouch_freq <- tl_final %>%
-  filter(layer %in% c("a", "c", "d", "e", "eor d")) %>%
-  group_by(layer) %>%
-  summarise(retFreq = sum(retouch == 1, na.rm = TRUE) / n(), .groups = 'drop')
-
-# Combining all the calculated frequencies, including retouch frequency, into one tibble
-final_table <- blank_freq %>%
-  inner_join(core_freq, by = "layer") %>%
-  inner_join(tool_div, by = "layer") %>%
-  inner_join(retouch_freq, by = "layer")
-
-final_table
-
-```
-
-```{r}
-# Joining calculated frequencies
-final_table <- blank_freq %>%
-  inner_join(core_freq, by = "layer") %>%
-  inner_join(tool_div, by = "layer") %>%
-  inner_join(retouch_freq, by = "layer")
-
-# Converting layer to row names
-final_table_rownames <- final_table %>%
-  column_to_rownames(var = "layer")
-
-```
-
-```{r}
-# Compute the PCA
-library(FactoMineR)
-
-res_pca <-
-  PCA(final_table_rownames,
-      graph = FALSE)
-
-```
-
-```{r}
-library(factoextra)
-
-# Visualize the contributions of variables to the first principal component
-fviz_contrib(res_pca,
-             axes = 1,   
-             choice = "var")
-
-```
-
-```{r}
-library(factoextra)
-
-# contributions of variables to the second principal component
-fviz_contrib(res_pca,
-             axes = 2,   
-             choice = "var")
-
-```
-
-```{r}
-library(factoextra)
-
-#  biplot
-fviz_pca_biplot(res_pca)
-```
-
-```{r}
-library(factoextra)
-fviz_screeplot(res_pca)
-```
-
-```{r}
-fviz_pca_var(res_pca)
-```
-
-```{r}
-
-library(FactoMineR)
-library(factoextra)
-
-# Run PCA
-res_pca <- PCA(final_table_rownames, graph = FALSE)
-
-# Visualize the contributions of variables to the first principal component
-fviz_contrib(res_pca, axes = 1, choice = "var")
-
-# Visualize the contributions of variables to the second principal component
-fviz_contrib(res_pca, axes = 2, choice = "var")
-
-# Visualize eigenvalues and explained variance
-fviz_screeplot(res_pca)
-
-# Create a biplot of the PCA
-fviz_pca_biplot(res_pca)
-
-# Visualize the variables on the principal components
-fviz_pca_var(res_pca)
-
-```
-
-```{r}
-# Run PCA
-library(FactoMineR)
-library(factoextra)
-
-res_pca <- PCA(final_table_rownames, graph = FALSE)
-
-# Get the coefficients of original variables for each principal component
-loadings_matrix <- res_pca$var$coord
-
-print(loadings_matrix)
-
-```
-
-```{r}
-#Table
-
-# Eigenvalues
-eigenvalues <- res_pca$eig[1:2, 1]
-
-# Variance percent
-variance_percent <- res_pca$eig[1:2, 2]
-
-# Cumulative variance percent
-cumulative_variance_percent <- res_pca$eig[1:2, 3]
-
-# Create the table
-df <- data.frame(Eigenvalue = eigenvalues,
-                 Variance_Percent = variance_percent,
-                 Cumulative_Variance_Percent = cumulative_variance_percent)
-
-rownames(df) <- c("Dim1", "Dim2")
-
-print(df)
-
-```
-
-```{r}
-#Table
-
-library(gt)
-
-# Create data frame
-df <- data.frame(Eigenvalue = eigenvalues,
-                 Variance_Percent = variance_percent,
-                 Cumulative_Variance_Percent = cumulative_variance_percent)
-rownames(df) <- c("Dim1", "Dim2")
-
-# Create the table using gt
-gt(df) %>%
-  tab_header(
-    title = "",
-    subtitle = "Eigenvalues and percentage of variance for each dimension of PCA"
-  ) %>%
-  cols_label(
-    Eigenvalue = "Eigenvalue",
-    Variance_Percent = "Variance (%)",
-    Cumulative_Variance_Percent = "Cumulative Variance (%)"
-  ) %>%
-  fmt_number(
-    columns = vars(Eigenvalue, Variance_Percent, Cumulative_Variance_Percent),
-    decimals = 2
-  )
-
-```
-
-feature frequency: d1 1 ta 1 tp1 0)
-
-```{r}
-library(dplyr)
-library(FactoMineR)
-library(tibble)
-
-# Define estimated area and feature frequency data
-area_data <- tibble(
-  area = c("d1", "ta", "tp1"),
-  feature_freq = c(1, 1, 0)
-)
-
-# Step 1: Number of artifacts by area
-artifact_counts <- tl_final_area %>%
-  group_by(area) %>%
-  tally()
-
-# Step 2: Calculate each variable
-
-# Retouch Frequency 
-retouch_freq <- tl_final_area %>%
-  group_by(area) %>%
-  summarise(retouchFreq = mean(!is.na(retouch)), .groups = 'drop')
-
-# Blank Frequency
-blank_freq <- tl_final_area %>%
-  group_by(area) %>%
-  summarise(blankFreq = mean(!is.na(blank)), .groups = 'drop')
-
-# Core Frequency
-core_freq <- tl_final_area %>%
-  group_by(area) %>%
-  summarise(coreFreq = mean(!is.na(`core-typology`)),.groups = 'drop')
-
-# Lithic density (Artifact counts / estimated area)
-lithic_density <- artifact_counts %>%
-  left_join(area_data, by = "area") %>%
-  mutate(lithic_density = n / estimated_area)
-
-# Tool Diversity using Menhinick's Index
-tool_div <- tl_final_area %>%
-  group_by(area) %>%
-  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(n()), .groups = 'drop')
-
-# Step 3: Combining all variables
-
-pca_final_table_area <- left_join(dimensions, retouch_freq, by = "area") %>%
-  left_join(blank_freq, by = "area") %>%
-  left_join(core_freq, by = "area") %>%
-  left_join(lithic_density, by = "area") %>%
-  left_join(tool_div, by = "area") %>%
-  select(area, new_dimensions, feature_freq, retouchFreq, blankFreq, coreFreq, lithic_density, toolDiv)
-
-```
-
-```{r}
-pca_final_table_area <- pca_final_table_area %>%
-    select(
-        area, 
-        estimated_area = estimated_area.x, 
-        feature_freq = feature_freq.x, 
-        retouchFreq, 
-        blankFreq, 
-        coreFreq, 
-        n, 
-        lithic_density, 
-        toolDiv
-    )
-
-```
-
-```{r}
-# Step 4: Perform PCA
-res_pca_area <- PCA(pca_final_table_area %>% select(-area, -n), graph = FALSE)
-
-# Show the summary of PCA
-print(res_pca_area)
-
-print(res_pca_area$eig)
-print(res_pca_area$var$coord)
-```
-
-```{r}
-# TABLE: Eigenvalues and percentage of variance for each dimension of PCA 
-library(kableExtra)
-
-# Define the values manually (you would get these from your PCA results)
-# BM: I prefer not to write values in the code like this. I think
-# this is bad practice and leads to mistakes
-eigenvalues <- c(4.411234, 2.588766)
-prop_var_explained <- c(63.02, 36.98)
-cumulative_var_explained <- c(63.02, 100.00)
-
-# Create the data frame
-results_df <- data.frame(
-  Dimension = c("Dim. 1", "Dim. 2"),  # Only keep the dimensions you want to display
-  Eigenvalue = eigenvalues,
-  `Variance / percent` = prop_var_explained,
-  `Cumulative / variance percent` = cumulative_var_explained
-)
-
-# Generate the table
-kable(results_df,
-      caption = "Table 3. Eigenvalues and percentage of variance for each dimension of PCA",
-      align = 'c',
-      col.names = c("Dimension", "Eigenvalue", "Variance (%)", "Cumulative Variance (%)")) %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 11)
-
-```
-
-```{r}
-#plotting Dim-1 and Dim-2
-library(FactoMineR)
-library(factoextra)
-
-# Visualize the contributions of variables to the first principal component
-fviz_contrib(res_pca_area, axes = 1, choice = "var")
-
-# Visualize the contributions of variables to the second principal component
-fviz_contrib(res_pca_area, axes = 2, choice = "var")
-
-```
-
-```{r}
-
-# having Dim1 and Dim2 plots next to each other
-
-library(gridExtra)
-
-# Create the first plot and assign it to p1
-p1 <- fviz_contrib(res_pca_area, axes = 1, choice = "var")
-
-# Create the second plot and assign it to p2
-p2 <- fviz_contrib(res_pca_area, axes = 2, choice = "var")
-
-# Use grid.arrange to place them next to each other
-grid.arrange(p1, p2, ncol = 2)
-
-```
-
-```{r}
-
-loadings <- res_pca_area$var$coord
-
-# 
-eigenvalues <- res_pca_area$eig
-
-#  Dim.1
-contrib_dim1 <- (loadings[,1]^2 / eigenvalues[1, 1]) * 100
-
-#  Dim.2
-contrib_dim2 <- (loadings[,2]^2 / eigenvalues[2, 1]) * 100
-
-contrib_df <- data.frame(
-  Variable = rownames(loadings),
-  Contrib_Dim1 = contrib_dim1,
-  Contrib_Dim2 = contrib_dim2
-)
-
-print(contrib_df)
-
-```
-
-```{r}
-# screeplot
-
-library(FactoMineR)
-library(factoextra)
-
-fviz_eig(res_pca_area)
-
-```
-
-```{r}
-# Correlation plots of variables for each of the tow relevant PCA dimensions
-
-fviz_pca_var(res_pca_area,
-
-             repel = TRUE) + 
-  theme(text = element_text(size = 10))
-
-```
-
-```{r}
-# Eigenvalues and percentage of variance
-eig.val <- res_pca_area$eig
-print(eig.val)
-
-# Variable coordinates (correlations with principal components)
-var.coord <- res_pca_area$var$coord
-print(var.coord)
-
-```
-
-```{r}
-#PCA biplot for Dimensions 1 and 2
-
-fviz_pca_biplot(res_pca_area, labelsize = 3, repel = TRUE)
-
-```
-
-```{r}
-# biplot 
-p <- fviz_pca_biplot(res_pca_area, label = "none", 
-                     repel = TRUE, labelsize = 4)
-
-p + geom_text(aes(x = res_pca_area$ind$coord[, 1], 
-                  y = res_pca_area$ind$coord[, 2], 
-                  label = pca_final_table_area$area), 
-             size = 3, nudge_x = 0.1, nudge_y = 0.1)
-
-```
-
-```{r}
-individual_coordinates = res_pca_area$ind$coord
-variable_coordinates = res_pca_area$var$coord
-
-```
-
-```{r}
-# For variables
-fviz_cos2(res_pca_area, choice = "var")
-
-# For individuals
-fviz_cos2(res_pca_area, choice = "ind")
-
-```
-
-DONE WITH PCA
-
-```{r}
-
-individual_coordinates <- res_pca_area$ind$coord
-
-individual_coordinates_df <- as.data.frame(individual_coordinates)
-individual_coordinates_df$area <- unique(area_data$area)
-
-print(individual_coordinates_df)
-
-```
-
-```{r}
-library(kableExtra)
-
-# Extract individual coordinates
-individual_coordinates <- res_pca_area$ind$coord
-
-# Convert to a data frame for better visibility and add the 'area' column back for context
-individual_coordinates_df <- as.data.frame(individual_coordinates)
-individual_coordinates_df$Area <- unique(area_data$area)  # Make sure the order matches
-
-# Use kable to create a beautiful table
-kable(individual_coordinates_df, 
-      col.names = c("Area", "Principal Component 1", "Principal Component 2"), 
-      caption = "PCA Coordinates for the Archaeological Areas excavation",
-      align = c('l', 'c', 'c'),
-      digits = 3) %>%
-  kable_styling("striped", full_width = F)
-
-```
-
-```{r}
-# Get the loadings (coefficients) of original variables for each principal component
-loadings_matrix <- res_pca$var$coord
-
-```
-
-```{r}
-# Eigenvalues
-eigenvalues <- res_pca$eig[1:2, 1]
-variance_percent <- res_pca$eig[1:2, 2]
-cumulative_variance_percent <- res_pca$eig[1:2, 3]
-
-# Create the table
-df <- data.frame(Eigenvalue = eigenvalues,
-                 Variance_Percent = variance_percent,
-                 Cumulative_Variance_Percent = cumulative_variance_percent)
-
-rownames(df) <- c("Dim1", "Dim2")
-
-# Print the table
-print(df)
-
-```
-
-```{r}
-library(gt)
-
-gt(df) %>%
-  tab_header(
-    title = "",
-    subtitle = "Eigenvalues and percentage of variance for each dimension of PCA"
-  ) %>%
-  cols_label(
-    Eigenvalue = "Eigenvalue",
-    Variance_Percent = "Variance (%)",
-    Cumulative_Variance_Percent = "Cumulative Variance (%)"
-  ) %>%
-  fmt_number(
-    columns = c("Eigenvalue", "Variance_Percent", "Cumulative_Variance_Percent"),
-    decimals = 2
-  )
-
-```
-
-```{r}
-
-artifact_counts <- tl_final %>%
-  filter(area %in% c("ta", "tp1", "d1")) %>%
-  group_by(area) %>%
-  tally()
-
-# Print the resulting table
-print(artifact_counts)
-```
-
-```{r}
-
-blank_freq <- tl_final_area %>%
-  filter(area %in% c("ta", "d1", "tp1"), !is.na(blank)) %>%
-  group_by(area) %>%
-  summarise(blankFreq = sum(blank %in% c("bladelet", "blade", "flake")) / sum(!is.na(blank)), .groups = 'drop')
-
-# Calculating Core Frequency
-core_freq <- tl_final %>%
-  filter(area %in% c("ta", "d1", "tp1"), !is.na(corefrag)) %>%
-  group_by(area) %>%
-  summarise(coreFreq = sum(corefrag %in% c("core fragment", "bullet", "flat", "pyramid", "cylinder/prismatic", "core rejuvenation piece", "shapeless", "core on flake", "bipolar")) / sum(!is.na(corefrag)), .groups = 'drop')
-
-# Calculating Tool Diversity using Menhinick's Index
-tool_div <- tl_final_area %>%
-  filter(area %in% c("ta", "d1", "tp1"), !is.na(tooltypegroup)) %>%
-  group_by(area) %>%
-  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(sum(!is.na(tooltypegroup))), .groups = 'drop')
-
-# Combining all the calculated frequencies into one tibble
-final_table <- blank_freq %>%
-  inner_join(core_freq, by = "area") %>%
-  inner_join(tool_div, by = "area")
-
-# Printing the final table
-print(final_table)
-
-```
-
-```{r}
-# Joining calculated frequencies
-final_table <- blank_freq %>%
-  inner_join(core_freq, by = "area") %>%
-  inner_join(tool_div, by = "area") %>%
-  inner_join(retouch_freq, by = "area")
-
-# Converting area to row names
-final_table_rownames <- final_table %>%
-  column_to_rownames(var = "area")
-
-```
-
-```{r}
-
-library(FactoMineR)
-library(factoextra)
-
-# Run PCA
-res_pca <- PCA(final_table_rownames, graph = FALSE)
-
-# Visualize the contributions of variables to the second principal component
-fviz_contrib(res_pca, axes = 2, choice = "var")
-
-```
-
-```{r}
-# Visualize the contributions of variables to the first principal component
-fviz_contrib(res_pca, axes = 1, choice = "var") 
-```
-
-```{r}
-# Visualize eigenvalues and explained variance
-fviz_screeplot(res_pca)
-```
-
-```{r}
-fviz_pca_biplot(res_pca, labelsize = 3, repel = TRUE)
-
-```
-
-```{r}
-# Visualize the variables on the principal components
-fviz_pca_var(res_pca)
-```
-
-```{r}
-# Get the loadings (coefficients) of original variables for each principal component
-loadings_matrix <- res_pca$var$coord
-
-# Print the loadings matrix
-print(loadings_matrix)
-
-```
-
-```{r}
-# Eigenvalues
-eigenvalues <- res_pca$eig[1:2, 1]
-variance_percent <- res_pca$eig[1:2, 2]
-cumulative_variance_percent <- res_pca$eig[1:2, 3]
-
-# Create the table
-df <- data.frame(Eigenvalue = eigenvalues,
-                 Variance_Percent = variance_percent,
-                 Cumulative_Variance_Percent = cumulative_variance_percent)
-
-rownames(df) <- c("Dim1", "Dim2")
-
-# Print the table
-print(df)
-
-```
-
-```{r}
-library(gt)
-
-# Create the table 
-gt(df) %>%
-  tab_header(
-    title = "",
-    subtitle = "Eigenvalues and percentage of variance for each dimension of PCA"
-  ) %>%
-  cols_label(
-    Eigenvalue = "Eigenvalue",
-    Variance_Percent = "Variance (%)",
-    Cumulative_Variance_Percent = "Cumulative Variance (%)"
-  ) %>%
-  fmt_number(
-    columns = c("Eigenvalue", "Variance_Percent", "Cumulative_Variance_Percent"),
-    decimals = 2
-  )
-
-```
-
-```{r}
-# Count the number of artifacts by area
-artifact_counts <- tl_final %>%
-  filter(area %in% c("ta", "tp1", "d1")) %>%
-  group_by(area) %>%
-  tally()
-
-# Print the resulting table
-print(artifact_counts)
-
-```
-
-```{r}
-
-volumes <- data.frame(
-  area = c("ta", "tp1", "d1"),
-  volume = c(52, 5.76, 2.1375)
-)
-
-# Count the number of artifacts by area
-artifact_counts <- tl_final %>%
-  filter(area %in% c("ta", "tp1", "d1")) %>%
-  group_by(area) %>%
-  tally()
-
-# Join artifact_counts with volumes by area
-lithic_density_df <- left_join(artifact_counts, volumes, by = "area")
-
-# Compute lithic density for each area
-lithic_density_df <- lithic_density_df %>%
-  mutate(lithic_density = n / volume)
-
-# Print the resulting table
-print(lithic_density_df[, c("area", "lithic_density")])
-
-```
-
-```{r}
-# Counting the number of cores by area
-core_counts <- tl_final %>%
-  filter(area %in% c("ta", "d1", "tp1") & !is.na(`core and core fragments`)) %>%
-  group_by(area) %>%
-  tally()
-
-print(core_counts)
-
-```
+
+```

### 2023-09-18 paper/scraps/scraps-tulai.qmd
edited and polished the codes. Good to render now.

@@
 ```
+
+PCA_al:
+
+```{r}
+#PCA: area_layer (al)
+#I think as we are going to calculate PCA to understand other factors that might be affected the duration of the occupation, it would be logical to calculate it for each layer, instead of level.
+we have 5 stratigraphical layers: e, d, c, b, a
+#d1:1.5*1.5/ ta: 4*5/ tp1:2*1.8
+#e: excavated area is A/ depth of excavation is 0-75(mean)
+#d: excavated area is A/ 0.75m-3m→ 2.25
+#c: d1 and TP → 30 cm and 40cm
+#b: TP/ 170 cm 
+#a: D1: 70cm
+
+# Add 'dimension' columns and relocate it after 'al' (based on Hole (1974))
+library(dplyr)
+
+summary_table_al_pca <- summary_table_al %>%
+  mutate(
+    length = case_when(
+      al == "D1_a" ~ 1.5,
+      al == "D1_c" ~ 1.5,
+      al == "TA_d" ~ 4,
+      al == "TP1_c" ~ 2,
+      al == "TA_e" ~ 4,
+      TRUE ~ NA_real_
+    ),
+    width = case_when(
+      al == "D1_a" ~ 1.5,
+      al == "D1_c" ~ 1.5,
+      al == "TA_d" ~ 5,
+      al == "TP1_c" ~ 1.8,
+       al == "TA_e" ~ 5,
+      TRUE ~ NA_real_
+    ),
+    thickness = case_when(
+      al == "D1_a" ~ 0.7,
+      al == "D1_c" ~ 0.3,
+      al == "TA_d" ~ 2.25,
+      al == "TP1_c" ~ 0.4,
+      al == "TA_e" ~ 0.75,
+      TRUE ~ NA_real_)
+  ) %>%
+
+  mutate(volume = length * width * thickness) %>%
+  relocate(length, width, thickness, volume, .after = al)
+# Show the updated data frame
+summary_table_al_pca
+
+```
+
+```{r}
+# Load required library
+library(dplyr)
+
+# Add a new column called 'blank_count'
+summary_table_al_pca <- summary_table_al_pca %>%
+  mutate(blank_count = blade_count + flake_count + bladelet_count) %>%
+  select(al, length, width, thickness, volume, blade_count, flake_count, bladelet_count, blank_count, everything())
+
+# Calculate core frequency, blank frequency, retouch frequency, lithic density 
+summary_table_al_pca <- summary_table_al_pca %>%
+  mutate(
+    lithic_density = round((core_count + blank_count + core_rejuvenation_count) / volume, 2),
+    core_freq = round(core_count / volume, 2),
+    blank_freq = round(blank_count / volume, 2),
+    retouch_freq = round(retouched_tool_count / volume, 2)
+  )
+
+# Calculate tool diversity using Menhinick's index
+summary_table_al_with_diversity <- tl_final %>% 
+  filter(retouch == 1) %>%
+  group_by(al, `tool-typology`) %>%  
+  summarise(n = n(), .groups = 'drop') %>% 
+  group_by(al) %>%  
+  summarise(
+    num_unique_tool_types = n_distinct(`tool-typology`),
+    total_retouched_tools = sum(n),
+    tool_diversity = round(num_unique_tool_types / sqrt(total_retouched_tools), 2),
+    .groups = 'drop'
+  )
+
+# Merge this new information into the existing summary_table_al_pca
+summary_table_al_pca <- left_join(summary_table_al_pca, summary_table_al_with_diversity, by = "al")
+
+# Remove the 'Total' row and select only the columns needed for PCA
+fin_summary_table_al_pca <- summary_table_al_pca %>%
+  filter(al != "Total") %>%
+  select(al, core_freq, blank_freq, lithic_density, retouch_freq, tool_diversity)
+
+# Show the final table
+print(fin_summary_table_al_pca)
+
+```
+
+```{r}
+library(FactoMineR)
+# Perform PCA
+res.pca <- PCA(fin_summary_table_al_pca[, -1], graph = FALSE)  # Exclude the 'al' column
+
+```
+
+```{r}
+## Got error: couldnt figure out the reason of error
+
+# Create a data frame to hold the rounded eigenvalues and the rounded percentage variance
+eigen_df <- data.frame(
+  Dimension = paste("Dim", 1:nrow(eigenvalues), sep = "."),
+  Eigenvalue = round(eigenvalues[, 1], 3),
+  `Variance percent` = round(eigenvalues[, 2], 3),
+  `Cumulative variance percent` = round(cumsum(eigenvalues[, 2]), 3)
+)
+
+# Print the table
+print(eigen_df)
+
+```
+
+```{r}
+# variables to PC1
+fviz_contrib(res.pca,  
+             choice = "var",  
+             axes = 1,  
+             top = 10)
+```
+
+```{r}
+fviz_contrib(res.pca,  
+             choice = "var",  
+             axes = 2,  
+             top = 10)
+
+```
+
+```{r}
+fviz_contrib(res.pca,  
+             choice = "var",  
+             axes = 3,  
+             top = 10)
+```
+
+```{r}
+fviz_contrib(res.pca,  
+             choice = "var",  
+             axes = 4,  
+             top = 10)
+```
+
+```{r}
+#Plot: Contribution of variables for each of the four relevant PCA dimensions
+library(factoextra)
+library(ggplot2)
+
+#save contribution plots into one objects
+p1 <- fviz_contrib(res.pca, 
+                   choice = "var", 
+                   axes = 1, 
+                   top = 10)
+p2 <- fviz_contrib(res.pca, 
+                   choice = "var",
+                   axes = 2, 
+                   top = 10)
+p3 <- fviz_contrib(res.pca, 
+                   choice = "var", 
+                   axes = 3, 
+                   top = 10)
+p4 <- fviz_contrib(res.pca, 
+                   choice = "var", 
+                   axes = 4, 
+                   top = 10)
+
+#combine individual plots into one plot 
+library(gridExtra)
+grid.arrange(p1, p2, p3, p4, ncol = 2)
+
+```
+
+```{r}
+# Extract contributions of variables to each principal component
+contributions <- res.pca$var$contrib
+
+# Compute average contribution for each dimension
+avg_contrib_dim1 <- mean(contributions[,1], na.rm = TRUE)
+avg_contrib_dim2 <- mean(contributions[,2], na.rm = TRUE)
+avg_contrib_dim3 <- mean(contributions[,3], na.rm = TRUE)
+avg_contrib_dim4 <- mean(contributions[,4], na.rm = TRUE)
+
+# Find variables that contribute more than average to each dimension
+high_contrib_dim1 <- contributions[contributions[,1] > avg_contrib_dim1, 1]
+high_contrib_dim2 <- contributions[contributions[,2] > avg_contrib_dim2, 2]
+high_contrib_dim3 <- contributions[contributions[,3] > avg_contrib_dim3, 3]
+high_contrib_dim4 <- contributions[contributions[,4] > avg_contrib_dim4, 4]
+
+# Sort them in descending order
+sorted_high_contrib_dim1 <- sort(high_contrib_dim1, decreasing = TRUE)
+sorted_high_contrib_dim2 <- sort(high_contrib_dim2, decreasing = TRUE)
+sorted_high_contrib_dim3 <- sort(high_contrib_dim3, decreasing = TRUE)
+sorted_high_contrib_dim4 <- sort(high_contrib_dim4, decreasing = TRUE)
+
+# Print or use for further analysis
+print("Significant contributions to Dim1:")
+print(sorted_high_contrib_dim1)
+
+print("Significant contributions to Dim2:")
+print(sorted_high_contrib_dim2)
+
+print("Significant contributions to Dim3:")
+print(sorted_high_contrib_dim3)
+
+print("Significant contributions to Dim4:")
+print(sorted_high_contrib_dim4)
+
+```
+
+```{r}
+# screeplot; inspect distribution of PCs
+library(factoextra)  
+fviz_screeplot(res.pca)
+```
+
+```{r}
+# Inspect eigenvalues
+print(res.pca$eig)
+
+```
+
+```{r}
+# Load the required package
+library(kableExtra)
+
+# Create a table with kable and kableExtra
+eigen_df %>%
+  kable("html") %>%
+  kable_styling("striped", full_width = F) %>%
+  column_spec(1, bold = T)
+```
+
+```{r}
+
+# Rounding the numerical columns to 3 decimal places
+eigen_df[, -1] <- round(eigen_df[, -1], 3)
+
+# Replacing 'comp' with 'Dim'
+eigen_df$Dimension <- gsub("comp", "Dim", eigen_df$Dimension)
+
+# Removing row names
+rownames(eigen_df) <- NULL
+
+# Creating the kable output
+kable_output <- kable(eigen_df, 
+                      "html", 
+                      align = 'c', 
+                      col.names = c("Dimension", 
+                                    "Eigenvalue", 
+                                    "Variance percent", 
+                                    "Cumulative variance percent")) %>%
+  kable_styling("striped", 
+                full_width = F) %>%
+  add_header_above(c(" " = 1, "Eigenvalues and percentage of variance for each dimension of PCA" = 3))
+
+htmltools::browsable(kable_output)
+
+```
+
+```{r}
+# Visualize biplot:1
+fviz_pca_biplot(res.pca,
+                axes = c(1, 2))
+rownames(res.pca$ind$coord) <- fin_summary_table_al_pca$al
+# Visualize biplot with row names
+fviz_pca_biplot(res.pca)
+
+```
+
+```{r}
+# Visualize biplot: 2
+fviz_pca_biplot(res.pca,
+                axes = c(3, 4))
+rownames(res.pca$ind$coord) <- fin_summary_table_al_pca$al
+# Visualize biplot with row names
+fviz_pca_biplot(res.pca)
+
+print(fviz_pca_biplot(res.pca, axes = c(3, 4)))
+
+```
+
+```{r}
+# BM: My preference would be to show the calculation of these 
+  # values from the raw data, rather than type them in 
+  # here, where we can't see where they came from, we can't easily
+  # verify these values as they are currently 
+##SS: fixed
+
+# Define the volumes for each area
+
+# add the dimensions for each area
+dimensions <- data.frame(
+  area = c("ta", "tp1", "d1"),
+  length = c(4, 2, 1.5),
+  width = c(5, 1.8, 1.5),
+  thickness = c(3, 1.6, 0.95)
+)
+
+# Calculate volume for each area based on the dimensions
+dimensions <- dimensions %>%
+  mutate(volume = length * width * thickness)
+
+# Count the number of artifacts by area
+artifact_counts <- tl_final %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  tally()
+
+# Join artifact_counts with dimensions (which now includes volume) by area
+lithic_density_df <- left_join(artifact_counts, dimensions, by = "area")
+
+# Compute lithic density for each area and round it to 2 decimal places
+lithic_density_df <- lithic_density_df %>%
+  mutate(lithic_density = round(n / volume, 2)) # number of artifacts / volume
+
+# Create the summary table
+summary_table2_area <- tl_final %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  summarise(
+    numb_retouched = sum(retouch == 1, na.rm = TRUE),
+    total_lithics = n(),
+    percent_retouch = round((numb_retouched / total_lithics), 2)
+  )
+
+# Merge the density data with the summary table
+summary_table2_area <- 
+  left_join(summary_table2_area, 
+            lithic_density_df[, c("area", "lithic_density")], 
+            by = "area") %>%
+  mutate(area = recode(area, 'ta' = 'TA', 
+                       'tp1' = 'TP1', 
+                       'd1' = 'D1'),
+         lithic_volumetric_density = lithic_density)
+
+# Remove the 'lithic_density' column
+summary_table2_area <- summary_table2_area %>% 
+  select(-lithic_density)
+
+# Print out the resulting data frame
+print(summary_table2_area)
+
+```
+
+```{r}
+# generate table 2 - area
+
+library(kableExtra)
+
+kable(summary_table2_area, 
+      caption = "Table 2. Lithic counts for each of the areas in Tula'i", 
+      align = 'c',
+      col.names = c("Area", 
+                    "Number of Retouched Lithics", 
+                    "Total Number of Lithics", 
+                    "Retouch frequency", 
+                    "Lithic Volumetric Density(pcs/(m^3)")) %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 11) %>%
+  column_spec(1, width = "1cm") %>% 
+  column_spec(2, width = "1cm") %>%
+
+  column_spec(3, width = "1cm") %>% 
+  column_spec(4, width = "1cm") %>% 
+  column_spec(5, width = "1cm")
+
+```
+
+------------------------------
+Done with the main plots
+------------------------------
+
+```{r}
+# WABI_LEVEL
+
+# BM: I am confused why we are getting the counts of artefacts 
+# and retouch pieces from the google sheet, after you wrote so many lines of
+# code above to prepare the data from the CSV file. Where did the lithic data
+# in the google sheet come from? My preference is to have 
+# only copy of the lithic data in one place, and then join it with 
+# other data using code. Currently it seems like we have the 
+# lithic data in both the google sheet and the CSV file here. I'm worried that 
+# will be confusing and be difficult to update and keep correct. I prefer that we 
+# omit the lithic data from the google sheet, then import the google sheet and 
+# join it to the lithic data from the CSV file. Then we can have a data frame
+# with the area/layer/level data and lithic data to work on for plotting.
+
+#SS: the googlesheet is the last thing I created.
+I created the summary table in R, and then for splitting up the depth (we have few depth range that was not possible to assign the layer to them, so, we splited them), I found it is easier to do so in Google sheet.
+
+filtered_data <- sum_depth_sheet %>% filter(lithic_dens > 0 & retouch_freq > 0)
+
+WABI_Level <- ggplot(filtered_data, 
+            aes(x = lithic_dens,
+                y = retouch_freq)) +
+  geom_point(color = "black", size = 2) +
+  geom_smooth(method = "lm", 
+              se = TRUE, 
+              color = "blue", 
+              size = 0.5) +
+  geom_text(aes(label = area_layer_level), 
+            vjust = 1.5, 
+            hjust = 0.5, 
+            size = 2,
+            check_overlap = TRUE) +
+  stat_poly_eq(aes(label = paste(stat(eq.label), 
+                                 stat(rr.label), 
+                                 stat(p.value.label), 
+                                 sep = "~~~")),
+               formula = y ~ x, 
+               parse = TRUE, 
+               size = 2) +
+  # Add titles and labels
+  labs(
+    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
+    x = "Lithic Volumetric Density",
+    y = "Retouch Frequency"
+  ) +
+
+  theme(
+    axis.text = element_text(size = 8),
+    axis.title = element_text(size = 10),
+    plot.title = element_text(size = 12)
+  )
+
+WABI_Level
+
+```
+
+```{r}
+# whole assemblage
+
+  # Summarize data by Area
+summary_table_tech <- tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1")) %>%
+  group_by(area) %>% 
+  summarise(
+   blade_count = sum(blank == "blade", na.rm = TRUE),
+    flake_count = sum(blank == "flake", na.rm = TRUE),
+    bladelet_count = sum(blank == "bladelet", na.rm = TRUE),
+    bullet_core_count = sum(`core-typology` == "bullet", na.rm = TRUE),
+    flat_core_count = sum(`core-typology` == "flat", na.rm = TRUE),
+    pyramid_core_count = sum(`core-typology` == "pyramid", na.rm = TRUE),
+    cylinder_core_count = sum(`core-typology` %in% c("prismatic", "cylinder"), na.rm = TRUE),
+    shapeless_core_count = sum(`core-typology` == "shapeless", na.rm = TRUE),
+
+    core_fragment_count = sum(`corefrag` == "core fragment", na.rm = TRUE),
+    core_rejuvention_count = sum(`corefrag` == "rejuvenation piece", na.rm = TRUE),
+    core_tablet_count = sum(`corefrag` == "core tablet", na.rm = TRUE),
+    core_crested_count = sum(`corefrag` == "crested bladelet", na.rm = TRUE),
+    core_primarypiece_count = sum(`corefrag`  %in% c("primary flake", "primary blade"), na.rm = TRUE),
+
+   total = n()
+  )
+```
+
+```{r}
+# plot_blank_type
+
+# Data filtering and plotting
+tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1")) %>% 
+  filter(blank %in% c('bladelet', 'blade', 'flake')) %>%
+  ggplot(aes(x = area, fill = blank)) +
+  geom_bar(position = "dodge") +
+  labs(
+    x = "Area",
+    y = "Number",
+    fill = "Blank Type"
+  ) +
+  theme_minimal()
+
+```
+
+```{r}
+#barplot_layer
+
+tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1")) %>%
+  filter(layer %in% c("a", "c", "d", "e", "eor d")) %>%
+  filter(!is.na(corefrag)) %>% 
+  select(layer, corefrag) %>%
+  ggplot() +
+  aes(x = corefrag) +
+  geom_bar() +  
+  labs(x = "Core and Core fragments", y = "Frequency") +
+  facet_wrap(~layer, ncol = 1) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
+
+```{r}
+#Core-barplot_area
+
+tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1")) %>%
+  filter(!is.na(corefrag)) %>% # Exclude rows where corefrag is NA
+  select(area, corefrag) %>%
+  ggplot() +
+  aes(x = corefrag) +
+  geom_bar() +
+  labs(x = "Core and Core fragments",
+       y = "Frequency") +
+  facet_wrap(~area, ncol = 1) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+```
+
+```{r}
+
+filtered_data <- tl_final %>%
+  filter(!is.na(`tool-typology`) & `tool-typology` != "" & `tool-typology` != "NA")
+ggplot(data = filtered_data) + 
+  aes(y = `tool-typology`) + 
+  geom_bar() +
+  labs(x = "Frequency",
+       y = "Tool Type") +
+  theme_minimal() +
+  theme(axis.text.y = element_text(angle = 20, hjust = 1, vjust = 0.5))
+
+```
+
+PCA-LAYER
+
+```{r}
+# prepare the data for PCA-layer
+
+# Calculating Blank Frequency
+blank_freq <- tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1")) %>%
+  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(blank)) %>%
+  group_by(layer) %>%
+  summarise(blankFreq = sum(blank %in% c("bladelet", "blade", "flake")) / sum(!is.na(blank)), .groups = 'drop')
+
+# Calculating Core Frequency
+core_freq <- tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1")) %>%
+  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(corefrag)) %>%
+  group_by(layer) %>%
+  summarise(coreFreq = sum(corefrag %in% c("core fragment", "bullet", "flat", "pyramid", "cylinder/prismatic", "core rejuvenation piece", "shapeless", "core on flake", "bipolar")) / sum(!is.na(corefrag)), .groups = 'drop')
+
+# Calculating Tool Diversity using Menhinick's Index
+tool_div <- tl_final %>%
+  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(tooltypegroup)) %>%
+  group_by(layer) %>%
+  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(sum(!is.na(tooltypegroup))), .groups = 'drop')
+
+# Combining all the calculated frequencies into one tibble
+final_table <- blank_freq %>%
+  inner_join(core_freq, by = "layer") %>%
+  inner_join(tool_div, by = "layer")
+
+# Printing the final table
+final_table
+
+```
+
+```{r}
+# Blank Frequency
+blank_freq <- tl_final %>%
+  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(blank)) %>%
+  group_by(layer) %>%
+  summarise(blankFreq = sum(blank %in% c("bladelet", "blade", "flake")) / sum(!is.na(blank)), .groups = 'drop')
+
+# Calculating Core Frequency
+core_freq <- tl_final %>%
+  filter(layer %in% c("a", "c", "d", "e", "e or d")) %>%
+  group_by(layer) %>%
+  summarise(coreFreq = sum(corefrag %in% c("core fragment", "bullet", "flat", "pyramid", "cylinder/prismatic", "core rejuvenation piece", "shapeless", "core on flake", "bipolar")) / n(), .groups = 'drop')
+
+# Calculating Tool Diversity using Menhinick's Index
+tool_div <- tl_final %>%
+  filter(layer %in% c("a", "c", "d", "e", "e or d"), !is.na(tooltypegroup)) %>%
+  group_by(layer) %>%
+  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(sum(!is.na(tooltypegroup))), .groups = 'drop')
+
+#------------------------------------------------------------------------------
+# BM: I wasn't able to run any code below this line
+##SS: fixed
+# -----------------------------------------------------------------------------
+
+# Calculating Retouch Frequency
+retouch_freq <- tl_final %>%
+  filter(layer %in% c("a", "c", "d", "e", "eor d")) %>%
+  group_by(layer) %>%
+  summarise(retFreq = sum(retouch == 1, na.rm = TRUE) / n(), .groups = 'drop')
+
+# Combining all the calculated frequencies, including retouch frequency, into one tibble
+final_table <- blank_freq %>%
+  inner_join(core_freq, by = "layer") %>%
+  inner_join(tool_div, by = "layer") %>%
+  inner_join(retouch_freq, by = "layer")
+
+final_table
+
+```
+
+```{r}
+# Joining calculated frequencies
+final_table <- blank_freq %>%
+  inner_join(core_freq, by = "layer") %>%
+  inner_join(tool_div, by = "layer") %>%
+  inner_join(retouch_freq, by = "layer")
+
+# Converting layer to row names
+final_table_rownames <- final_table %>%
+  column_to_rownames(var = "layer")
+
+```
+
+```{r}
+# Compute the PCA
+library(FactoMineR)
+
+res_pca <-
+  PCA(final_table_rownames,
+      graph = FALSE)
+
+```
+
+```{r}
+library(factoextra)
+
+# Visualize the contributions of variables to the first principal component
+fviz_contrib(res_pca,
+             axes = 1,   
+             choice = "var")
+
+```
+
+```{r}
+library(factoextra)
+
+# contributions of variables to the second principal component
+fviz_contrib(res_pca,
+             axes = 2,   
+             choice = "var")
+
+```
+
+```{r}
+library(factoextra)
+
+#  biplot
+fviz_pca_biplot(res_pca)
+```
+
+```{r}
+library(factoextra)
+fviz_screeplot(res_pca)
+```
+
+```{r}
+fviz_pca_var(res_pca)
+```
+
+```{r}
+
+library(FactoMineR)
+library(factoextra)
+
+# Run PCA
+res_pca <- PCA(final_table_rownames, graph = FALSE)
+
+# Visualize the contributions of variables to the first principal component
+fviz_contrib(res_pca, axes = 1, choice = "var")
+
+# Visualize the contributions of variables to the second principal component
+fviz_contrib(res_pca, axes = 2, choice = "var")
+
+# Visualize eigenvalues and explained variance
+fviz_screeplot(res_pca)
+
+# Create a biplot of the PCA
+fviz_pca_biplot(res_pca)
+
+# Visualize the variables on the principal components
+fviz_pca_var(res_pca)
+
+```
+
+```{r}
+# Run PCA
+library(FactoMineR)
+library(factoextra)
+
+res_pca <- PCA(final_table_rownames, graph = FALSE)
+
+# Get the coefficients of original variables for each principal component
+loadings_matrix <- res_pca$var$coord
+
+print(loadings_matrix)
+
+```
+
+```{r}
+#Table
+
+# Eigenvalues
+eigenvalues <- res_pca$eig[1:2, 1]
+
+# Variance percent
+variance_percent <- res_pca$eig[1:2, 2]
+
+# Cumulative variance percent
+cumulative_variance_percent <- res_pca$eig[1:2, 3]
+
+# Create the table
+df <- data.frame(Eigenvalue = eigenvalues,
+                 Variance_Percent = variance_percent,
+                 Cumulative_Variance_Percent = cumulative_variance_percent)
+
+rownames(df) <- c("Dim1", "Dim2")
+
+print(df)
+
+```
+
+```{r}
+#Table
+
+library(gt)
+
+# Create data frame
+df <- data.frame(Eigenvalue = eigenvalues,
+                 Variance_Percent = variance_percent,
+                 Cumulative_Variance_Percent = cumulative_variance_percent)
+rownames(df) <- c("Dim1", "Dim2")
+
+# Create the table using gt
+gt(df) %>%
+  tab_header(
+    title = "",
+    subtitle = "Eigenvalues and percentage of variance for each dimension of PCA"
+  ) %>%
+  cols_label(
+    Eigenvalue = "Eigenvalue",
+    Variance_Percent = "Variance (%)",
+    Cumulative_Variance_Percent = "Cumulative Variance (%)"
+  ) %>%
+  fmt_number(
+    columns = vars(Eigenvalue, Variance_Percent, Cumulative_Variance_Percent),
+    decimals = 2
+  )
+
+```
+
+feature frequency: d1 1 ta 1 tp1 0)
+
+```{r}
+library(dplyr)
+library(FactoMineR)
+library(tibble)
+
+# Define estimated area and feature frequency data
+area_data <- tibble(
+  area = c("d1", "ta", "tp1"),
+  feature_freq = c(1, 1, 0)
+)
+
+# Step 1: Number of artifacts by area
+artifact_counts <- tl_final_area %>%
+  group_by(area) %>%
+  tally()
+
+# Step 2: Calculate each variable
+
+# Retouch Frequency 
+retouch_freq <- tl_final_area %>%
+  group_by(area) %>%
+  summarise(retouchFreq = mean(!is.na(retouch)), .groups = 'drop')
+
+# Blank Frequency
+blank_freq <- tl_final_area %>%
+  group_by(area) %>%
+  summarise(blankFreq = mean(!is.na(blank)), .groups = 'drop')
+
+# Core Frequency
+core_freq <- tl_final_area %>%
+  group_by(area) %>%
+  summarise(coreFreq = mean(!is.na(`core-typology`)),.groups = 'drop')
+
+# Lithic density (Artifact counts / estimated area)
+lithic_density <- artifact_counts %>%
+  left_join(area_data, by = "area") %>%
+  mutate(lithic_density = n / estimated_area)
+
+# Tool Diversity using Menhinick's Index
+tool_div <- tl_final_area %>%
+  group_by(area) %>%
+  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(n()), .groups = 'drop')
+
+# Step 3: Combining all variables
+
+pca_final_table_area <- left_join(dimensions, retouch_freq, by = "area") %>%
+  left_join(blank_freq, by = "area") %>%
+  left_join(core_freq, by = "area") %>%
+  left_join(lithic_density, by = "area") %>%
+  left_join(tool_div, by = "area") %>%
+  select(area, new_dimensions, feature_freq, retouchFreq, blankFreq, coreFreq, lithic_density, toolDiv)
+
+```
+
+```{r}
+pca_final_table_area <- pca_final_table_area %>%
+    select(
+        area, 
+        estimated_area = estimated_area.x, 
+        feature_freq = feature_freq.x, 
+        retouchFreq, 
+        blankFreq, 
+        coreFreq, 
+        n, 
+        lithic_density, 
+        toolDiv
+    )
+
+```
+
+```{r}
+# Step 4: Perform PCA
+res_pca_area <- PCA(pca_final_table_area %>% select(-area, -n), graph = FALSE)
+
+# Show the summary of PCA
+print(res_pca_area)
+
+print(res_pca_area$eig)
+print(res_pca_area$var$coord)
+```
+
+```{r}
+# TABLE: Eigenvalues and percentage of variance for each dimension of PCA 
+library(kableExtra)
+
+# Define the values manually (you would get these from your PCA results)
+# BM: I prefer not to write values in the code like this. I think
+# this is bad practice and leads to mistakes
+eigenvalues <- c(4.411234, 2.588766)
+prop_var_explained <- c(63.02, 36.98)
+cumulative_var_explained <- c(63.02, 100.00)
+
+# Create the data frame
+results_df <- data.frame(
+  Dimension = c("Dim. 1", "Dim. 2"),  # Only keep the dimensions you want to display
+  Eigenvalue = eigenvalues,
+  `Variance / percent` = prop_var_explained,
+  `Cumulative / variance percent` = cumulative_var_explained
+)
+
+# Generate the table
+kable(results_df,
+      caption = "Table 3. Eigenvalues and percentage of variance for each dimension of PCA",
+      align = 'c',
+      col.names = c("Dimension", "Eigenvalue", "Variance (%)", "Cumulative Variance (%)")) %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 11)
+
+```
+
+```{r}
+#plotting Dim-1 and Dim-2
+library(FactoMineR)
+library(factoextra)
+
+# Visualize the contributions of variables to the first principal component
+fviz_contrib(res_pca_area, axes = 1, choice = "var")
+
+# Visualize the contributions of variables to the second principal component
+fviz_contrib(res_pca_area, axes = 2, choice = "var")
+
+```
+
+```{r}
+
+# having Dim1 and Dim2 plots next to each other
+
+library(gridExtra)
+
+# Create the first plot and assign it to p1
+p1 <- fviz_contrib(res_pca_area, axes = 1, choice = "var")
+
+# Create the second plot and assign it to p2
+p2 <- fviz_contrib(res_pca_area, axes = 2, choice = "var")
+
+# Use grid.arrange to place them next to each other
+grid.arrange(p1, p2, ncol = 2)
+
+```
+
+```{r}
+
+loadings <- res_pca_area$var$coord
+
+# 
+eigenvalues <- res_pca_area$eig
+
+#  Dim.1
+contrib_dim1 <- (loadings[,1]^2 / eigenvalues[1, 1]) * 100
+
+#  Dim.2
+contrib_dim2 <- (loadings[,2]^2 / eigenvalues[2, 1]) * 100
+
+contrib_df <- data.frame(
+  Variable = rownames(loadings),
+  Contrib_Dim1 = contrib_dim1,
+  Contrib_Dim2 = contrib_dim2
+)
+
+print(contrib_df)
+
+```
+
+```{r}
+# screeplot
+
+library(FactoMineR)
+library(factoextra)
+
+fviz_eig(res_pca_area)
+
+```
+
+```{r}
+# Correlation plots of variables for each of the tow relevant PCA dimensions
+
+fviz_pca_var(res_pca_area,
+
+             repel = TRUE) + 
+  theme(text = element_text(size = 10))
+
+```
+
+```{r}
+# Eigenvalues and percentage of variance
+eig.val <- res_pca_area$eig
+print(eig.val)
+
+# Variable coordinates (correlations with principal components)
+var.coord <- res_pca_area$var$coord
+print(var.coord)
+
+```
+
+```{r}
+#PCA biplot for Dimensions 1 and 2
+
+fviz_pca_biplot(res_pca_area, labelsize = 3, repel = TRUE)
+
+```
+
+```{r}
+# biplot 
+p <- fviz_pca_biplot(res_pca_area, label = "none", 
+                     repel = TRUE, labelsize = 4)
+
+p + geom_text(aes(x = res_pca_area$ind$coord[, 1], 
+                  y = res_pca_area$ind$coord[, 2], 
+                  label = pca_final_table_area$area), 
+             size = 3, nudge_x = 0.1, nudge_y = 0.1)
+
+```
+
+```{r}
+individual_coordinates = res_pca_area$ind$coord
+variable_coordinates = res_pca_area$var$coord
+
+```
+
+```{r}
+# For variables
+fviz_cos2(res_pca_area, choice = "var")
+
+# For individuals
+fviz_cos2(res_pca_area, choice = "ind")
+
+```
+
+DONE WITH PCA
+
+```{r}
+
+individual_coordinates <- res_pca_area$ind$coord
+
+individual_coordinates_df <- as.data.frame(individual_coordinates)
+individual_coordinates_df$area <- unique(area_data$area)
+
+print(individual_coordinates_df)
+
+```
+
+```{r}
+library(kableExtra)
+
+# Extract individual coordinates
+individual_coordinates <- res_pca_area$ind$coord
+
+# Convert to a data frame for better visibility and add the 'area' column back for context
+individual_coordinates_df <- as.data.frame(individual_coordinates)
+individual_coordinates_df$Area <- unique(area_data$area)  # Make sure the order matches
+
+# Use kable to create a beautiful table
+kable(individual_coordinates_df, 
+      col.names = c("Area", "Principal Component 1", "Principal Component 2"), 
+      caption = "PCA Coordinates for the Archaeological Areas excavation",
+      align = c('l', 'c', 'c'),
+      digits = 3) %>%
+  kable_styling("striped", full_width = F)
+
+```
+
+```{r}
+# Get the loadings (coefficients) of original variables for each principal component
+loadings_matrix <- res_pca$var$coord
+
+```
+
+```{r}
+# Eigenvalues
+eigenvalues <- res_pca$eig[1:2, 1]
+variance_percent <- res_pca$eig[1:2, 2]
+cumulative_variance_percent <- res_pca$eig[1:2, 3]
+
+# Create the table
+df <- data.frame(Eigenvalue = eigenvalues,
+                 Variance_Percent = variance_percent,
+                 Cumulative_Variance_Percent = cumulative_variance_percent)
+
+rownames(df) <- c("Dim1", "Dim2")
+
+# Print the table
+print(df)
+
+```
+
+```{r}
+library(gt)
+
+gt(df) %>%
+  tab_header(
+    title = "",
+    subtitle = "Eigenvalues and percentage of variance for each dimension of PCA"
+  ) %>%
+  cols_label(
+    Eigenvalue = "Eigenvalue",
+    Variance_Percent = "Variance (%)",
+    Cumulative_Variance_Percent = "Cumulative Variance (%)"
+  ) %>%
+  fmt_number(
+    columns = c("Eigenvalue", "Variance_Percent", "Cumulative_Variance_Percent"),
+    decimals = 2
+  )
+
+```
+
+```{r}
+
+artifact_counts <- tl_final %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  tally()
+
+# Print the resulting table
+print(artifact_counts)
+```
+
+```{r}
+
+blank_freq <- tl_final_area %>%
+  filter(area %in% c("ta", "d1", "tp1"), !is.na(blank)) %>%
+  group_by(area) %>%
+  summarise(blankFreq = sum(blank %in% c("bladelet", "blade", "flake")) / sum(!is.na(blank)), .groups = 'drop')
+
+# Calculating Core Frequency
+core_freq <- tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1"), !is.na(corefrag)) %>%
+  group_by(area) %>%
+  summarise(coreFreq = sum(corefrag %in% c("core fragment", "bullet", "flat", "pyramid", "cylinder/prismatic", "core rejuvenation piece", "shapeless", "core on flake", "bipolar")) / sum(!is.na(corefrag)), .groups = 'drop')
+
+# Calculating Tool Diversity using Menhinick's Index
+tool_div <- tl_final_area %>%
+  filter(area %in% c("ta", "d1", "tp1"), !is.na(tooltypegroup)) %>%
+  group_by(area) %>%
+  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(sum(!is.na(tooltypegroup))), .groups = 'drop')
+
+# Combining all the calculated frequencies into one tibble
+final_table <- blank_freq %>%
+  inner_join(core_freq, by = "area") %>%
+  inner_join(tool_div, by = "area")
+
+# Printing the final table
+print(final_table)
+
+```
+
+```{r}
+# Joining calculated frequencies
+final_table <- blank_freq %>%
+  inner_join(core_freq, by = "area") %>%
+  inner_join(tool_div, by = "area") %>%
+  inner_join(retouch_freq, by = "area")
+
+# Converting area to row names
+final_table_rownames <- final_table %>%
+  column_to_rownames(var = "area")
+
+```
+
+```{r}
+
+library(FactoMineR)
+library(factoextra)
+
+# Run PCA
+res_pca <- PCA(final_table_rownames, graph = FALSE)
+
+# Visualize the contributions of variables to the second principal component
+fviz_contrib(res_pca, axes = 2, choice = "var")
+
+```
+
+```{r}
+# Visualize the contributions of variables to the first principal component
+fviz_contrib(res_pca, axes = 1, choice = "var") 
+```
+
+```{r}
+# Visualize eigenvalues and explained variance
+fviz_screeplot(res_pca)
+```
+
+```{r}
+fviz_pca_biplot(res_pca, labelsize = 3, repel = TRUE)
+
+```
+
+```{r}
+# Visualize the variables on the principal components
+fviz_pca_var(res_pca)
+```
+
+```{r}
+# Get the loadings (coefficients) of original variables for each principal component
+loadings_matrix <- res_pca$var$coord
+
+# Print the loadings matrix
+print(loadings_matrix)
+
+```
+
+```{r}
+# Eigenvalues
+eigenvalues <- res_pca$eig[1:2, 1]
+variance_percent <- res_pca$eig[1:2, 2]
+cumulative_variance_percent <- res_pca$eig[1:2, 3]
+
+# Create the table
+df <- data.frame(Eigenvalue = eigenvalues,
+                 Variance_Percent = variance_percent,
+                 Cumulative_Variance_Percent = cumulative_variance_percent)
+
+rownames(df) <- c("Dim1", "Dim2")
+
+# Print the table
+print(df)
+
+```
+
+```{r}
+library(gt)
+
+# Create the table 
+gt(df) %>%
+  tab_header(
+    title = "",
+    subtitle = "Eigenvalues and percentage of variance for each dimension of PCA"
+  ) %>%
+  cols_label(
+    Eigenvalue = "Eigenvalue",
+    Variance_Percent = "Variance (%)",
+    Cumulative_Variance_Percent = "Cumulative Variance (%)"
+  ) %>%
+  fmt_number(
+    columns = c("Eigenvalue", "Variance_Percent", "Cumulative_Variance_Percent"),
+    decimals = 2
+  )
+
+```
+
+```{r}
+# Count the number of artifacts by area
+artifact_counts <- tl_final %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  tally()
+
+# Print the resulting table
+print(artifact_counts)
+
+```
+
+```{r}
+
+volumes <- data.frame(
+  area = c("ta", "tp1", "d1"),
+  volume = c(52, 5.76, 2.1375)
+)
+
+# Count the number of artifacts by area
+artifact_counts <- tl_final %>%
+  filter(area %in% c("ta", "tp1", "d1")) %>%
+  group_by(area) %>%
+  tally()
+
+# Join artifact_counts with volumes by area
+lithic_density_df <- left_join(artifact_counts, volumes, by = "area")
+
+# Compute lithic density for each area
+lithic_density_df <- lithic_density_df %>%
+  mutate(lithic_density = n / volume)
+
+# Print the resulting table
+print(lithic_density_df[, c("area", "lithic_density")])
+
+```
+
+```{r}
+# Counting the number of cores by area
+core_counts <- tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1") & !is.na(`core and core fragments`)) %>%
+  group_by(area) %>%
+  tally()
+
+print(core_counts)
+
+```

### 2023-09-19 paper/paper_tulai.qmd
PCA-Color points

@@
 
+summary_table_area_filtered <- summary_table_area %>%
+  filter(area %in% c("ta", "d1", "tp1"))
+
+library(kableExtra)
+
+kable(summary_table_area_filtered, 
+      caption = "Table: Summary by Area for 'ta', 'd1', and 'tp1'", 
+      align = 'c') %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
+                full_width = F, 
+                font_size = 12) %>%
+  row_spec(0, bold = TRUE, font_size = 14)
+
 ```
@@
 ```{r}
+library(ggpmisc)
+
 # Create the plot: WABI_Level_logscale
@@
 
-#PCA_level: for PCA Bicho and Cascalheira considered these variables: estimated area (as we calculated based on levels, I consider the thickness; is that OK?), core frequency, blank frequency, chip frequency: artefacts smaller than 1 cm (as we work on a neolithic asssemblage, we have so many small artefacts; so, I do not consider it), feature frequency (we have some stone alignments, but we do not know the exact level, we know the area of those features), retouch frequency, tool diversity (diversity of tool types within each assemblage, calculated using Menhinick’s index);
+#PCA_level: for PCA Bicho and Cascalheira considered these variables: estimated area (as we calculated based on levels, should I consider the thickness?), core frequency, blank frequency, chip frequency: artefacts smaller than 1 cm (as we work on a neolithic asssemblage, we have so many small artefacts; so, I do not consider it), feature frequency (we have some stone alignments, but we do not know the exact level, we know the area of those features), retouch frequency, tool diversity (diversity of tool types within each assemblage, calculated using Menhinick’s index);
   ##core frequency
@@
                    top = 10)
+p5 <- fviz_contrib(res.pca, 
+                   choice = "var", 
+                   axes = 5, 
+                   top = 10)
 
@@
 library(gridExtra)
-grid.arrange(p1, p2, p3, p4, ncol = 2)
+grid.arrange(p1, p2, p3, p4, p5, ncol = 2)
 
@@
 avg_contrib_dim4 <- mean(contributions[,4], na.rm = TRUE)
+avg_contrib_dim5 <- mean(contributions[,5], na.rm = TRUE)
 
@@
 high_contrib_dim4 <- contributions[contributions[,4] > avg_contrib_dim4, 4]
+high_contrib_dim5 <- contributions[contributions[,5] > avg_contrib_dim4, 5]
 
@@
 sorted_high_contrib_dim4 <- sort(high_contrib_dim4, decreasing = TRUE)
+sorted_high_contrib_dim5 <- sort(high_contrib_dim5, decreasing = TRUE)
 
@@
 print(sorted_high_contrib_dim4)
+
+print("Significant contributions to Dim5:")
+print(sorted_high_contrib_dim5)
 
@@
                 )
+
 rownames(res.pca$ind$coord) <- sum_depth_sheet$label
+
+```
+
+```{r}
+library(ggrepel)
+
+# Generate the base biplot
+biplot_base <- fviz_pca_biplot(res.pca,
+                               axes = c(1, 2),
+                               labelsize = 2
+                              )
+
+# Modify row names of coordinates
+rownames(res.pca$ind$coord) <- sum_depth_sheet$label
+
+# Add geom_text_repel() to the base plot
+biplot_modified <- biplot_base +
+                   geom_text_repel(data = as.data.frame(res.pca$ind$coord),
+                                   aes(x = Dim.1, 
+                                       y = Dim.2, 
+                                       label = rownames(res.pca$ind$coord)
+                                       ),
+                                   box.padding = 0.5,
+                                   point.padding = 0.5,
+                                   size = 2
+                                  )
+
+# Display the plot
+biplot_modified
+
+```
+
+```{r}
+# Base biplot without points
+biplot_base <- fviz_pca_biplot(res.pca,
+                               axes = c(1, 2),
+                               geom = "none",  
+                               labelsize = 2
+                               )
+
+# Add points and text labels manually
+biplot_modified <- biplot_base +
+  geom_point(data = coord_df, aes(x = Dim.1, y = Dim.2, color = group), size = 1) +
+  geom_text_repel(data = coord_df,
+                  aes(x = Dim.1,
+                      y = Dim.2,
+                      label = label),
+                  box.padding = 0.5,
+                  point.padding = 0.5,
+                  size = 2)
+
+# Display the plot
+biplot_modified
 

### 2023-09-19 paper/paper_tulai.qmd
move chunks to make summary_table_area before we use it
exploring PCA

@@
 
-```{r}
-library(dplyr)
-
-# Filter the data to include only the areas 'ta', 'd1', and 'tp1'
-tl_final_area <- tl_final %>%
-  filter(area %in% c("ta", "d1", "tp1"))
-
-summary_table_area_filtered <- summary_table_area %>%
-  filter(area %in% c("ta", "d1", "tp1"))
-
-library(kableExtra)
-
-kable(summary_table_area_filtered, 
-      caption = "Table: Summary by Area for 'ta', 'd1', and 'tp1'", 
-      align = 'c') %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
-                full_width = F, 
-                font_size = 12) %>%
-  row_spec(0, bold = TRUE, font_size = 14)
-
-```
-
 Summary tables: 1. AREA
@@
       caption = "Table: Summary by Area", 
+      align = 'c') %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
+                full_width = F, 
+                font_size = 12) %>%
+  row_spec(0, bold = TRUE, font_size = 14)
+
+```
+
+```{r}
+library(dplyr)
+
+# Filter the data to include only the areas 'ta', 'd1', and 'tp1'
+tl_final_area <- tl_final %>%
+  filter(area %in% c("ta", "d1", "tp1"))
+
+summary_table_area_filtered <- summary_table_area %>%
+  filter(area %in% c("ta", "d1", "tp1"))
+
+library(kableExtra)
+
+kable(summary_table_area_filtered, 
+      caption = "Table: Summary by Area for 'ta', 'd1', and 'tp1'", 
       align = 'c') %>%
@@
 ```{r}
-
-pca_data <- sum_depth_sheet %>% 
-  select(lithic_dens, core_freq, blank_freq, retouch_freq, tool_diversity) %>%
-  na.omit()  # Remove rows with NA values
-
-```
-
-```{r}
 # performin PCA
@@
 
-pca_data <- sum_depth_sheet[, c('lithic_dens', 'core_freq', 'blank_freq', 'retouch_freq', 'tool_diversity')]
+sum_depth_sheet_subset <- 
+ sum_depth_sheet %>%
+  drop_na(tool_diversity) #%>% 
+  #filter(!label %in% c("tp1_c_3_a",
+ #                      "d1_a_2_b",
+  #                     "d1_c_3_b"))
+
+pca_data <- 
+sum_depth_sheet_subset %>% 
+  select(lithic_dens, 
+         core_freq, 
+         blank_freq, 
+         retouch_freq, 
+         tool_diversity)
 
@@
 #Plot: Contribution of variables for each of the four relevant PCA dimensions
-library(factoextra)
-library(ggplot2)
 
 #save contribution plots into one objects
-p1 <- fviz_contrib(res.pca, 
+
+contribution_plots <- 
+map(1:5, ~fviz_contrib(res.pca, 
                    choice = "var", 
-                   axes = 1, 
-                   top = 10)
-p2 <- fviz_contrib(res.pca, 
-                   choice = "var",
-                   axes = 2, 
-                   top = 10)
-p3 <- fviz_contrib(res.pca, 
-                   choice = "var", 
-                   axes = 3, 
-                   top = 10)
-p4 <- fviz_contrib(res.pca, 
-                   choice = "var", 
-                   axes = 4, 
-                   top = 10)
-p5 <- fviz_contrib(res.pca, 
-                   choice = "var", 
-                   axes = 5, 
-                   top = 10)
+                   axes = .x, 
+                   top = 10))
 
 #combine individual plots into one plot 
-library(gridExtra)
-grid.arrange(p1, p2, p3, p4, p5, ncol = 2)
+library(cowplot)
+plot_grid(plotlist = contribution_plots,
+                   ncol = 2)
 
@@
 print(sorted_high_contrib_dim5)
-
 ```
@@
 # screeplot; inspect distribution of PCs
-library(factoextra)  
 fviz_screeplot(res.pca)
@@
 # Inspect eigenvalues
-print(res.pca$eig)
-
+res.pca$eig
 ```
@@
 # Create a new column named "Dimension"
-eigen_df$Dimension <- paste("Dim", seq_len(nrow(eigen_df)), sep = "")
+eigen_df$Dimension <- 
+  paste("Dim", seq_len(nrow(eigen_df)), sep = "")
 
@@
 # Display the table
-htmltools::browsable(kable_output)
-
-```
-
-```{r}
-# Visualize biplot:1
-fviz_pca_biplot(res.pca,
-                axes = c(1, 2),
-                labelsize = 2
-                )
-
-rownames(res.pca$ind$coord) <- sum_depth_sheet$label
+kable_output
 
@@
 library(ggrepel)
+
+rownames(res.pca$ind$coord) <- sum_depth_sheet_subset$label
 
@@
                                axes = c(1, 2),
-                               labelsize = 2
+                               labelsize = 2,
+                               label = "var",
+                               addEllipses = TRUE,
+                               col.ind = sum_depth_sheet_subset$area
                               )
@@
 # Modify row names of coordinates
-rownames(res.pca$ind$coord) <- sum_depth_sheet$label
+rownames(res.pca$ind$coord) <- sum_depth_sheet_subset$label
+
+pca_label_text <- 
+res.pca$ind$coord %>% 
+  as_tibble() %>% 
+  mutate(labels = sum_depth_sheet_subset$label,
+         area =   sum_depth_sheet_subset$area)
 
@@
 biplot_modified <- biplot_base +
-                   geom_text_repel(data = as.data.frame(res.pca$ind$coord),
+                   geom_text_repel(data = pca_label_text,
                                    aes(x = Dim.1, 
                                        y = Dim.2, 
-                                       label = rownames(res.pca$ind$coord)
-                                       ),
-                                   box.padding = 0.5,
-                                   point.padding = 0.5,
-                                   size = 2
+                                       label = labels,
+                                       colour = area),
+                                   segment.color = "grey80",
+                                   size = 2,
+                                   force = 10
                                   )
@@
 biplot_modified <- biplot_base +
-  geom_point(data = coord_df, aes(x = Dim.1, y = Dim.2, color = group), size = 1) +
+  geom_point(data = coord_df, 
+             aes(x = Dim.1, 
+                 y = Dim.2, 
+                 color = group), size = 1) +
   geom_text_repel(data = coord_df,
@@
 rownames(res.pca$ind$coord) <- sum_depth_sheet$label
-
-```
+```

### 2023-09-20 paper/paper_tulai.qmd
edit to simplify code up to the point where we read in the google sheet

@@
 library(ggbeeswarm)
+library(cowplot)
 ```
@@
 
+p1 <- 
 tl1 %>% 
@@
   geom_col()
-```
-
-```{r}
-# exploring Data
-
+
+p2 <- 
 tl1 %>% 
@@
   geom_col()
+
+plot_grid(p1, p2, ncol = 1)
 ```
@@
       Breakage %in% c("0", "1", NA_character_) ~ Blank, 
-
       TRUE ~ Breakage  
@@
   )
-
 ```
@@
 
-rev(sort(table(tl2$Blank)))
-
-# BM: still many typos and unusual categories in here, why is that? Can they be fixed?
+# rev(sort(table(tl2$Blank)))
+
+# BM: still many typos and unusual categories in here, why is that? Can they be fixed? This might increase our sample size.
 
@@
 na_rows <- which(is.na(tl2$Blank))
-print(na_rows)
 
 zero_rows <- which(tl2$Blank == "0")
-print(zero_rows)
-
 one_rows <- which(tl2$Blank == "1")
-print(one_rows)
-
-rev(sort(table(tl3$Blank2))) # BM: still a lot of typos on the categories here, can you fix them?
+
+# rev(sort(table(tl3$Blank2)))
 
@@
 ```{r}
-# Remove rows based on row names
-# tl3 <- tl3[!(rownames(tl3) %in% c("1030", "644", "543")), ]
-
-# BM: why are we removing these rows? A comment here would be good to explain. A simpler way, by referencing the row numbers directly:
 # SS: they are natural rocks, and don't want to count them as lithic
-
-tl3 <- tl3[-c(1030, 644, 543), ]
+# BM: better to remove them by their ID than by row number, since row number can easily change, but ID cannot
+
+tl3 <- tl3 %>% 
+  filter(!lithic_id %in% c("3154", "1741", "1680"))
 ```
@@
 tl4 <- tl4 %>%
-  # BM: the problem with this combination of mutate_all and ifelse that you use below is that they 
-  # coerce all columns to be the same data type, so you'll see that all the numeric
-  # columns are now character columns after you run these lines. So you need to
-  # coerce those number columns back to numeric type so you can plot and compute
-  # with them
-  mutate_all(~ ifelse(is.na(.x) | .x == "", NA, .x)) %>% 
-  # BM: what does this next line do?
-  #SS: I wanted to remove any asterisks, whitespace, or *, ?, or !. is that wrong?
-
-  mutate_all(~ str_replace(., "(?<=\\D)\\*(?=\\D)|\\s+|[*?!]+", "")) %>%
-  # BM: what is this SHINE variable? I'm not familiar with it
-
+  # BM: I updated this to use across
+  mutate(across(everything(), 
+                ~ifelse(is.na(.x)|.x == ""|.x == "-", NA, .x))) %>% 
+  #SS:  remove any asterisks, whitespace, or *, ?, or !. is that wrong?
+   mutate(across(everything(),
+                 ~str_replace(., "(?<=\\D)\\*(?=\\D)|\\s+|[*?!]+", ""))) %>%
   ##SS: we meant sickle shine, this is one of the criteria based on which Hole claimed the site is not related to agriculture (similar to ALi Kosh located in Dehluran)
@@
                    "L:semiparallel/R:scaled",
+                   "L:semiparallel/R: Scaled",
                    "Semiparallel", 
@@
 # BM: I think this is a better way to inspect:
-rev(sort(table(tl4$SHINE)))
-rev(sort(table(tl4$shine2)))
-
-```
-
-```{r}
-
-# Convert 'NA' and 'na' strings to actual NA
-tl4$`Core Typology` <- na_if(tl4$`Core Typology`, "NA")
-tl4$`Core Typology` <- na_if(tl4$`Core Typology`, "na")
+# rev(sort(table(tl4$SHINE)))
+# rev(sort(table(tl4$shine2)))
+
+```
+
+```{r}
 
@@
 
-# Convert 'NA' and 'na' in new columns to actual NA
-tl4$`core-typology` <- na_if(tl4$`core-typology`, "na")
-tl4$`core-technology` <- na_if(tl4$`core-technology`, "na")
-
 # Check the unique values again
-# BM: still many typos, are they important?
-rev(sort(table((tl4$`core-typology`))))
-rev(sort(table((tl4$`core-technology`))))
+# rev(sort(table((tl4$`core-typology`))))
+# rev(sort(table((tl4$`core-technology`))))
 
@@
 # BM: still some typos in there
-rev(sort(table((tl4$`tool-typology`))))
-rev(sort(table((tl4$`core-fragment`))))
-```
-
-```{r}
-# Exploring data
-
-tl4 %>% 
-  # BM: we need this because of the mutate_all ifelse combo you have above
-  mutate(midpoint = parse_number(midpoint)) %>% 
-  group_by(midpoint) %>% 
-  tally() %>% 
-  drop_na(midpoint) %>% 
-ggplot() + 
-  aes(midpoint, n) +
-  geom_col()
-
+# rev(sort(table((tl4$`tool-typology`))))
+# rev(sort(table((tl4$`core-fragment`))))
 ```
@@
       `sickle shine` == 1 ~ "tool",
-
       TRUE ~ `typology-tool-core`      
@@
 # BM: take a look at tool typology
-rev(sort(table(tl_final$`tool-typology`)))
+# rev(sort(table(tl_final$`tool-typology`)))
 
@@
 # Convert each column to numeric
-for (col_name in columns_to_convert) {
-  # Replace non-numeric characters with NA
-  # BM: this is quite an unusual method for doing this, 
-  # where did you find it?
-  tl_final[[col_name]][tl_final[[col_name]] %in% c("", "NA", "N/A")] <- NA
-  # Convert to numeric
-  tl_final[[col_name]] <- as.numeric(tl_final[[col_name]])
-}
 
@@
 # I think it's easier to read and requires less typing
-# tl_final %>% 
-#  mutate(across(all_of(columns_to_convert), parse_number))
+tl_final <- 
+tl_final %>% 
+ mutate(across(all_of(columns_to_convert), 
+               parse_number))
 
 # check the structure to see if they have been converted
-str(tl_final)
+# str(tl_final)
 ```
@@
   rowwise() %>%
-  mutate(utilization = ifelse(is.na(`sickle shine`) | is.na(utilization), utilization,
-                              ifelse(`sickle shine` == 1 & utilization == 1, NA_real_, utilization))) %>%
-  ungroup()
-
-```
-
-```{r}
-#number of `sickle shine` in clumn `sickle shine`=1 and tooltypegroup are not match: 
-rev(sort(table(tl_final$tooltypegroup)))
-rev(sort(table(tl_final$`sickle shine`)))
-
-#troubleshoot:
-tl_final %>% filter(`sickle shine` == 1) %>% 
-  select(`tool-typology`, 
-         `sickle shine`)
-
-tl_final %>% filter(`tool-typology` == "sickle shine") %>% 
-  select(`tool-typology`, `sickle shine`)
-
-tl_final <- tl_final %>%
+  mutate(utilization = ifelse(is.na(`sickle shine`) | is.na(utilization),
+                              utilization,
+                              ifelse(`sickle shine` == 1 & utilization == 1, 
+                                     NA_real_, 
+                                     utilization))) %>%
+  ungroup() %>%
+  # make the number of `sickle shine` in column `sickle shine` match:
   mutate(
@@
 
-rev(sort(table(tl_final$`tool-typology`)))
 ```
@@
 # Checking the unique values of tooltypegroup
-rev(sort(table(tl_final$tooltypegroup)))
-
-```
-
-```{r}
-
-library(dplyr)
-library(stringr)
+# rev(sort(table(tl_final$tooltypegroup)))
+
+```
+
+```{r}
 
@@
   mutate(retouch = ifelse(str_detect(tooltypegroup, 
-                                     str_c(patterns, collapse = "|")), 
+                                     str_c(patterns, 
+                                           collapse = "|")), 
                           1, retouch))
@@
 # View the first few rows to confirm
-head(tl_final)
-rev(sort(table(tl_final$tooltypegroup)))
-rev(sort(table(tl_final$retouch)))
+# head(tl_final)
+# rev(sort(table(tl_final$tooltypegroup)))
+# rev(sort(table(tl_final$retouch)))
 ```
@@
 # take a look
-rev(sort(table(tl_final$core)))
-rev(sort(table(tl_final$corefrag)))
-rev(sort(table(tl_final$`core-fragment`)))
-rev(sort(table(tl_final$blank)))
-```
-
-Summary tables: 1. AREA
+# rev(sort(table(tl_final$core)))
+# rev(sort(table(tl_final$corefrag)))
+# rev(sort(table(tl_final$`core-fragment`)))
+# rev(sort(table(tl_final$blank)))
+```
+
+### Summary tables: AREA
 
@@
 
-# BM: looks like this is area and layer, is that right? I see c1, d1, d2, d3, d4 etc.
-# SS: those are different excavated areas, we mainly consider ta, tp1, and d1.
-knitr::kable(summary_table_area)
-
-```
-
-```{r}
-library(kableExtra)
-
-kable(summary_table_area, 
-      caption = "Table: Summary by Area", 
-      align = 'c') %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
-                full_width = F, 
-                font_size = 12) %>%
-  row_spec(0, bold = TRUE, font_size = 14)
-
-```
-
-```{r}
-library(dplyr)
+# SS: we mainly consider ta, tp1, and d1.
+knitr::kable(summary_table_area,
+             caption = "Summary of lithics in all areas")
+```
+
+```{r}
 
@@
 
-library(kableExtra)
-
-kable(summary_table_area_filtered, 
-      caption = "Table: Summary by Area for 'ta', 'd1', and 'tp1'", 
-      align = 'c') %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
-                full_width = F, 
-                font_size = 12) %>%
-  row_spec(0, bold = TRUE, font_size = 14)
-
-```
-
-2.LAYER
+knitr::kable(summary_table_area_filtered,
+             caption = "Summary of lithics in the main areas")
+
+```
+
+### Summary tables: LAYER
 
@@
 kable(summary_table_layer, 
-      caption = "Table: Summary by Layer", 
-      align = 'c') %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
-                full_width = F, 
-                font_size = 12) %>%
-  row_spec(0, bold = TRUE, font_size = 14)
-
-```
-
-AREA-LAYER:
+      caption = "Summary of lithics by Layer", 
+      align = 'c')
+
+```
+
+### Summary tables: AREA-LAYER:
 
@@
 kable(summary_table_al, 
-      caption = "Table: Summary by Al", 
-      align = 'c') %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
-                full_width = F, 
-                font_size = 12) %>%
-  row_spec(0, bold = TRUE, font_size = 14)
-
-```
-
-Techno-Typo:
+      caption = "Summary of lithics by Area and layer")
+
+```
+
+### Summary tables: Techno-Typology
 
@@
 
-rev(sort(table(tl_final$tooltypegroup)))
+# rev(sort(table(tl_final$tooltypegroup)))
 ```
@@
   summarise(
-
-  )
-
-```
-
-```{r}
-#generate a table
+  )
+```
+
+```{r}
+# generate a table
 
@@
   ),
-  caption = "Table: Summary ",
-  align = 'c'
-) %>%
-  kable_styling(
-    bootstrap_options = c("striped", 
-                          "hover", 
-                          "condensed"),
-    full_width = FALSE,
-    font_size = 12
-  ) %>%
-  row_spec(0, 
-           bold = TRUE, 
-           font_size = 14)
+  caption = "Summary of lithics by techno-typological classification")
 
@@
   col.names = c("Tools", "D1", "TA", "TP1", "Total"),
-  caption = "Table: Tool Typology in three different main areas.",
-  align = 'c'
-) %>%
-  kable_styling(
-    bootstrap_options = c("striped", "hover", "condensed"),
-    full_width = F,
-    font_size = 12
-  ) %>%
-  column_spec(1, extra_css = "text-align:left;") %>%
-  row_spec(0, bold = TRUE, font_size = 14)
+  caption = "Tool Typology in the three main areas")
 
@@
 # Generate the table
-library(kableExtra)
-
 kable(summary_table_depth_area_layer, 
-      caption = "Table: Summary by Depth, Area, and Layer", 
-      align = 'c') %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
-                full_width = F, 
-                font_size = 12) %>%
-  row_spec(0, bold = TRUE, font_size = 14)
+      caption = "Summary of lithics by Depth, Area, and Layer")
 ```
@@
 # Generate the table
-library(kableExtra)
 
 kable(summary_table_depth_area_layer, 
-      caption = "Table: Summary by Depth, Area, and Layer", 
-      align = 'c') %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
-                full_width = F, 
-                font_size = 12) %>%
-  row_spec(0, bold = TRUE, font_size = 14)
+      caption = "Summary of lithics by Depth, Area, and Layer")
 
@@
 # Generate the table
-library(kableExtra)
-
 kable(summary_table_depth_area_layer, 
-      caption = "Table: Summary by Depth, Area, and Layer", 
-      align = 'c') %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
-                full_width = F, 
-                font_size = 12) %>%
-  row_spec(0, bold = TRUE, font_size = 14)
-
-```
-
-```{r}
-
-#Here, I've imported the table 'Summary by Depth, Area, Layer' into Google Sheets for easier management.
+      caption = "Summary of lithics by Depth, Area, and Layer")
+
+```
+
+```{r}
+
+# import table with depths that we have subdivided
+depths_divided_tbl <- read_csv(here("data/Lookup Table - Sheet1.csv"))
+
+# join with lithic data
+tl_final %>% 
+  left_join(depths_divided_tbl) %>% View
+
+```
+
+```{r}
+#| eval: false
+
+# Here, I've imported the table 'Summary by Depth, Area, Layer' into Google Sheets for easier management.
 #I thought this format would simplify the process of those depths needed to be splitting up.

### 2023-09-20 paper/paper_tulai.qmd
stopping work to wait for updates from SS

@@
 library(cowplot)
+library(knitr)
 ```
@@
 
+summary_table_depth_area_layer_to_join <- 
+  summary_table_depth_area_layer
+
 # Generate the table
@@
 # Specify the exact columns want to keep
-desired_columns <- c("TA", "depth", "area", "layer", "upper", "lower", 
-                     "Cores (n)", "Core rejuvenations pieces (n)", 
-                     "Blade (n)", "Flake (n)", "Bladelet (n)", 
-                     "Retouched Tools (n)", "Utilized (n)", "Sickle Shine (n)",
-                     "tool_type")  # Only keep this 'tool_type'
+desired_columns <-
+  c(
+    "TA",
+    "depth",
+    "area",
+    "layer",
+    "upper",
+    "lower",
+    "Cores (n)",
+    "Core rejuvenations pieces (n)",
+    "Blade (n)",
+    "Flake (n)",
+    "Bladelet (n)",
+    "Retouched Tools (n)",
+    "Utilized (n)",
+    "Sickle Shine (n)",
+    "tool_type"
+  )  # Only keep this 'tool_type'
 
@@
       caption = "Summary of lithics by Depth, Area, and Layer")
-
 ```
@@
 # import table with depths that we have subdivided
-depths_divided_tbl <- read_csv(here("data/Lookup Table - Sheet1.csv"))
-
-# join with lithic data
-tl_final %>% 
-  left_join(depths_divided_tbl) %>% View
+depths_divided_tbl <- 
+  read_csv(here("data/Lookup Table - Sheet1.csv")) %>% 
+  mutate(division_id = rep(c("a", "b"),
+                           nrow(.)/2))
+
+# google sheet has 42 rows, and cols for 
+# area, layer, level, depth. What is the equivalent above? tp1_c_3_a
+
+# This 'depths_divided_tbl' needs more information, for example, some depth ranges
+
+# join with lithic data that 
+summary_table_depth_area_layer_to_join %>% 
+  left_join(depths_divided_tbl) %>% 
+  mutate(label = paste(area, 
+                       layer, 
+                       # level, # need this
+                       division_id,
+                       sep = "_")) %>% View
 
@@
 
-library(tidyverse)
 library(googlesheets4)

### 2023-09-21 paper/paper_tulai.qmd
added grouped_level column. edited depths and chronological units.

@@
   mutate(midpoint = mean(c(upper, lower))) %>%
-# create 'layer' column
-  mutate(layer = case_when(
+# create 'chronological unit' column
+  mutate(chronological_unit = case_when(
       area == "tp1" & depth == "0-10 cm" ~ "c",
@@
       area == "tp1" & depth == "50-100 cm" ~ "c",
-      area == "tp1" & depth == "50-60 cm" ~ "c",
-      area == "tp1" & depth == "50-70 cm" ~ "c",
-      area == "tp1" & depth == "80-90 cm" ~ "c",
-      area == "tp1" & depth == "90-100 cm" ~ "c",
-      area == "tp1" & depth == "100-110 cm" ~ "c",
-      area == "tp1" & depth == "110-120 cm" ~ "c",
-      area == "tp1" & depth == "120-130 cm" ~ "c",
-      area == "tp1" & depth == "120-140 cm" ~ "c",
+      area == "tp1" & depth == "50-60 cm" ~ "b",
+      area == "tp1" & depth == "50-70 cm" ~ "b",
+      area == "tp1" & depth == "80-90 cm" ~ "b",
+      area == "tp1" & depth == "90-100 cm" ~ "b",
+      area == "tp1" & depth == "100-110 cm" ~ "b",
+      area == "tp1" & depth == "110-120 cm" ~ "b",
+      area == "tp1" & depth == "120-130 cm" ~ "b",
+      area == "tp1" & depth == "120-140 cm" ~ "b",
       area == "d1" & depth == "0-20 cm" ~ "c",
@@
       TRUE ~ NA_character_
-  ))
+  )) %>%
+
+  mutate(
+ grouped_level = case_when(
+    area == "tp1" & depth == "0-10 cm" ~ "3",
+      area == "tp1" & depth == "0-20 cm" ~ "3",
+      area == "tp1" & depth == "10-30 cm" ~ "3",
+      area == "tp1" & depth == "10-90 cm" ~ "3",
+      area == "tp1" & depth == "20-30 cm" ~ "3",
+      area == "tp1" & depth == "30-40 cm" ~ "3",
+      area == "tp1" & depth == "40-50 cm" ~ "3",
+      area == "tp1" & depth == "50-100 cm" ~ "3",
+      area == "tp1" & depth == "50-60 cm" ~ "2",
+      area == "tp1" & depth == "50-70 cm" ~ "2",
+      area == "tp1" & depth == "80-90 cm" ~ "2",
+      area == "tp1" & depth == "90-100 cm" ~ "2",
+      area == "tp1" & depth == "100-110 cm" ~ "1",
+      area == "tp1" & depth == "110-120 cm" ~ "1",
+      area == "tp1" & depth == "120-130 cm" ~ "1",
+      area == "tp1" & depth == "120-140 cm" ~ "1",
+      area == "d1" & depth == "0-20 cm" ~ "3",
+      area == "d1" & depth == "20-30 cm" ~ "3",
+      area == "d1" & depth == "40-60 cm" ~ "2",
+      area == "d1" & depth == "50-60 cm" ~ "2",
+      area == "d1" & depth == "60-70 cm" ~ "1",
+      area == "d1" & depth == "70-80 cm" ~ "1",
+      area == "ta" & depth == "0-30 cm" ~ "6",
+      area == "ta" & depth == "30-40 cm" ~ "6",
+      area == "ta" & depth == "40-50 cm" ~ "5",
+      area == "ta" & depth == "50-60 cm" ~ "5",
+      area == "ta" & depth == "60-70 cm" ~ "5",
+      area == "ta" & depth == "70-120 cm" ~ "4",
+      area == "ta" & depth == "120-150 cm" ~ "4",
+      area == "ta" & depth == "150-180 cm" ~ "3",
+      area == "ta" & depth == "180-190 cm" ~ "3",
+      area == "ta" & depth == "190-200 cm" ~ "2",
+      area == "ta" & depth == "200-210 cm" ~ "2",
+      area == "ta" & depth == "210-220 cm" ~ "2",
+      area == "ta" & depth == "220-230 cm" ~ "2",
+      area == "ta" & depth == "230-240 cm" ~ "2",
+      area == "ta" & depth == "230-250 cm" ~ "1",
+      area == "ta" & depth == "250-260 cm" ~ "1",
+      area == "ta" & depth == "260-270 cm" ~ "1",
+
+TRUE ~ NA_character_
+  )
+)
 
@@
 tl1 %>% 
-  group_by(layer) %>% 
+  group_by(chronological_unit) %>% 
   tally() %>% 
-  drop_na(layer) %>% 
+  drop_na(chronological_unit) %>% 
 ggplot() + 
-  aes(layer, n) +
+  aes(chronological_unit, n) +
   geom_col()
@@
 tl3 <- tl3 %>% 
-  filter(!lithic_id %in% c("3154", "1741", "1680"))
+  filter(!lithic_id %in% c("3154", "1741", "1680", "طبیعی"))
 ```
@@
     cortex =               `Cortex (%)`)  %>%
-  mutate(al = paste0(toupper(area), "_", layer)) %>%
-  relocate(al, .after = layer)   %>%
+  mutate(area_unit = paste0(toupper(area), "_", chronological_unit)) %>%
+  relocate(area_unit, .after = chronological_unit)   %>%
+  relocate(grouped_level, .after = area_unit)   %>%
   select(lithic_id, 
@@
          area, 
-         layer,
-         al,
+         chronological_unit,
+         area_unit,
+         grouped_level,
          cortex,
@@
 
-### Summary tables: LAYER
+### Summary tables: chronological unit
 
@@
 # Create a summary table
-summary_table_layer <- tl_final %>%
-  group_by(layer) %>% 
+summary_table_unit <- tl_final %>%
+  group_by(chronological_unit) %>% 
   summarise(
@@
 # Generate the table
-kable(summary_table_layer, 
-      caption = "Summary of lithics by Layer", 
+kable(summary_table_unit, 
+      caption = "Summary of lithics by Chronological Units", 
       align = 'c')
@@
 
-### Summary tables: AREA-LAYER:
-
-```{r}
-# Create a summary table based on the 'al' column
-summary_table_al <- tl_final %>%
-  group_by(al) %>%  
+### Summary tables: AREA-UNIT:
+
+```{r}
+# Create a summary table based on the 'area_unit' column
+summary_table_au <- tl_final %>%
+  group_by(area_unit) %>%  
   summarise(
@@
 kable(summary_table_al, 
-      caption = "Summary of lithics by Area and layer")
+      caption = "Summary of lithics by Area and chronological unit")
 
@@
     shine_tool_count =         sum(.$shine_tool_count, na.rm = TRUE),
-    total =                    sum(.$total, na.rm = TRUE)  # Explicitly setting the total
+
   )
@@
 ```{r}
-# Prepare data for Summary table-area_layer(al)
-summary_table_al <- tl_final %>%
+# Prepare data for Summary table-area_unit(au)
+summary_table_au <- tl_final %>%
   filter(area %in% c("ta", "d1", "tp1")) %>% 
-  group_by(al) %>% 
+  group_by(area_unit) %>% 
   summarise(
@@
   add_row(
-    al = "Total",
+    area_unit = "Total",
     blade_count =              sum(.$blade_count, na.rm = TRUE),
@@
 summary_table_al <- summary_table_al %>%
-  filter(!(al %in% c("D1_NA", "TP1_NA")))
+  filter(!(area_unit %in% c("D1_NA", "TP1_NA")))
 
@@
   col.names = c(
-    "Area-Layer",
+    "Area-unit",
     "Blade",
@@
 ```{r}
-# Create the summary table based on layer
-summary_table1_layer <- tl_final %>%
+# Create the summary table based on chronological_unit
+summary_table1_unit <- tl_final %>%
   filter(area %in% c("ta", "d1", "tp1")) %>%
-  group_by(layer) %>%
+  group_by(chronological_unit) %>%
   summarise(
@@
 ```{r}
-# Create a summary table based on the 'depth', 'area', and 'layer' columns
+# Create a summary table based on the 'depth', 'area', and 'chronological unit' columns
 tl_final <- tl_final %>%
@@
 
-summary_table_depth_area_layer <- tl_final %>%
-  group_by(TA, depth, area, layer, upper, lower) %>%  
+summary_table_depth_area_unit <- tl_final %>%
+  group_by(TA, depth, area, chronological_unit, grouped_level, upper, lower) %>%  
   summarise(
@@
 
-summary_table_depth_area_layer_to_join <- 
-  summary_table_depth_area_layer
+summary_table_depth_area_unit_to_join <- 
+  summary_table_depth_area_unit
 
 # Generate the table
-kable(summary_table_depth_area_layer, 
-      caption = "Summary of lithics by Depth, Area, and Layer")
+kable(summary_table_depth_area_unit, 
+      caption = "Summary of lithics by Depth, Area, and Chronological Unit")
 ```
@@
 # Create a full summary table by joining it back to the original summary table based on 'depth' and 'area'
-summary_table_depth_area_layer <- summary_table_depth_area_layer %>%
+summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
   left_join(summary_table_depth_tool_type, by = c("area", "depth"))
@@
 # Remove the existing 'tool_type' if it exists
-if ("tool_type" %in% names(summary_table_depth_area_layer)) {
-  summary_table_depth_area_layer <- summary_table_depth_area_layer %>%
+if ("tool_type" %in% names(summary_table_depth_area_unit)) {
+  summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
     select(-tool_type)
@@
 # Join again based on 'depth', this time being specific about what we want
-summary_table_depth_area_layer <- 
-  summary_table_depth_area_layer %>%
+summary_table_depth_area_unit <- 
+  summary_table_depth_area_unit %>%
   left_join(summary_table_depth_tool_type %>% select(depth, tool_type), 
@@
 
-kable(summary_table_depth_area_layer, 
-      caption = "Summary of lithics by Depth, Area, and Layer")
+kable(summary_table_depth_area_unit, 
+      caption = "Summary of lithics by Depth, Area, and chronological unit")
 
@@
     "area",
-    "layer",
+    "chronological_unit",
+    "grouped_level",
     "upper",
@@
     "tool_type"
-  )  # Only keep this 'tool_type'
+  )
 
 # Filter the tl_final dataframe
-summary_table_depth_area_layer <- summary_table_depth_area_layer %>%
+summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
   filter(area %in% c('ta', 'd1', 'tp1'))
@@
 # Generate the table
-kable(summary_table_depth_area_layer, 
-      caption = "Summary of lithics by Depth, Area, and Layer")
+kable(summary_table_depth_area_unit, 
+      caption = "Summary of lithics by Depth, Area, and chronological_unit")
 ```
@@
 # google sheet has 42 rows, and cols for 
-# area, layer, level, depth. What is the equivalent above? tp1_c_3_a
+# area, chronological_unit, level, depth. What is the equivalent above? tp1_c_3_a
 
@@
 # join with lithic data that 
-summary_table_depth_area_layer_to_join %>% 
+summary_table_depth_area_unit_to_join %>% 
   left_join(depths_divided_tbl) %>% 
   mutate(label = paste(area, 
-                       layer, 
+                       chronological_unit, 
                        # level, # need this
@@
 
-# Here, I've imported the table 'Summary by Depth, Area, Layer' into Google Sheets for easier management.
+# Here, I've imported the table 'Summary by Depth, Area, chronological_unit' into Google Sheets for easier management.
 #I thought this format would simplify the process of those depths needed to be splitting up.
@@
 sum_depth_sheet <- sum_depth_sheet %>%
-  mutate(label = paste(area, layer, level, sep = "_"))
+  mutate(label = paste(area, chronological_unit, grouped_level, sep = "_"))
 
@@
 
+```
+
+#try with summary_table_depth_area_unit
+```{r}
+# Create the 'label' column
+summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
+  mutate(label = paste(area, chronological_unit, grouped_level, sep = "_"))
+
+# Add a suffix only to duplicates
+summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
+  group_by(label) %>%
+  mutate(suffix = case_when(
+    n() == 1 ~ NA_character_,
+    TRUE ~ letters[row_number()]
+  )) %>%
+  ungroup()
+
+# Append the suffix to 'label'
+summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
+  mutate(label = if_else(is.na(suffix), label, paste0(label, "_", suffix))) %>%
+  select(-suffix)
+
+# Relocate 'label' column after 'level'
+summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
+  relocate(label, .after = grouped_level)
+
+------------------------------------------
+# working up to here. I need to add length, width, and thickness columns
+------------------------------------------
+
+# Separate `excavation area` into `length` and `width` and calculate `volume`
+summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
+  separate(`excavation area`, into = c("length", "width"), sep = "\\*") %>%
+  mutate(
+    length = as.numeric(length),
+    width = as.numeric(width),
+    `thickness` = `thickness` / 100  # Convert from cm to m
+  ) %>%
+  mutate(
+    volume = length * width * `thickness`  # Calculate volume in m^3
+  )
+
+# Calculate lithic volumetric density 
+summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
+  mutate(
+    `lithic_dens` = round(total_lithics / volume, 2)
+  )
+
+# Calculate retouch frequency 
+summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
+  mutate(
+    retouch_freq = round(retouch_tool / volume, 2)  
+  )
 ```

### 2023-09-22 paper/paper_tulai.qmd
Major edits to remove the google sheet and replace it with a local CSV. Recomputed the values that were in the google sheet using the raw lithic data. Removed many tables that are not mentioned in the manuscript.

@@
 title-block-published: "last updated"
-editor: visual
 ---
@@
 
-Cleaning Data:
+Import the data
 
@@
 tl <- read_csv(here("data/Tulai_Lithic_Assamblage.csv"))
+
+# this file is prepared from the google sheet here
+# https://docs.google.com/spreadsheets/d/1Q0QZESk412ZQLE24yPs6Rg-7Y9OMX63DHoXeIPNAIFM/edit#gid=0
+# and simplified using code in the scraps.R file
+excavation_data <- read_csv(here("data/excavation_data.csv"))
+```
+
+Clean the data
+
+```{r}
 
@@
 # Define the patterns to search for
-patterns <- c("denticulate-notch", "retouched piece", "scraper",
-              "Perforator", "serrated scraper", "backed pieces",
-              "truncated pieces", "burin", "geometric", "scraper-notch",
+patterns <- c("denticulate-notch", 
+              "retouched piece", 
+              "scraper",
+              "Perforator", 
+              "serrated scraper", 
+              "backed pieces",
+              "truncated pieces", 
+              "burin", 
+              "geometric", 
+              "scraper-notch",
               "microburin")
@@
 
-### Summary tables: AREA
+### Summary tables of the lithic assemblages
+
+#### Basic typological classes by excavation area
 
@@
 
-# SS: we mainly consider ta, tp1, and d1.
-knitr::kable(summary_table_area,
-             caption = "Summary of lithics in all areas")
-```
-
-```{r}
-
 # Filter the data to include only the areas 'ta', 'd1', and 'tp1'
@@
 
-summary_table_area_filtered <- summary_table_area %>%
+summary_table_area_filtered <- 
+  summary_table_area %>%
   filter(area %in% c("ta", "d1", "tp1"))
 
+# BM: this is table 1 in the manuscript
 knitr::kable(summary_table_area_filtered,
@@
 
-### Summary tables: chronological unit
-
-```{r}
-# Create a summary table
-summary_table_unit <- tl_final %>%
-  group_by(chronological_unit) %>% 
-  summarise(
-    `Cores (n)` = sum(!is.na(core)),  # Count non-NA core entries
-    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag)),  # Count non-NA corefrag entries
-    `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
-    `Flake (n)` = sum(blank == "flake", na.rm = TRUE),
-    `Bladelet (n)` = sum(blank == "bladelet", na.rm = TRUE),
-    `Retouched Tools (n)` = sum(retouch == 1, na.rm = TRUE),
-    `Utilized (n)` = sum(utilization == 1, na.rm = TRUE),
-    `Sickle Shine (n)` = sum(`sickle shine` == 1, na.rm = TRUE)
-  )
-
-```
-
-```{r}
-# Generate the table
-kable(summary_table_unit, 
-      caption = "Summary of lithics by Chronological Units", 
-      align = 'c')
-
-```
-
-### Summary tables: AREA-UNIT:
-
-```{r}
-# Create a summary table based on the 'area_unit' column
-summary_table_au <- tl_final %>%
-  group_by(area_unit) %>%  
-  summarise(
-    `Cores (n)` = sum(!is.na(core)),  
-    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag)),  
-    `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
-    `Flake (n)` = sum(blank == "flake", na.rm = TRUE),
-    `Bladelet (n)` = sum(blank == "bladelet", na.rm = TRUE),
-    `Retouched Tools (n)` = sum(retouch == 1, na.rm = TRUE),
-    `Utilized (n)` = sum(utilization == 1, na.rm = TRUE),
-    `Sickle Shine (n)` = sum(`sickle shine` == 1, na.rm = TRUE)
-  )
-
-```
-
-```{r}
-# Generate the table
-
-kable(summary_table_al, 
-      caption = "Summary of lithics by Area and chronological unit")
-
-```
-
-### Summary tables: Techno-Typology
-
-```{r}
-
-# Prepare data for Summary table-area
-summary_table_area <- tl_final %>%
-  filter(area %in% c("ta", "d1", "tp1")) %>%
-  group_by(area) %>%
-  summarise(
-    blade_count = sum(blank == "blade", na.rm = TRUE),
-    flake_count = sum(blank == "flake", na.rm = TRUE),
-    bladelet_count = sum(blank == "bladelet", na.rm = TRUE),
-    tool_count = sum(retouch == 1, na.rm = TRUE) + sum(utilization == 1, na.rm = TRUE),
-    core_count = sum(!is.na(`core-typology`), na.rm = TRUE),
-    core_rejuvenation_count = sum(!is.na(`corefrag`), na.rm = TRUE),
-    retouched_tool_count = sum(retouch == 1, na.rm = TRUE),
-    utilized_tool_count = sum(utilization == 1, na.rm = TRUE),
-    shine_tool_count = sum(`sickle shine` == 1, na.rm = TRUE),
-    total = n()
-  )
-
-summary_table_area <- summary_table_area %>%
-  add_row(
-    area = "Total",
-    blade_count =              sum(.$blade_count, na.rm = TRUE),
-    flake_count =              sum(.$flake_count, na.rm = TRUE),
-    bladelet_count =           sum(.$bladelet_count, na.rm = TRUE),
-    tool_count =               sum(.$tool_count, na.rm = TRUE),
-    core_count =               sum(.$core_count, na.rm = TRUE),
-    core_rejuvenation_count =  sum(.$core_rejuvenation_count, na.rm = TRUE),
-    retouched_tool_count =     sum(.$retouched_tool_count, na.rm = TRUE),
-    utilized_tool_count =      sum(.$utilized_tool_count, na.rm = TRUE),
-    shine_tool_count =         sum(.$shine_tool_count, na.rm = TRUE),
-
-  )
-
-# Adding a new row with total counts
-summary_table_area <- summary_table_area %>%
-  add_row(
-    area = "Total",
-    blade_count =              sum(.$blade_count, na.rm = TRUE),
-    flake_count =              sum(.$flake_count, na.rm = TRUE),
-    bladelet_count =           sum(.$bladelet_count, na.rm = TRUE),
-    tool_count =               sum(.$tool_count, na.rm = TRUE),
-    core_count =               sum(.$core_count, na.rm = TRUE),
-    core_rejuvenation_count =  sum(.$core_rejuvenation_count, na.rm = TRUE),
-    retouched_tool_count =     sum(.$retouched_tool_count, na.rm = TRUE),
-    utilized_tool_count =      sum(.$utilized_tool_count, na.rm = TRUE),
-    shine_tool_count =         sum(.$shine_tool_count, na.rm = TRUE),
-    total =                    sum(.$total, na.rm = TRUE)  # Explicitly setting the total
-  )
-
-```
-
-```{r}
-# Prepare data for Summary table-area_unit(au)
-summary_table_au <- tl_final %>%
-  filter(area %in% c("ta", "d1", "tp1")) %>% 
-  group_by(area_unit) %>% 
-  summarise(
-    blade_count = sum(blank == "blade", na.rm = TRUE),
-    flake_count = sum(blank == "flake", na.rm = TRUE),
-    bladelet_count = sum(blank == "bladelet", na.rm = TRUE),
-    tool_count = sum(retouch == 1, na.rm = TRUE) + sum(utilization == 1, na.rm = TRUE),
-    core_count = sum(!is.na(`core-typology`), na.rm = TRUE),
-    core_rejuvenation_count = sum(!is.na(`corefrag`), na.rm = TRUE),
-    retouched_tool_count = sum(retouch == 1, na.rm = TRUE),
-    utilized_tool_count = sum(utilization == 1, na.rm = TRUE),
-    shine_tool_count = sum(`sickle shine` == 1, na.rm = TRUE),
-    total = n()
-  )
-
-# Adding a new row with total counts
-summary_table_al <- summary_table_al %>% 
-  add_row(
-    area_unit = "Total",
-    blade_count =              sum(.$blade_count, na.rm = TRUE),
-    flake_count =              sum(.$flake_count, na.rm = TRUE),
-    bladelet_count =           sum(.$bladelet_count, na.rm = TRUE),
-    tool_count =               sum(.$tool_count, na.rm = TRUE),
-    core_count =               sum(.$core_count, na.rm = TRUE),
-    core_rejuvenation_count =  sum(.$core_rejuvenation_count, na.rm = TRUE),
-    retouched_tool_count =     sum(.$retouched_tool_count, na.rm = TRUE),
-    utilized_tool_count =      sum(.$utilized_tool_count, na.rm = TRUE),
-    shine_tool_count =         sum(.$shine_tool_count, na.rm = TRUE),
-    total =                    sum(.$total, na.rm = TRUE)  
-  )
-
-# rev(sort(table(tl_final$tooltypegroup)))
-```
-
-```{r}
-#explore tool type group
-
-summary_table_tooltype <- tl_final %>%
-  filter(tooltypegroup != 'utilized tool') %>%  # Exclude 'utilized tool'
-  group_by(tooltypegroup) %>% 
-  summarise(
-  )
-```
-
-```{r}
-# generate a table
-
-summary_table_al <- summary_table_al %>%
-  filter(!(area_unit %in% c("D1_NA", "TP1_NA")))
-
-kable(
-  summary_table_al,
-  col.names = c(
-    "Area-unit",
-    "Blade",
-    "Flake",
-    "Bladelet",
-    "Tools",
-    "Cores",
-    "Core Rejuvenations",
-    "Retouched Tools",
-    "Utilized Tools",
-    "Sickle Shine",
-    "Total"
-  ),
-  caption = "Summary of lithics by techno-typological classification")
-
-```
+#### Techno-typological classes by excavation area
 
@@
   arrange(area, tooltypegroup) %>%  # Arrange by area and tooltypegroup
-  spread(key = area, value = n, fill = 0)  # Spread the 'area' column into multiple columns
+  spread(key = area, 
+         value = n, 
+         fill = 0)  # Spread the 'area' column into multiple columns
 
@@
 
-```{r}
-# Create the summary table based on chronological_unit
-summary_table1_unit <- tl_final %>%
-  filter(area %in% c("ta", "d1", "tp1")) %>%
-  group_by(chronological_unit) %>%
+### Combine lithic data with excavation data to prepare for WABI
+
+```{r}
+# Create the 'label' column
+tl_final_depth_area_unit <- 
+  tl_final %>%
+  mutate(label1 = paste(area, 
+                       chronological_unit, 
+                       grouped_level, 
+                       sep = "_")) %>% 
+  mutate(label2 = paste(area, 
+                       chronological_unit, 
+                       grouped_level, 
+                       upper,
+                       lower,
+                       sep = "_")) %>% 
+  # we don't care about lithics with missing and incomplete 
+  # excavation provenance, so drop those
+ filter(!str_detect(label1, "NA")) %>% 
+  # so we can join this column ok 
+  mutate(grouped_level = parse_number(grouped_level))
+
+# BM: For the individual lithic data, what area/chrono-unit/grouped level 
+# combinations do we have? We only want to match those with the 
+# excavation data that we can calclate volume with.
+
+lithics_unique_labels <- 
+tl_final_depth_area_unit %>% 
+  pull(label2) %>% 
+  unique() %>% 
+  sort()
+
+# BM: prepare excavation data here
+excavation_data_volume <-
+  excavation_data %>% 
+  mutate(thickness_of_excavation_unit  = lower_depth - upper_depth) %>% 
+  mutate(excavation_volume = excavation_dimension_1 * 
+           excavation_dimension_2 * 
+           thickness_of_excavation_unit / 100)
+
+# BM: for the WABI plot, we have our primary analytical unit as
+# excavation_area / chronological_unit / depth which is our label2
+# variable, each row in the excavation_data_volume table is a 
+# unique combination of excavation_area / chronological_unit / depth
+
+# BM: Now we subset the excavation data so we only look at the 
+# data that we need for the lithics, ie. we drop excavation data 
+# that has no corresponding lithics
+
+tl_final_excavation_data_volume  <- 
+excavation_data_volume %>% 
+  filter(label2 %in% lithics_unique_labels) %>% 
+  # now we join the excavation data to the individual lithic records
+  right_join(tl_final_depth_area_unit)
+
+# get the counts of artefact types per ex-area/chrono-unit/depth
+
+tl_final_excavation_data_volume_artefact_counts <- 
+tl_final_excavation_data_volume %>% 
+  group_by(label2) %>% 
    summarise(
-    numb_retouched = sum(retouch == 1, na.rm = TRUE),
-    numb_cores = sum(!is.na(`core-typology`), na.rm = TRUE),
-    total_lithics = n(),
-    percent_retouch = round((numb_retouched / total_lithics) * 100, 2)
-  )
-
-```
-
-```{r}
-# Create a summary table based on the 'depth', 'area', and 'chronological unit' columns
-tl_final <- tl_final %>%
-  mutate(depth = gsub("cm", "", depth))
-
-summary_table_depth_area_unit <- tl_final %>%
-  group_by(TA, depth, area, chronological_unit, grouped_level, upper, lower) %>%  
-  summarise(
-    `Cores (n)` = sum(!is.na(core)),  
-    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag)),  
-    `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
-    `Flake (n)` = sum(blank == "flake", na.rm = TRUE),
-    `Bladelet (n)` = sum(blank == "bladelet", na.rm = TRUE),
-    `Retouched Tools (n)` = sum(retouch == 1, na.rm = TRUE),
-    `Utilized (n)` = sum(utilization == 1, na.rm = TRUE),
-    `Sickle Shine (n)` = sum(`sickle shine` == 1, na.rm = TRUE)
-  )
-
-summary_table_depth_area_unit_to_join <- 
-  summary_table_depth_area_unit
-
-# Generate the table
-kable(summary_table_depth_area_unit, 
-      caption = "Summary of lithics by Depth, Area, and Chronological Unit")
-```
-
-```{r}
-# Generate a summary table by grouping by 'depth' and 'area'
-summary_table_depth_tool_type <- tl_final %>%
-  group_by(area, depth) %>%
-  summarise(tool_type = n_distinct(tooltypegroup, na.rm = TRUE), .groups = 'drop')
-
-# Create a full summary table by joining it back to the original summary table based on 'depth' and 'area'
-summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
-  left_join(summary_table_depth_tool_type, by = c("area", "depth"))
-
-# Remove the existing 'tool_type' if it exists
-if ("tool_type" %in% names(summary_table_depth_area_unit)) {
-  summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
-    select(-tool_type)
-}
-
-# Join again based on 'depth', this time being specific about what we want
-summary_table_depth_area_unit <- 
-  summary_table_depth_area_unit %>%
-  left_join(summary_table_depth_tool_type %>% select(depth, tool_type), 
-            by = "depth")
-
-# Generate the table
-
-kable(summary_table_depth_area_unit, 
-      caption = "Summary of lithics by Depth, Area, and chronological unit")
-
-```
-
-```{r}
-# Specify the exact columns want to keep
-desired_columns <-
-  c(
-    "TA",
-    "depth",
-    "area",
-    "chronological_unit",
-    "grouped_level",
-    "upper",
-    "lower",
-    "Cores (n)",
-    "Core rejuvenations pieces (n)",
-    "Blade (n)",
-    "Flake (n)",
-    "Bladelet (n)",
-    "Retouched Tools (n)",
-    "Utilized (n)",
-    "Sickle Shine (n)",
-    "tool_type"
-  )
-
-# Filter the tl_final dataframe
-summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
-  filter(area %in% c('ta', 'd1', 'tp1'))
-
-# Generate the table
-kable(summary_table_depth_area_unit, 
-      caption = "Summary of lithics by Depth, Area, and chronological_unit")
-```
-
-```{r}
-
-# import table with depths that we have subdivided
-depths_divided_tbl <- 
-  read_csv(here("data/Lookup Table - Sheet1.csv")) %>% 
-  mutate(division_id = rep(c("a", "b"),
-                           nrow(.)/2))
-
-# google sheet has 42 rows, and cols for 
-# area, chronological_unit, level, depth. What is the equivalent above? tp1_c_3_a
-
-# This 'depths_divided_tbl' needs more information, for example, some depth ranges
-
-# join with lithic data that 
-summary_table_depth_area_unit_to_join %>% 
-  left_join(depths_divided_tbl) %>% 
-  mutate(label = paste(area, 
-                       chronological_unit, 
-                       # level, # need this
-                       division_id,
-                       sep = "_")) %>% View
-
-```
-
-```{r}
-#| eval: false
-
-# Here, I've imported the table 'Summary by Depth, Area, chronological_unit' into Google Sheets for easier management.
-#I thought this format would simplify the process of those depths needed to be splitting up.
-
-# BM: looks like you need this data starting from around here in the document
-
-library(googlesheets4)
-
-google_sheet_url <- 
-  "https://docs.google.com/spreadsheets/d/1Q0QZESk412ZQLE24yPs6Rg-7Y9OMX63DHoXeIPNAIFM/edit#gid=0"
-sum_depth_sheet <- read_sheet(google_sheet_url)
-
-# Create the 'label' column
-sum_depth_sheet <- sum_depth_sheet %>%
-  mutate(label = paste(area, chronological_unit, grouped_level, sep = "_"))
-
-# Add a suffix only to duplicates
-sum_depth_sheet <- sum_depth_sheet %>%
-  group_by(label) %>%
-  mutate(suffix = case_when(
-    n() == 1 ~ NA_character_,
-    TRUE ~ letters[row_number()]
-  )) %>%
-  ungroup()
-
-# Append the suffix to 'label'
-sum_depth_sheet <- sum_depth_sheet %>%
-  mutate(label = if_else(is.na(suffix), label, paste0(label, "_", suffix))) %>%
-  select(-suffix)
-
-# Relocate 'label' column after 'level'
-sum_depth_sheet <- sum_depth_sheet %>%
-  relocate(label, .after = level)
-
-# BM: from my perspective, what would be ideal here is to have the 
-# excavation area and volume calculations done here in R. I feel anxious about
-# the possibilities of typos when these values are calculated somewhere I can't 
-# see and pasted into the google sheet. I prefer for the workflow of calculations
-# to be as transparent and traceable as possible. From this point in the qmd I
-# think we can use the volume values from this google sheet. Another option is to 
-# export the google sheet as a CSV, and put that in the data folder in this project. 
-# Then we can use read_csv to import it.
-
-#SS: is that true?: (I have a weird number in row 5, cl Depth and couldn't find out what happened)
-
-#### previously I had a column'excavation area'; I seprated them in 2 columns in order to calculate the colume; but now when I wanted to run the code it said "!
-object 'excavation area' not found"; would it be OK to delete these lines?
-
-# Separate `excavation area` into `length` and `width` and calculate `volume`
-sum_depth_sheet <- sum_depth_sheet %>%
-  separate(`excavation area`, into = c("length", "width"), sep = "\\*") %>%
-  mutate(
-    length = as.numeric(length),
-    width = as.numeric(width),
-    `thickness` = `thickness` / 100  # Convert from cm to m
+    blade_count =              sum(blank == "blade", na.rm = TRUE),
+    flake_count =              sum(blank == "flake", na.rm = TRUE),
+    bladelet_count =           sum(blank == "bladelet", na.rm = TRUE),
+    tool_count =               sum(retouch == 1, 
+                                   na.rm = TRUE) + sum(utilization == 1, 
+                                                       na.rm = TRUE),
+    core_count =               sum(!is.na(`core-typology`), na.rm = TRUE),
+    core_rejuvenation_count =  sum(!is.na(`corefrag`), 
+                                   na.rm = TRUE),
+    retouched_tool_count =     sum(retouch == 1, 
+                                   na.rm = TRUE),
+    utilized_tool_count =      sum(utilization == 1, 
+                                   na.rm = TRUE),
+    shine_tool_count =         sum(`sickle shine` == 1, 
+                                   na.rm = TRUE),
+    tool_type_count = n_distinct(tooltypegroup),
+    excavation_volume = unique(excavation_volume)
+
   ) %>% 
-  mutate(
-    volume = length * width * `thickness`  # Calculate volume in m^3
-  )
-
-# Calculate lithic volumetric density 
-sum_depth_sheet <- sum_depth_sheet %>%
-  mutate(
-    `lithic_dens` = round(total_lithics / volume, 2)
-  )
-
-# Calculate retouch frequency 
-sum_depth_sheet <- sum_depth_sheet %>%
-  mutate(
-    retouch_freq = round(retouch_tool / volume, 2)  
-  )
-
-```
-
-#try with summary_table_depth_area_unit
-```{r}
-# Create the 'label' column
-summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
-  mutate(label = paste(area, chronological_unit, grouped_level, sep = "_"))
-
-# Add a suffix only to duplicates
-summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
-  group_by(label) %>%
-  mutate(suffix = case_when(
-    n() == 1 ~ NA_character_,
-    TRUE ~ letters[row_number()]
-  )) %>%
-  ungroup()
-
-# Append the suffix to 'label'
-summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
-  mutate(label = if_else(is.na(suffix), label, paste0(label, "_", suffix))) %>%
-  select(-suffix)
-
-# Relocate 'label' column after 'level'
-summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
-  relocate(label, .after = grouped_level)
-
-------------------------------------------
-# working up to here. I need to add length, width, and thickness columns
-------------------------------------------
-
-# Separate `excavation area` into `length` and `width` and calculate `volume`
-summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
-  separate(`excavation area`, into = c("length", "width"), sep = "\\*") %>%
-  mutate(
-    length = as.numeric(length),
-    width = as.numeric(width),
-    `thickness` = `thickness` / 100  # Convert from cm to m
-  ) %>%
-  mutate(
-    volume = length * width * `thickness`  # Calculate volume in m^3
-  )
-
-# Calculate lithic volumetric density 
-summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
-  mutate(
-    `lithic_dens` = round(total_lithics / volume, 2)
-  )
-
-# Calculate retouch frequency 
-summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
-  mutate(
-    retouch_freq = round(retouch_tool / volume, 2)  
-  )
-```
-
-```{r}
+ # BM:  "total lithics" is everything except sickle shine, 
+ # "retouch frequency" is retouched_tool_count / "total lithics"
+  rowwise() %>% 
+  mutate(total_lithics = sum(blade_count,
+                             flake_count,
+                             bladelet_count,
+                             tool_count,
+                             core_count,
+                             core_rejuvenation_count,
+                             retouched_tool_count,
+                             utilized_tool_count)) %>% 
+  mutate(retouch_frequency =  retouched_tool_count / total_lithics)
+
+# BM: get the total number of artefacts, number of retouched pieces, and 
+# artefact density. I think this gets us to something similar to the google
+# sheet, the values are not exact, but I have no way to check how how
+# calculated the values in the google sheet so 
+sum_depth_sheet <- 
+tl_final_excavation_data_volume_artefact_counts %>% 
+  # some ex-area/chrono-unit/depth have no volume, so drop those
+  drop_na(excavation_volume) %>% 
+  mutate(lithic_density = total_lithics / excavation_volume)
+```
+
+### Plot the WABI
+
+```{r}
+#| label: fig-wabi
+#| fig-cap: "Relationship between Retouch Frequency and Lithic Volumetric Density"
+
 library(ggpmisc)
+library(ggrepel)
 
@@
 # here I do not consider those levels with a value 0 of retouch frequency. 
-filtered_data <- sum_depth_sheet %>% 
-  filter(lithic_dens > 0 & retouch_freq > 0)
+filtered_data <- 
+  sum_depth_sheet %>% 
+  filter(lithic_density > 0 & retouch_frequency > 0)
 
 WABI_Level <- ggplot(filtered_data,  
-            aes(x = lithic_dens,
-                y = retouch_freq)) +
-  geom_point(color = "black", size = 1.5) +
+            aes(x = lithic_density,
+                y = retouch_frequency)) +
+  geom_point(color = "black", 
+             size = 1.5) +
   geom_smooth(method = "lm", 
@@
   # Add text labels based on 'label' column
-  geom_text(aes(label = label), 
-            vjust = 1.5, 
-            hjust = 0.5, 
-            check_overlap = TRUE) +
-
+  geom_text_repel(aes(label = label2)) +
   stat_poly_eq(aes(label = paste(stat(eq.label), 
@@
   labs(
-    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
     x = "Lithic Volumetric Density",
@@
 # Print the plot to display it
-print(WABI_Level)
-
-```
-
-```{r}
-#PCA_Level: preparing data
-
-#PCA_level: for PCA Bicho and Cascalheira considered these variables: estimated area (as we calculated based on levels, should I consider the thickness?), core frequency, blank frequency, chip frequency: artefacts smaller than 1 cm (as we work on a neolithic asssemblage, we have so many small artefacts; so, I do not consider it), feature frequency (we have some stone alignments, but we do not know the exact level, we know the area of those features), retouch frequency, tool diversity (diversity of tool types within each assemblage, calculated using Menhinick’s index);
-  ##core frequency
-  ##blank frequency
-  ##retouch frequency
-  ##tool diversity
-  ##thickness of deposit (???)
-  ##lithic density
-
-#Calculate Core Frequency
-sum_depth_sheet <- sum_depth_sheet %>%
-  mutate(
-    core_freq = round(cores / volume, 2)  
-  )
-
-# calculate Blank Frequency
-sum_depth_sheet <- sum_depth_sheet %>%
-  mutate(
-    blank_freq = round((flake + blade + bladelet) / volume, 2)  
-  )
-
+WABI_Level
+
+```
+
+### Compute PCA
+
+```{r}
+# PCA_Level: preparing data
+
+# PCA_level: for PCA Bicho and Cascalheira considered these variables: 
+# - estimated area (as we calculated based on levels, should I consider the thickness?),
+# - core frequency: core_count / total_lithics
+# - blank frequency: (flake + blade + bladelet) / total_lithics
+# - chip frequency: artefacts smaller than 1 cm (as we work on a neolithic asssemblage, we have so many small artefacts; so, I do not consider it), 
+# - feature frequency (we have some stone alignments, but we do not know the exact level, we know the area of those features), 
+# - retouch frequency: retouch_frequency
+# - tool diversity (diversity of tool types within each assemblage, calculated using Menhinick’s index) BM: tool diversity from B & C: "calculated using the Menhinick’s index in which the number of tool types represented is divided by the square root of the total number of retouched tool".
+
+sum_depth_sheet_for_pca <- 
+ sum_depth_sheet %>%
+  mutate(
+      core_frequency = core_count / total_lithics,
+      blank_frequency = (flake_count + blade_count + bladelet_count) / total_lithics,
+      total_tools = retouched_tool_count + utilized_tool_count + shine_tool_count,
     # Calculate tool diversity using Menhinick's index
-sum_depth_sheet <- sum_depth_sheet %>%
-  mutate(
-    total_tools = retouch_tool + utilized + sickle_shine,
-    tool_diversity = ifelse(total_tools > 0, n_distinct(tool_type) / sqrt(total_tools), NA)
-  )
-
-```
-
-```{r}
-# performin PCA
+      tool_diversity = tool_type_count / sqrt(total_tools)) %>% 
+   # drop rows that have no tools 
+  filter(total_tools != 0)
+
+# performing PCA
 
@@
 sum_depth_sheet_subset <- 
- sum_depth_sheet %>%
+ sum_depth_sheet_for_pca %>%
   drop_na(tool_diversity) #%>% 
-  #filter(!label %in% c("tp1_c_3_a",
+  # filter(!label %in% c("tp1_c_3_a",
   #                      "d1_a_2_b",
@@
 sum_depth_sheet_subset %>% 
-  select(lithic_dens, 
-         core_freq, 
-         blank_freq, 
-         retouch_freq, 
+  select(lithic_density, 
+         core_frequency, 
+         blank_frequency, 
+         retouch_frequency, 
          tool_diversity)
@@
 # View the summary results
-summary(res.pca)
-
-```
-
-```{r}
-#Plot: Contribution of variables for each of the four relevant PCA dimensions
-
-#save contribution plots into one objects
+# summary(res.pca)
+```
+
+```{r}
+#| label: pca-contrib
+#| fig-cap: "PCA contribution plots"
+#| fig-height: 10
+
+# Plot: Contribution of variables for each of the four relevant PCA dimensions
+# save contribution plots into one objects
 
@@
                    ncol = 2)
-
-```
-
-```{r}
-# Extract contributions of variables to each principal component
-contributions <- res.pca$var$contrib
-
-# Compute average contribution for each dimension
-avg_contrib_dim1 <- mean(contributions[,1], na.rm = TRUE)
-avg_contrib_dim2 <- mean(contributions[,2], na.rm = TRUE)
-avg_contrib_dim3 <- mean(contributions[,3], na.rm = TRUE)
-avg_contrib_dim4 <- mean(contributions[,4], na.rm = TRUE)
-avg_contrib_dim5 <- mean(contributions[,5], na.rm = TRUE)
-
-# Find variables that contribute more than average to each dimension
-high_contrib_dim1 <- contributions[contributions[,1] > avg_contrib_dim1, 1]
-high_contrib_dim2 <- contributions[contributions[,2] > avg_contrib_dim2, 2]
-high_contrib_dim3 <- contributions[contributions[,3] > avg_contrib_dim3, 3]
-high_contrib_dim4 <- contributions[contributions[,4] > avg_contrib_dim4, 4]
-high_contrib_dim5 <- contributions[contributions[,5] > avg_contrib_dim4, 5]
-
-# Sort them in descending order
-sorted_high_contrib_dim1 <- sort(high_contrib_dim1, decreasing = TRUE)
-sorted_high_contrib_dim2 <- sort(high_contrib_dim2, decreasing = TRUE)
-sorted_high_contrib_dim3 <- sort(high_contrib_dim3, decreasing = TRUE)
-sorted_high_contrib_dim4 <- sort(high_contrib_dim4, decreasing = TRUE)
-sorted_high_contrib_dim5 <- sort(high_contrib_dim5, decreasing = TRUE)
-
-# Print or use for further analysis
-print("Significant contributions to Dim1:")
-print(sorted_high_contrib_dim1)
-
-print("Significant contributions to Dim2:")
-print(sorted_high_contrib_dim2)
-
-print("Significant contributions to Dim3:")
-print(sorted_high_contrib_dim3)
-
-print("Significant contributions to Dim4:")
-print(sorted_high_contrib_dim4)
-
-print("Significant contributions to Dim5:")
-print(sorted_high_contrib_dim5)
 ```
@@
 # Inspect eigenvalues
-res.pca$eig
-```
-
-```{r}
+
 eigen_df <- data.frame(
@@
 
-```
-
-```{r}
-
 library(kableExtra)
-library(htmltools)
 
@@
 # Re-order the columns based on the corrected names
-eigen_df <- eigen_df[, c("Dimension", "eigenvalue", "percentage.of.variance", "cumulative.percentage.of.variance")]
+eigen_df <- eigen_df[, c("Dimension", 
+                         "eigenvalue", 
+                         "percentage.of.variance",
+                         "cumulative.percentage.of.variance")]
 
 # Round the numerical columns to 3 decimal places
-eigen_df[, 2:4] <- round(eigen_df[, 2:4], 3)
+eigen_df[, 2:4] <- round(eigen_df[, 2:4], 2)
 
@@
 kable_output <- kable(eigen_df, 
-                      "html", 
-                      align = 'c', 
                       col.names = c("Dimension", 
@@
                                     "Variance Percent", 
-                                    "Cumulative Variance Percent")) %>%
-  kable_styling("striped", 
-                full_width = F) %>%
-  add_header_above(c(" " = 1, "Eigenvalues and percentage of variance for each dimension of PCA" = 3))
+                                    "Cumulative Variance Percent"),
+  caption = "Eigenvalues and percentage of variance for each dimension of PCA")
 
@@
 kable_output
-
 ```
@@
 
-rownames(res.pca$ind$coord) <- sum_depth_sheet_subset$label
+rownames(res.pca$ind$coord) <- 
+  sum_depth_sheet_subset$label2
+
+# get excavation area so we can use it for grouping
+excavation_area <- 
+  str_extract(sum_depth_sheet_subset$label2, 
+              "[^_]+")
 
@@
                                addEllipses = TRUE,
-                               col.ind = sum_depth_sheet_subset$area
+                               col.ind = excavation_area
                               )
@@
 # Modify row names of coordinates
-rownames(res.pca$ind$coord) <- sum_depth_sheet_subset$label
+rownames(res.pca$ind$coord) <- sum_depth_sheet_subset$label2
 
@@
   as_tibble() %>% 
-  mutate(labels = sum_depth_sheet_subset$label,
-         area =   sum_depth_sheet_subset$area)
+  mutate(labels = sum_depth_sheet_subset$label2,
+         area =   excavation_area)
 
@@
 ```
-
-```{r}
-# Base biplot without points
-biplot_base <- fviz_pca_biplot(res.pca,
-                               axes = c(1, 2),
-                               geom = "none",  
-                               labelsize = 2
-                               )
-
-# Add points and text labels manually
-biplot_modified <- biplot_base +
-  geom_point(data = coord_df, 
-             aes(x = Dim.1, 
-                 y = Dim.2, 
-                 color = group), size = 1) +
-  geom_text_repel(data = coord_df,
-                  aes(x = Dim.1,
-                      y = Dim.2,
-                      label = label),
-                  box.padding = 0.5,
-                  point.padding = 0.5,
-                  size = 2)
-
-# Display the plot
-biplot_modified
-
-```
-
-```{r}
-# Visualize biplot: 2
-fviz_pca_biplot(res.pca,
-                axes = c(3, 4),
-                labelsize = 2
-                )
-rownames(res.pca$ind$coord) <- sum_depth_sheet$label
-```

### 2023-09-22 paper/scraps/scraps.R

@@ -1,5 +1,82 @@
 
 #---------------------------------------------------------------
+# BM: we need a single sheet that has only the excavation data
+# such as excavation area, chronological level, depth, area of excavation
+# only that. The we join it to the lithic data. We only want the lithic
+# data to exist in one sheet. Not multiple sheets, that will lead to errors.
+# This is called 'normalisation' of the data.
+
+# get data from google sheet
+library(googlesheets4)
+
+google_sheet_url <-
+  "https://docs.google.com/spreadsheets/d/1Q0QZESk412ZQLE24yPs6Rg-7Y9OMX63DHoXeIPNAIFM/edit#gid=0"
+
+sum_depth_sheet <-
+  read_sheet(google_sheet_url) %>%
+  mutate(level = unlist(level)) %>%
+  mutate(level = ifelse(level == "clay", 1, level)) %>%
+  mutate(label1 = paste(area,
+                       layer,
+                       level,
+                       sep = "_")) %>%
+  mutate(label2 = paste(area,
+                        layer,
+                        level,
+                        upper,
+                        lower,
+                        sep = "_")) %>%
+  select(label1,
+         label2,
+         upper_depth = upper,
+         lower_depth = lower,
+         excavation_area = area,
+         chronological_unit = layer,
+         grouped_level = level,
+         `excavation area`) %>%
+  separate_wider_delim(`excavation area`,
+           names = c("excavation_dimension_1",
+                    "excavation_dimension_2"),
+           delim = "*")
+
+write_csv(sum_depth_sheet,
+          "data/excavation_data.csv")
+
+sum_depth_sheet$label
+
+summary_table_depth_area_unit$label
+
+
+
+
+
+
+
+#---------------------------------------------------------------
+
+# import table with depths that we have subdivided
+depths_divided_tbl <-
+  read_csv(here("data/Lookup Table - Sheet1.csv")) %>%
+  mutate(division_id = rep(c("a", "b"),
+                           nrow(.)/2))
+
+# google sheet has 42 rows, and cols for
+# area, chronological_unit, level, depth. What is the equivalent above? tp1_c_3_a
+
+# This 'depths_divided_tbl' needs more information, for example, some depth ranges
+
+# join with lithic data that
+summary_table_depth_area_unit_to_join %>%
+  left_join(depths_divided_tbl) %>%
+  mutate(label = paste(area,
+                       chronological_unit,
+                       # level, # need this
+                       division_id,
+                       sep = "_")) %>% View
[…]

### 2023-09-22 paper/scraps/scraps.R
activate renv and docker for CI on Github

@@ -46,12 +46,6 @@ sum_depth_sheet$label
 
 summary_table_depth_area_unit$label
 
-
-
-
-
-
-
 #---------------------------------------------------------------
 
 # import table with depths that we have subdivided
@@ -183,15 +177,6 @@ summary_table2_area <- summary_table2_area %>%
 
 print(summary_table2_area)
 
-```
-
-
-
-
-
-
-
-```{r}
 
 
 ##### DO NOT RUN ######
@@ -220,8 +205,6 @@ density_data_layer <- data.frame(
 summary_table2_layer <- merge(summary_table2_layer, density_data_layer, by = "layer")
 
 
-
-```
 
 # regression plot-area
 

### 2023-09-23 paper/paper.qmd

--- 2023-09-22 paper/paper_tulai.qmd
@@
 # Read the data
-tl <- read_csv(here("data/Tulai_Lithic_Assamblage.csv"))
+tl <- read_csv(here("data/raw-data/Tulai_Lithic_Assamblage.csv"))
 
@@
 # and simplified using code in the scraps.R file
-excavation_data <- read_csv(here("data/excavation_data.csv"))
+excavation_data <- read_csv(here("data//raw-data/excavation_data.csv"))
 ```
@@
 
+# BM: export this table as a CSV to paste into the 
+# drafting document to prepare for submission
+
+write_csv(summary_table_area_filtered,
+          here("data/derived-data/tbl-001-summary_table_area_filtered.csv"))
+
 # BM: this is table 1 in the manuscript
@@
              caption = "Summary of lithics in the main areas")
-
 ```
@@
 
-```
-
-```{r}
+# BM: export this table as a CSV to paste into the 
+# drafting document to prepare for submission
+
+write_csv(summary_table_tool_area,
+          here("data/derived-data/tbl-002-summary_table_tool_area.csv"))
 
@@
   sum_depth_sheet %>% 
-  filter(lithic_density > 0 & retouch_frequency > 0)
+  filter(lithic_density > 0 & retouch_frequency > 0)  %>% 
+  mutate(excavation_area =  str_extract(label2, "[^_]+"))
 
@@
                 y = retouch_frequency)) +
-  geom_point(color = "black", 
-             size = 1.5) +
+  geom_point(aes(color = excavation_area,
+                 shape = excavation_area), 
+             size = 4) +
   geom_smooth(method = "lm", 
-              se = TRUE, 
+              se = FALSE, 
               color = "blue", 
@@
   # Add text labels based on 'label' column
-  geom_text_repel(aes(label = label2)) +
+  # BM: this is useful for investigating and
+  # preparing, but not really informative 
+  # for the reader, so we omit for publication
+  # geom_text_repel(aes(label = label2)) +
   stat_poly_eq(aes(label = paste(stat(eq.label), 
@@
                parse = TRUE, 
-               size = 4) +
+               size = 5,
+               label.x = "right",
+               label.y = "bottom") +
   labs(
     x = "Lithic Volumetric Density",
-    y = "Retouch Frequency"
+    y = "Retouch Frequency",
+    colour = "Excavation area",
+    shape = "Excavation area"
   ) +
@@
   scale_y_log10() +
+  theme_minimal() +
   theme(
+    legend.position = c(.8,.3),
     axis.text = element_text(size = 7),
@@
 WABI_Level
+
+# BM: save this figure to paste into the 
+# drafting document to prepare for submission
+
+ggsave(here("figures/fig-wabi-plot.jpg"),
+       h = 6, 
+       w = 8)
 
@@
  sum_depth_sheet_for_pca %>%
-  drop_na(tool_diversity) #%>% 
-  # filter(!label %in% c("tp1_c_3_a",
-  #                      "d1_a_2_b",
-  #                     "d1_c_3_b"))
+  drop_na(tool_diversity) %>% 
+  # explore the effect of outliers on the PCA space
+ filter(!label2 %in% c("tp1_b_1_120_140",
+                        "tp1_c_3_0_10"))
 
@@
 contribution_plots <- 
-map(1:5, ~fviz_contrib(res.pca, 
+map(1:4, ~fviz_contrib(res.pca, 
                    choice = "var", 
@@
                    ncol = 2)
+
+# BM: save this figure to paste into the 
+# drafting document to prepare for submission
+
+ggsave(here("figures/fig-contribution-plots.jpg"),
+       h = 6, 
+       w = 8)
+
 ```
@@
 
-eigen_df <- data.frame(
-   eigenvalue = res.pca$eig[,1],
-  `percentage of variance` = res.pca$eig[,2],
-  `cumulative percentage of variance` = res.pca$eig[,3]
+eigen_df <- tibble(
+   Eigenvalue = res.pca$eig[,1],
+  `Percentage of Variance` = res.pca$eig[,2],
+  `Cumulative Percentage of Variance` = res.pca$eig[,3]
 )
@@
 eigen_df <- eigen_df[, c("Dimension", 
-                         "eigenvalue", 
-                         "percentage.of.variance",
-                         "cumulative.percentage.of.variance")]
+                         "Eigenvalue", 
+                         "Percentage of Variance",
+                         "Cumulative Percentage of Variance")]
 
@@
 
-# Removing row names
-rownames(eigen_df) <- NULL
+# BM: export this table as a CSV to paste into the 
+# drafting document to prepare for submission
+
+write_csv(eigen_df,
+          here("data/derived-data/tbl-003-pca-eigenvalues.csv"))
 
@@
                                addEllipses = TRUE,
-                               col.ind = excavation_area
+                               col.ind = excavation_area,
+                               title = ""
                               )
@@
                                    force = 10
-                                  )
+                                  ) +
+  labs(colour = "",
+       shape = "",
+       fill = "") +
+  theme(legend.position = c(0.9, 0.2))
 
@@
 
-```
+# BM: save this figure to paste into the 
+# drafting document to prepare for submission
+
+ggsave(here("figures/fig-pca-biplot.jpg"),
+       h = 6, 
+       w = 8)
+
+```

### 2023-09-25 paper/paper.qmd
changed calculation for 'total_lithics' and "core_count"

@@
              `Core Typology` == "prussurebullet core" ~ "bullet-pressure",
-             `Core Typology` == "prismaticcore" ~ "prismatic-percussion",
+             `Core Typology` == "prismaticcore" ~ "cylinder/prismatic-percussion",
              `Core Typology` == "pyramidalpressure core" ~ "pyramid-pressure",
@@
              `Core Typology` == "corefragment" ~ "core fragment-NA",
-             `Core Typology` == "cilandrical/pressure" ~ "cylinder-pressure",
+             `Core Typology` == "cilandrical/pressure" ~ "cylinder/prismatic-pressure",
              `Core Typology` == "bullet/pressure" ~ "bullet-pressure",
@@
 
-    `Core Typology` == "bullet" ~ "bullet-pressure",
+    `Core Typology` == "0" ~ "",
+    `Core Typology` == "na" ~ "",
+    `Core Typology` == "na" ~ "",
 
@@
       `Typology` == "unidirectionalflat bladelet core" ~ "core/flat",
-      `Typology` == "" ~ "NA-NA",
-      `Typology` == "na" ~ "NA-NA",
+      `Typology` == "" ~ "",
+      `Typology` == "na" ~ "",
       TRUE ~ `Typology`
@@
       `core-fragment` == "rejuvention piece-side" ~ "rejuvenation piece",
+      `core-fragment` == "rejuvenation piece-side" ~ "rejuvenation piece",
       `core-fragment` == "core preparation piece-platform" ~ "core tablet",
+      `core-fragment` == "core preparation piece-platform" ~ "core tablet",
+      `core-fragment` == "core platform rejuvention" ~ "core tablet",
+      `core-fragment` == "core platform rejuvention " ~ "core tablet",
       `core-fragment` == "core preparation" ~ "rejuvenation piece",
@@
       `core-fragment` == "prismatic" ~ "cylinder/prismatic",
-      `core-fragment` == "shapeless" ~ "cylinder/prismatic",
+      `core-fragment` == "shapeless" ~ "shapeless",
       `core-fragment` == "multidirectional core fragment" ~ "shapeless",
@@
       `core-fragment` == "bullet" ~ "bullet",
+      `core-fragment` == "NA" ~ "",
       TRUE ~ `core-fragment`
@@
         "pyramid",
-        "pyramid core",
-        "pyramid-unidirectional",
-        "core fragment-pyramid",
         "cylinder/prismatic",
-        "shapeless",
-        "multidirectional core fragment",
-        "multidirectional flake core",
-        "multidirectional bladelet core",
+        "core fragment",
+        "cylinder/prismatic",
+        "bullet",
+        "flat",
+        "core on flake",
+        "bipolar",
         "mixed core",
@@
         "core on flake",
-        "bullet"
+        "bullet",
+        "pyramid",
+        "flat",
+        "cylinder/prismatic ",
+        "bipolar ",
+        "shapeless ",
+        "bullet "
+
       ) ~ `core-fragment`,
@@
         "rejuvenation piece",
-        "rejuvention piece-side",
-        "core preparation piece-platform",
-        "core preparation",
-        "core preparation piece",
+        "core tablet",
+        "primary flake",
+        "primary blade",
+        "crested bladelet",
         "core preperation",
@@
         "primary flake" ,
-        "primary blade"
+        "primary blade",
+        "core fragment "
+
       ) ~ `core-fragment`,
@@
       c(
+        "shapeless",
         "pyramid",
@@
 # take a look
-# rev(sort(table(tl_final$core)))
-# rev(sort(table(tl_final$corefrag)))
-# rev(sort(table(tl_final$`core-fragment`)))
-# rev(sort(table(tl_final$blank)))
+ rev(sort(table(tl_final$core)))
+ rev(sort(table(tl_final$corefrag)))
+ rev(sort(table(tl_final$`core-fragment`)))
+ rev(sort(table(tl_final$`core-typology`)))
+ rev(sort(table(tl_final$blank)))
+```
+
+```{r}
+
 ```
@@
   summarise(
-    `Cores (n)` = sum(!is.na(core) & core != ""),
+   `Cores (n)` = sum((!is.na(`core-typology`) & `core-typology` != "") |
+                      (!is.na(core) & core != ""), na.rm = TRUE),  # Summing variables from both columns
     `Core rejuvenations pieces (n)` = sum(!is.na(corefrag) &
@@
              caption = "Summary of lithics in the main areas")
+```
+
+```{r}
+
 ```
@@
   caption = "Tool Typology in the three main areas")
+
+```
+
+```{r}
+# Manually calculate the totals for each area and grand total
+total_d1 <- sum(summary_table_tool_area$d1, na.rm = TRUE)
+total_ta <- sum(summary_table_tool_area$ta, na.rm = TRUE)
+total_tp1 <- sum(summary_table_tool_area$tp1, na.rm = TRUE)
+grand_total <- total_d1 + total_ta + total_tp1
+
+# Calculate the percentages for each cell based on the total count
+summary_table_tool_area_percentage <- summary_table_tool_area %>%
+  mutate(
+    percent_d1 = round(d1 / total_d1 * 100, 1),
+    percent_ta = round(ta / total_ta * 100, 1),
+    percent_tp1 = round(tp1 / total_tp1 * 100, 1)
+  ) %>%
+  # Add the total_percent column, calculating based on grand total
+  mutate(total_percent = round((d1 + ta + tp1) / grand_total * 100, 1))
+
+# Select only the necessary columns (tooltypegroup, the percentage columns, and total_percent)
+summary_table_tool_area_percentage <- summary_table_tool_area_percentage %>%
+  select(tooltypegroup, percent_d1, percent_ta, percent_tp1, total_percent)
+
+# write the percentage table to a CSV file
+write_csv(summary_table_tool_area_percentage,
+          here("data/derived-data/tbl-002-summary_table_tool_area_percentage.csv"))
+
+kable(
+  summary_table_tool_area_percentage,
+  col.names = c("Tools", "D1 (%)", "TA (%)", "TP1 (%)", "Total (%)"),
+  caption = "Tool Typology in the three main areas (Percentage)"
+)
 
@@
                                                        na.rm = TRUE),
-    core_count =               sum(!is.na(`core-typology`), na.rm = TRUE),
+    core_count =                sum(!is.na(`core-typology`), na.rm = TRUE) + sum(!is.na(core), na.rm = TRUE),
+
     core_rejuvenation_count =  sum(!is.na(`corefrag`), 
@@
  # "retouch frequency" is retouched_tool_count / "total lithics"
+# SS: I think "total lithics" would be the total of blade, flake,bladelet, core and core rejuvenation counts (as we already cound the tools in bladelet, blade, or flake) 
   rowwise() %>% 
@@
                              bladelet_count,
-                             tool_count,
                              core_count,
-                             core_rejuvenation_count,
-                             retouched_tool_count,
-                             utilized_tool_count)) %>% 
+                             core_rejuvenation_count
+
+                             )) %>% 
   mutate(retouch_frequency =  retouched_tool_count / total_lithics)

### 2023-09-29 paper/paper.qmd
add some chi-square tests and residual plots

@@
 library(knitr)
+library(ggcorrplot)
+library(ggpubr)
+library(ggbiplot)
 ```
@@
 
-```
+# chi-square test for different frequencies of tool type groups 
+# by area
+  summary_table_area_filtered_for_chi_square <- 
+  summary_table_area_filtered %>% 
+  pivot_longer(-area) %>% 
+  pivot_wider(names_from = "area") %>% 
+  select(-name)
+
+chisq <- 
+  chisq.test(summary_table_area_filtered_for_chi_square)
+
+rownames(chisq$residuals) <- 
+   summary_table_area_filtered %>% 
+  pivot_longer(-area) %>% 
+  pivot_wider(names_from = "area") %>% 
+  pull(name)  %>% 
+  str_remove("\\(n\\)")
+
+ggcorrplot(chisq$residuals,
+           method = 'circle') + 
+  ggplot2::scale_fill_gradient2(
+    low = 'red',
+    high = 'royalblue',
+    mid = 'white',
+    midpoint = 0,
+    limit = c(-7, 11),
+    space = "Lab",
+    name = "Residual"
+  ) + 
+  coord_flip()
+
+```
+
+A Chi-square test on the frequency of major lithic types by excavation area reveals a significant difference between the areas (`r apastats::describe.chi(summary_table_area_filtered_for_chi_square)`).
+Figure XXX shows that there are fewer utlized pieces and flakes in d1 than expected, and more flakes and fewer cores and blades in tp1 than expected assuming a null hypothesis of no difference between the areas.
 
@@
   caption = "Tool Typology in the three main areas")
-
-```
+```
+
+```{r}
+
+# chi-square test for different frequencies of tool type groups 
+# by area
+  summary_table_tool_area_for_chi_square <- 
+  summary_table_tool_area %>% 
+  filter(!tooltypegroup == "Total") %>% 
+  select(d1, ta, tp1)
+
+chisq <- 
+  chisq.test(summary_table_tool_area_for_chi_square)
+
+rownames(chisq$residuals) <- 
+  summary_table_tool_area %>% 
+  filter(!tooltypegroup == "Total") %>% 
+  pull(tooltypegroup)
+
+ggcorrplot(chisq$residuals,
+           method = 'circle') + 
+  ggplot2::scale_fill_gradient2(
+    low = 'red',
+    high = 'royalblue',
+    mid = 'white',
+    midpoint = 0,
+    limit = c(-7, 11),
+    space = "Lab",
+    name = "Residual"
+  ) + 
+  coord_flip()
+
+```
+
+A Chi-square test on the frequency of tool type groups by excavation area reveals a significant difference between the areas (`r apastats::describe.chi(summary_table_tool_area_for_chi_square)`).
+Figure XXX shows that there are more geomeric pieces and burins in d1 than expected, and more backed and retouched pieces in tp1 than expected assuming a null hypothesis of no difference between the areas.
 

### 2023-09-29 paper/paper.qmd
avoid function name conflicts with ::, replace PCA ellipses with convex hulls

@@
 # Read the data
-tl <- read_csv(here("data/raw-data/Tulai_Lithic_Assamblage.csv"))
+tl <- read_csv(here::here("data/raw-data/Tulai_Lithic_Assamblage.csv"))
 
@@
 # and simplified using code in the scraps.R file
-excavation_data <- read_csv(here("data//raw-data/excavation_data.csv"))
+excavation_data <- read_csv(here::here("data//raw-data/excavation_data.csv"))
 ```
@@
     ),
-    Utilization = if_else(Utilization == "0", NA_character_, Utilization),
-    Retouch =     if_else(Retouch == "0", NA_character_, Retouch),
-    SHINE =       if_else(SHINE == "0", NA_character_, SHINE),
-    shine2 =      if_else(shine2 == "0", NA_character_, shine2),
-    Blank2 =      if_else(Blank2 == "0", NA_character_, Blank2)
+    Utilization = if_else(Utilization == "0", NA, Utilization),
+    Retouch =     if_else(Retouch ==     "0", NA, Retouch),
+    SHINE =       if_else(SHINE ==       "0", NA, SHINE),
+    shine2 =      if_else(shine2 ==      "0", NA, shine2),
+    Blank2 =      if_else(Blank2 ==      "0", NA, Blank2)
   ) %>% 
@@
 tl_final <- tl4 %>%
-  rename(
+  dplyr::rename(
     blank =                Blank2,
@@
   group_by(area) %>%
-  summarise(
+  dplyr::summarise(
    `Cores (n)` = sum((!is.na(`core-typology`) & `core-typology` != "") |
@@
   summary_table_area %>%
-  filter(area %in% c("ta", "d1", "tp1"))
+  dplyr::filter(area %in% c("ta", "d1", "tp1"))
 
@@
 write_csv(summary_table_area_filtered,
-          here("data/derived-data/tbl-001-summary_table_area_filtered.csv"))
+          here::here("data/derived-data/tbl-001-summary_table_area_filtered.csv"))
 
@@
   select(-tooltypegroup) %>%
-  summarise(across(everything(), \(x) sum(x, na.rm = TRUE))) %>%
+  dplyr::summarise(across(everything(), \(x) sum(x, na.rm = TRUE))) %>%
   mutate(tooltypegroup = "Total")
@@
 write_csv(summary_table_tool_area,
-          here("data/derived-data/tbl-002-summary_table_tool_area.csv"))
+          here::here("data/derived-data/tbl-002-summary_table_tool_area.csv"))
 
@@
 write_csv(summary_table_tool_area_percentage,
-          here("data/derived-data/tbl-002-summary_table_tool_area_percentage.csv"))
+          here::here("data/derived-data/tbl-002-summary_table_tool_area_percentage.csv"))
 
@@
   group_by(label2) %>% 
-   summarise(
+   dplyr::summarise(
     blade_count =              sum(blank == "blade", na.rm = TRUE),
@@
 
-ggsave(here("figures/fig-wabi-plot.jpg"),
+ggsave(here::here("figures/fig-wabi-plot.jpg"),
        h = 6, 
@@
 
-ggsave(here("figures/fig-contribution-plots.jpg"),
+ggsave(here::here("figures/fig-contribution-plots.jpg"),
        h = 6, 
@@
 write_csv(eigen_df,
-          here("data/derived-data/tbl-003-pca-eigenvalues.csv"))
+          here::here("data/derived-data/tbl-003-pca-eigenvalues.csv"))
 
@@
               "[^_]+")
-
-# Generate the base biplot
-biplot_base <- fviz_pca_biplot(res.pca,
-                               axes = c(1, 2),
-                               labelsize = 2,
-                               label = "var",
-                               addEllipses = TRUE,
-                               col.ind = excavation_area,
-                               title = ""
-                              )
-
-# Modify row names of coordinates
-rownames(res.pca$ind$coord) <- sum_depth_sheet_subset$label2
-
-pca_label_text <- 
-res.pca$ind$coord %>% 
-  as_tibble() %>% 
-  mutate(labels = sum_depth_sheet_subset$label2,
-         area =   excavation_area)
-
-# Add geom_text_repel() to the base plot
-biplot_modified <- biplot_base +
-                   geom_text_repel(data = pca_label_text,
-                                   aes(x = Dim.1, 
-                                       y = Dim.2, 
-                                       label = labels,
-                                       colour = area),
-                                   segment.color = "grey80",
-                                   size = 2,
-                                   force = 10
-                                  ) +
-  labs(colour = "",
-       shape = "",
-       fill = "") +
-  theme(legend.position = c(0.9, 0.2))
+pca_data.pca <- prcomp(pca_data, scale. = TRUE)
+
+biplot_modified <- 
+ggbiplot(
+  pca_data.pca,
+  obs.scale = 1,
+  var.scale = 1,
+  groups = excavation_area,
+  ellipse = FALSE,
+  circle = TRUE) +
+  stat_chull(aes(color = excavation_area,
+                 fill = excavation_area),
+             alpha = 0.1,
+             geom = "polygon") +
+  scale_colour_brewer(palette = "Set1",
+                      name = '',
+                      guide = 'none') +
+  scale_fill_brewer(palette = "Set1",
+                    name = '')  +
+  theme_minimal()
 
@@
 
-ggsave(here("figures/fig-pca-biplot.jpg"),
+ggsave(here::here("figures/fig-pca-biplot.jpg"),
        h = 6, 

### 2023-09-29 paper/scraps/scraps.R

@@ -1,4 +1,34 @@
 
+# BM: PCA plot with convex hulls (I thought we did this already
+# in our meeting because I can see the plot in the google doc
+# but seems we didn't save-commit-push it?)
+
+library(ggpubr)
+library(ggbiplot)
+data(wine)
+pca_data.pca <- prcomp(pca_data, scale. = TRUE)
+
+ggbiplot(
+  pca_data.pca,
+  obs.scale = 1,
+  var.scale = 1,
+  groups = excavation_area,
+  ellipse = FALSE,
+  circle = TRUE
+) +
+  stat_chull(aes(color = excavation_area,
+                 fill = excavation_area),
+             alpha = 0.1,
+             geom = "polygon") +
+  scale_colour_brewer(palette = "Set1",
+                      name = '',
+                      guide = 'none') +
+  scale_fill_brewer(palette = "Set1",
+                    name = '')  +
+  theme_minimal()
+
+
+
 #---------------------------------------------------------------
 # BM: we need a single sheet that has only the excavation data
 # such as excavation area, chronological level, depth, area of excavation

### 2023-09-29 paper/paper.qmd
remove function conflicts where multiple packages have functions of the same name

@@
 # load libraries
-library(tidyverse)
 library(here)
@@
 library(ggbiplot)
+library(tidyverse) # load this last to minimize conflicts
 ```
@@
 # and simplified using code in the scraps.R file
-excavation_data <- read_csv(here::here("data//raw-data/excavation_data.csv"))
+excavation_data <- 
+  read_csv(here::here("data//raw-data/excavation_data.csv"))
 ```
@@
 # take a look
- rev(sort(table(tl_final$core)))
- rev(sort(table(tl_final$corefrag)))
- rev(sort(table(tl_final$`core-fragment`)))
- rev(sort(table(tl_final$`core-typology`)))
- rev(sort(table(tl_final$blank)))
-```
-
-```{r}
-
+ # rev(sort(table(tl_final$core)))
+ # rev(sort(table(tl_final$corefrag)))
+ # rev(sort(table(tl_final$`core-fragment`)))
+ # rev(sort(table(tl_final$`core-typology`)))
+ # rev(sort(table(tl_final$blank)))
 ```

### 2023-09-29 paper/scraps/scraps.R
update renv

@@ -1,4 +1,21 @@
 
+# BM: new method to me! I found https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8858582/
+
+library(PCAtest)
+
+result <-
+  PCAtest(pca_data,
+          100,
+          100,
+          0.05,
+          varcorr=FALSE,
+          counter=FALSE,
+          plot=TRUE)
+#>
+
+
+
+#---------------------------------------------------------------
 # BM: PCA plot with convex hulls (I thought we did this already
 # in our meeting because I can see the plot in the google doc
 # but seems we didn't save-commit-push it?)
@@ -26,6 +43,8 @@ ggbiplot(
   scale_fill_brewer(palette = "Set1",
                     name = '')  +
   theme_minimal()
+
+
 
 
 

### 2023-09-29 paper/scraps/scraps.R
exploring post-hoc tests for the PCA

@@ -11,9 +11,26 @@ result <-
           varcorr=FALSE,
           counter=FALSE,
           plot=TRUE)
-#>
 
 
+# PerMANOVA - partitioning the euclidean distance matrix by species
+# from https://stackoverflow.com/a/20267537/1036500
+library(vegan)
+
+pca_data_for_post_hoc_testing <-
+pca_data %>%
+  add_column(excavation_area)
+
+PerMANOVA_results <-
+adonis2(pca_data[, 1:5] ~ excavation_area,
+        data = pca_data_for_post_hoc_testing,
+        method='eu')
+
+
+library(pairwiseAdonis)
+pairwise.adonis2_results <-
+pairwise.adonis2(pca_data[, 1:5] ~ excavation_area,
+                data = pca_data_for_post_hoc_testing)
 
 #---------------------------------------------------------------
 # BM: PCA plot with convex hulls (I thought we did this already

### 2023-09-29 paper/scraps/scraps.R
add a few more comments

@@ -21,12 +21,15 @@ pca_data_for_post_hoc_testing <-
 pca_data %>%
   add_column(excavation_area)
 
+# Is there are significant difference in the groups that
+# we see in the PCA?
 PerMANOVA_results <-
 adonis2(pca_data[, 1:5] ~ excavation_area,
         data = pca_data_for_post_hoc_testing,
         method='eu')
 
-
+# Which specific pairs of groups differ from each other?
+# https://github.com/pmartinezarbizu/pairwiseAdonis
 library(pairwiseAdonis)
 pairwise.adonis2_results <-
 pairwise.adonis2(pca_data[, 1:5] ~ excavation_area,

### 2023-09-29 paper/paper.qmd
stash

--- 2023-09-29 paper/paper.qmd
@@
 library(knitr)
-library(ggcorrplot)
-library(ggpubr)
-library(ggbiplot)
 ```
@@
 
-# chi-square test for different frequencies of tool type groups 
-# by area
-  summary_table_area_filtered_for_chi_square <- 
-  summary_table_area_filtered %>% 
-  pivot_longer(-area) %>% 
-  pivot_wider(names_from = "area") %>% 
-  select(-name)
-
-chisq <- 
-  chisq.test(summary_table_area_filtered_for_chi_square)
-
-rownames(chisq$residuals) <- 
-   summary_table_area_filtered %>% 
-  pivot_longer(-area) %>% 
-  pivot_wider(names_from = "area") %>% 
-  pull(name)  %>% 
-  str_remove("\\(n\\)")
-
-ggcorrplot(chisq$residuals,
-           method = 'circle') + 
-  ggplot2::scale_fill_gradient2(
-    low = 'red',
-    high = 'royalblue',
-    mid = 'white',
-    midpoint = 0,
-    limit = c(-7, 11),
-    space = "Lab",
-    name = "Residual"
-  ) + 
-  coord_flip()
-
-```
-
-A Chi-square test on the frequency of major lithic types by excavation area reveals a significant difference between the areas (`r apastats::describe.chi(summary_table_area_filtered_for_chi_square)`).
-Figure XXX shows that there are fewer utlized pieces and flakes in d1 than expected, and more flakes and fewer cores and blades in tp1 than expected assuming a null hypothesis of no difference between the areas.
+```
 
@@
   caption = "Tool Typology in the three main areas")
-```
-
-```{r}
-
-# chi-square test for different frequencies of tool type groups 
-# by area
-  summary_table_tool_area_for_chi_square <- 
-  summary_table_tool_area %>% 
-  filter(!tooltypegroup == "Total") %>% 
-  select(d1, ta, tp1)
-
-chisq <- 
-  chisq.test(summary_table_tool_area_for_chi_square)
-
-rownames(chisq$residuals) <- 
-  summary_table_tool_area %>% 
-  filter(!tooltypegroup == "Total") %>% 
-  pull(tooltypegroup)
-
-ggcorrplot(chisq$residuals,
-           method = 'circle') + 
-  ggplot2::scale_fill_gradient2(
-    low = 'red',
-    high = 'royalblue',
-    mid = 'white',
-    midpoint = 0,
-    limit = c(-7, 11),
-    space = "Lab",
-    name = "Residual"
-  ) + 
-  coord_flip()
-
-```
-
-A Chi-square test on the frequency of tool type groups by excavation area reveals a significant difference between the areas (`r apastats::describe.chi(summary_table_tool_area_for_chi_square)`).
-Figure XXX shows that there are more geomeric pieces and burins in d1 than expected, and more backed and retouched pieces in tp1 than expected assuming a null hypothesis of no difference between the areas.
+
+```
 
@@
 # write the percentage table to a CSV file
-write_csv(summary_table_tool_area_percentage,
-          here("data/derived-data/tbl-002-summary_table_tool_area_percentage.csv"))
-
-kable(
-  summary_table_tool_area_percentage,
-  col.names = c("Tools", "D1 (%)", "TA (%)", "TP1 (%)", "Total (%)"),
-  caption = "Tool Typology in the three main areas (Percentage)"
-)
 
@@
 
-# Generate the base biplot
-biplot_base <- fviz_pca_biplot(res.pca,
-                               axes = c(1, 2),
-                               labelsize = 2,
-                               label = "var",
-                               addEllipses = TRUE,
-                               col.ind = excavation_area,
-                               title = ""
-                              )
-
-# Modify row names of coordinates
-rownames(res.pca$ind$coord) <- sum_depth_sheet_subset$label2
-
-pca_label_text <- 
-res.pca$ind$coord %>% 
-  as_tibble() %>% 
-  mutate(labels = sum_depth_sheet_subset$label2,
-         area =   excavation_area)
-
-# Add geom_text_repel() to the base plot
-biplot_modified <- biplot_base +
-                   geom_text_repel(data = pca_label_text,
-                                   aes(x = Dim.1, 
-                                       y = Dim.2, 
-                                       label = labels,
-                                       colour = area),
-                                   segment.color = "grey80",
-                                   size = 2,
-                                   force = 10
+library(ggpubr)
+library(ggbiplot)
+
+pca_data.pca <- prcomp(pca_data, scale. = TRUE)
+
+biplot_modified <- 
+ggbiplot(
+  pca_data.pca,
+  obs.scale = 1,
+  var.scale = 1,
+  groups = excavation_area,
+  ellipse = FALSE,
+  circle = TRUE
 ) +
-  labs(colour = "",
-       shape = "",
-       fill = "") +
-  theme(legend.position = c(0.9, 0.2))
+  stat_chull(aes(color = excavation_area,
+                 fill = excavation_area),
+             alpha = 0.1,
+             geom = "polygon") +
+  scale_colour_brewer(palette = "Set1",
+                      name = '',
+                      guide = 'none') +
+  scale_fill_brewer(palette = "Set1",
+                    name = '')  +
+  theme_minimal()
 
@@
 
-ggsave(here("figures/fig-pca-biplot.jpg"),
+ggsave(plot = biplot_modified,
+       filename = here::here("figures/fig-pca-biplot.jpg"),
        h = 6, 

### 2023-09-29 paper/scraps/scraps.R

@@ -1,4 +1,40 @@
 
+#BM: explore PCA with hulls
+# library(devtools)
+# install_github("vqv/ggbiplot")
+library(ggpubr)
+library(ggbiplot)
+data(wine)
+wine.pca <- prcomp(pca_data, scale. = TRUE)
+
+ggbiplot(
+  wine.pca,
+  obs.scale = 1,
+  var.scale = 1,
+  groups = excavation_area,
+  ellipse = FALSE,
+  circle = TRUE
+) +
+  stat_chull(aes(color = excavation_area,
+                 fill = excavation_area),
+             alpha = 0.1,
+             geom = "polygon") +
+  scale_colour_brewer(palette = "Set1",
+                      name = '',
+                      guide = 'none') +
+  scale_fill_brewer(palette = "Set1",
+                    name = '')  +
+  theme_minimal()
+
+
+
+# BM: some discrepancy found in artefact counts for
+
+ rev(sort(table(tl_final_area$area)))
+
+ rev(sort(table(tl_final_depth_area_unit$area)))
+
+
 #---------------------------------------------------------------
 # BM: we need a single sheet that has only the excavation data
 # such as excavation area, chronological level, depth, area of excavation

### 2023-09-29 paper/paper.qmd
review chi-square test and PCA post-hoc

--- 2023-09-29 paper/paper.qmd
@@
 
-kable(
-  summary_table_tool_area_percentage,
-  col.names = c("Tools", "D1 (%)", "TA (%)", "TP1 (%)", "Total (%)"),
-  caption = "Tool Typology in the three main areas (Percentage)"
-)
-
 ```
@@
 
+library(ggpubr)
+library(ggbiplot)
+
+pca_data.pca <- prcomp(pca_data, scale. = TRUE)
+
 biplot_modified <- 
@@
   ellipse = FALSE,
-  circle = TRUE) +
+) +
   stat_chull(aes(color = excavation_area,
@@
 
-ggsave(here::here("figures/fig-pca-biplot.jpg"),
+ggsave(plot = biplot_modified,
+       filename = here::here("figures/fig-pca-biplot.jpg"),
        h = 6, 

### 2023-09-29 paper/scraps/scraps.R

@@ -1,16 +1,4 @@
 
-<<<<<<< HEAD
-#BM: explore PCA with hulls
-# library(devtools)
-# install_github("vqv/ggbiplot")
-library(ggpubr)
-library(ggbiplot)
-data(wine)
-wine.pca <- prcomp(pca_data, scale. = TRUE)
-
-ggbiplot(
-  wine.pca,
-=======
 # BM: new method to me! I found https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8858582/
 
 library(PCAtest)
@@ -23,7 +11,6 @@ result <-
           varcorr=FALSE,
           counter=FALSE,
           plot=TRUE)
-
 
 # PerMANOVA - partitioning the euclidean distance matrix by species
 # from https://stackoverflow.com/a/20267537/1036500
@@ -54,12 +41,10 @@ pairwise.adonis2(pca_data[, 1:5] ~ excavation_area,
 
 library(ggpubr)
 library(ggbiplot)
-data(wine)
 pca_data.pca <- prcomp(pca_data, scale. = TRUE)
 
 ggbiplot(
   pca_data.pca,
->>>>>>> af55b91edd04c2dc750f6f8fc8ab9d8a7c9f59ab
   obs.scale = 1,
   var.scale = 1,
   groups = excavation_area,
@@ -78,15 +63,12 @@ ggbiplot(
   theme_minimal()
 
 
-
-<<<<<<< HEAD
 # BM: some discrepancy found in artefact counts for
 
  rev(sort(table(tl_final_area$area)))
 
  rev(sort(table(tl_final_depth_area_unit$area)))
-=======
->>>>>>> af55b91edd04c2dc750f6f8fc8ab9d8a7c9f59ab
+
 
 
 #---------------------------------------------------------------

### 2023-10-03 paper/paper.qmd
add PCA test

@@
 ```{r}
+# PCA test
+# BM: new method to me! I found https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8858582/
+
+library(PCAtest)
+
+# compute the tests and save the plot
+jpeg( filename = here::here("figures/fig-pca-tests.jpg"),
+      height  = 6, 
+      width = 8,
+      units = "in", 
+      res = 900)
+
+result <-
+  PCAtest(pca_data,
+          1000,
+          1000,
+          0.05,
+          varcorr=FALSE,
+          counter=FALSE,
+          plot=TRUE)
+
+dev.off()
+
+knitr::include_graphics(here::here("figures/fig-pca-tests.jpg"))
+
+```
+
+```{r}
 #| label: pca-contrib

### 2023-10-03 paper/paper.qmd
add pairwise tests for PCA results

@@
 ```
+
+```{r}
+#| label: tbl-permanova
+
+# PerMANOVA - partitioning the euclidean distance matrix by species
+# from https://stackoverflow.com/a/20267537/1036500
+library(vegan)
+
+pca_data_for_post_hoc_testing <-
+pca_data %>%
+  add_column(excavation_area)
+
+# Is there are significant difference in the groups that
+# we see in the PCA?
+PerMANOVA_results <-
+adonis2(pca_data[, 1:5] ~ excavation_area,
+        data = pca_data_for_post_hoc_testing,
+        method='eu')
+
+PerMANOVA_results_inline <- 
+  paste0("_F_(",
+         PerMANOVA_results$Df[1],
+         ", ",
+         PerMANOVA_results$Df[2],
+         ") = ",
+         round(PerMANOVA_results$F[1], 3),
+         " _p_ = ",
+         PerMANOVA_results$`Pr(>F)`[1]
+         )
+
+# Which specific pairs of groups differ from each other?
+# https://github.com/pmartinezarbizu/pairwiseAdonis
+library(pairwiseAdonis)
+pairwise.adonis2_results <-
+pairwise.adonis2(pca_data[, 1:5] ~ excavation_area,
+                data = pca_data_for_post_hoc_testing)
+
+# turn this output into a nice table
+pairwise.adonis2_results_tbl <- 
+tibble(
+  # names
+  Pair =   c(names(pairwise.adonis2_results[2]),
+             names(pairwise.adonis2_results[3]),
+             names(pairwise.adonis2_results[4])),
+  # F stat
+       F = c(pairwise.adonis2_results[2][[1]]$F[1],
+             pairwise.adonis2_results[3][[1]]$F[1],
+             pairwise.adonis2_results[4][[1]]$F[1]), 
+  # p-values
+       p = c(pairwise.adonis2_results[2][[1]]$`Pr(>F)`[1],
+             pairwise.adonis2_results[3][[1]]$`Pr(>F)`[1],
+             pairwise.adonis2_results[4][[1]]$`Pr(>F)`[1])
+) %>% 
+  mutate(Pair = str_replace_all(Pair, "_", " "))
+
+knitr::kable(pairwise.adonis2_results_tbl,
+             caption = "Summary of pairwise post-hoc tests on the excavation areas")
+
+```
+
+A permutation MANOVA indicates that there is a significant difference in the lithic assemblage composition across the three excavation areas (`rPerMANOVA_results_inline`).
+We investigated this further using pairwise post-hoc tests and found that this significant difference is between TA and TP1, with the other two areas being statisically indistinguishable (@tbl-permanova )

### 2023-10-06 paper/paper.qmd
resolve merge conflict

@@
 ```{r}
+# SS: get the contributions of each variable to the first two principal components, which we can then use in the text.
+
+res.pca <- PCA(pca_data)
+
+# Contributions of variables to dimension 1 and 2:
+contrib_dim1 <- res.pca$var$contrib[, 1]
+contrib_dim2 <- res.pca$var$contrib[, 2]
+contrib_dim3 <- res.pca$var$contrib[, 3]
+contrib_dim4 <- res.pca$var$contrib[, 4]
+contrib_dim5 <- res.pca$var$contrib[, 5]
+
+# To view the contributions:
+contrib_dim1
+contrib_dim2
+contrib_dim3
+contrib_dim4
+contrib_dim5
+
+```
+
+```{r}
 # screeplot; inspect distribution of PCs

### 2023-10-06 paper/scraps/scraps.R

@@ -1,0 +1,321 @@
+
+# BM: new method to me! I found https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8858582/
+
+library(PCAtest)
+
+result <-
+  PCAtest(pca_data,
+          100,
+          100,
+          0.05,
+          varcorr=FALSE,
+          counter=FALSE,
+          plot=TRUE)
+
+# PerMANOVA - partitioning the euclidean distance matrix by species
+# from https://stackoverflow.com/a/20267537/1036500
+library(vegan)
+
+pca_data_for_post_hoc_testing <-
+pca_data %>%
+  add_column(excavation_area)
+
+# Is there are significant difference in the groups that
+# we see in the PCA?
+PerMANOVA_results <-
+adonis2(pca_data[, 1:5] ~ excavation_area,
+        data = pca_data_for_post_hoc_testing,
+        method='eu')
+
+# Which specific pairs of groups differ from each other?
+# https://github.com/pmartinezarbizu/pairwiseAdonis
+library(pairwiseAdonis)
+pairwise.adonis2_results <-
+pairwise.adonis2(pca_data[, 1:5] ~ excavation_area,
+                data = pca_data_for_post_hoc_testing)
+
+#---------------------------------------------------------------
+# BM: PCA plot with convex hulls (I thought we did this already
+# in our meeting because I can see the plot in the google doc
+# but seems we didn't save-commit-push it?)
+
+library(ggpubr)
+library(ggbiplot)
+pca_data.pca <- prcomp(pca_data, scale. = TRUE)
+
+ggbiplot(
+  pca_data.pca,
+  obs.scale = 1,
+  var.scale = 1,
+  groups = excavation_area,
+  ellipse = FALSE,
+  circle = TRUE
+) +
+  stat_chull(aes(color = excavation_area,
+                 fill = excavation_area),
+             alpha = 0.1,
+             geom = "polygon") +
+  scale_colour_brewer(palette = "Set1",
+                      name = '',
+                      guide = 'none') +
+  scale_fill_brewer(palette = "Set1",
+                    name = '')  +
+  theme_minimal()
+
+
+# BM: some discrepancy found in artefact counts for
+
+ rev(sort(table(tl_final_area$area)))
+
+ rev(sort(table(tl_final_depth_area_unit$area)))
+
+
+
+#---------------------------------------------------------------
+# BM: we need a single sheet that has only the excavation data
+# such as excavation area, chronological level, depth, area of excavation
[…]

### 2023-10-06 paper/paper.qmd
resolve merge conflict

@@
 ```{r}
+
 # SS: get the contributions of each variable to the first two principal components, which we can then use in the text.
@@
 ```{r}
+
 # screeplot; inspect distribution of PCs
@@
 # Display the table
+
 kable_output
+
 ```

### 2023-10-27 paper/paper.qmd
stash

--- 2023-10-03 paper/paper.qmd
@@
 
-A permutation MANOVA indicates that there is a significant difference in the lithic assemblage composition across the three excavation areas (`rPerMANOVA_results_inline`).
+A permutation MANOVA indicates that there is a significant difference in the lithic assemblage composition across the three excavation areas (`r PerMANOVA_results_inline`).
 We investigated this further using pairwise post-hoc tests and found that this significant difference is between TA and TP1, with the other two areas being statisically indistinguishable (@tbl-permanova )

### 2023-11-05 paper/paper.qmd
add map of sites, update stats in text, update table of post-hoc tests, update PCA plot

@@
 tl <- read_csv(here::here("data/raw-data/Tulai_Lithic_Assamblage.csv"))
+site_locations <- read_csv(here::here("data/raw-data/Sites-mentioned-in-the-text.csv"))
 
@@
 excavation_data <- 
-  read_csv(here::here("data//raw-data/excavation_data.csv"))
+  read_csv(here::here("data/raw-data/excavation_data.csv"))
+```
+
+Draw the map and save it
+
+```{r}
+#| eval: false
+
+# get map tiles from the internet, we only need to do this once and save them
+
+library(ggmap) # devtools::install_github("stadiamaps/ggmap")
+
+# using http://bboxfinder.com/ to get this
+bbox = c(
+  42, # left
+  30, # lower
+  51, # right
+  38) # upper
+
+# this takes a few minutes and needs an internet connection
+map <-
+  get_stadiamap(bbox = bbox,
+                zoom = 7)
+
+saveRDS(map, file = here::here("data/raw-data/map_tiles.rds"))
+
+# map with site names next to the points
+map_of_sites <-
+  ggmap(map)  +
+  geom_point(data = site_locations,
+             aes(y = `Longitude DD` ,
+                 x = `Latitude DD`),
+             colour = "red",
+             size = 2) +
+  # label points with site name
+  geom_text_repel(data = site_locations,
+             aes(y = `Longitude DD` ,
+                 x = `Latitude DD`,
+                      label = Site),
+                  size = 2,
+                  bg.color = "white",
+                  bg.r = 0.15) +
+  theme_minimal(base_size = 6) +
+  labs(x = "Longitute",
+       y = "Latitude") +
+  theme(axis.title.x = element_blank(),
+        axis.title.y = element_blank())
+
+# inset world map, from
+# https://r-spatial.org/r/2018/10/25/ggplot2-sf.html
+library(sf)
+library("rnaturalearth")
+library("rnaturalearthdata")
+
+world <- ne_countries(scale = "medium", returnclass = "sf")
+
+worldplot <- 
+ggplot(data = world) +
+     geom_sf(color = "black", 
+             fill = "white",
+             linewidth = 0.05) +
+    annotate("rect",
+           xmin = bbox[3],
+           xmax = bbox[1],
+           ymin = bbox[2],
+           ymax = bbox[4],
+           fill = NA,
+           colour = "red",
+           size = 1.5
+  ) +
+  theme_void() +
+  theme(panel.background = element_rect(fill = "white"))
+
+# combine site map and inset map
+library(cowplot)
+ggdraw() +
+  draw_plot(map_of_sites) +
+  draw_plot(plot = worldplot,
+            x = -0.02, # x location of inset placement
+            y = -0.06, # y location of inset placement
+            width = .6, # Inset width
+            height = .4, # Inset height
+            scale = 0.75 # Inset scale
+  ) +
+  theme(panel.background = element_rect(fill='white', colour="white"),
+        plot.background = element_rect(fill='white', colour="white"))
+
+ggsave(here::here("figures/site-map.png"),
+       width = 4.45,
+       height = 4.75,
+       units = "in")
+
+```
+
+Display the map
+
+```{r}
+#| label: fig-map
+#| out-width: "400px"
+knitr::include_graphics(here::here("figures/site-map.png"))
 ```
@@
 ```{r}
+#| label: fig-pca-biplot
+
 library(ggrepel)
@@
               "[^_]+")
-pca_data.pca <- prcomp(pca_data, scale. = TRUE)
 
@@
 
-pca_data.pca <- prcomp(pca_data, scale. = TRUE)
+names(pca_data) <- str_replace_all(names(pca_data), "_", " ")
+
+pca_data.pca <- prcomp(pca_data, scale = TRUE)
 
@@
                     name = '')  +
-  theme_minimal()
+  scale_x_continuous(limit = c(-6, 3.5)) +
+  scale_y_continuous(limit = c(-3, 3)) +
+  theme_minimal() +
+  theme(legend.position = c(0.1, 0.9),
+        legend.background = element_rect(fill = "white", linewidth = 0))
 
@@
 #| label: tbl-permanova
+
+set.seed(0411)
 

### 2023-11-06 paper/paper.qmd
update chi-sq figs

@@
            colour = "red",
-           size = 1.5
+           linewidth = 1.5
   ) +
@@
 
+# BM: save this figure to paste into the 
+# drafting document to prepare for submission
+
+ggsave(here::here("figures/fig-chi-sq-residuals-tool-type-freqs1.jpg"),
+       h = 4, 
+       w = 4)
+
 ```
@@
 ```
+
+BM: I don't understand what is the difference between the two chi-square tests where have here (above and below). Can you please clarify?
 
@@
   coord_flip()
+
+# BM: save this figure to paste into the 
+# drafting document to prepare for submission
+
+ggsave(here::here("figures/fig-chi-sq-residuals-tool-type-freqs2.jpg"),
+       h = 4, 
+       w = 4)
 

### 2023-11-18 paper/paper.qmd
edited map
Modify the plot with uppercase x axis and lowercase y axis for chi-square test

@@
   42, # left
-  30, # lower
-  51, # right
-  38) # upper
+  31, # lower
+  50, # right
+  37) # upper
 
@@
            method = 'circle') + 
+  scale_y_discrete(labels = function(y) str_to_upper(y)) +  
+  scale_x_discrete(labels = function(x) str_to_lower(x)) +   
   ggplot2::scale_fill_gradient2(
@@
            method = 'circle') + 
+  scale_y_discrete(labels = function(y) str_to_upper(y)) +  # Convert x axis labels to uppercase
+  scale_x_discrete(labels = function(x) str_to_lower(x)) +  # Convert x axis labels to uppercase
   ggplot2::scale_fill_gradient2(

### 2023-11-20 paper/paper.qmd
improve number spacing on map scale bar, make all text Arial font, fix typo in Qmd, remove duplicate map files

@@
 A permutation MANOVA indicates that there is a significant difference in the lithic assemblage composition across the three excavation areas (`r PerMANOVA_results_inline`).
-We investigated this further using pairwise post-hoc tests and found that this significant difference is between TA and TP1, with the other two areas being statisically indistinguishable (@tbl-permanova )
+We investigated this further using pairwise post-hoc tests and found that this significant difference is between TA and TP1, with the other two areas being statistically indistinguishable (@tbl-permanova )

### 2023-11-20 paper/paper.qmd
update the regional map

@@
            colour = "red",
-           linewidth = 1.5
+           linewidth = 0.5
   ) +
@@
             x = -0.02, # x location of inset placement
-            y = -0.06, # y location of inset placement
+            y = 0.02, # y location of inset placement
             width = .6, # Inset width

### 2024-10-16 paper/paper.qmd
log-transform data before PCA to normalise

@@
 ---
-title: ""
-author: ""
+title: "Tulai"
+author:
+  - name: First Author
+    orcid: 0000-0000-0000-0000
+  - name: Second Author
+    orcid: 0000-0000-0000-0001
 date: now
@@
     self-contained: true
+    title-block-published: "last updated"
 execute: 
@@
   fig-dpi: 600
-title-block-published: "last updated"
+
 ---
+
+The manuscript with figures generated here is at this URL: https://docs.google.com/document/d/1t7oICQn-sWIFlSJSL92ZVy-iKNMAE7kAJ6MvFc0vKVI/edit?usp=sharing
 
@@
 ```{r}
+# we need this pkg for the inline R code below this chunk
+# devtools::install_github('achetverikov/apastats')
 
@@
 
+artefact_types_corrplot <- 
 ggcorrplot(chisq$residuals,
@@
   ) + 
-  coord_flip()
+  coord_flip() +
+  guides(fill = "none") +
+  theme(axis.text.x  = element_text(angle = 0,
+                                    hjust = 0.5))
 
@@
 
+tool_types_corrplot <- 
 ggcorrplot(chisq$residuals,
            method = 'circle') + 
-  scale_y_discrete(labels = function(y) str_to_upper(y)) +  # Convert x axis labels to uppercase
-  scale_x_discrete(labels = function(x) str_to_lower(x)) +  # Convert x axis labels to uppercase
+  # Convert x axis labels to uppercase
+  scale_y_discrete(labels = function(y) str_to_upper(y)) +  
+  # Convert x axis labels to uppercase
+  scale_x_discrete(labels = function(x) str_to_lower(x)) +  
   ggplot2::scale_fill_gradient2(
@@
   ) + 
-  coord_flip()
+  coord_flip() +
+  theme(axis.text.x  = element_text(angle = 0,
+                                    hjust = 0.5))
 
@@
        w = 4)
+
+```
+
+```{r}
+# combine corrplots
+library(cowplot)
+
+plot_grid(
+  artefact_types_corrplot,
+  tool_types_corrplot,
+  labels = "AUTO"
+  )
+
+ggsave(here::here("figures/fig-chi-sq-residuals-both_plots.jpg"),
+       h = 6, 
+       w = 8)
 
@@
                  shape = excavation_area), 
-             size = 4) +
+             size = 5) +
   geom_smooth(method = "lm", 
@@
                label.x = "right",
-               label.y = "bottom") +
+               label.y = "top") +
   labs(
@@
     legend.position = c(.8,.3),
-    axis.text = element_text(size = 7),
-    axis.title = element_text(size = 10),
-    plot.title = element_text(size = 12)
+    axis.text = element_text(size = 14),
+    axis.title = element_text(size = 14),
+    plot.title = element_text(size = 14)
   )
@@
 sum_depth_sheet_subset %>% 
-  select(lithic_density, 
-         core_frequency, 
-         blank_frequency, 
-         retouch_frequency, 
-         tool_diversity)
+   mutate(lithic_density = log(lithic_density + 0.001),  # Adding 1 to avoid log(0)
+         core_frequency = log(core_frequency + 0.001),
+         blank_frequency = log(blank_frequency + 0.001),
+         retouch_frequency = log(retouch_frequency + 0.001),
+         tool_diversity = log(tool_diversity + 0.001)) %>%
+
+  select(`lithic\ndensity` = lithic_density, 
+         `core\nfrequency` = core_frequency, 
+         `blank\nfrequency` = blank_frequency, 
+         `retouch\nfrequency` = retouch_frequency, 
+         `tool\ndiversity` = tool_diversity)
 
@@
 # BM: new method to me! I found https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8858582/
-
+# devtools::install_github("arleyc/PCAtest")
 library(PCAtest)
@@
                    axes = .x, 
-                   top = 10))
+                   top = 10) +
+  theme(axis.text =  element_text(size = 5),
+        axis.text.x = element_text(size = 8,
+                                   angle = 0,
+                                   hjust = 0.5, 
+                                   vjust = 1),
+        axis.title = element_text(size = 6),
+        plot.title = element_text(size = 6)))
 
@@
 library(cowplot)
+contribution_plots_panel <- 
 plot_grid(plotlist = contribution_plots,
-                   ncol = 2)
+                   ncol = 1)
 
@@
 
-names(pca_data) <- str_replace_all(names(pca_data), "_", " ")
+names(pca_data) <- str_replace_all(names(pca_data), "\n", " ")
 
@@
 ```{r}
+# combine PCA plots
+plot_grid(biplot_modified,
+          contribution_plots_panel,
+          ncol = 1,
+       labels = "AUTO")
+
+ggsave(
+       filename = here::here("figures/fig-pca-biplot-and-contribution.jpg"),
+       h = 10, 
+       w = 6)
+
+```
+
+```{r}
 #| label: tbl-permanova
@@
 # https://github.com/pmartinezarbizu/pairwiseAdonis
+# install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")
 library(pairwiseAdonis)

### 2024-10-23 paper/paper.qmd
CA analysis added

@@
 
+```{r}
+# Set the environment variable for GDAL
+Sys.setenv(GDAL_CONFIG = "/path/to/your/conda/env/bin/gdal-config")
+
+# Install the 'terra' package
+install.packages("terra")
+
+```
+
 ### Compute PCA
@@
        w = 8)
+
+```
+
+```{r}
 
@@
 ```{r}
+
+```
+
+```{r}
 #| label: tbl-permanova
@@
 We investigated this further using pairwise post-hoc tests and found that this significant difference is between TA and TP1, with the other two areas being statistically indistinguishable (@tbl-permanova )
+
+# Correspondence Analysis (CA)
+```{r}
+# Example of a contingency table
+lithic_data <- matrix(c(30, 20, 50, 
+                        15, 25, 35,
+                        45, 10, 30), 
+                      nrow = 3, byrow = TRUE)
+colnames(lithic_data) <- c("Cores", "Blades", "Flakes")
+rownames(lithic_data) <- c("Unit1", "Unit2", "Unit3")
+
+```
+
+```{r}
+# Perform Correspondence Analysis
+res.ca <- CA(lithic_data, graph = FALSE)
+
+# View the summary results
+summary(res.ca)
+
+```
+
+```{r}
+# Plot the CA biplot
+fviz_ca_biplot(res.ca, repel = TRUE)
+
+```
+
+# contribution plots
+```{r}
+# Example lithic data matrix
+lithic_data <- matrix(c(30, 20, 50, 
+                        15, 25, 35,
+                        45, 10, 30), 
+                      nrow = 3, byrow = TRUE)
+colnames(lithic_data) <- c("Cores", "Blades", "Flakes")
+rownames(lithic_data) <- c("Unit1", "Unit2", "Unit3")
+
+# Perform Correspondence Analysis (CA)
+res.ca <- CA(lithic_data, graph = FALSE)
+
+```
+
+```{r}
+# Contribution of rows to Dimension 1
+fviz_contrib(res.ca, choice = "row", axes = 1, top = 3)
+
+# Contribution of rows to Dimension 2
+fviz_contrib(res.ca, choice = "row", axes = 2, top = 3)
+```
+
+```{r}
+# Contribution of columns to Dimension 1
+fviz_contrib(res.ca, choice = "col", axes = 1, top = 3)
+
+# Contribution of columns to Dimension 2
+fviz_contrib(res.ca, choice = "col", axes = 2, top = 3)
+
+```
+
+```{r}
+# Combine the top contributing rows for both Dim1 and Dim2
+fviz_contrib(res.ca, choice = "row", axes = 1:2, top = 3)
+
+# Combine the top contributing columns for both Dim1 and Dim2
+fviz_contrib(res.ca, choice = "col", axes = 1:2, top = 3)
+
+```
+
+```{r}
+# Save the contribution plot for rows
+ggsave("contribution_plot_rows_Dim1.png", plot = fviz_contrib(res.ca, choice = "row", axes = 1, top = 3))
+
+# Save the contribution plot for columns
+ggsave("contribution_plot_cols_Dim1.png", plot = fviz_contrib(res.ca, choice = "col", axes = 1, top = 3))
+
+```

### 2024-10-23 paper/paper.qmd
map edited

@@
 library(ggmap) # devtools::install_github("stadiamaps/ggmap")
+library(ggmap)
+ggmap::register_stadiamaps(key = "ca45118a-734f-480b-8c3c-8d82e5a361e3")
 
@@
              colour = "red",
-             size = 2) +
+             size = 0.5) +
   # label points with site name
@@
                   bg.r = 0.15) +
-  theme_minimal(base_size = 6) +
+  theme_minimal(base_size = 10) +
   labs(x = "Longitute",
@@
     annotate("rect",
-           xmin = bbox[3],
-           xmax = bbox[1],
+           xmin = bbox[1],
+           xmax = bbox[3],
            ymin = bbox[2],
@@
        height = 4.75,
-       units = "in")
+       units = "in", 
+       dpi = 300
+       )
 
@@
 
-```
-
-```{r}
-
 # SS: get the contributions of each variable to the first two principal components, which we can then use in the text.

### 2025-07-28 paper/paper.qmd
map revised
stash
Save local changes before merging

@@
 library(ggmap) # devtools::install_github("stadiamaps/ggmap")
-library(ggmap)
 ggmap::register_stadiamaps(key = "ca45118a-734f-480b-8c3c-8d82e5a361e3")
@@
        units = "in", 
-       dpi = 300
+       dpi = 600
        )

### 2025-07-28 paper/paper.qmd
minor edits to readme, update renv

@@
 ```{r}
+# BM: What are these doing here? Not appropriate for a literate programming context. These are for interactive use, in the console.
+
 # Set the environment variable for GDAL
-Sys.setenv(GDAL_CONFIG = "/path/to/your/conda/env/bin/gdal-config")
+# Sys.setenv(GDAL_CONFIG = "/path/to/your/conda/env/bin/gdal-config")
 
 # Install the 'terra' package
-install.packages("terra")
+# install.packages("terra")
 
@@
 ```{r}
-
-```
-
-```{r}
 #| label: tbl-permanova
@@
 A permutation MANOVA indicates that there is a significant difference in the lithic assemblage composition across the three excavation areas (`r PerMANOVA_results_inline`).
-We investigated this further using pairwise post-hoc tests and found that this significant difference is between TA and TP1, with the other two areas being statistically indistinguishable (@tbl-permanova )
+We investigated this further using pairwise post-hoc tests and found that this significant difference is between TA and TP1, with the other two areas being statistically indistinguishable (@tbl-permanova)
 
 # Correspondence Analysis (CA)
+
 ```{r}
@@
 
-# contribution plots
+# Contribution plots
+
 ```{r}
